---
title: "Xenium panel resolution benchmark"
format: html
editor: source
---

Determines the finest level of the snRNA-seq cell type taxonomy that can be
recovered from the 425-gene Xenium panel (design `Z2XBPV`) alone. This is the
prerequisite for label transfer onto the Xenium sections: it sets the label
granularity every downstream spatial analysis is allowed to assume.

The reference snRNA-seq data are subset to the panel genes and binomially
thinned to the transcript depth actually observed in the Xenium run, producing a
"pseudo-Xenium" dataset with known labels. Held-out cells are then classified
under cross-validation, at three levels of the taxonomy, with two independent
classifiers.

```{r}
library(tidyverse)
library(Seurat)
library(BPCells)   # snRNA counts layers are on-disk BPCells matrices
library(qs2)
library(Matrix)
library(cowplot)
library(here)

source(here::here("config/paths.R"))
select = dplyr::select
```

```{r}
script_name = "panel_resolution_benchmark"
out_dir = here::here("xenium/label_transfer", script_name)
dir.create(out_dir, recursive = T, showWarnings = F)

ref_obj_fname = file.path(SNRNA_SEURAT_DIR, "obj_clustered.qs2")

## CellBender-corrected counts (default RNA assay of the snRNA object)
assay_to_use = "RNA"
label_col = "cluster"

## Cross-validation
n_folds = 5
min_cells_per_label = 30    # labels below this are excluded and reported
max_cells_per_label = 500   # downsample large clusters for tractability
npcs = 30                   # < 425 features, so PCA is well posed

## A cluster is judged "recoverable" at this F1 against held-out truth.
f1_threshold = 0.5

## Labels that must survive as themselves. Merging one of these away would
## destroy a planned comparison, so a failing label is never merged into them —
## it is reported as unresolved instead. Glut-NC-1 is the comparator for
## Glut-HVC-1; the song nuclei types are the subject of the spatial analysis.
protected_labels = c("Glut-NC-1", "Glut-HVC-1", "Glut-HVC-1a", "Glut-HVC-2",
                     "Glut-RA", "Glut-LMAN-1", "Glut-LMAN-2", "Glut-Arco-1")

## Minimum share of a failing label's cells that must be confused with a
## candidate partner before the two are merged. Below this the confusion is
## diffuse rather than a specific collision, and merging would degrade a
## perfectly good label to rescue a bad one; the failing label is left
## unresolved instead.
min_merge_frac = 0.15

seed = 10
set.seed(seed)
options(future.globals.maxSize = Inf)
```

## Panel and observed Xenium depth

```{r}
panel = read_csv(XENIUM_PANEL_CSV, show_col_types = FALSE)
panel_genes = panel$Gene
length(panel_genes)
```

Target depth for thinning is the median transcripts per cell actually achieved
across the ten Xenium sections, not a nominal value.

```{r}
metrics_fnames = list.files(XENIUM_RAW_DIR, pattern = "metrics_summary.csv",
                            recursive = TRUE, full.names = TRUE)

xenium_metrics = map_dfr(metrics_fnames, read_csv, show_col_types = FALSE) %>%
  select(cassette_name, region_name, num_cells_detected,
         median_transcripts_per_cell, median_genes_per_cell,
         fraction_transcripts_assigned, adjusted_negative_control_probe_rate)

xenium_metrics

target_depth = median(xenium_metrics$median_transcripts_per_cell)
target_depth
```

## Reference

```{r}
obj = qs2::qs_read(ref_obj_fname)

## Same filter as snrna/deg/all_deg.qmd: drop cells with ambiguous sample assignment
cells = Cells(obj)[!grepl("/", obj$assignment)]
obj = subset(obj, cells = cells)

counts_all = LayerData(obj, assay = assay_to_use, layer = "counts")
md = obj@meta.data
dim(counts_all)
```

Panel coverage of the reference feature space. All 425 genes are expected to be
present — the panel was designed against this dataset — so any shortfall here is
a symbol-mismatch bug, not a biological result.

```{r}
panel_in_ref = intersect(panel_genes, rownames(counts_all))
panel_missing = setdiff(panel_genes, rownames(counts_all))

tibble(n_panel = length(panel_genes),
       n_in_reference = length(panel_in_ref),
       n_missing = length(panel_missing))

panel_missing
```

```{r}
tibble(gene = panel_genes,
       in_reference = panel_genes %in% rownames(counts_all)) %>%
  write_csv(file.path(out_dir, "panel_genes_in_reference.csv"))
```

## Pseudo-Xenium construction

Subset to panel genes, then binomially thin each cell to the observed Xenium
depth. Thinning matters: reference cells carry far more counts over these genes
than a Xenium cell does, and an un-thinned benchmark would report an accuracy
the real data cannot reach.

```{r}
thin_counts = function(m, target, seed = 1) {
  set.seed(seed)
  m = as(m, "CsparseMatrix")
  tot = Matrix::colSums(m)
  p = pmin(1, target / pmax(tot, 1))
  col_idx = rep.int(seq_len(ncol(m)), diff(m@p))
  m@x = as.numeric(rbinom(length(m@x), size = as.integer(round(m@x)), prob = p[col_idx]))
  Matrix::drop0(m)
}

## Materialize the 425-gene slice out of the on-disk BPCells matrix
counts_panel = as(counts_all[panel_in_ref, ], "dgCMatrix")
counts_thin = thin_counts(counts_panel, target_depth, seed = seed)

tibble(stage = c("reference, panel genes", "thinned", "Xenium (observed)"),
       median_counts_per_cell = c(median(Matrix::colSums(counts_panel)),
                                  median(Matrix::colSums(counts_thin)),
                                  target_depth),
       median_genes_per_cell = c(median(Matrix::colSums(counts_panel > 0)),
                                 median(Matrix::colSums(counts_thin > 0)),
                                 median(xenium_metrics$median_genes_per_cell)))
```

## Label hierarchy

Three nested levels. `cluster` is the published 50-type taxonomy; `subclass`
collapses numbered variants of a type; `class` collapses to the major
populations.

```{r}
to_subclass = function(x) {
  map_chr(x, function(nm) {
    tok = str_split(nm, "-")[[1]]
    if (tok[1] %in% c("Astro", "Oligo")) return(tok[1])
    if (length(tok) >= 3) return(paste(tok[1], tok[2], sep = "-"))
    nm
  })
}

## Glut, GABA, and each non-neuronal population (Astro, Oligo, OPC, Micro, Endo, Epen)
to_class = function(x) str_split_i(x, "-", 1)

labels_tbl = tibble(cell = rownames(md),
                    cluster = as.character(md[[label_col]])) %>%
  filter(!is.na(cluster)) %>%
  mutate(subclass = to_subclass(cluster),
         class = to_class(cluster))

labels_tbl %>% count(class, subclass, cluster) %>% print(n = Inf)
```

Exclude clusters too small to evaluate, and cap the large ones. Capping
deliberately flattens cluster abundance so the benchmark measures the panel's
discriminative information rather than the prior.

```{r}
label_n = labels_tbl %>% count(cluster, name = "n_cells")

labels_dropped = label_n %>% filter(n_cells < min_cells_per_label)
labels_dropped

set.seed(seed)
bench_tbl = labels_tbl %>%
  semi_join(label_n %>% filter(n_cells >= min_cells_per_label), by = "cluster") %>%
  group_by(cluster) %>%
  slice_sample(n = max_cells_per_label) %>%
  ungroup()

nrow(bench_tbl)
n_distinct(bench_tbl$cluster)
```

```{r}
counts_bench = counts_thin[, bench_tbl$cell]

## Cells left with no transcripts after thinning cannot be classified by any
## method; Xenium QC would remove them too.
keep = Matrix::colSums(counts_bench) > 0
bench_tbl = bench_tbl[keep, ]
counts_bench = counts_bench[, keep]
sum(!keep)
```

## Cross-validation folds

Stratified by cluster, so every fold sees every label.

```{r}
set.seed(seed)
bench_tbl = bench_tbl %>%
  group_by(cluster) %>%
  mutate(fold = sample(rep_len(seq_len(n_folds), n()))) %>%
  ungroup()

table(bench_tbl$fold)
```

## Classifier 1 — multinomial naive Bayes

Fast, calibrated for count data, and free of the assumptions of anchor-based
integration. Uniform class prior, Laplace smoothing. This is the reference
point for how much information the panel carries.

```{r}
nb_fit = function(counts, labels) {
  labels = factor(labels)
  ind = Matrix::sparse.model.matrix(~ 0 + labels)
  colnames(ind) = levels(labels)
  s = as.matrix(counts %*% ind)
  theta = sweep(s + 1, 2, colSums(s) + nrow(counts), "/")
  log(theta)
}

nb_predict = function(counts, log_theta) {
  ll = as.matrix(Matrix::crossprod(counts, log_theta))
  colnames(log_theta)[max.col(ll, ties.method = "first")]
}

nb_res = map_dfr(seq_len(n_folds), function(k) {
  train = bench_tbl$fold != k
  lt = nb_fit(counts_bench[, train, drop = FALSE], bench_tbl$cluster[train])
  tibble(cell = bench_tbl$cell[!train],
         truth = bench_tbl$cluster[!train],
         pred = nb_predict(counts_bench[, !train, drop = FALSE], lt),
         fold = k, method = "naive_bayes")
})

nrow(nb_res)
```

## Classifier 2 — Seurat transfer anchors

Mirrors the workflow that would actually be applied to the Xenium data, so its
gap from naive Bayes indicates method loss rather than panel loss.

```{r}
make_obj = function(counts, labels = NULL) {
  o = CreateSeuratObject(counts = counts)
  if (!is.null(labels)) o$label = labels
  o = NormalizeData(o, scale.factor = median(Matrix::colSums(counts)), verbose = FALSE)
  o
}

transfer_res = map_dfr(seq_len(n_folds), function(k) {
  train = bench_tbl$fold != k

  ref = make_obj(counts_bench[, train, drop = FALSE], bench_tbl$cluster[train])
  ref = ScaleData(ref, features = panel_in_ref, verbose = FALSE)
  ref = RunPCA(ref, features = panel_in_ref, npcs = npcs, verbose = FALSE)

  qry = make_obj(counts_bench[, !train, drop = FALSE])

  anchors = FindTransferAnchors(reference = ref, query = qry,
                                features = panel_in_ref,
                                normalization.method = "LogNormalize",
                                reference.reduction = "pca",
                                dims = 1:npcs, k.filter = NA, verbose = FALSE)

  td = TransferData(anchorset = anchors, refdata = ref$label,
                    dims = 1:npcs, verbose = FALSE)

  tibble(cell = bench_tbl$cell[!train],
         truth = bench_tbl$cluster[!train],
         pred = td$predicted.id,
         score = td$predicted.id.score,
         fold = k, method = "transfer_anchors")
})

nrow(transfer_res)
```

```{r}
cv_res = bind_rows(nb_res, transfer_res) %>%
  left_join(labels_tbl %>% select(cluster, subclass, class) %>% distinct(),
            by = c("truth" = "cluster")) %>%
  rename(truth_subclass = subclass, truth_class = class) %>%
  mutate(pred_subclass = to_subclass(pred),
         pred_class = to_class(pred))

qs2::qs_save(cv_res, file.path(out_dir, "cv_predictions.qs2"))
```

## Metrics

```{r}
prf = function(truth, pred) {
  lv = union(truth, pred)
  truth = factor(truth, levels = lv); pred = factor(pred, levels = lv)
  cm = table(truth = truth, pred = pred)
  tp = diag(cm)
  tibble(label = lv,
         n_true = as.integer(rowSums(cm)),
         n_pred = as.integer(colSums(cm)),
         tp = as.integer(tp)) %>%
    mutate(recall = if_else(n_true > 0, tp / n_true, NA_real_),
           precision = if_else(n_pred > 0, tp / n_pred, NA_real_),
           f1 = if_else(recall + precision > 0,
                        2 * recall * precision / (recall + precision), 0))
}

levels_tbl = tribble(
  ~level,      ~truth_col,       ~pred_col,
  "cluster",   "truth",          "pred",
  "subclass",  "truth_subclass", "pred_subclass",
  "class",     "truth_class",    "pred_class"
)

per_label = pmap_dfr(levels_tbl, function(level, truth_col, pred_col) {
  cv_res %>%
    group_by(method) %>%
    group_modify(~ prf(.x[[truth_col]], .x[[pred_col]])) %>%
    ungroup() %>%
    mutate(level = level)
})

summary_tbl = per_label %>%
  group_by(method, level) %>%
  summarize(n_labels = n(),
            macro_f1 = mean(f1),
            balanced_accuracy = mean(recall, na.rm = TRUE),
            .groups = "drop") %>%
  left_join(cv_res %>%
              group_by(method) %>%
              summarize(accuracy_cluster = mean(truth == pred), .groups = "drop"),
            by = "method") %>%
  arrange(method, factor(level, levels = c("cluster", "subclass", "class")))

summary_tbl
write_csv(per_label, file.path(out_dir, "per_label_metrics.csv"))
write_csv(summary_tbl, file.path(out_dir, "summary_metrics.csv"))
```

Per-cluster performance, worst first. These are the clusters that cannot be
mapped onto the Xenium sections as distinct types.

```{r}
per_label %>%
  filter(level == "cluster", method == "naive_bayes") %>%
  arrange(f1) %>%
  select(label, n_true, recall, precision, f1) %>%
  print(n = Inf)
```

```{r}
per_label %>%
  filter(level == "cluster") %>%
  select(method, label, f1) %>%
  pivot_wider(names_from = method, values_from = f1) %>%
  ggplot(aes(naive_bayes, transfer_anchors, label = label)) +
  geom_abline(slope = 1, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.7) +
  ggrepel::geom_text_repel(size = 2.5, max.overlaps = 20) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "F1, naive Bayes", y = "F1, transfer anchors",
       title = "Per-cluster recoverability from the 425-gene panel") +
  theme_cowplot()

ggsave(file.path(out_dir, "f1_by_method.pdf"), width = 8, height = 8)
```

## Confusion structure

Row-normalized: what each true cluster is called. Off-diagonal mass shows which
types collapse into which — that is what determines the sensible label set, not
the F1 values alone.

```{r}
plot_confusion = function(df, truth_col, pred_col, title) {
  lv = sort(union(df[[truth_col]], df[[pred_col]]))
  df %>%
    count(truth = .data[[truth_col]], pred = .data[[pred_col]]) %>%
    group_by(truth) %>%
    mutate(frac = n / sum(n)) %>%
    ungroup() %>%
    mutate(truth = factor(truth, levels = lv),
           pred = factor(pred, levels = lv)) %>%
    ggplot(aes(pred, truth, fill = frac)) +
    geom_tile() +
    scale_fill_viridis_c(limits = c(0, 1)) +
    labs(x = "predicted", y = "true", title = title) +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
          axis.text.y = element_text(size = 6))
}

for (m in unique(cv_res$method)) {
  d = cv_res %>% filter(method == m)
  p1 = plot_confusion(d, "truth", "pred", paste(m, "- cluster"))
  p2 = plot_confusion(d, "truth_subclass", "pred_subclass", paste(m, "- subclass"))
  ggsave(file.path(out_dir, paste0("confusion_cluster_", m, ".pdf")), p1, width = 11, height = 10)
  ggsave(file.path(out_dir, paste0("confusion_subclass_", m, ".pdf")), p2, width = 9, height = 8)
  print(p1); print(p2)
}
```

## Recommended label set

Collapsing a failing cluster onto its position in the taxonomy would only be a
rename unless the labels it is *actually* confused with come with it, so the
recommended set is derived from the confusion structure instead of the naming
hierarchy: repeatedly merge the worst-performing label with the label carrying
the most confusion mass with it, until every merged label clears
`f1_threshold`. The result is the finest partition of the taxonomy that the
panel supports, which may cut across the `cluster` → `subclass` → `class`
hierarchy.

Two guards keep the merge from doing more harm than good. Labels in
`protected_labels` are never merged into, so a planned comparison can never be
destroyed to rescue a failing label. And a merge requires the partner to account
for at least `min_merge_frac` of the failing label's cells: below that the
confusion is diffuse rather than a specific collision between two types, and
merging would degrade a perfectly good label for no gain.

A failing label that clears neither guard is marked **unresolved** and left as
is. Unresolved labels should not be interpreted as distinct populations in the
spatial data, but they also do not contaminate anything else.

The merge sequence is derived from naive Bayes — the estimate of what the panel
can support in principle — and the resulting label set is then scored under both
classifiers.

```{r}
merge_until_resolved = function(truth, pred, threshold, protected = character(),
                                min_frac = 0, max_iter = 100) {
  lv = union(truth, pred)
  members = setNames(as.list(lv), lv)          # merged label -> original clusters
  map_lab = setNames(lv, lv)                   # original cluster -> merged label
  unresolved = character()
  log = tibble()

  for (i in seq_len(max_iter)) {
    t2 = map_lab[truth]; p2 = map_lab[pred]
    cur = union(t2, p2)
    m = prf(t2, p2)
    worst = m %>%
      filter(f1 < threshold, !label %in% unresolved, !label %in% protected) %>%
      arrange(f1, desc(n_true)) %>%
      slice(1)
    if (nrow(worst) == 0) break

    cm = table(factor(t2, levels = cur), factor(p2, levels = cur))
    conf = cm[worst$label, ] + cm[, worst$label]
    conf[c(worst$label, intersect(protected, names(conf)))] = -1

    ## No unprotected partner left, or the best one is only diffusely confused
    if (max(conf) <= 0 || max(conf) / worst$n_true < min_frac) {
      unresolved = c(unresolved, worst$label)
      log = bind_rows(log, tibble(iteration = i, merged = worst$label,
                                  f1 = worst$f1,
                                  with = if (max(conf) > 0) names(which.max(conf)) else NA_character_,
                                  best_partner_frac = max(conf) / worst$n_true,
                                  into = NA_character_, action = "unresolved"))
      next
    }
    partner = names(which.max(conf))

    new_members = c(members[[worst$label]], members[[partner]])
    subcl = unique(to_subclass(new_members))
    new_name = if (length(subcl) == 1) subcl else paste(sort(subcl), collapse = "/")
    ## keep names unique if a merge lands on an existing group name
    while (new_name %in% setdiff(names(members), c(worst$label, partner))) {
      new_name = paste0(new_name, "*")
    }

    log = bind_rows(log, tibble(iteration = i, merged = worst$label,
                                f1 = worst$f1, with = partner,
                                best_partner_frac = max(conf) / worst$n_true,
                                into = new_name, action = "merged"))

    members[[worst$label]] = NULL; members[[partner]] = NULL
    members[[new_name]] = new_members
    map_lab[new_members] = new_name
  }

  list(map = map_lab, members = members, log = log, unresolved = unresolved)
}

merge_res = cv_res %>%
  filter(method == "naive_bayes") %>%
  { merge_until_resolved(.$truth, .$pred, f1_threshold,
                         protected = protected_labels, min_frac = min_merge_frac) }

merge_res$log
merge_res$unresolved
```

```{r}
cv_rec = cv_res %>%
  mutate(truth_rec = merge_res$map[truth],
         pred_rec = merge_res$map[pred])

per_label_rec = cv_rec %>%
  group_by(method) %>%
  group_modify(~ prf(.x$truth_rec, .x$pred_rec)) %>%
  ungroup()

rec_summary = per_label_rec %>%
  group_by(method) %>%
  summarize(n_labels = n(),
            macro_f1 = mean(f1),
            balanced_accuracy = mean(recall, na.rm = TRUE),
            .groups = "drop") %>%
  left_join(cv_rec %>%
              group_by(method) %>%
              summarize(accuracy = mean(truth_rec == pred_rec), .groups = "drop"),
            by = "method")

rec_summary

per_label_rec %>%
  filter(method == "naive_bayes") %>%
  arrange(f1) %>%
  select(label, n_true, recall, precision, f1) %>%
  print(n = Inf)
```

Labels still below threshold after merging — protected labels, and labels left
unresolved because every confusion partner was protected. These require a
judgement call rather than an automatic rule, and should not be interpreted as
distinct populations in the spatial data.

```{r}
per_label_rec %>%
  filter(f1 < f1_threshold) %>%
  mutate(status = case_when(label %in% merge_res$unresolved ~ "unresolved",
                            label %in% protected_labels ~ "protected",
                            TRUE ~ "below threshold")) %>%
  arrange(f1)
```

```{r}
recommended = tibble(cluster = sort(unique(labels_tbl$cluster))) %>%
  mutate(subclass = to_subclass(cluster),
         class = to_class(cluster),
         evaluated = cluster %in% bench_tbl$cluster,
         recommended_label = unname(merge_res$map[cluster]),
         resolved_at_cluster = evaluated & recommended_label == cluster,
         protected = cluster %in% protected_labels,
         status = case_when(!evaluated ~ "not evaluated",
                            recommended_label %in% merge_res$unresolved ~ "unresolved",
                            recommended_label != cluster ~ "merged",
                            TRUE ~ "resolved"))

recommended %>% print(n = Inf)
write_csv(recommended, file.path(out_dir, "recommended_label_set.csv"))
write_csv(per_label_rec, file.path(out_dir, "per_label_metrics_recommended.csv"))
write_csv(rec_summary, file.path(out_dir, "summary_metrics_recommended.csv"))
write_csv(merge_res$log, file.path(out_dir, "merge_log.csv"))
```

```{r}
p = plot_confusion(cv_rec %>% filter(method == "naive_bayes"),
                   "truth_rec", "pred_rec", "recommended label set - naive Bayes")
ggsave(file.path(out_dir, "confusion_recommended.pdf"), p, width = 10, height = 9)
p
```

```{r}
sessionInfo()
```
