---
title: "Xenium spatial maps — proseg RCTD calls, song-system cell types"
format: html
editor: source
---

Spatial location of confident RCTD calls (proseg segmentation, all ten
sections) for the ten glutamatergic types named as the primary targets of this
analysis: `Glut-HVC-1`, `Glut-HVC-2`, `Glut-NC-1`, `Glut-NC-4`, `Glut-RA`,
`Glut-Arco-1`, `Glut-LMAN-1`, `Glut-LMAN-2`, `Glut-NR-1`, `Glut-NR-4`.

Two products: one combined figure with all ten types overlaid (distinct
colors, faceted by section), and ten single-type figures (same layout, one
type highlighted at a time) — the single-type views are what should be used to
judge any one type's distribution, since with ten overlapping colors the
combined figure is dominated by whichever type happens to be plotted last in
a given location.

Every panel also shows every profiled cell in light grey, regardless of
`spot_class` or type, so the tissue outline and section orientation are always
visible — without it, the highlighted points alone give no anatomical context.

```{r}
library(tidyverse)
library(Seurat)   # only for DiscretePalette — no Seurat object involved here
library(cowplot)
library(here)

source(here::here("config/paths.R"))
select = dplyr::select
```

```{r}
script_name = "proseg_spatial_maps"
hpc_dir = path.expand("~/hdd/rstudio/xenium/260811_brainard_adult-425g/hpc_rctd_proseg")
out_dir = here::here("xenium/label_transfer", script_name)
dir.create(out_dir, recursive = T, showWarnings = F)

song_types = c("Glut-HVC-1", "Glut-HVC-2", "Glut-NC-1", "Glut-NC-4", "Glut-RA",
              "Glut-Arco-1", "Glut-LMAN-1", "Glut-LMAN-2", "Glut-NR-1", "Glut-NR-4")

section_order = c("OR52YW26_1_4", "OR52YW26_1_7", "OR52YW26_2_2", "OR52YW26_2_4", "OR52YW26_2_7",
                  "OR69PU4_1_4", "OR69PU4_1_7", "OR69PU4_2_2", "OR69PU4_2_4", "OR69PU4_2_7")

## confident = the same threshold used throughout xenium_label_transfer.qmd and
## the proseg validation: RCTD singlet or confident doublet
confident_classes = c("singlet", "doublet_certain")
```

## Load

```{r}
rctd = read_csv(file.path(hpc_dir, "rctd_all.csv.gz"), show_col_types = FALSE) %>%
  select(cell, spot_class, first_type)

md = read_csv(file.path(hpc_dir, "proseg_cell_metadata.csv.gz"), show_col_types = FALSE) %>%
  select(cell, section_id, x_centroid, y_centroid)

## Per-cell transcript count (panel nUMI), for the depth-filtered set built
## further down. Already computed once by RCTD (`spatialRNA@nUMI` inside every
## chunk object, sliced from `colSums()` at query-build time) — read back out
## of the chunks rather than resumming raw proseg counts.
ncount = read_csv(file.path(hpc_dir, "proseg_ncount.csv.gz"), show_col_types = FALSE)

d = md %>%
  left_join(rctd, by = "cell") %>%
  left_join(ncount, by = "cell") %>%
  mutate(section_id = factor(section_id, levels = section_order),
         confident = spot_class %in% confident_classes)

nrow(d)
d %>% count(section_id)
summary(d$nCount)
```

## Common orientation — posterior left, dorsal up

Per-section rotation/flip so every panel shares the same anatomical
orientation, applied to `x_centroid`/`y_centroid` before any plotting (raw
coordinates elsewhere in the pipeline — ROI polygons, RCTD itself — are
untouched; this is cosmetic to this figure set only). Rotations are in the
standard mathematical sense (x right, y up): CCW90 is `(x,y) -> (-y,x)`, CW90 is
`(x,y) -> (y,-x)`; a flip negates x, applied after any rotation.
`OR52YW26_1_4` and `OR69PU4_1_4` are not listed and are left as-is.

```{r}
section_transform = tribble(
  ~section_id,     ~rotate,  ~flip_h,
  "OR52YW26_1_4",  "none",   FALSE,
  "OR52YW26_1_7",  "ccw90",  FALSE,
  "OR52YW26_2_2",  "ccw90",  TRUE,
  "OR52YW26_2_4",  "ccw90",  FALSE,
  "OR52YW26_2_7",  "cw90",   TRUE,
  "OR69PU4_1_4",   "none",   FALSE,
  "OR69PU4_1_7",   "ccw90",  TRUE,
  "OR69PU4_2_2",   "cw90",   FALSE,
  "OR69PU4_2_4",   "ccw90",  FALSE,
  "OR69PU4_2_7",   "ccw90",  FALSE,
)
stopifnot(setequal(section_transform$section_id, section_order))

d = d %>%
  left_join(section_transform, by = "section_id") %>%
  mutate(
    x_rot = case_when(rotate == "none"  ~ x_centroid,
                      rotate == "ccw90" ~ -y_centroid,
                      rotate == "cw90"  ~ y_centroid),
    y_rot = case_when(rotate == "none"  ~ y_centroid,
                      rotate == "ccw90" ~ x_centroid,
                      rotate == "cw90"  ~ -x_centroid),
    x_plot = if_else(flip_h, -x_rot, x_rot),
    y_plot = y_rot) %>%
  select(-rotate, -flip_h, -x_rot, -y_rot)
```

```{r}
stopifnot(all(song_types %in% unique(d$first_type)))

highlight = d %>% filter(confident, first_type %in% song_types) %>%
  mutate(first_type = factor(first_type, levels = song_types))

highlight %>% count(first_type) %>% arrange(desc(n))
```

## Palette

`Seurat::DiscretePalette(..., "polychrome")`'s first two colors are `#5A5156`
(dark grey) and `#E4E1E3` (near-white) — both nearly invisible against the
grey background / white page used here, and Glut-HVC-1/Glut-HVC-2 are exactly
the two types that most need to stay legible. Picked by hand instead: fully
saturated, mutually distinguishable, none close to white or grey.

Glut-HVC-1's original color (`#E6194B`) was replaced with black — it was
nearly identical to `position_colors["ra"]` (`#e31a1c`) in
`/opt/colquittlab/utils/R/common_aesthetics.R`, the lab's standard color for
the RA position, which risked being misread as RA in any figure shown
alongside that convention. Black is unused elsewhere in this palette and
reads clearly against the grey background. The same fix (that red -> black)
is applied everywhere else in this document a single fixed highlight color is
used: `single_type_colour`, `pair_song_colour`, and `pre_colors["Glut-Pre-1"]`.

```{r}
type_colors = setNames(
  c("#000000", "#3CB44B", "#4363D8", "#F58231", "#911EB4",
    "#42D4F4", "#F032E6", "#BFEF45", "#000075", "#9A6324"),
  song_types)

scales::show_col(type_colors)
```

## Plot helper

`coord_equal()` (needed so tissue isn't visually stretched) cannot be combined
with `facet_wrap(scales = "free")` in current ggplot2 — sections differ enough
in extent and aspect that a shared scale would shrink most of them to a sliver.
Built as one `coord_equal()` plot per section instead, then assembled with
`cowplot::plot_grid()`, which gives the same layout without that restriction.

```{r}
theme_spatial = theme_cowplot() +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.line = element_blank(), axis.ticks = element_blank(),
        plot.title = element_text(size = 10, hjust = 0.5))

section_plot = function(sid, hl, point_colour = NULL) {
  bg = d %>% filter(section_id == sid)
  hl_sid = hl %>% filter(section_id == sid)

  p = ggplot() +
    geom_point(data = bg, aes(x_plot, y_plot),
               colour = "grey85", size = 0.03, alpha = 0.4) +
    coord_equal() + labs(title = sid) + theme_spatial

  if (nrow(hl_sid) == 0) return(p)

  if (is.null(point_colour)) {
    p + geom_point(data = hl_sid, aes(x_plot, y_plot, colour = first_type),
                   size = 0.25, alpha = 0.85) +
      scale_colour_manual(values = type_colors, name = NULL, limits = song_types,
                          guide = "none")
  } else {
    p + geom_point(data = hl_sid, aes(x_plot, y_plot),
                   colour = point_colour, size = 0.3, alpha = 0.85)
  }
}
```

## Aggregate — all ten types, one figure

```{r}
panels = map(section_order, section_plot, hl = highlight)

## shared legend, built once from a dummy plot carrying every type level
legend_plot = ggplot(tibble(first_type = factor(song_types, levels = song_types)),
                     aes(1, 1, colour = first_type)) +
  geom_point(size = 3) +
  scale_colour_manual(values = type_colors, name = NULL)
legend = cowplot::get_legend(legend_plot)

grid = cowplot::plot_grid(plotlist = panels, ncol = 5)
p_agg = cowplot::plot_grid(grid, legend, ncol = 2, rel_widths = c(1, 0.12))

ggsave(file.path(out_dir, "spatial_song_types_aggregate.png"), p_agg,
       width = 22, height = 9, dpi = 250, bg = "white")
p_agg
```

## Individual — one type per figure

```{r}
for (ty in song_types) {
  hl = highlight %>% filter(first_type == ty)
  panels = map(section_order, section_plot, hl = hl, point_colour = type_colors[[ty]])
  p = cowplot::plot_grid(plotlist = panels, ncol = 5) +
    cowplot::draw_figure_label(paste0(ty, "  (n = ", nrow(hl), " confident cells)"),
                               position = "top", size = 14)

  fname = paste0("spatial_", str_replace_all(ty, "-", "_"), ".png")
  ggsave(file.path(out_dir, fname), p, width = 20, height = 9.5, dpi = 250, bg = "white")
}

list.files(out_dir, pattern = "^spatial_")
```

## Song vs. non-song comparator pairs

Five song-nucleus / adjacent-comparator pairs, each pair in one figure, song
type drawn on top of its comparator so it isn't obscured where the comparator
is more abundant. Fixed colors reused across all five pairs (song = red,
comparator = blue) rather than a distinct palette per pair.

```{r}
pair_list = tribble(
  ~song,          ~nonsong,
  "Glut-HVC-1",   "Glut-NC-1",
  "Glut-HVC-2",   "Glut-NC-4",
  "Glut-LMAN-1",  "Glut-NR-1",
  "Glut-LMAN-2",  "Glut-NR-4",
  "Glut-RA",      "Glut-Arco-1",
)
stopifnot(all(c(pair_list$song, pair_list$nonsong) %in% song_types))

pair_song_colour = "#000000"
pair_nonsong_colour = "#4363D8"

pair_section_plot = function(sid, song_ty, nonsong_ty, hl_data = highlight) {
  bg = d %>% filter(section_id == sid)
  hl_non  = hl_data %>% filter(section_id == sid, first_type == nonsong_ty)
  hl_song = hl_data %>% filter(section_id == sid, first_type == song_ty)

  ggplot() +
    geom_point(data = bg, aes(x_plot, y_plot), colour = "grey85", size = 0.03, alpha = 0.4) +
    geom_point(data = hl_non, aes(x_plot, y_plot), colour = pair_nonsong_colour,
               size = 0.25, alpha = 0.8) +
    geom_point(data = hl_song, aes(x_plot, y_plot), colour = pair_song_colour,
               size = 0.25, alpha = 0.85) +
    coord_equal() + labs(title = sid) + theme_spatial
}

for (i in seq_len(nrow(pair_list))) {
  song_ty = pair_list$song[i]; nonsong_ty = pair_list$nonsong[i]
  panels = map(section_order, pair_section_plot, song_ty = song_ty, nonsong_ty = nonsong_ty)
  p = cowplot::plot_grid(plotlist = panels, ncol = 5) +
    cowplot::draw_figure_label(
      paste0(song_ty, " (red)  vs.  ", nonsong_ty, " (blue)"),
      position = "top", size = 14)

  fname = paste0("spatial_pair_", str_replace_all(song_ty, "-", "_"),
                 "_vs_", str_replace_all(nonsong_ty, "-", "_"), ".png")
  ggsave(file.path(out_dir, fname), p, width = 20, height = 9.5, dpi = 250, bg = "white")
}

list.files(out_dir, pattern = "^spatial_pair_")
```

## Individual — every cell call in the taxonomy

All 48 labels present in the RCTD output, not just the ten glutamatergic
targets above. One fixed colour throughout (no per-type palette, since only
one type is ever shown per figure) — kept as a separate `spatial_all_*` file
series so the distinctly-coloured ten figures above are untouched.

```{r}
all_types = sort(unique(d$first_type))
length(all_types)

single_type_colour = "#000000"

for (ty in all_types) {
  hl = d %>% filter(confident, first_type == ty)
  panels = map(section_order, section_plot, hl = hl, point_colour = single_type_colour)
  p = cowplot::plot_grid(plotlist = panels, ncol = 5) +
    cowplot::draw_figure_label(paste0(ty, "  (n = ", nrow(hl), " confident cells)"),
                               position = "top", size = 14)

  ## "/" in Glut-NC/Glut-Nido would otherwise break the path
  safe_name = str_replace_all(ty, "[/-]", "_")
  ggsave(file.path(out_dir, paste0("spatial_all_", safe_name, ".png")), p,
         width = 20, height = 9.5, dpi = 250, bg = "white")
}

list.files(out_dir, pattern = "^spatial_all_")
```

## Aggregate — Glut-Pre-1/2/3

Same construction as the ten-type aggregate above, scoped to the three
`Glut-Pre-*` types with their own small palette.

```{r}
pre_types = c("Glut-Pre-1", "Glut-Pre-2", "Glut-Pre-3")
stopifnot(all(pre_types %in% unique(d$first_type)))

pre_colors = setNames(c("#000000", "#3CB44B", "#4363D8"), pre_types)

pre_highlight = d %>% filter(confident, first_type %in% pre_types) %>%
  mutate(first_type = factor(first_type, levels = pre_types))
pre_highlight %>% count(first_type)

pre_section_plot = function(sid, hl_data = pre_highlight) {
  bg = d %>% filter(section_id == sid)
  hl_sid = hl_data %>% filter(section_id == sid)

  p = ggplot() +
    geom_point(data = bg, aes(x_plot, y_plot), colour = "grey85", size = 0.03, alpha = 0.4) +
    coord_equal() + labs(title = sid) + theme_spatial

  if (nrow(hl_sid) == 0) return(p)

  p + geom_point(data = hl_sid, aes(x_plot, y_plot, colour = first_type),
                 size = 0.25, alpha = 0.85) +
    scale_colour_manual(values = pre_colors, name = NULL, limits = pre_types, guide = "none")
}

panels = map(section_order, pre_section_plot)

legend_plot = ggplot(tibble(first_type = factor(pre_types, levels = pre_types)),
                     aes(1, 1, colour = first_type)) +
  geom_point(size = 3) +
  scale_colour_manual(values = pre_colors, name = NULL)
legend = cowplot::get_legend(legend_plot)

grid = cowplot::plot_grid(plotlist = panels, ncol = 5)
p_pre = cowplot::plot_grid(grid, legend, ncol = 2, rel_widths = c(1, 0.12))

ggsave(file.path(out_dir, "spatial_glut_pre_aggregate.png"), p_pre,
       width = 22, height = 9, dpi = 250, bg = "white")
p_pre
```

## Depth-filtered parallel set (nCount >= 400)

Every plot above uses only `confident` (RCTD singlet or confident doublet) with
no floor on transcript count beyond RCTD's own `UMI_min = 20` — permissive,
since a cell can be a confident *call* while still resting on very little
signal. `xenium_label_transfer.qmd`'s depth analysis found composition below
roughly 400 counts is increasingly depth-driven rather than identity-driven
(e.g. `Endo` runs 49% of calls under 100 counts, 0.5% above 1600); this section
repeats every figure above with that same `nCount >= 400` floor added on top of
`confident`, so the two can be compared directly. The grey background (all
profiled cells, any depth) is left unfiltered in both, so only the colored
overlay differs — the tissue silhouette is a fixed reference, not part of the
comparison. Files carry a `_gte400` suffix.

```{r}
deep_min_counts = 400

highlight_gte400 = highlight %>% filter(nCount >= deep_min_counts)
pre_highlight_gte400 = pre_highlight %>% filter(nCount >= deep_min_counts)

highlight_gte400 %>% count(first_type) %>% arrange(desc(n))
pre_highlight_gte400 %>% count(first_type)
```

### Aggregate (gte400)

```{r}
panels = map(section_order, section_plot, hl = highlight_gte400)
grid = cowplot::plot_grid(plotlist = panels, ncol = 5)
p_agg_deep = cowplot::plot_grid(grid, legend, ncol = 2, rel_widths = c(1, 0.12))

ggsave(file.path(out_dir, "spatial_song_types_aggregate_gte400.png"), p_agg_deep,
       width = 22, height = 9, dpi = 250, bg = "white")
p_agg_deep
```

### Individual — one type per figure (gte400)

```{r}
for (ty in song_types) {
  hl = highlight_gte400 %>% filter(first_type == ty)
  panels = map(section_order, section_plot, hl = hl, point_colour = type_colors[[ty]])
  p = cowplot::plot_grid(plotlist = panels, ncol = 5) +
    cowplot::draw_figure_label(paste0(ty, "  (n = ", nrow(hl), " confident cells, nCount >= ",
                                      deep_min_counts, ")"),
                               position = "top", size = 14)

  fname = paste0("spatial_", str_replace_all(ty, "-", "_"), "_gte400.png")
  ggsave(file.path(out_dir, fname), p, width = 20, height = 9.5, dpi = 250, bg = "white")
}
```

### Song vs. non-song comparator pairs (gte400)

```{r}
for (i in seq_len(nrow(pair_list))) {
  song_ty = pair_list$song[i]; nonsong_ty = pair_list$nonsong[i]
  panels = map(section_order, pair_section_plot, song_ty = song_ty, nonsong_ty = nonsong_ty,
              hl_data = highlight_gte400)
  p = cowplot::plot_grid(plotlist = panels, ncol = 5) +
    cowplot::draw_figure_label(
      paste0(song_ty, " (red)  vs.  ", nonsong_ty, " (blue) — nCount >= ", deep_min_counts),
      position = "top", size = 14)

  fname = paste0("spatial_pair_", str_replace_all(song_ty, "-", "_"),
                 "_vs_", str_replace_all(nonsong_ty, "-", "_"), "_gte400.png")
  ggsave(file.path(out_dir, fname), p, width = 20, height = 9.5, dpi = 250, bg = "white")
}
```

### Individual — every cell call in the taxonomy (gte400)

```{r}
all_confident_gte400 = d %>% filter(confident, nCount >= deep_min_counts)

for (ty in all_types) {
  hl = all_confident_gte400 %>% filter(first_type == ty)
  panels = map(section_order, section_plot, hl = hl, point_colour = single_type_colour)
  p = cowplot::plot_grid(plotlist = panels, ncol = 5) +
    cowplot::draw_figure_label(paste0(ty, "  (n = ", nrow(hl), " confident cells, nCount >= ",
                                      deep_min_counts, ")"),
                               position = "top", size = 14)

  safe_name = str_replace_all(ty, "[/-]", "_")
  ggsave(file.path(out_dir, paste0("spatial_all_", safe_name, "_gte400.png")), p,
         width = 20, height = 9.5, dpi = 250, bg = "white")
}

list.files(out_dir, pattern = "_gte400\\.png$") %>% length()
```

### Aggregate — Glut-Pre-1/2/3 (gte400)

```{r}
panels = map(section_order, pre_section_plot, hl_data = pre_highlight_gte400)
grid = cowplot::plot_grid(plotlist = panels, ncol = 5)
p_pre_deep = cowplot::plot_grid(grid, legend, ncol = 2, rel_widths = c(1, 0.12))

ggsave(file.path(out_dir, "spatial_glut_pre_aggregate_gte400.png"), p_pre_deep,
       width = 22, height = 9, dpi = 250, bg = "white")
p_pre_deep
```

```{r}
sessionInfo()
```
