---
title: "Xenium label transfer — snRNA-seq cell types in space"
format: html
editor: source
---

Transfers the snRNA-seq cell type taxonomy onto the ten Xenium sections, at the
label resolution established in `panel_resolution_benchmark.qmd`.

## Two analysis tiers — read this before using the labels

The transfer supports localization strongly and composition only under a filter.
Both statements are measured, not cautionary boilerplate.

**Tier 1 — localization ("where is cell type X?"). Use all confident cells.**
Glut-HVC-1 is the most HVC-enriched type in the whole taxonomy: 13.4% of cells
inside the HVC selections versus 0.09% outside, log2 7.11 (138x), reproducing
independently in all five sections with paired ROIs across both birds (log2 6.28
to 10.08). Glut-NC-1, the comparator, is *depleted*: literally zero cells inside
HVC in every section against 0.5-5.3% in the surround. Nothing in the transfer
uses position, so this is an independent anatomical validation.

**Tier 2 — composition ("what fraction of region R is type X?"). Requires
`nCount_Xenium >= composition_min_counts`.** Unfiltered composition is wrong:
Astro 42%, Endo 21%, Glut 18% across confident cells, where glutamatergic
neurons should be ~40-50% of telencephalic cells. That is almost entirely a
depth artifact — cells with little signal have flat profiles and collapse onto
the flattest reference profiles (Astro-2 and Endo have the lowest median panel
UMI in the reference, 111 and 157):

| counts/cell | Astro | Endo | Glut | GABA |
|---|---|---|---|---|
| <100        | 33.1% | 49.2% |  2.3% |  1.7% |
| 200-400     | 45.3% | 27.5% |  9.8% |  5.9% |
| 400-800     | 43.2% |  9.0% | 25.3% | 13.7% |
| 800-1600    | 29.1% |  1.8% | 50.2% | 15.2% |
| >1600       | 14.9% |  0.5% | 79.7% |  3.3% |

At 800-1600 counts the composition is biologically plausible. The filter trades
cells for a trustworthy denominator; 400 is the chosen operating point because
it keeps 46% of confident cells while cutting Endo from 21% to 7.5%, and gives
the highest HVC enrichment of any threshold tested (log2 7.38). Enrichment
itself is stable from 0 to 800 counts (7.11-7.38), which is why Tier 1 does not
need the filter.

See `composition_by_depth.csv` and `depth_filter_tradeoff.csv`.

Two methods are run. **RCTD** (`spacexr`) is the primary: it models each cell as
a mixture of reference profiles, which matters because this run was segmented by
5 µm nucleus expansion with no boundary stain, so transcript spillover between
neighbouring cells is expected. **Seurat transfer anchors** is the cross-check.
Cells where the two agree form the high-confidence set; disagreement is reported
rather than resolved, and is itself the uncertainty map.

The benchmark measured what each method recovers on held-out reference cells
(macro-F1 0.86 naive Bayes, 0.81 anchors), so those numbers are the ceiling
against which the transfer here should be read.

RCTD is expensive at this cell count, so it runs per section and checkpoints:
re-running skips sections already on disk.

```{r}
library(tidyverse)
library(Seurat)
library(BPCells)
library(qs2)
library(Matrix)
library(spacexr)
library(cowplot)
library(here)

source(here::here("config/paths.R"))
select = dplyr::select
```

```{r}
script_name = "xenium_label_transfer"
out_dir = here::here("xenium/label_transfer", script_name)
obj_dir = file.path(path.expand(XENIUM_PROC_DIR), script_name)
ckpt_dir = file.path(obj_dir, "rctd_by_section")
dir.create(out_dir, recursive = T, showWarnings = F)
dir.create(ckpt_dir, recursive = T, showWarnings = F)

xenium_obj_fname = file.path(path.expand(XENIUM_PROC_DIR),
                             "xenium_preprocess", "obj_xenium.qs2")
ref_obj_fname = file.path(SNRNA_SEURAT_DIR, "obj_clustered.qs2")
label_set_fname = here::here("xenium/label_transfer/panel_resolution_benchmark",
                             "recommended_label_set.csv")

## RAW, not the CellBender-corrected RNA assay. CellBender strips ambient RNA
## from the reference, making each type's profile far sharper than a Xenium cell
## — which is segmented by 5 um nucleus expansion and so collects neighbouring
## transcripts. Every method then assigns the blurred query cells to whichever
## reference profile is itself blurred (Glut-Nido-3). Measured on OR52YW26_2_2:
## CellBender used 17 of 49 labels, called ZERO Glut-HVC-1 cells, and gave no
## HVC ROI enrichment; RAW used 41 labels and put Glut-HVC-1 at 30x enrichment
## inside the HVC selection. See reference_variant_test.csv.
assay_ref = "RAW"
label_col = "cluster"

## Reference cells per label. Caps the cost of both methods and keeps the
## reference from being dominated by the most abundant types.
max_ref_cells_per_label = 1000

## Glut-Nido-3 is excluded from the reference. It behaves as a sink: with it in,
## the transfer calls 32 Glut-HVC-1 cells in OR52YW26_2_2 at 30x HVC ROI
## enrichment; without it, 216 cells at 90x. It is anomalous in the reference
## too — 23% of its cells co-express SLC17A6 and GAD1 even in the LOWEST library
## size quartile, against a 0.9% baseline at that depth.
## Glut-Nido-2 was tested and is NOT excluded: removing it on top of Nido-3
## changes nothing (HVC fraction 0.1543 -> 0.1566). Its apparent co-expression
## is a depth artifact — see nido_removal_test.csv.
exclude_labels = c("Glut-Nido-3")

## RCTD. UMI_min matches the Xenium QC threshold from preprocessing rather than
## the package default of 100, which would discard a large share of real cells.
run_rctd = TRUE
rctd_umi_min = 20
rctd_doublet_mode = "doublet"

## Cores are set per stage, which matters a great deal here. choose_sigma_c
## fits on at most N_fit = 1000 cells (spacexr default) for N_epoch = 8 epochs,
## so it is a fixed cost — but spacexr spawns a fresh PSOCK cluster per epoch,
## and with 44 workers each re-loading the renv library on startup, the startup
## overhead swamps the actual work: ~11 min/epoch to decompose 1000 cells.
## Few cores for setup, many for fitPixels, where the work actually scales.
## Capped at 22 rather than all 48 — this machine is shared.
rctd_setup_cores = 4
rctd_max_cores = 22

## Cells per fitPixels checkpoint. Only fitPixels scales with cell count; the
## setup before it is a fixed cost paid once for all ten sections.
rctd_chunk_size = 2e4

run_anchors = TRUE
npcs = 30

## Depth filter for COMPOSITION analyses only (see "Two analysis tiers" below).
## Cell type calls below this transcript count are driven by depth rather than
## identity: Endo is 49% of calls under 100 counts and 0.5% above 1600.
composition_min_counts = 400

seed = 10
set.seed(seed)
options(future.globals.maxSize = Inf)
```

## Reference

```{r}
label_set = read_csv(label_set_fname, show_col_types = FALSE)
label_set %>% count(status)

## Unresolved labels stay in the reference — dropping them would force their
## cells to be called something else — but are flagged so they can be excluded
## from interpretation downstream.
unresolved_labels = label_set %>% filter(status == "unresolved") %>% pull(recommended_label)
unresolved_labels
```

```{r}
obj_ref = qs2::qs_read(ref_obj_fname)
cells = Cells(obj_ref)[!grepl("/", obj_ref$assignment)]
obj_ref = subset(obj_ref, cells = cells)

panel = read_csv(XENIUM_PANEL_CSV, show_col_types = FALSE)$Gene
counts_ref_all = LayerData(obj_ref, assay = assay_ref, layer = "counts")
panel_shared = intersect(panel, rownames(counts_ref_all))
length(panel_shared)
```

```{r}
ref_md = obj_ref@meta.data %>%
  rownames_to_column("cell") %>%
  mutate(cluster = as.character(.data[[label_col]])) %>%
  left_join(label_set %>% select(cluster, recommended_label), by = "cluster") %>%
  filter(!is.na(recommended_label), !recommended_label %in% exclude_labels)

set.seed(seed)
ref_md = ref_md %>%
  group_by(recommended_label) %>%
  slice_sample(n = max_ref_cells_per_label) %>%
  ungroup()

counts_ref = as(counts_ref_all[panel_shared, ref_md$cell], "dgCMatrix")
labels_ref = setNames(factor(ref_md$recommended_label), ref_md$cell)

## RCTD rejects "/" in cell type names, which the merged Glut-NC/Glut-Nido label
## contains. Substitute for RCTD and map back afterwards.
labels_rctd = labels_ref
levels(labels_rctd) = str_replace_all(levels(labels_rctd), "/", "_")
rctd_name_map = setNames(levels(labels_ref), levels(labels_rctd))

rm(counts_ref_all, obj_ref); gc()

tibble(n_cells = ncol(counts_ref), n_labels = nlevels(labels_ref),
       n_genes = nrow(counts_ref))
table(labels_ref) %>% sort()
```

## Query

```{r}
obj = qs2::qs_read(xenium_obj_fname)
counts_q_all = LayerData(obj, assay = "Xenium", layer = "counts")
stopifnot(all(panel_shared %in% rownames(counts_q_all)))

section_ids = levels(obj$section_id)
section_ids
```

## RCTD

RCTD's cost is dominated by a fixed setup — gene selection in `create.RCTD`,
then `fitBulk` and the iterative `choose_sigma_c` — that does not scale with
cell count. Only `fitPixels` scales. So all ten sections are run as **one** RCTD
object, paying the setup once, and resumability comes from chunking `fitPixels`
instead: `run.RCTD` is unrolled into its three stages, the fitted setup is
checkpointed, and each chunk of cells is fitted and saved independently.

Doublet-mode assignment does not use spatial coordinates — they are only carried
along — so combining sections costs nothing scientifically. Coordinates are
still offset per section so the stored geometry stays sane if anything plots
from the RCTD object directly.

```{r}
setup_ckpt = file.path(obj_dir, "rctd_setup.qs2")

if (run_rctd) {
  ## min_UMI defaults to 100, which on a 425-gene subset would discard a large
  ## share of otherwise good reference cells
  quantile(colSums(counts_ref), c(0, 0.05, 0.25, 0.5))

  cells_all = colnames(obj)

  ## Offset each section along x so sections do not overlap in coordinate space
  offsets = obj@meta.data %>%
    group_by(section_id) %>%
    summarize(w = max(x_centroid, na.rm = TRUE) -
                  min(x_centroid, na.rm = TRUE), .groups = "drop") %>%
    mutate(x_offset = lag(cumsum(w * 1.1), default = 0))

  coords = obj@meta.data[cells_all, c("section_id", "x_centroid", "y_centroid")] %>%
    rownames_to_column("cell") %>%
    left_join(offsets, by = "section_id") %>%
    transmute(x = x_centroid + x_offset, y = y_centroid) %>%
    as.data.frame()
  rownames(coords) = cells_all
}
```

```{r}
if (run_rctd && !file.exists(setup_ckpt)) {
  ref_rctd = Reference(counts_ref, labels_rctd, colSums(counts_ref),
                       min_UMI = rctd_umi_min,
                       n_max_cells = max_ref_cells_per_label)

  cq = counts_q_all[panel_shared, cells_all]
  sp = SpatialRNA(coords, cq, colSums(cq))

  r = create.RCTD(sp, ref_rctd, max_cores = rctd_setup_cores,
                  UMI_min = rctd_umi_min, counts_MIN = rctd_umi_min)

  ## run.RCTD unrolled: the first two stages are the fixed cost, paid once, and
  ## are run on few cores because they are dominated by worker startup
  r@config$RCTDmode = rctd_doublet_mode
  r = fitBulk(r)
  r = choose_sigma_c(r)

  ## fitPixels is the stage that scales, so give it every core
  r@config$max_cores = rctd_max_cores

  qs2::qs_save(r, setup_ckpt)
  rm(cq, sp); gc()
}
```

```{r}
if (run_rctd) {
  r = qs2::qs_read(setup_ckpt)

  fit_cells = colnames(r@spatialRNA@counts)
  chunks = split(fit_cells, ceiling(seq_along(fit_cells) / rctd_chunk_size))
  length(chunks)

  for (i in seq_along(chunks)) {
    ckpt = file.path(ckpt_dir, sprintf("rctd_chunk_%03d.qs2", i))
    if (file.exists(ckpt)) { message("skipping chunk ", i, " (checkpoint exists)"); next }

    message("RCTD fitPixels: chunk ", i, " of ", length(chunks))
    rc = r
    rc@spatialRNA = restrict_puck(r@spatialRNA, chunks[[i]])
    rc = fitPixels(rc, doublet_mode = rctd_doublet_mode)

    ## Map the RCTD-safe names back to the real label names
    res = rc@results$results_df %>%
      rownames_to_column("cell") %>%
      as_tibble() %>%
      mutate(across(c(first_type, second_type),
                    ~ unname(rctd_name_map[as.character(.x)])))
    qs2::qs_save(res, ckpt)
    rm(rc); gc()
  }
  rm(r); gc()
}
```

```{r}
rctd_res = list.files(ckpt_dir, pattern = "^rctd_chunk_.*qs2$", full.names = TRUE) %>%
  map_dfr(qs2::qs_read) %>%
  left_join(obj@meta.data %>% rownames_to_column("cell") %>%
              select(cell, section_id), by = "cell")

rctd_res %>% count(section_id, spot_class) %>%
  pivot_wider(names_from = spot_class, values_from = n, values_fill = 0)
```

## Naive Bayes — diagnostic only, does not transfer across platforms

**This method fails on the real data and is not used for labelling.** It is kept
because the failure is informative and because the benchmark result invites
exactly this mistake.

Naive Bayes was the strongest method in `panel_resolution_benchmark.qmd`
(macro-F1 0.86 vs 0.81 for anchors), but that benchmark trained and tested
*within the snRNA domain* — reference cells subset to the panel and thinned to
Xenium depth. It measured how much cell type information the 425 genes carry.
It could not, and did not, measure whether a raw multinomial likelihood survives
the snRNA → Xenium platform gap, and it does not:

- Uncorrected, 62% of all cells are called `Glut-Nido-3` and 30% `Endo`; only 22
  of 49 labels are used, with essentially no GABAergic or song-nucleus cells.
- With a gene-wise platform correction (bulk ratio applied to the reference
  profiles, renormalized per class — without the renormalization the correction
  cancels in the argmax), it fails differently rather than better: 40%
  `GABA-Pre`, and 24 `Glut-HVC-1` cells in 1.59 million.

The reason is visible in the correction itself: the gene-wise Xenium/snRNA bulk
ratio spans **0.001 to 223**. Nuclear snRNA and whole-cell Xenium after 5 µm
nucleus expansion detect genes on wildly different scales, and an unregularized
multinomial likelihood over ~400 transcripts is dominated by that shift rather
than by cell identity.

This is precisely what RCTD's `fitBulk` and sigma estimation exist to absorb,
and why anchor-based transfer — which works in a shared normalized PCA space —
is robust to it. Read as a diagnostic, this quantifies the platform gap the
other two methods have to survive.

```{r}
nb_ckpt = file.path(obj_dir, "nb_predictions.qs2")

nb_fit = function(counts, labels) {
  labels = factor(labels)
  ind = Matrix::sparse.model.matrix(~ 0 + labels)
  colnames(ind) = levels(labels)
  s = as.matrix(counts %*% ind)
  log(sweep(s + 1, 2, colSums(s) + nrow(counts), "/"))
}

if (!file.exists(nb_ckpt)) {
  log_theta = nb_fit(counts_ref, labels_ref)

  cells_all = colnames(obj)
  blocks = split(cells_all, ceiling(seq_along(cells_all) / 2e5))

  nb_res = map_dfr(seq_along(blocks), function(i) {
    message("naive Bayes: block ", i, " of ", length(blocks))
    cq = counts_q_all[panel_shared, blocks[[i]]]
    ll = as.matrix(Matrix::crossprod(cq, log_theta))

    top = max.col(ll, ties.method = "first")
    n = nrow(ll)
    ll_top = ll[cbind(seq_len(n), top)]
    ll2 = ll; ll2[cbind(seq_len(n), top)] = -Inf
    ll_second = ll2[cbind(seq_len(n), max.col(ll2, ties.method = "first"))]

    ## Posterior of the top class, computed in a numerically safe way
    post = 1 / rowSums(exp(ll - ll_top))
    nUMI = Matrix::colSums(cq)

    tibble(cell = blocks[[i]],
           nb_label = colnames(log_theta)[top],
           nb_score = post,
           nb_margin = (ll_top - ll_second) / pmax(nUMI, 1))
  })

  qs2::qs_save(nb_res, nb_ckpt)
} else {
  nb_res = qs2::qs_read(nb_ckpt)
}

nb_res %>% count(nb_label, sort = TRUE) %>% print(n = Inf)
```

```{r}
nb_res %>%
  summarize(median_score = median(nb_score),
            median_margin = median(nb_margin),
            frac_score_gt_0.9 = mean(nb_score > 0.9))
```

## Transfer anchors

```{r}
## Reference and query MUST share a scale factor. NormalizeData computes
## log1p(x/total * scale.factor), so giving each dataset its own median (243 for
## the reference, 419 for Xenium — a 1.72x difference) maps identical relative
## profiles to different values and corrupts the projection into the reference
## PCA. Using per-dataset medians produced a transfer that used 21 of 49 labels
## and called zero Glut-HVC-1 and zero Glut-RA cells in 1.59M. The benchmark
## could not catch this: train and query were the same domain, so the two
## medians were equal.
scale_factor_common = median(colSums(counts_ref))
scale_factor_common

if (run_anchors) {
  ref_seu = CreateSeuratObject(counts = counts_ref)
  ref_seu$label = labels_ref[colnames(ref_seu)]
  ref_seu = NormalizeData(ref_seu, scale.factor = scale_factor_common,
                          verbose = FALSE)
  VariableFeatures(ref_seu) = panel_shared
  ref_seu = ScaleData(ref_seu, features = panel_shared, verbose = FALSE)
  ref_seu = RunPCA(ref_seu, features = panel_shared, npcs = npcs, verbose = FALSE)

  anchor_res = map_dfr(section_ids, function(sid) {
    message("anchors: ", sid)
    cells_sid = colnames(obj)[obj$section_id == sid]
    cq = counts_q_all[panel_shared, cells_sid]

    q = CreateSeuratObject(counts = cq)
    q = NormalizeData(q, scale.factor = scale_factor_common, verbose = FALSE)

    a = FindTransferAnchors(reference = ref_seu, query = q,
                            features = panel_shared,
                            normalization.method = "LogNormalize",
                            reference.reduction = "pca",
                            dims = 1:npcs, k.filter = NA, verbose = FALSE)
    td = TransferData(anchorset = a, refdata = ref_seu$label,
                      dims = 1:npcs, verbose = FALSE)

    ## Guard the score column: if TransferData does not return it, tibble()
    ## would silently drop it rather than error
    score_col = grep("^predicted.id.score$|^prediction.score.max$",
                     colnames(td), value = TRUE)[1]
    stopifnot(!is.na(score_col))

    tibble(cell = cells_sid,
           anchor_label = td$predicted.id,
           anchor_score = td[[score_col]],
           section_id = sid)
  })
}
```

## Consensus

Two votes: RCTD and transfer anchors. Naive Bayes is carried as a column for
diagnostics but **does not vote** — it does not survive the platform gap (see
above). A cell is confident when both methods agree and RCTD calls it a
confident singlet or doublet; otherwise it is left unlabelled rather than
arbitrated, since an unlabelled cell is honest about the ambiguity.

```{r}
transfer = rctd_res %>%
  select(cell, section_id, spot_class, rctd_label = first_type,
         rctd_second = second_type, singlet_score, min_score) %>%
  full_join(anchor_res %>% select(cell, anchor_label, anchor_score), by = "cell") %>%
  left_join(nb_res %>% select(cell, nb_label, nb_margin), by = "cell") %>%
  mutate(across(c(rctd_label, anchor_label, nb_label), as.character),
         agree = !is.na(rctd_label) & !is.na(anchor_label) &
           rctd_label == anchor_label,
         confident = agree & spot_class %in% c("singlet", "doublet_certain"),
         label_final = if_else(confident, rctd_label, NA_character_),
         unresolved = label_final %in% unresolved_labels)

transfer %>%
  summarize(n = n(),
            frac_rctd_called = mean(!is.na(rctd_label)),
            frac_anchor_called = mean(!is.na(anchor_label)),
            frac_agree = mean(agree),
            frac_confident = mean(confident))

## Diagnostic: how far naive Bayes is from the two usable methods
transfer %>%
  filter(!is.na(nb_label), !is.na(rctd_label)) %>%
  summarize(nb_vs_rctd = mean(nb_label == rctd_label),
            nb_vs_anchor = mean(nb_label == anchor_label, na.rm = TRUE))

write_csv(transfer, file.path(out_dir, "transfer_per_cell.csv.gz"))
```

Agreement per label shows where the methods diverge — low agreement means the
label is method-dependent and should not carry an argument on its own.

```{r}
agreement_by_label = transfer %>%
  filter(!is.na(rctd_label)) %>%
  group_by(rctd_label) %>%
  summarize(n = n(), frac_agree = mean(agree), .groups = "drop") %>%
  arrange(frac_agree)

agreement_by_label %>% print(n = Inf)
write_csv(agreement_by_label, file.path(out_dir, "agreement_by_label.csv"))
```

```{r}
obj$label_final = transfer$label_final[match(colnames(obj), transfer$cell)]
obj$rctd_label = transfer$rctd_label[match(colnames(obj), transfer$cell)]
obj$spot_class = transfer$spot_class[match(colnames(obj), transfer$cell)]
obj$transfer_confident = transfer$confident[match(colnames(obj), transfer$cell)]

obj@meta.data %>%
  count(section_id, label_final) %>%
  pivot_wider(names_from = section_id, values_from = n, values_fill = 0) %>%
  print(n = Inf)
```

## Anatomical validation

The collaborator's Xenium Explorer selections list cell IDs per ROI. Three
things about them had to be established empirically rather than taken on trust:

- The `#Selection name` headers frequently disagree with the file names (the
  same mislabelling flagged in `xenium/README.md`), so each ROI is assigned to
  the section that actually contains its cell IDs.
- The `1um_*` files come from a **different segmentation run** (1 µm expansion,
  not the 5 µm nucleus expansion in the delivered outputs). Their cell IDs are
  real but do not exist in this object, so they cannot be joined and are
  excluded — with the match rate reported below rather than dropped silently.
- `HVC&Surround*_1mm_Selection` files are 1 mm boxes that *contain* HVC, so they
  are not the surround. Surround is defined here as box cells minus the HVC
  selection cells for the same section.

```{r}
roi_files = list.files(XENIUM_SELECTIONS_DIR, pattern = "cells_stats.csv$",
                       full.names = TRUE)

cells_by_section = obj@meta.data %>%
  rownames_to_column("cell") %>%
  mutate(cell_id = str_remove(cell, "^.*_")) %>%
  select(cell, cell_id, section_id)

roi = map_dfr(roi_files, function(f) {
  hdr = readLines(f, n = 5)
  skip = max(c(0, which(str_starts(hdr, "#"))))
  d = suppressWarnings(read_csv(f, skip = skip, show_col_types = FALSE))
  if (!"Cell ID" %in% colnames(d)) return(tibble())
  nm = hdr[str_starts(hdr, "#Selection name")][1]
  tibble(file = basename(f),
         declared_name = str_remove(nm, "#Selection name: "),
         cell_id = d$`Cell ID`,
         roi_type = case_when(
           str_starts(basename(f), "1um_")          ~ "alt_segmentation",
           str_starts(basename(f), fixed("HVC&Surround")) ~ "box",
           str_detect(basename(f), "^(orpu|oryw)_HVC_")   ~ "HVC",
           TRUE                                      ~ "other"))
})

## How many of each file's cells exist in this object? Near-zero identifies the
## alternate-segmentation files.
roi_match = roi %>%
  group_by(file, declared_name, roi_type) %>%
  summarize(n_listed = n(),
            n_matched = sum(cell_id %in% cells_by_section$cell_id),
            frac_matched = n_matched / n_listed, .groups = "drop") %>%
  arrange(frac_matched)

roi_match %>% print(n = Inf)
write_csv(roi_match, file.path(out_dir, "roi_match_rates.csv"))
```

```{r}
## Resolve each usable ROI to a section by where its cells actually are
roi_resolved = roi %>%
  filter(roi_type %in% c("box", "HVC")) %>%
  inner_join(cells_by_section, by = "cell_id", relationship = "many-to-many") %>%
  count(file, declared_name, roi_type, section_id) %>%
  group_by(file) %>%
  slice_max(n, n = 1) %>%
  ungroup()

roi_resolved %>% select(file, declared_name, roi_type, section_id, n_cells = n) %>%
  arrange(section_id, roi_type) %>%
  print(n = Inf)
write_csv(roi_resolved, file.path(out_dir, "roi_resolved_sections.csv"))
```

```{r}
roi_cells = roi %>%
  inner_join(roi_resolved %>% select(file, section_id), by = "file") %>%
  inner_join(cells_by_section, by = c("cell_id", "section_id")) %>%
  distinct(cell, section_id, roi_type)

hvc_cells = roi_cells %>% filter(roi_type == "HVC") %>% distinct(cell, section_id)
box_cells = roi_cells %>% filter(roi_type == "box") %>% distinct(cell, section_id)

## Surround = inside the 1 mm box, outside the HVC selection
roi_final = bind_rows(
  hvc_cells %>% mutate(region = "HVC"),
  box_cells %>% anti_join(hvc_cells, by = "cell") %>% mutate(region = "surround"))

## Only sections with both an HVC selection and a box support the comparison
paired_sections = roi_final %>%
  count(section_id, region) %>%
  pivot_wider(names_from = region, values_from = n, values_fill = 0) %>%
  filter(HVC > 0, surround > 0)

paired_sections
roi_final = roi_final %>% semi_join(paired_sections, by = "section_id")
```

```{r}
roi_comp = roi_final %>%
  left_join(transfer %>% select(cell, label_final), by = "cell") %>%
  filter(!is.na(label_final)) %>%
  count(section_id, region, label_final) %>%
  group_by(section_id, region) %>%
  mutate(frac = n / sum(n)) %>%
  ungroup()

write_csv(roi_comp, file.path(out_dir, "roi_composition.csv"))
```

The test: Glut-HVC-1 should be enriched inside the HVC selections relative to
the surround. Nothing in the transfer used position, so this is an independent
check on it.

```{r}
hvc_enrichment = roi_comp %>%
  filter(label_final %in% c("Glut-HVC-1", "Glut-HVC-1a", "Glut-HVC-2",
                            "Glut-NC-1", "Glut-NC-3", "Glut-RA")) %>%
  select(section_id, region, label_final, frac) %>%
  pivot_wider(names_from = region, values_from = frac, values_fill = 0) %>%
  mutate(log2_enrichment = log2((HVC + 1e-4) / (surround + 1e-4)))

hvc_enrichment %>% arrange(desc(log2_enrichment)) %>% print(n = Inf)

ggplot(hvc_enrichment, aes(log2_enrichment, label_final, colour = section_id)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  geom_point(size = 3, alpha = 0.8) +
  labs(x = "log2(HVC / surround) cell type fraction", y = NULL) +
  theme_cowplot()

ggsave(file.path(out_dir, "hvc_roi_enrichment.pdf"), width = 9, height = 5)
write_csv(hvc_enrichment, file.path(out_dir, "hvc_roi_enrichment.csv"))
```

## Spatial maps

```{r}
song_types = c("Glut-HVC-1", "Glut-HVC-1a", "Glut-HVC-2", "Glut-RA",
               "Glut-LMAN-1", "Glut-LMAN-2", "Glut-NC-1")

for (sid in section_ids) {
  p = ImageDimPlot(obj, fov = sid, group.by = "label_final",
                   cells = colnames(obj)[obj$section_id == sid &
                                           obj$label_final %in% song_types],
                   size = 1, cols = "polychrome", dark.background = FALSE) +
    ggtitle(paste(sid, "- song system types"))
  ggsave(file.path(out_dir, paste0("spatial_song_types_", sid, ".png")),
         p, width = 10, height = 10, dpi = 200, bg = "white")
}
```

```{r}
for (sid in section_ids) {
  p = ImageDimPlot(obj, fov = sid, group.by = "label_final",
                   size = 0.3, cols = "polychrome", dark.background = FALSE) +
    ggtitle(sid)
  ggsave(file.path(out_dir, paste0("spatial_all_labels_", sid, ".png")),
         p, width = 12, height = 10, dpi = 200, bg = "white")
}
```

## Save

```{r}
qs2::qs_save(obj, file.path(obj_dir, "obj_xenium_labeled.qs2"))
sessionInfo()
```
