---
title: "MAFB EMX2 OE Chicken Analysis - Run 1"
format: html
editor: source
---

```{r setup}
library(tidyverse)
library(Seurat)
library(scCustomize)
library(rtracklayer)
library(cowplot)
library(GenomicRanges)
library(Rsamtools)
library(ComplexHeatmap)
library(qs)
library(Nebulosa)
library(RRHO2)
library(parallel)
# library(extrafont)

options(future.globals.maxSize = Inf)
 
source("../scripts/scRNA.R")
source(here::here("config/paths.R"))
source("../scripts/gene_lists.R")
source("../scripts/scenicplus.R")

plot_theme <- function() {
  theme(
    text             = element_text(family = "sans"),
    axis.text        = element_text(size = 5),
    axis.title       = element_text(size = 5), 
    legend.text      = element_text(size = 5),
    legend.title     = element_text(size = 5),
    legend.key.size  = unit(3, "pt"),
    plot.title       = element_text(size = 6),
    strip.text       = element_text(size = 5)
  )
}

umap_theme <- function() {
  plot_theme() +
  theme(
    axis.text        = element_blank(),
    axis.ticks       = element_blank(),
    axis.title       = element_blank(),
    axis.line        = element_blank()
  )
}

set.seed(10)
```

```{r setup-paths}
# Define input/output paths

samples <- c("MAFBn", "MAFBp", "EMX2p")
data_dirs <- setNames(
  c(
    file.path(PIPSEQ_MAFB_EMX2_DIR, "bGalGal1.mat.mafb_v2", "MAFBn"),
    file.path(PIPSEQ_MAFB_EMX2_DIR, "bGalGal1.mat.mafb_v2", "MAFBp"),
    file.path(PIPSEQ_MAFB_EMX2_DIR, "bGalGal1.mat.emx2",    "EMX2p")
  ),
  samples
)

in_files <- map(data_dirs, function(x) {
  list.files(x, pattern = "scRNA\\.filtered", full.names = TRUE)
})

out_dir <- here::here("overexpression/chick_mafb-emx2_run1")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

obj_fname <- file.path(out_dir, "obj.qs")
obj_filter_fname <- file.path(out_dir, "obj_filter.qs")
```

## Gene Annotation

```{r load-annotation}
load_gene_annotation <- function(gtf_path) {
  annotation <- import(gtf_path)
  annotation$gene_biotype <- if_else(is.na(annotation$gene_biotype), "", annotation$gene_biotype)
  annotation$gene_name <- annotation$gene
  annotation$gene_id <- str_replace(annotation$gene_id, "_", "-")
  
  # Extract mitochondrial genes
  annotation_mt <- annotation[seqnames(annotation) == "chrMT", ]
  genes_mt <- na.omit(annotation_mt$Name)
  
  list(
    annotation = annotation,
    annotation_mt = annotation_mt,
    genes_mt = genes_mt
  )
}

annot_data <- load_gene_annotation(
  GTF_CHICKEN_MAFB_EMX2
)

# Check for transgenes
as.data.frame(annot_data$annotation) %>% 
  filter(grepl("plasmid", Name))
```

## Process snRNA-seq Data

```{r qc-filtering}
apply_qc_filters <- function(obj, mito_features, max_mito = 2, 
                             max_count = 20000, min_count = 1000) {
  obj <- PercentageFeatureSet(obj, features = mito_features, col.name = "perc_mito")
  
  # Plot before filtering
  print(VlnPlot_scCustom(obj, c("perc_mito", "nCount_RNA"), pt.size = 0))
  
  # Apply filters
  obj_filt <- subset(
    obj,
    subset = perc_mito < max_mito & 
             nCount_RNA < max_count & 
             nCount_RNA > min_count
  )
  
  # Plot after filtering
  gg <- VlnPlot_scCustom(
    obj_filt, 
    features = c("nCount_RNA", "perc_mito"),
    pt.size = 0
  )
  print(gg)
  
  return(obj_filt)
}
```

```{r process-seurat}
redo <- T

if (redo) {
  # Load and merge data
  mats <- map(in_files, ~ReadMtx(mtx = .x[3], cells = .x[1], features = .x[2]))

  # Pad matrices to the union of all rownames so transgene rows present in one
  # condition (e.g. MAFB-tg, EMX2-tg) are represented as zeros in the other
  all_genes <- Reduce(union, map(mats, rownames))
  mats <- map(mats, function(mat) {
    missing <- setdiff(all_genes, rownames(mat))
    if (length(missing) == 0) return(mat)
    zero_rows <- Matrix::Matrix(0L, nrow = length(missing), ncol = ncol(mat),
                                sparse = TRUE,
                                dimnames = list(missing, colnames(mat)))
    rbind(mat, zero_rows)[all_genes, ]
  })

  objs <- imap(mats, function(mat, sample_name) {
    obj_tmp <- CreateSeuratObject(mat)
    obj_tmp$sample <- sample_name
    obj_tmp
  })
  
  obj <- Merge_Seurat_List(objs, add.cell.ids = names(in_files))
  
  # QC filtering
  mito_features <- intersect(rownames(obj), annot_data$annotation_mt$gene_id)
  obj <- apply_qc_filters(obj, mito_features, max_count = 20000, min_count = 100)
  
  # Process RNA - standard and harmony
  obj <- process_rna(
    obj, 
    cluster_prefix = "cluster", 
    rna_dims_list = 30, 
    run_sct = TRUE, 
    run_parallel = FALSE
  )
  
  obj <- process_rna(
    obj, 
    cluster_prefix = "cluster", 
    rna_dims_list = 30, 
    run_sct = TRUE, 
    run_parallel = FALSE, 
    use_harmony = TRUE
  )
  
  qsave(obj, obj_fname)
} else {
  obj <- qread(obj_fname)
}
```

## Visualization

```{r plot-functions}
plot_umap_by_category <- function(obj, categories, reduction, out_dir) {
  walk(categories, function(cat) {
    # Standard plot
    gg <- DimPlot(obj, group.by = cat, reduction = reduction) +
      coord_equal() + umap_theme()
    print(gg)
    save_plot(
      file.path(out_dir, sprintf("umap_by_%s_%s.pdf", reduction, cat)),
      gg, base_height = 5, base_width = 5
    )

    # Split plot
    gg_split <- DimPlot(obj, group.by = cat, split.by = cat, reduction = reduction) +
      coord_equal() + umap_theme()
    print(gg_split)
    save_plot(
      file.path(out_dir, sprintf("umap_split-by_%s_%s.pdf", reduction, cat)),
      gg_split, base_height = 5, base_width = 5
    )
  })
}

plot_clusters <- function(obj, cluster_pattern, reduction, out_dir,
                         exclude_pattern = NULL) {
  cluster_cols <- grep(cluster_pattern, colnames(obj@meta.data), value = TRUE)

  if (!is.null(exclude_pattern)) {
    cluster_cols <- grep(exclude_pattern, cluster_cols, value = TRUE, invert = TRUE)
  }

  walk(cluster_cols, function(res) {
    gg <- DimPlot(obj, group.by = res, label = TRUE, reduction = reduction) +
      labs(title = res) + umap_theme()
    print(gg)
    save_plot(file.path(out_dir, sprintf("umap_%s.pdf", res)), gg,
              base_height = 5, base_width = 5)
  })
}

plot_features <- function(obj, features, reduction, out_dir) {
  features_present <- intersect(features, rownames(obj))
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  walk(features_present, function(feat) {
    gg <- FeaturePlot_scCustom(obj, features = feat, reduction = reduction) +
      coord_equal() + umap_theme()
    print(gg)
    save_plot(file.path(out_dir, sprintf("umap_%s.pdf", feat)), gg,
              base_height = 5, base_width = 5)
  })
}

plot_density_features <- function(obj, features, reduction, out_dir) {
  features_present <- intersect(features, rownames(obj))
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  walk(features_present, function(feat) {
    gg <- plot_density(obj, features = feat, reduction = reduction, size = 0.1) +
      coord_equal() + umap_theme()
    print(gg)
    save_plot(file.path(out_dir, sprintf("umap_density_%s.pdf", feat)), gg,
              base_height = 5, base_width = 5)
  })
}
```

```{r generate-plots}
redo_plot <- TRUE
reduction <- "umap_rna_30_mindist0.3nn30"

if (redo_plot) {
  # Plot by sample
  plot_umap_by_category(obj, c("sample"), reduction, out_dir)
  
  # Plot clusters
  plot_clusters(obj, "cluster", reduction, out_dir, exclude_pattern = "harmony")
  
  # Plot features
  features <- c(
    unlist(ct_markers),
    "FOXP2", "NPY", "PVALB", "PENK", "LHX8",
    "NXPH1", "DACH2", "NR2F2", "LHX9", "NR4A2", "MAFB", "EMX2", "mScarlet", "EMX2-tg", "MAFB-tg", "mNeonGreen"
  )
  plot_features(obj, features, reduction, out_dir)
  #plot_density_features(obj, features, reduction, file.path(out_dir, "density"))
}
```

```{r}
# Dot plot

features <- c(
  unlist(ct_markers), 
  "FOXP2", "NPY", "PVALB", "PENK", "LHX8", 
  "NXPH1", "DACH2", "NR2F2", "LHX9", "NR4A2", "MAFB", "EMX2", "mScarlet", "EMX2-tg", "MAFB-tg", "mNeonGreen"
)

gg <- DotPlot_scCustom(
  obj,
  features = intersect(features, rownames(obj)),
  group.by = "cluster_pca_30_0.5",
  flip_axes = TRUE
) + plot_theme()
print(gg)
save_plot(file.path(out_dir, "dotplot_cluster.pdf"), gg,
          base_height = 1.5, base_width = 1.5)
```

```{r}
gg = FeaturePlot_scCustom(obj, c("MAFB-tg"), split.by="sample") + umap_theme()
print(gg)
save_plot(file.path(out_dir, "umap_MAFB-tg_split-sample.pdf"), gg,
          base_height = 5, base_width = 5)
```

```{r}
# Plot features
features <- c(
  "RCAN2", "M1AP", "MAFB-tg", "DOLK", "GSDMA", "SLC13A1","FKBP9", "CDHR2", "MGAT4C", "CACNA1C"
)
plot_features(obj, features, reduction, out_dir)

walk(features, function(feat) {
  gg = FeaturePlot_scCustom(obj, feat, split.by="sample") + umap_theme()
  print(gg)
  save_plot(file.path(out_dir, sprintf("umap_%s_split-sample.pdf", feat)), gg,
            base_height = 5, base_width = 5)
})
```

## Determine number of transgene reads and cells

```{r}
obj <- JoinLayers(obj, assay="RNA")
dat <- GetAssayData(obj, assay = "RNA", layer = "counts")
num_reads = sum(dat["MAFB-tg",])
num_cells = sum(dat["MAFB-tg",]>=1)
perc_cells = round(num_cells / length(Cells(obj)) * 100, 2)
message(sprintf("# MAFB-tg reads: %s", num_reads))
message(sprintf("# MAFB-tg+ cells: %s (%s %% of total)", num_cells, perc_cells))

mafb_pos_cells = dat["MAFB-tg",]>=1
emx2_pos_cells = dat["EMX2-tg",]>=1
obj$tg_pos = case_when(mafb_pos_cells ~ "MAFB",
                       emx2_pos_cells ~ "EMX2",
                       .default="none")
```

```{r}
res_to_use = "cluster_pca_30_0.5"
idents_comb = c(`0`="Glut-1",
                `1`="PC-1",
                `2`="PC-2",
                `3`="PC-3",
                `4`="GABA-1",
                `5`="Glut-2",
                `6`="GABA-2",
                `7`="Epen",
                `8`="GABA-3",
                `9`="Endo",
                `10`="PC-4",
                `11`="GABA-4",
                `12`="OPC",
                `13`="Micro"
                )

out_cluster_cur = str_interp("celltype")
out_cluster_cur_sym = as.symbol(out_cluster_cur)
obj@meta.data[,out_cluster_cur] <- plyr::mapvalues(x = obj@meta.data[,res_to_use], from = names(idents_comb), to = idents_comb)
```

```{r}
plot_umap_by_category(obj, c("celltype"), reduction, out_dir)
```

```{r add-variables}
obj$celltype_tg_pos = paste(obj$celltype, obj$tg_pos, sep="_")
celltype_vec = as.character(obj$celltype)
obj$cell_class = case_when(grepl("^PC", celltype_vec) ~ "PC",
                           grepl("Glut", celltype_vec) ~ "Glut",
                           grepl("GABA", celltype_vec) ~ "GABA",
                           .default=celltype_vec)
obj$cell_class_tg_pos = paste(obj$cell_class, obj$tg_pos, sep="_")
obj$cell_class_sample_tg_pos = paste(obj$cell_class, obj$sample, obj$tg_pos, sep="_")
```

## DEG

```{r}
# ortho_fname = OMA_ORTHOLOGS_ZF_CK
# oma = read_delim(ortho_fname)
```


```{r}
markers_pos_fname = file.path(out_dir, "markers_pos.rds")

tg_genes = c("MAFB", "EMX2")
cell_classes = c("Glut", "GABA", "PC")
redo = T
if (redo) {
  obj = PrepSCTFindMarkers(obj, assay = "SCT", verbose = TRUE)
  
  features_to_test = grep("^ENS", Features(obj), invert = T, value=T )

  markers_pos = map(cell_classes, function(cc_cur) {
    map(tg_genes, function(tg_gene) {
      markers_cur = FindMarkers(obj, features = features_to_test,  logfc.threshold = .1,
                                ident.1 = sprintf("%s_%s", cc_cur, tg_gene),
                                ident.2 = sprintf("%s_none", cc_cur),
                                group.by = "cell_class_tg_pos", recorrect_umi=FALSE)
      markers_cur$gene = rownames(markers_cur)
      markers_cur
    }) %>% set_names(tg_genes) %>% bind_rows(.id="tg")
  }) %>% set_names(cell_classes) %>% bind_rows(.id="cell_class")
  
  saveRDS(markers_pos, markers_pos_fname)
} else {
  markers_pos = readRDS(markers_pos_fname)
}
```

## Load scenicplus results

<!-- convert_category_to_factor and extract_matrix are defined in scenicplus.R -->

```{r}
# Helper: load eRegulon data from a SCENIC+ config directory
load_ereg_data <- function(config_dir) {
  ereg_file = file.path(config_dir, "outs", "scplusmdata.h5mu")
  cis_types = c("direct", "extended")

  ereg = map(cis_types, \(cis_type) {
    rhdf5::h5read(ereg_file, name = str_interp("/uns/${cis_type}_e_regulon_metadata")) %>%
      do.call(cbind.data.frame, .)
  }) %>% bind_rows() %>% mutate(class = "Glut")

  gene_auc = map(cis_types, \(cis_type) {
    extract_matrix(str_interp("/mod/${cis_type}_gene_based_AUC"), ereg_file)
  }) %>% bind_cols() %>% rownames_to_column("_index")

  region_auc = map(cis_types, \(cis_type) {
    extract_matrix(str_interp("/mod/${cis_type}_region_based_AUC"), ereg_file)
  }) %>% bind_cols() %>% rownames_to_column("_index")

  group_set = rhdf5::h5read(ereg_file, name = str_interp("/mod/scRNA_counts/obs"))
  group_set = group_set[c("_index", "cluster")]
  factor_ind = which(map_lgl(group_set, ~length(.x)==2))
  for (fi in factor_ind) {
    group_set[[fi]] = convert_category_to_factor(group_set[[fi]])
  }
  group_df = cbind.data.frame(group_set)

  labels_order = levels(group_df$cluster)
  labels_order = c(sort(grep("HVC", labels_order, value=T)),
                   sort(grep("NC", labels_order, value=T)),
                   sort(grep("Nido", labels_order, value=T)),
                   sort(grep("RA", labels_order, value=T)),
                   sort(grep("Arco", labels_order, value=T)),
                   sort(grep("Glut-Pre", labels_order, value=T), decreasing = T),
                   sort(grep("GABA", labels_order, value=T)),
                   sort(grep("GABA-Pre", labels_order, value=T)),
                   sort(grep("Astro", labels_order, value=T)),
                   sort(grep("Epen", labels_order, value=T)),
                   sort(grep("Oligo", labels_order, value=T)),
                   sort(grep("OPC", labels_order, value=T)),
                   sort(grep("Micro", labels_order, value=T)),
                   sort(grep("Endo", labels_order, value=T)))
  labels_order = unique(labels_order)

  group_df = group_df %>% mutate(cluster = factor(cluster, levels=labels_order))
  gene_auc   = gene_auc   %>% left_join(group_df) %>% filter(!is.na(cluster))
  region_auc = region_auc %>% left_join(group_df) %>% filter(!is.na(cluster))

  ereg = ereg %>%
    mutate(importance_x_abs_rho_TF2G = importance_TF2G * abs(rho_TF2G),
           importance_x_rho_TF2G = importance_TF2G * rho_TF2G) %>%
    mutate(importance_x_rho_log_TF2G = log(importance_x_rho_TF2G),
           importance_x_abs_rho_log_TF2G = log(importance_x_abs_rho_TF2G),
           importance_log_TF2G = log(importance_TF2G))

  ereg_sizes = ereg %>%
    mutate(size_genes = as.integer(str_extract(Gene_signature_name, "\\(([0-9]+)g\\)", group=T))) %>%
    distinct(size_genes, TF, eRegulon_name)

  ereg_un = ereg %>% distinct(Gene, importance_TF2G, importance_x_rho_TF2G, rho_TF2G, TF, eRegulon_name)

  list(ereg = ereg, ereg_sizes = ereg_sizes, ereg_un = ereg_un,
       gene_auc = gene_auc, region_auc = region_auc, group_df = group_df)
}

# Load separate configs for MAFB and EMX2
ereg_mafb = load_ereg_data(SCENICPLUS_MAFB_CONFIG_DIR)
ereg_emx2 = load_ereg_data(SCENICPLUS_EMX2_CONFIG_DIR)
```

```{r}
eregulon_n = 20
mafb_pattern <- "MAFB_.*_\\+\\/\\+"
emx2_pattern = "EMX2_.*_\\+\\/\\+"

walk(tg_genes, function(tg_cur) {
  out_dir_tg = file.path(out_dir, tg_cur)
  dir.create(out_dir_tg, showWarnings = FALSE)

  # Select config data for this TG
  ereg_data   = if (tg_cur == "MAFB") ereg_mafb else ereg_emx2
  ereg_un     = ereg_data$ereg_un
  ereg_sizes  = ereg_data$ereg_sizes

  markers_pos_tg = markers_pos %>% filter(tg == tg_cur) %>%
    filter(cell_class == "Glut")

  top_ereg_tfs <- ereg_sizes %>%
    filter(grepl("\\+/\\+", eRegulon_name), !TF %in% c("MAFB", "EMX2")) %>%
    group_by(TF) %>%
    summarise(size_genes = max(size_genes, na.rm = TRUE)) %>%
    arrange(desc(size_genes)) %>%
    slice_head(n = eregulon_n) %>%
    pull(TF)

  to_select_list <- c(mafb_pattern, emx2_pattern,
                      map_chr(top_ereg_tfs, ~sprintf("%s_.*_\\+\\/\\+", str_escape(.x))))

  ereg_un_sel_base = ereg_un %>%
    dplyr::filter(grepl(paste(to_select_list, collapse="|"), eRegulon_name)) %>%
    dplyr::rename(gene = Gene) %>%
    distinct(gene, TF, .keep_all=T)

  marker_sets = list(markers_pos = markers_pos_tg)
  ks_df = imap_dfr(marker_sets, function(mset, mset_name) {
    ereg_un_sel = ereg_un_sel_base %>%
      left_join(mset) %>%
      filter(!is.na(p_val)) %>%
      distinct(TF, gene, .keep_all=T) %>%
      group_by(TF) %>%
      arrange(desc(importance_TF2G)) %>%
      slice_head(n = 50) %>%
      ungroup()

    baseline_df = ereg_un_sel %>%
      filter(!TF %in% tg_genes) %>%
      group_by(gene) %>%
      summarise(avg_log2FC = mean(avg_log2FC, na.rm = TRUE), .groups = "drop")
    focal_df = ereg_un_sel %>% filter(TF == tg_cur)

    focal_fc = focal_df$avg_log2FC
    other_fc = baseline_df$avg_log2FC
    ks_res = ks.test(focal_fc, other_fc, alternative = "two.sided")

    gg = ggplot() +
      geom_density(data = baseline_df, aes(avg_log2FC),
                   fill = "gray70", color = "gray70", alpha = 0.5) +
      geom_density(data = focal_df, aes(avg_log2FC, color = TF)) +
      annotate("text", x = Inf, y = Inf, hjust = 1, vjust = 1, size = 2,
               label = sprintf("KS D=%.3f, p=%.3g", ks_res$statistic, ks_res$p.value)) +
      theme_bw() + plot_theme()
    print(gg)
    save_plot(file.path(out_dir_tg, sprintf("ereg_density_avg_log2FC_%s.pdf", mset_name)), gg,
              base_height = 1.5, base_width = 3)

    tibble(tg = tg_cur, marker_set = mset_name,
           ks_D = ks_res$statistic, ks_p = ks_res$p.value,
           n_focal = length(focal_fc), n_other = length(other_fc))
  })
  write_csv(ks_df, file.path(out_dir_tg, "ereg_ks_test_avg_log2FC.csv"))
  print(ks_df)
})
```

## Load songbird DEG

```{r}
markers_sns = readRDS(file.path(SNRNA_SEURAT_DIR, "deg/song-surround_deg_glut_v2/marker_genes_cluster_SCT.rds"))

markers_sns = markers_sns %>%
  dplyr::rename(p_val_sns = p_val,
         avg_log2FC_sns = avg_log2FC,
         pct.1_sns = pct.1,
         pct.2_sns = pct.2,
         p_val_adj_sns = p_val_adj)
markers_sns_base = markers_sns  # save before join, used for RRHO
markers_sns = markers_sns %>%
  left_join(markers_pos)
```

## RRHO Analysis

```{r rrho-helpers}
# Returns a two-column data.frame (gene, value) sorted descending, ready for RRHO2
make_rrho_list <- function(df, rank_var) {
  if (rank_var == "avg_log2FC") {
    df %>%
      transmute(gene, value = avg_log2FC)
  } else {
    df %>%
      mutate(value = -log10(p_val_adj + 1e-300) * sign(avg_log2FC)) %>%
      dplyr::select(gene, value)
  }
}

make_rrho_list_sns <- function(df, rank_var) {
  if (rank_var == "avg_log2FC") {
    df %>%
      transmute(gene, value = avg_log2FC_sns)
  } else {
    df %>%
      mutate(value = -log10(p_val_adj_sns + 1e-300) * sign(avg_log2FC_sns)) %>%
      dplyr::select(gene, value)
  }
}
```

```{r rrho-run}
oe_marker_sets  <- list(markers_pos = markers_pos)
sns_comparisons <- unique(markers_sns_base$compare)
rank_vars       <- c("avg_log2FC", "p_val_adj")

rrho_dir <- file.path(out_dir, "rrho")
dir.create(rrho_dir, showWarnings = FALSE, recursive = TRUE)

# Build flat list of (oe_name, tg_cur, cc_cur) combinations to parallelise over
rrho_jobs <- cross_df(list(
  oe_name  = names(oe_marker_sets),
  tg_cur   = tg_genes,
  cc_cur   = cell_classes
))

rrho_results <- mclapply(seq_len(nrow(rrho_jobs)), function(i) {
  oe_name <- rrho_jobs$oe_name[i]
  tg_cur  <- rrho_jobs$tg_cur[i]
  cc_cur  <- rrho_jobs$cc_cur[i]
  oe_df   <- oe_marker_sets[[oe_name]]

  rrho_dir_tg <- file.path(rrho_dir, tg_cur)
  dir.create(rrho_dir_tg, showWarnings = FALSE, recursive = TRUE)

  map_dfr(sns_comparisons, function(cmp_cur) {
    map_dfr(rank_vars, function(rv) {

      oe_cur  <- oe_df %>% filter(tg == tg_cur, cell_class == cc_cur) %>%
        filter(!duplicated(gene), !is.na(avg_log2FC), !is.na(p_val_adj),
               abs(avg_log2FC) > 0.1)
      sns_cur <- markers_sns_base %>% filter(compare == cmp_cur) %>%
        filter(!duplicated(gene), !is.na(avg_log2FC_sns), !is.na(p_val_adj_sns))

      common_genes <- intersect(oe_cur$gene, sns_cur$gene)
      if (length(common_genes) < 10)
        return(tibble(tg = tg_cur, oe_set = oe_name, cell_class = cc_cur,
                      compare = cmp_cur, rank_var = rv,
                      n_common = length(common_genes), max_log_pval = NA_real_))

      l1 <- make_rrho_list(oe_cur %>% filter(gene %in% common_genes), rv) %>%
        arrange(desc(value)) %>% as.data.frame()
      l2 <- make_rrho_list_sns(sns_cur %>% filter(gene %in% common_genes), rv) %>%
        arrange(desc(value)) %>% as.data.frame()

      rrho_obj <- RRHO2_initialize(l1, l2,
        labels = c(sprintf("%s_%s_%s", oe_name, tg_cur, cc_cur),
                   sprintf("sns_%s", cmp_cur)), method = "hyper", log10.ind = TRUE)

      fname_stem <- gsub("[^A-Za-z0-9_-]", "_",
        sprintf("rrho_%s_%s_%s_%s_%s", oe_name, tg_cur, cc_cur, cmp_cur, rv))
      pdf(file.path(rrho_dir_tg, sprintf("%s.pdf", fname_stem)),
          width = 7, height = 7)
      RRHO2_heatmap(rrho_obj)
      dev.off()

      nr      <- nrow(rrho_obj$hypermat)
      nc      <- ncol(rrho_obj$hypermat)
      split_r <- round(nr * mean(l1[, 2] > 0))
      split_c <- round(nc * mean(l2[, 2] > 0))
      quad_idx <- list(
        uu = list(r = seq_len(split_r),            c = seq_len(split_c)),
        ud = list(r = seq_len(split_r),            c = (split_c+1):nc),
        du = list(r = (split_r+1):nr,              c = seq_len(split_c)),
        dd = list(r = (split_r+1):nr,              c = (split_c+1):nc)
      )
      for (quad in names(quad_idx)) {
        idx      <- quad_idx[[quad]]
        masked   <- matrix(NA_real_, nrow = nr, ncol = nc)
        masked[idx$r, idx$c] <- rrho_obj$hypermat[idx$r, idx$c]
        vals <- masked[!is.na(masked)]
        if (length(vals) == 0 || diff(range(vals)) == 0) next
        rrho_quad           <- rrho_obj
        rrho_quad$hypermat  <- masked
        pdf(file.path(rrho_dir_tg, sprintf("%s_%s.pdf", fname_stem, quad)),
            width = 7, height = 7)
        tryCatch(RRHO2_heatmap(rrho_quad), error = function(e) NULL)
        dev.off()
      }

      for (quad in c("uu", "dd", "ud", "du")) {
        gl <- rrho_obj[[paste0("genelist_", quad)]]
        if (length(gl) > 0)
          write_lines(gl, file.path(rrho_dir_tg,
            sprintf("%s_%s_genes.txt", fname_stem, quad)))
      }

      max_lp <- max(abs(as.numeric(rrho_obj$hypermat)), na.rm = TRUE)

      tibble(tg = tg_cur, oe_set = oe_name, cell_class = cc_cur,
             compare = cmp_cur, rank_var = rv,
             n_common = length(common_genes), max_log_pval = max_lp)
    })
  })
}, mc.cores = length(rrho_jobs), mc.set.seed = FALSE)

rrho_summary <- bind_rows(rrho_results)

walk(tg_genes, function(tg_cur) {
  rrho_summary_tg <- rrho_summary %>% filter(tg == tg_cur)
  write_csv(rrho_summary_tg, file.path(rrho_dir, tg_cur, sprintf("rrho_summary_%s.csv", tg_cur)))
  print(rrho_summary_tg)
})
```

## Load snRNA-seq object

```{r}
sn_fname = file.path(SNRNA_SEURAT_DIR, "reduction_viz/combined_all_umap/obj_clustered.qs")
sn_obj = qread(sn_fname)
sn_glut = subset(sn_obj, cells = Cells(sn_obj)[grepl("Glut", sn_obj$cluster)])
```

### Plot DEG expression in songbird

```{r plot-deg-songbird-sn}
sn_reduction = "dims40nn30mindist0.3"
n_top = 20
 
walk(tg_genes, function(tg_cur) {
  out_dir_tg = file.path(out_dir, tg_cur)
  dir.create(out_dir_tg, showWarnings = FALSE)

  top_markers_pos = markers_pos %>%
    filter(tg == tg_cur, cell_class == "Glut") %>%
    filter(gene %in% Features(sn_obj)) %>%
    slice_min(p_val_adj, n = n_top) %>%
    slice_max(avg_log2FC, n = n_top) %>%
    pull(gene) %>% unique()

  plot_density_features(sn_obj, top_markers_pos, sn_reduction,
                        file.path(out_dir_tg, "sn_top_markers_pos", "density"))
})
```

#### Heatmaps

```{r plot-deg-songbird-sn-heatmap}
# Song neuron → paired non-song neuron
song_pairs = list(
  "Glut-RA"    = "Glut-Arco-1",
  "Glut-HVC-1" = "Glut-NC-1",
  "Glut-HVC-2" = "Glut-NC-4"
)

all_cts      = c(names(song_pairs), unname(unlist(song_pairs)))
adjust_factor = 0.08

n_top = 20

walk(tg_genes, function(tg_cur) {
  out_dir_tg = file.path(out_dir, tg_cur)

  top_markers_pos = markers_pos %>%
    filter(tg == tg_cur, cell_class == "Glut") %>%
    filter(gene %in% Features(sn_obj)) %>%
    slice_min(p_val_adj, n = n_top) %>%
    slice_max(avg_log2FC, n = n_top) %>%
    pull(gene) %>% unique()

  gene_sets = list(
    list(genes = top_markers_pos, stem = "sn_top_markers_pos", oe_src = markers_pos)
  )
  for (gene_set in gene_sets) {
    
    out_dir_cur = file.path(out_dir_tg, gene_set$stem)
    dir.create(out_dir_cur, showWarnings = F, recursive = T)
    
    out_fname = file.path(out_dir_cur, "heatmap_glut_normed.pdf")
    
    feats = intersect(gene_set$genes, Features(sn_glut))
    if (length(feats) == 0) next
    avg_mat = log2(AggregateExpression(sn_obj, features = feats, group.by = "cluster", assay = "SCT")$SCT+1)
    avg_mat = avg_mat[, all_cts]

    # Normalize each song neuron against its non-song pair (log-scale difference)
    norm_mat = do.call(cbind, map(names(song_pairs), function(song_ct) {
      avg_mat[, song_ct] - avg_mat[, song_pairs[[song_ct]]]
    }))
    colnames(norm_mat) = names(song_pairs)
    norm_scaled = norm_mat
    norm_scaled = norm_scaled[complete.cases(norm_scaled), , drop = FALSE]

    val_limit = ceiling(quantile(abs(norm_scaled), 0.95, na.rm = TRUE))
    norm_scaled = pmin(pmax(norm_scaled, -val_limit), val_limit)
    message(sprintf("[%s / %s] val_limit = %g  (data range: %g – %g)",
                    tg_cur, gene_set$stem, val_limit,
                    min(norm_scaled, na.rm = TRUE), max(norm_scaled, na.rm = TRUE)))
    col = circlize::colorRamp2(breaks = c(-1 * val_limit, 0, val_limit),
                               colors = c("darkblue", "white", "darkred"))
    if (nrow(norm_scaled) == 0) next

    ht_song = Heatmap(norm_scaled, name = "log2\n(song - non-song)",
                      col = col,
                      cluster_rows    = TRUE,
                      clustering_method_rows = "ward.D",
                      show_row_dend   = FALSE,
                      cluster_columns = FALSE,
                      border          = "black",
                      row_names_gp    = gpar(fontsize = 5, fontfamily = "sans"),
                      column_names_gp = gpar(fontsize = 5, fontfamily = "sans"),
                      height = unit(nrow(norm_scaled) * adjust_factor, "in"),
                      width  = unit(ncol(norm_scaled) * adjust_factor, "in"),
                      heatmap_legend_param = list(
                        at         = c(-val_limit, 0, val_limit),
                        labels     = c(sprintf("%.1f", -val_limit), "0",
                                       sprintf("%.1f",  val_limit)),
                        labels_gp  = gpar(fontsize = 5),
                        title_gp   = gpar(fontsize = 5),
                        direction  = "horizontal",
                        legend_width = unit(15, "mm"),
                        grid_height  = unit(2, "mm")))

    ht = ht_song

    print(ht)
    pdf(out_fname,
        width  = ncol(norm_scaled) * 0.25 + 1.5,
        height = nrow(norm_scaled) * 0.2 + 1)
    draw(ht)
    dev.off()
  }
})
```

#### Heatmaps — row-normalized expression

```{r plot-deg-songbird-sn-heatmap-rownorm}
ct_cols = c("Glut-RA", "Glut-Arco-1", "Glut-HVC-1", "Glut-NC-1", "Glut-HVC-2", "Glut-NC-4")

walk(tg_genes, function(tg_cur) {
  out_dir_tg = file.path(out_dir, tg_cur)

  top_markers_pos = markers_pos %>%
    filter(tg == tg_cur, cell_class == "Glut") %>%
    filter(gene %in% Features(sn_obj)) %>%
    slice_min(p_val_adj, n = n_top) %>%
    slice_max(avg_log2FC, n = n_top) %>%
    pull(gene) %>% unique()

  feats = intersect(top_markers_pos, Features(sn_glut))
  if (length(feats) == 0) return(invisible(NULL))

  avg_mat = log2(AggregateExpression(sn_obj, features = feats, group.by = "cluster", assay = "SCT")$SCT + 1)
  avg_mat = avg_mat[feats, ct_cols, drop = FALSE]

  rn_mat = t(scale(t(avg_mat)))
  rn_mat = rn_mat[complete.cases(rn_mat), , drop = FALSE]
  if (nrow(rn_mat) == 0) return(invisible(NULL))

  val_limit = ceiling(quantile(abs(rn_mat), 0.95, na.rm = TRUE))
  rn_mat = pmin(pmax(rn_mat, -val_limit), val_limit)

  col = circlize::colorRamp2(
    breaks = c(-val_limit, 0, val_limit),
    colors = c("darkblue", "white", "darkred")
  )

  out_dir_cur = file.path(out_dir_tg, "sn_top_markers_pos")
  dir.create(out_dir_cur, showWarnings = FALSE, recursive = TRUE)
  out_fname = file.path(out_dir_cur, "heatmap_glut_rownorm.pdf")

  ht = Heatmap(rn_mat, name = "row z-score",
               col = col,
               cluster_rows           = TRUE,
               clustering_method_rows = "ward.D",
               show_row_dend          = FALSE,
               cluster_columns        = FALSE,
               border          = "black",
               row_names_gp    = gpar(fontsize = 5, fontfamily = "sans"),
               column_names_gp = gpar(fontsize = 5, fontfamily = "sans"),
               height = unit(nrow(rn_mat) * adjust_factor, "in"),
               width  = unit(ncol(rn_mat) * adjust_factor, "in"),
               heatmap_legend_param = list(
                 at           = c(-val_limit, 0, val_limit),
                 labels       = c(sprintf("%.1f", -val_limit), "0",
                                  sprintf("%.1f",  val_limit)),
                 labels_gp    = gpar(fontsize = 5),
                 title_gp     = gpar(fontsize = 5),
                 direction    = "horizontal",
                 legend_width = unit(15, "mm"),
                 grid_height  = unit(2, "mm")))

  print(ht)
  pdf(out_fname,
      width  = ncol(rn_mat) * 0.25 + 1.5,
      height = nrow(rn_mat) * 0.2 + 1)
  draw(ht)
  dev.off()
})
```

#### Scatter plots — OE DE vs song/non-song DE

```{r plot-deg-songbird-sn-scatter}

n_top = 50
walk(tg_genes, function(tg_cur) {
  out_dir_tg = file.path(out_dir, tg_cur)
  dir.create(out_dir_tg, showWarnings = FALSE)

  markers_tg = markers_pos %>%
    filter(tg == tg_cur, cell_class == "Glut") %>%
    filter(gene %in% Features(sn_obj)) %>%
    slice_min(p_val_adj, n = n_top) %>%
    slice_max(avg_log2FC, n = n_top)

  feats = unique(markers_tg$gene)
  if (length(feats) == 0) return(invisible(NULL))

  avg_mat = log2(
    AggregateExpression(sn_obj, features = feats, group.by = "cluster", assay = "SCT")$SCT + 1
  )

  label_genes = switch(tg_cur,
    MAFB = c("SLC13A1", "ATP6V1C2", "GSDMA"),
    EMX2 = c("LIG4", "FCF1", "CNTD1"),
    character(0)
  )

  plots = imap(song_pairs, function(nonsong_ct, song_ct) {
    df = tibble(
      gene      = feats,
      oe_log2fc = markers_tg$avg_log2FC[match(feats, markers_tg$gene)],
      song_diff = avg_mat[feats, song_ct] - avg_mat[feats, nonsong_ct]
    ) %>% filter(!is.na(oe_log2fc), !is.na(song_diff))

    df_label = df %>% filter(gene %in% label_genes)

    ggplot(df, aes(x = oe_log2fc, y = song_diff)) +
      geom_hline(yintercept = 0, linewidth = 0.3, linetype = "dashed", color = "gray50") +
      geom_point(size = 0.5, alpha = 0.4,
                 position = position_jitter(width = 0.02, height = 0, seed = 42)) +
      ggrepel::geom_text_repel(data = df_label, aes(label = gene),
                               size = 2, max.overlaps = Inf) +
      labs(x = sprintf("log2FC (%s OE)", tg_cur),
           y = sprintf("log2 expr\n(%s − %s)", song_ct, nonsong_ct)) +
      theme_bw() + plot_theme()
  })

  gg_combined = plot_grid(plotlist = plots, nrow = 1)
  print(gg_combined)
  save_plot(
    file.path(out_dir_tg, "scatter_oe_vs_song-nonsong.pdf"),
    gg_combined,
    base_height = 2,
    base_width  = 2 * length(song_pairs)
  )
})
```

#### Heatmaps — combined MAFB + EMX2 top genes

```{r plot-deg-songbird-sn-heatmap-combined}
# Select top DE genes for both transgenes and plot song neuron DE alongside
# both transgene DE columns in a single heatmap per gene set.

gene_sets_combined = list(
  list(stem = "sn_top_markers_pos", oe_src = markers_pos)
)

out_dir_combined = file.path(out_dir, "combined")
dir.create(out_dir_combined, showWarnings = FALSE)

for (gene_set in gene_sets_combined) {
  # Union of top genes across both transgenes
  all_top = map(tg_genes, function(tg_cur) {
    gene_set$oe_src %>%
      filter(tg == tg_cur, cell_class == "Glut") %>%
      filter(gene %in% Features(sn_obj)) %>%
      slice_min(p_val_adj, n = n_top) %>%
          slice_max(avg_log2FC, n = n_top) %>%
      pull(gene) %>% unique()
  }) %>% unlist() %>% unique()

  feats = intersect(all_top, Features(sn_glut))
  if (length(feats) == 0) next

  avg_mat = log2(AggregateExpression(sn_obj, features = feats, group.by = "cluster", assay = "SCT")$SCT + 1)
  avg_mat = avg_mat[, all_cts]

  # Normalize each song neuron against its paired non-song neuron (log-scale difference)
  norm_mat = do.call(cbind, map(names(song_pairs), function(song_ct) {
    avg_mat[, song_ct] - avg_mat[, song_pairs[[song_ct]]]
  }))
  colnames(norm_mat) = names(song_pairs)
  norm_mat = norm_mat[complete.cases(norm_mat), , drop = FALSE]
  if (nrow(norm_mat) == 0) next

  val_limit = ceiling(quantile(abs(norm_mat), 0.95, na.rm = TRUE))
  norm_scaled = pmin(pmax(norm_mat, -val_limit), val_limit)
  col = circlize::colorRamp2(breaks = c(-1 * val_limit, 0, val_limit),
                             colors = c("darkblue", "white", "darkred"))

  g = rownames(norm_scaled)

  # OE matrix: one column per transgene
  oe_mat = do.call(cbind, map(tg_genes, function(tg_cur) {
    oe_df = gene_set$oe_src %>% filter(tg == tg_cur, cell_class == "Glut")
    oe_fc = setNames(oe_df$avg_log2FC, oe_df$gene)
    matrix(oe_fc[g], ncol = 1, dimnames = list(g, sprintf("%s+/-", tg_cur)))
  }))

  oe_lim = ceiling(max(abs(oe_mat), na.rm = TRUE))
  oe_pal = scales::brewer_pal(type = "div", palette = 4, direction = -1)
  oe_col = circlize::colorRamp2(seq(-oe_lim, oe_lim, length.out = 9), oe_pal(9))

  ht_song = Heatmap(norm_scaled, name = "log2\n(song - non-song)",
                    col = col,
                    cluster_rows    = TRUE,
                    show_row_dend   = FALSE,
                    cluster_columns = FALSE,
                    border          = "black",
                    row_names_gp    = gpar(fontsize = 5, fontfamily = "sans"),
                    column_names_gp = gpar(fontsize = 5, fontfamily = "sans"),
                    height = unit(nrow(norm_scaled) * adjust_factor, "in"),
                    width  = unit(ncol(norm_scaled) * adjust_factor, "in"),
                    heatmap_legend_param = list(
                      at           = c(-val_limit, 0, val_limit),
                      labels       = c(sprintf("%.1f", -val_limit), "0",
                                       sprintf("%.1f",  val_limit)),
                      labels_gp    = gpar(fontsize = 5),
                      title_gp     = gpar(fontsize = 5),
                      direction    = "horizontal",
                      legend_width = unit(15, "mm"),
                      grid_height  = unit(2, "mm")))

  ht_oe = Heatmap(oe_mat, name = "log2FC\n(OE)",
                  col = oe_col,
                  na_col = "gray90",
                  cluster_rows    = FALSE,
                  cluster_columns = FALSE,
                  border          = "black",
                  row_names_gp    = gpar(fontsize = 5, fontfamily = "sans"),
                  column_names_gp = gpar(fontsize = 5, fontfamily = "sans"),
                  height = unit(nrow(oe_mat) * adjust_factor, "in"),
                  width  = unit(ncol(oe_mat) * adjust_factor, "in"),
                  heatmap_legend_param = list(
                    at           = c(-oe_lim, 0, oe_lim),
                    labels       = c(sprintf("%.1f", -oe_lim), "0",
                                     sprintf("%.1f",  oe_lim)),
                    labels_gp    = gpar(fontsize = 5),
                    title_gp     = gpar(fontsize = 5),
                    direction    = "horizontal",
                    legend_width = unit(15, "mm"),
                    grid_height  = unit(2, "mm")))

  ht = ht_song + ht_oe
  print(ht)
  pdf(file.path(out_dir_combined, sprintf("heatmap_glut_normed_%s.pdf", gene_set$stem)),
      width  = (ncol(norm_scaled) + ncol(oe_mat)) * 0.25 + 1.5,
      height = nrow(norm_scaled) * 0.2 + 1)
  draw(ht)
  dev.off()
}
```

## DEG Expression — Violin and Box Plots

```{r deg-vln-box}
n_top_vln <- 20

tg_comparisons <- list(
  MAFB = c("Glut_MAFB", "Glut_none"),
  EMX2 = c("Glut_EMX2", "Glut_none")
)

walk(tg_genes, function(tg_cur) {
  out_dir_tg <- file.path(out_dir, tg_cur)
  dir.create(out_dir_tg, showWarnings = FALSE)

  groups <- tg_comparisons[[tg_cur]]

  top_genes <- markers_pos %>%
    filter(tg == tg_cur, cell_class == "Glut") %>%
    filter(avg_log2FC > 0) %>%
    slice_min(p_val_adj, n = n_top_vln) %>%
    slice_max(avg_log2FC, n = n_top_vln) %>%
    pull(gene) %>%
    unique() %>%
    intersect(Features(obj))

  if (length(top_genes) == 0) return(invisible(NULL))

  cells_use <- Cells(obj)[obj$cell_class_tg_pos %in% groups]
  obj_sub   <- subset(obj, cells = cells_use)
  obj_sub$cell_class_tg_pos <- factor(obj_sub$cell_class_tg_pos, levels = groups)

  # Violin + box plot grid
  gg_vln <- VlnPlot(
    obj_sub,
    features    = top_genes,
    group.by    = "cell_class_tg_pos",
    pt.size     = 1,
    combine     = FALSE
  )
  gg_vln <- map(gg_vln, function(p) {
    p + geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white", alpha = 0.6) +
      plot_theme() +
      theme(axis.title.x = element_blank(), legend.position = "none")
  })
  gg_vln_combined <- plot_grid(plotlist = gg_vln, ncol = 5)
  print(gg_vln_combined)
  save_plot(
    file.path(out_dir_tg, "vln_top_deg_glut.pdf"),
    gg_vln_combined,
    base_height = ceiling(length(top_genes) / 5) * 1.5,
    base_width  = 7.5
  )

  # Box plot grid
  expr_mat <- GetAssayData(obj_sub, assay = "SCT", layer = "data")[top_genes, , drop = FALSE]
  expr_df  <- as.data.frame(t(as.matrix(expr_mat))) %>%
    rownames_to_column("cell") %>%
    mutate(group = obj_sub$cell_class_tg_pos) %>%
    pivot_longer(cols = all_of(top_genes), names_to = "gene", values_to = "expression") %>%
    mutate(gene = factor(gene, levels = top_genes),
           group = factor(group, levels = groups))

  gg_box <- ggplot(expr_df, aes(x = group, y = expression, fill = group)) +
    geom_violin(trim = TRUE, scale = "width", alpha = 0.6) +
    geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white", alpha = 0.7) +
    facet_wrap(~gene, ncol = 5, scales = "free_y") +
    scale_fill_manual(values = setNames(
      scales::hue_pal()(length(groups)), groups
    )) +
    theme_bw() + plot_theme() +
    theme(axis.text.x  = element_text(angle = 30, hjust = 1),
          axis.title.x = element_blank(),
          legend.position = "none")
  print(gg_box)
  save_plot(
    file.path(out_dir_tg, "box_top_deg_glut.pdf"),
    gg_box,
    base_height = ceiling(length(top_genes) / 5) * 1.8,
    base_width  = 7.5
  )
})
```

