---
title: "Apply composite chicken-integration labels"
format: html
editor: source
---

Applies the real cross-species integration (`snrna/integration/composite_gg_adult`, run directly
against this project's own reference object -- not the self-generated `Ex_`-style aliases from
`glut_ex_style_naming.qmd`, which this supersedes for every cluster confidently matched here) as the
new naming for glutamatergic clusters, per specific rules. Independent of the preprocessing notebook
by the same convention as the other naming notebooks: reads the backup object read-only, writes its
own copy, never touches the canonical file.

**Confirmed before anything else**: `composite_calls_summary.csv`'s `query` column is an exact 1:1
match to this backup object's `cluster` levels (no rows extra on either side), so the integration was
run against this same object and the lookup below can join on `cluster` directly.

**Later additions to the rules below, in the order they came in:**
- `Glut-Pre-1/2/3` are no longer left unchanged -- renamed to the developmental stage they actually
  represent (`Glut-NSC` -> `Glut-NB` -> `Glut-Im`), not a chicken correspondence.
- `GABA-Pre` reverted to unchanged after initially being relabelled like every other GABA cluster
  (the same "leave the precursor alone" treatment `Glut-Pre-*` already had, restored) -- then later
  renamed again, this time to `GABA-Im`, matching the immature-interneuron developmental stage
  `Glut-Im` already uses on the Glut side, rather than staying an unchanged "Pre" label.
- `Glut-Arco-4` dropped as a confirmed artifact (elevated mito/complexity relative to its Arco
  siblings; see the QC section below), joining `Glut-Nido-3`.
- Every disambiguated cluster now takes a real distinguishing gene instead of a bare `-1`/`-2`
  wherever `naming_filters()` found one, via a nested pseudobulk DE within just that collision
  group -- see *A second disambiguation strategy* below for a specificity-score cross-check on
  these picks, and *Where are the DE results stored* for the actual per-gene tables.
- `_like_` dropped from the GABA-side names (Zaremba's ganglionic-eminence-of-origin naming,
  `Inh_LGE_like_FOXP2` -> `GABA-LGE_FOXP2`) -- readability only, not a change in which chicken
  population a cluster was matched to.
- The five (now more) pseudobulk DESeq2 fits are cached (`redo$disambiguate`), since quarto
  re-executes the whole document on every render and this is the slow part by a wide margin.

## The rules, as given, and what they mean concretely

- **Song-region clusters (HVC, RA, LMAN) keep their existing region-based names.** These are
  genuinely region-specific song nuclei; no chicken correspondence is substituted for them, matched
  or not.
- **`Glut-Pre-1/2/3` stay exactly as they are**, regardless of what the composite calls for them
  (some are high-confidence, e.g. `Glut-Pre-2` -> `Ex_Pre_KCNH7` at 0.69, but they are precursor
  populations and out of scope here either way).
- **`Glut-Nido-1` folds into `Glut-NC-1`, and `Glut-NC-1` itself now also takes its own composite
  name** (confirmed: rename `Glut-NC-1` -> `Glut-DACH2_MGAT4C`). Both clusters' top composite call is
  `Ex_DACH2_MGAT4C` (`Glut-NC-1` at confidence 0.465, `Glut-Nido-1` at 0.539) -- the integration found
  them to be the same chicken-matched population independently of the fold instruction. Because
  `Glut-NC-1` is no longer exempted, it re-enters the general eligible pool below, where it now
  collides with `Glut-NR-3` (also `Ex_DACH2_MGAT4C`) and needs the same numeric-suffix treatment as
  the Arco trio -- so the fold target is computed programmatically (whatever `Glut-NC-1` ends up
  named after de-duplication) rather than hardcoded, so the two stay in sync if the composite table
  or the suffix order ever changes.
- **`Glut-Nido-2` becomes `Glut-GABA`** -- a standalone label, not derived from the composite call at
  all. Independently consistent with the composite result: its own top calls are barely above
  chance (confidence 0.264) and split between an excitatory and an *inhibitory* chicken type
  (`Ex_DACH2_MGAT4C` vs `Inh_LGE_like_FOXP1_DGKH`) with zero of the four underlying methods agreeing
  with each other -- exactly what a real dual-transmitter population that matches no single-identity
  reference type would look like.
- **`Glut-Nido-3` is untouched** -- already excluded upstream as a confirmed artifact (the
  manuscript-revision fix in `snrna_seurat_cellbender_preprocess.qmd`); this notebook does not
  reintroduce it via a composite label. Its own composite call is incidentally the weakest in the
  whole glutamatergic set (confidence 0.141), which is further consistent with that decision but was
  not the basis for it.
- **Every remaining glutamatergic cluster (Arco, NC, NR, now including NC-1) takes its
  `top_rank_agg` composite call**, prefix swapped from `Ex_` to `Glut-` (literal substitution, e.g.
  `Ex_CACNA1H_LHX9` -> `Glut-CACNA1H_LHX9`) -- `top_rank_agg` because it is the rank-aggregation
  consensus across all four underlying methods (gsi, samap, saturn, cca), with its own dedicated
  score/margin columns immediately following it, whereas `top_borda`/`top_zaremba` are single-method
  comparisons shown alongside it. Where more than one cluster shares the same winner, a `_1`/`_2`/...
  suffix is appended, ordered by the original cluster's own label, so distinct populations the
  notebook's own clustering resolved are not collapsed into one name just because Zaremba's chicken
  atlas did not resolve a matching split. Two such collisions this time: `Ex_CACNA1H_LHX9`
  (`Glut-Arco-1/3/5`) and, newly, `Ex_DACH2_MGAT4C` (`Glut-NC-1`/`Glut-NR-3`, now that NC-1 is no
  longer exempted).
- **GABA clusters get the identical treatment, except `GABA-Pre`**, exempted from the composite match
  and renamed `GABA-Im` instead (same exemption as `Glut-Pre-*`, same developmental-stage relabel as
  `Glut-Im`) -- which also means `GABA-7` is no longer part of a collision, since `GABA-Pre` was its
  only `Inh_MGE_like_ST18` partner, and gets the clean unsuffixed name.
- **`Glut-Arco-4` is dropped** -- excluded the same way `Glut-Nido-3` already is, not merely left
  unlabelled. Consistent with the QC check below (elevated mito and complexity relative to its Arco
  siblings, though not a dramatic outlier) and with the Xenium evidence that Arco-1/3/5 occupy
  distinct spatial territory (below), which argues the *opposite* of Arco-4's problem: those three
  are being kept apart precisely because they are real, distinct populations, not folded together
  just because one chicken cluster matched all three.
- **Every remaining duplicate-winner group gets a marker-based disambiguator, not a bare number.**
  Where the composite integration assigns the same chicken population to more than one of this
  project's own clusters, that is chicken's resolution running out, not evidence the clusters are
  the same thing -- confirmed directly for the Arco trio by the Xenium spatial result. So each
  group (Arco-1/3/5 on `Ex_CACNA1H_LHX9`; `Glut-NC-1`+folded-`Glut-Nido-1`/`Glut-NR-3` on
  `Ex_DACH2_MGAT4C`; three GABA pairs) gets its own nested pseudobulk DE -- the same
  `cluster_de()`/`select_naming_genes()` machinery as `glut_ex_style_naming.qmd`, restricted to just
  that group's own clusters -- to find a gene that tells its members apart, appended after a dash
  (`Glut-CACNA1H_LHX9-<gene>`). A numeral is the fallback only where no marker clears
  `naming_filters()`, not the default.
- **Pseudobulk here uses `sample_name` + `assignment`, not `individual`.** This backup object
  predates the souporcell reprocessing (no `individual`/`batch` columns), and the earlier notebooks
  in this session refused to substitute anything for that reason -- raw `assignment` alone is a
  *per-library* souporcell index, so the same value in two different libraries is two different
  birds, and pooling on it directly would silently mix animals. Qualifying it by library first
  (`replicate_lib = paste(sample_name, assignment)`) is what makes using it here valid: within one
  library, distinct `assignment` values are genuinely distinct known birds (that library's own
  souporcell run), which is exactly the biological-replicate structure pseudobulk DE needs. Built
  once, below, and used for every group.
- **Non-neuronal clusters (Astro, Oligo, OPC, Micro, Endo, Epen) are untouched.**

## Packages

```{r}
#| label: packages
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(knitr)

source(here::here("scripts/cluster_curation.R"))
source(here::here("scripts/cluster_naming.R"))

## Excludes annotation placeholders (`^LOC[0-9]` NCBI-style loci, `^nbis-`/`-gr[0-9]+$` NBIS-pipeline
## loci -- which is also every chrM gene in this genome's annotation, confirmed directly by filtering
## `GTF_LONSTR_BIOTYPE`'s transcripts to `seqnames == "chrM"`: all 37 get one of these placeholder IDs,
## none a real symbol) and ribosomal protein genes (`^RP[SL][0-9]`/`^MRP[SL][0-9]`, 146 genes, checked
## to spare the `RPS6K*` kinase family) from candidacy as a *naming* gene -- used both for the
## DESeq2-based disambiguation below and the specificity-score cross-check, so a housekeeping or
## contamination-associated gene cannot become a cluster's name via either route.
is_named_gene <- function(gene) {
  !grepl("^LOC[0-9]", gene) & !grepl("^nbis-", gene, ignore.case = TRUE) & !grepl("-gr[0-9]+$", gene) &
    !grepl("^RP[SL][0-9]+[A-Z]?[0-9]?$", gene) & !grepl("^MRP[SL][0-9]+[A-Z]?[0-9]?$", gene)
}
```

## Parameters

```{r}
#| label: params
## EXPLORATORY -- not production. See ../README.md; the production filtering and naming live in
## ../hybrid_division_naming.qmd.
## The main clustered object -- the same one the composite integration was run against (confirmed
## 1:1 against composite_calls_summary.csv$query below). Read-only, never written back to.
in_fname <- here::here("snrna/clustering/snrna_seurat_cellbender_preprocess",
                       "obj_clustered.qs2")
composite_fname <- here::here("snrna/integration/composite_gg_adult", "composite_calls_summary.csv")
cluster_col <- "cluster"

out_dir <- here::here("snrna/naming/exploratory", "apply_composite_labels")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
lookup_fname <- file.path(out_dir, "composite_label_lookup.csv")
obj_out_fname <- file.path(out_dir, "obj_composite_labels.qs2")
disambig_cache_fname <- file.path(out_dir, "disambiguation_cache.qs2")

## quarto re-executes the whole document on every render (no chunk caching configured), and the
## disambiguation step below is five separate pseudobulk DESeq2 fits over the full gene set --
## the actual slow part of this notebook by a wide margin. Cached manually rather than via
## knitr's `cache: true`: that cache invalidates on the chunk's own code text (or explicit
## `cache.extra`), not on upstream objects like `obj` or `composite` changing, which would risk a
## silently stale result exactly where wrongness is hardest to notice (a label that looks fine but
## was disambiguated against last week's data). A `redo` flag makes the recompute decision
## explicit instead, matching the convention already used elsewhere in this project
## (snrna_souporcell_clustering.qmd, zaremba_naming_validation.qmd, ...).
redo <- list(disambiguate = FALSE)

## Excluded regardless: confirmed artifacts, dropped upstream of any naming exercise.
excluded_clusters <- c("Glut-Nido-3", "Glut-Arco-4")
song_region_clusters <- c("Glut-HVC-1", "Glut-HVC-1a", "Glut-HVC-2", "Glut-LMAN-1", "Glut-LMAN-2", "Glut-RA")
## Developmental-stage renames, not "unchanged" -- NSC -> NB -> Im is the maturation order the old
## Pre-1/2/3 numbering never made explicit. precursor_clusters (the original names) is still what
## excludes these from composite eligibility below; precursor_overrides is the label they land on.
precursor_overrides <- c("Glut-Pre-1" = "Glut-NSC", "Glut-Pre-2" = "Glut-NB", "Glut-Pre-3" = "Glut-Im")
precursor_clusters <- names(precursor_overrides)
gaba_precursor_overrides <- c("GABA-Pre" = "GABA-Im")
gaba_precursor_clusters <- names(gaba_precursor_overrides)
## Glut-Nido-1's target is filled in below, once Glut-NC-1's own composite-derived name is known --
## NC-1 is no longer exempted, so it is not necessarily still called "Glut-NC-1" itself.
nido2_override <- c("Glut-Nido-2" = "Glut-GABA")

## naming_filters() for the within-group disambiguation DE, below. max_genes = 1: the point is one
## short, readable disambiguator per cluster, not a full combination name.
disambig_filters <- naming_filters(max_genes = 1)
```

## Load

```{r}
#| label: load
stopifnot(file.exists(in_fname), file.exists(composite_fname))
obj <- qs_read(in_fname, nthreads = 8)
composite <- read_csv(composite_fname, show_col_types = FALSE)

stopifnot(
  "composite_calls_summary.csv$query is not an exact match to the object's cluster levels" =
    setequal(composite$query, levels(factor(obj@meta.data[[cluster_col]])))
)

## Library-qualified souporcell assignment -- see the prose note above for why raw `assignment`
## alone is not safe to pool across libraries. Built once here, used by every within-group
## disambiguation DE call below.
obj$replicate_lib <- paste(obj$sample_name, obj$assignment, sep = "_")
```

## Build the lookup

```{r}
#| label: lookup
all_clusters <- levels(factor(obj@meta.data[[cluster_col]]))
gaba_clusters <- all_clusters[grepl("^GABA", all_clusters)]
nonneuronal_clusters <- setdiff(
  all_clusters,
  c(all_clusters[grepl("^Glut", all_clusters)], gaba_clusters)
)

## Shared logic for both the Glut and GABA sides: strip the chicken-side prefix, swap in this
## project's own prefix, and suffix any winner shared by more than one of *this eligible set's*
## clusters (a shared winner against something exempted elsewhere is not a naming collision, since
## the exempted cluster never takes the composite string at all).
## body_cleanup runs *before* duplicate detection, not after -- so if a cleanup ever collapsed two
## distinct raw chicken calls into the same cleaned string, that would correctly be caught as a
## collision needing disambiguation, rather than silently missed because the check ran on the
## pre-cleanup string.
assign_composite_names <- function(composite, eligible_clusters, old_prefix, new_prefix,
                                   body_cleanup = identity) {
  elig <- composite %>%
    dplyr::filter(query %in% eligible_clusters) %>%
    dplyr::transmute(cluster = query, body = body_cleanup(sub(paste0("^", old_prefix), "", top_rank_agg)),
                     confidence, confidence_tier)
  dup <- elig %>% dplyr::count(body) %>% dplyr::filter(n > 1) %>% dplyr::pull(body)
  elig %>%
    dplyr::arrange(body, cluster) %>%
    dplyr::group_by(body) %>%
    dplyr::mutate(
      suffix = dplyr::if_else(body %in% dup, paste0("_", dplyr::row_number()), ""),
      new_label = paste0(new_prefix, body, suffix)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(cluster, new_label, confidence, confidence_tier)
}

## Both Nido-1 and Nido-2 are excluded here regardless of what their final labels turn out to be --
## Nido-1's target is derived from Glut-NC-1's own result below, not computed independently, so it
## must not also get a composite name of its own here (that would create a second, conflicting row
## for the same cluster and trip the "exactly once" check further down).
glut_eligible <- setdiff(
  all_clusters[grepl("^Glut", all_clusters)],
  c(song_region_clusters, precursor_clusters, "Glut-Nido-1", names(nido2_override), excluded_clusters)
)
glut_named <- assign_composite_names(composite, glut_eligible, "Ex_", "Glut-")

gaba_eligible <- setdiff(gaba_clusters, gaba_precursor_clusters)
## "_like_" dropped from Zaremba's ganglionic-eminence-of-origin naming (Inh_LGE_like_FOXP2 ->
## LGE_FOXP2, etc.) -- a readability call, not a change in which chicken population each cluster
## was actually matched to.
gaba_named <- assign_composite_names(composite, gaba_eligible, "Inh_", "GABA-",
                                     body_cleanup = function(x) gsub("_like_", "_", x))

## Replace the bare numeric suffix with a real distinguishing gene, wherever one clears
## naming_filters() -- a shared chicken match is chicken's resolution running out, not evidence the
## clusters are the same population (confirmed directly for Arco-1/3/5 by the Xenium spatial
## result), so "_1"/"_2" is a fallback here, not the intended final name.
##
## Returns list(named, de_tables): `de_tables` is the full per-gene cluster_de() output for every
## collision group, keyed by the shared composite body (e.g. "Glut-CACNA1H_LHX9") -- previously
## computed inside this function's loop, used once by select_naming_genes(), and discarded, so
## there was nowhere to go look at *why* a cluster got the gene it did, or why one got nothing
## (e.g. Glut-Arco-1). Kept now specifically so that question has an answer.
disambiguate_with_markers <- function(obj, cluster_col, named, filters, replicate_col = "replicate_lib") {
  base_body <- sub("_[0-9]+$", "", named$new_label)
  dup_bodies <- unique(base_body[duplicated(base_body) | duplicated(base_body, fromLast = TRUE)])
  named$marker_disambiguated <- FALSE
  de_tables <- list()

  for (b in dup_bodies) {
    grp_clusters <- named$cluster[base_body == b]
    cells <- colnames(obj)[obj@meta.data[[cluster_col]] %in% grp_clusters]
    obj_grp <- subset(obj, cells = cells)
    obj_grp@meta.data[[cluster_col]] <- droplevels(factor(obj_grp@meta.data[[cluster_col]]))
    DefaultAssay(obj_grp) <- "SCT"
    de_grp <- cluster_de(obj_grp, cluster_col, replicate_col = replicate_col)
    de_tables[[b]] <- de_grp
    ## Placeholder/mitochondrial/ribosomal genes excluded from candidacy here, not only in the
    ## specificity-score cross-check below -- a real miss the first time round: AZ255-gr01 (a
    ## mitochondrial gene, see *A second disambiguation strategy*) had cleared naming_filters() and
    ## become Glut-Arco-3's actual DESeq2-picked name before this exclusion existed.
    excl <- unique(de_grp$feature)[!is_named_gene(unique(de_grp$feature))]
    nm_grp <- select_naming_genes(de_grp, filters = filters, exclude_genes = excl)

    for (cl in grp_clusters) {
      gene <- nm_grp$naming_genes[nm_grp$cluster == cl]
      row <- which(named$cluster == cl)
      if (length(gene) == 1 && !is.na(gene)) {
        named$new_label[row] <- paste0(b, "_", gene)
        named$marker_disambiguated[row] <- TRUE
      }
    }
  }
  list(named = named, de_tables = de_tables)
}

## The five pseudobulk DESeq2 fits, cached: see the note on `redo` in the params chunk. A bare
## file-existence check would happily hand back a stale result if the exclusion lists ever change
## (e.g. Glut-Arco-4 un-dropped) without redo$disambiguate being flipped -- exactly the silent-
## staleness failure mode knitr's own cache was rejected for above, so this cache is not exempt
## from the same check: it also stores which *input* clusters it was computed over and refuses
## itself if that no longer matches.
## Two distinct ways this cache can be stale, checked separately: the *input* changed (cluster
## sets differ -- guarded before), or the *schema* changed (an older cache predates a field the
## current code now expects to read, e.g. de_tables, added after some caches already existed).
## The second one bit immediately: an old cache validated on cluster sets alone, silently handed
## back a NULL de_tables, and the first read of it errored several chunks later with
## "no applicable method for 'filter' applied to ... NULL" -- a schema check catches this at the
## cache itself instead of wherever the missing field happens to first get used.
required_fields <- c("glut_named", "gaba_named", "de_tables", "glut_clusters", "gaba_clusters")
cache_valid <- FALSE
if (!redo$disambiguate && file.exists(disambig_cache_fname)) {
  cached <- qs_read(disambig_cache_fname, nthreads = 8)
  schema_ok <- all(required_fields %in% names(cached))
  cache_valid <- schema_ok &&
    setequal(cached$glut_clusters, glut_named$cluster) &&
    setequal(cached$gaba_clusters, gaba_named$cluster)
  if (!schema_ok) {
    warning("Disambiguation cache predates a field the current code expects (",
            paste(setdiff(required_fields, names(cached)), collapse = ", "), ") -- recomputing.",
            call. = FALSE)
  } else if (!cache_valid) {
    warning("Disambiguation cache does not match the current eligible cluster sets -- recomputing.",
            call. = FALSE)
  }
}

if (cache_valid) {
  glut_named <- cached$glut_named
  gaba_named <- cached$gaba_named
  de_tables <- cached$de_tables
} else {
  glut_clusters_in <- glut_named$cluster
  gaba_clusters_in <- gaba_named$cluster
  glut_result <- disambiguate_with_markers(obj, cluster_col, glut_named, disambig_filters)
  gaba_result <- disambiguate_with_markers(obj, cluster_col, gaba_named, disambig_filters)
  glut_named <- glut_result$named
  gaba_named <- gaba_result$named
  ## Keyed by collision group (e.g. "Glut-CACNA1H_LHX9"), one full gene x cluster DE table each --
  ## this is where they live now; previously nowhere, discarded inside the function that computed
  ## them. Read this to see *why* a cluster got the gene it did, or why one (Glut-Arco-1) got none.
  de_tables <- c(glut_result$de_tables, gaba_result$de_tables)
  qs_save(list(glut_named = glut_named, gaba_named = gaba_named, de_tables = de_tables,
              glut_clusters = glut_clusters_in, gaba_clusters = gaba_clusters_in),
          disambig_cache_fname, nthreads = 8)
}
```

### Where are the DE results stored?

In `de_tables` (in-session) and in `disambig_cache_fname` on disk (`disambiguation_cache.qs2`,
loaded via `qs2::qs_read()`) -- a named list, one entry per collision group, each the full
`cluster_de()` output (every gene x every cluster in that group: `logFC`, `padj`, `pct_in`,
`pct_out`). Before this change they were computed inside `disambiguate_with_markers()`'s loop, fed
straight into `select_naming_genes()`, and discarded -- nothing to go back and look at. Example,
Glut-Arco-1's own group, to see why nothing cleared `naming_filters()` for it specifically:

```{r}
#| label: de-tables-example
de_tables[["Glut-CACNA1H_LHX9"]] %>%
  filter(cluster == "Glut-Arco-1") %>%
  arrange(padj) %>%
  head(10) %>%
  kable(digits = 3, caption = "Glut-Arco-1's own top candidates by padj, none clearing naming_filters()")
```

```{r}
#| label: disambig-report

kable(bind_rows(glut_named, gaba_named) %>% filter(grepl("_[0-9]+$", new_label) | marker_disambiguated) %>%
        select(cluster, new_label, marker_disambiguated, confidence_tier),
      caption = paste(
        "Every cluster that started in a collision group: marker_disambiguated = TRUE means a real",
        "gene replaced the numeral; FALSE means no marker cleared naming_filters() and the numeric",
        "fallback stands."
      ))

## Glut-Nido-1's fold target: whatever Glut-NC-1 itself was just assigned (now possibly a
## marker-disambiguated name, not a hardcoded string) -- keeps the two in sync if the composite
## table, the suffix/marker choice, or NC-1's own exemption ever changes again.
nc1_new_label <- glut_named$new_label[glut_named$cluster == "Glut-NC-1"]
stopifnot("Glut-NC-1 not found in glut_named -- check glut_eligible" = length(nc1_new_label) == 1)
manual_overrides <- c("Glut-Nido-1" = nc1_new_label, nido2_override)

lookup <- bind_rows(
  tibble(cluster = song_region_clusters, new_label = song_region_clusters, reason = "song region, unchanged"),
  tibble(cluster = names(precursor_overrides), new_label = unname(precursor_overrides),
         reason = "developmental stage (NSC -> NB -> Im), manual override"),
  tibble(cluster = names(gaba_precursor_overrides), new_label = unname(gaba_precursor_overrides),
         reason = "developmental stage (immature interneuron, matches Glut-Im), manual override"),
  tibble(cluster = excluded_clusters, new_label = NA_character_, reason = "already excluded upstream (confirmed artifact)"),
  tibble(cluster = names(manual_overrides), new_label = unname(manual_overrides), reason = "manual override"),
  glut_named %>%
    mutate(reason = if_else(marker_disambiguated,
                            paste0("composite top_rank_agg, disambiguated by own marker (", confidence_tier, ", conf=", confidence, ")"),
                            paste0("composite top_rank_agg (", confidence_tier, ", conf=", confidence, ")"))) %>%
    select(cluster, new_label, reason),
  gaba_named %>%
    mutate(reason = if_else(marker_disambiguated,
                            paste0("composite top_rank_agg, disambiguated by own marker (", confidence_tier, ", conf=", confidence, ")"),
                            paste0("composite top_rank_agg (", confidence_tier, ", conf=", confidence, ")"))) %>%
    select(cluster, new_label, reason),
  tibble(cluster = nonneuronal_clusters, new_label = nonneuronal_clusters, reason = "non-neuronal, unchanged")
) %>%
  arrange(cluster)

stopifnot(
  "lookup does not cover every cluster in the object exactly once" =
    setequal(lookup$cluster, all_clusters) && !anyDuplicated(lookup$cluster)
)

kable(lookup, caption = "Full old -> new label lookup, with the reason for every row")
```

## Where clusters merge, and where names are newly ambiguous or newly shared

```{r}
#| label: merges
## The fold: which original clusters now share one label.
lookup %>%
  filter(!is.na(new_label)) %>%
  count(new_label, name = "n_clusters") %>%
  filter(n_clusters > 1) %>%
  left_join(lookup %>% filter(!is.na(new_label)) %>% group_by(new_label) %>%
              summarise(from = paste(cluster, collapse = ", "), .groups = "drop"),
            by = "new_label") %>%
  kable(caption = paste(
    "Labels now shared by more than one original cluster. Expected: only the Glut-NC-1/Glut-Nido-1",
    "fold. The Arco trio and NC-1/NR-3 collisions should each appear as their own distinct",
    "marker- (or, failing that, numeral-) disambiguated labels here, never merged into one shared",
    "string -- a shared name for any of those would mean the disambiguation step missed a collision."
  ))

## Which collisions still fall back to a bare numeral -- i.e. no marker cleared naming_filters()
## for at least one member of that group. Not necessarily a problem (a real absence of a
## distinguishing gene is itself informative), but worth knowing before trusting the name as more
## than a positional tag.
lookup %>%
  filter(grepl("_[0-9]+$", new_label)) %>%
  kable(caption = "Clusters still on a numeric fallback -- no marker distinguished them from their group")
```

## A second disambiguation strategy: a specificity score, not a significance test

`disambiguate_with_markers()` above picks a gene by clearing `naming_filters()` on `cluster_de()`'s
pseudobulk padj/logFC -- a real hypothesis test, but one with as few as 2-3 replicates per cluster in
some of these groups, so a real distinguishing gene can fail to reach significance purely from
underpowered replication (`Glut-Arco-1`, with no marker at all, is the case to watch). `stats.R`'s
`calc_gene_specificity()` (Yanai et al. 2005's tissue-specificity index) is a genuinely different
kind of check: no p-value, no dispersion model, no replicate structure at all -- just how
concentrated a gene's mean expression is on one cluster among the group, computed directly from
average expression. It cannot replace the DESeq2 pick as the basis for a real claim (a specificity
score with no test behind it will always find *a* top gene, whether or not the difference is real),
but it is not gated by the same statistical power problem, so it is used here only as a cross-check --
agreement with the DESeq2 pick is reassuring, and a candidate where DESeq2 found nothing at all is a
lead worth a closer look, not an answer to substitute in silently.

**The first version of this had no detection-rate floor, and the result was unusable rather than
merely noisy**: every 2-cluster group's top candidate scored exactly 0.698 and every 3-cluster
group's exactly 1.102, regardless of which gene or which cluster -- the tell that the score had
saturated at its theoretical maximum (`log(n_clusters)`), which any gene detected in a single cell
of one cluster and nowhere else reaches by chance, because the score has no term for how many
cells support the pattern. Confirmed directly: `Glut-Arco-1`'s "winner" was detected in 1 of 1168
cells. `pct_min` below (mirroring `naming_filters()`'s own `pct_in_min`) excludes a gene from
ranking in a cluster unless it clears a real detection-rate floor there, which is what makes the
comparison below meaningful rather than a list of sequencing noise.

With the floor in place, the top candidates were real signal but still frequently unnamed
annotation placeholders (`LOC110472734`, `nbis-12`) -- not useful as a label even though the
detection rate behind them is genuine. `is_named_gene()` (defined in the `packages` chunk, so both
this cross-check and the DESeq2 disambiguation above can share one definition) excludes the three
placeholder patterns confirmed directly against this genome's annotation (`^LOC[0-9]` NCBI-style
loci, 3737 of 15523 genes; `^nbis-` NBIS-pipeline-annotated loci/pseudogenes, 45 genes; `-gr[0-9]+$`
a handful of uncharacterized `AZ255-gr01/02`-style entries) -- checked to not collide with real gene
symbols that happen to contain hyphens or digits (`NKX2-1`, `TRNAQ-CUG`, etc. all pass through
untouched).

`Glut-Arco-3`'s remaining top candidates after that filter were still dominated by housekeeping
genes (`ACTB`, `LDHB`, `UBB`, `HSPA8`...), which lines up with a real finding from the QC section
below: this cluster's %mito (0.65%) sits between the clean Arco clusters (0.05-0.08%) and the two
excluded artifact clusters (0.80%, 1.09%), and its pseudobulk profile correlates with both excluded
clusters more than the other Arco clusters do -- consistent with ambient/mitochondrial contamination
diluting real signal with generic "stressed cell" expression. `is_named_gene()` also excludes
ribosomal protein genes (`^RP[SL][0-9]` cytoplasmic and `^MRP[SL][0-9]` mitochondrial-ribosomal,
146 genes total -- checked to spare the `RPS6K*` kinase family, which the naive prefix would have
caught).

**The placeholder filter was added here first, and only here, which was a real miss**: this same
`Glut-Arco-3` had already been named `Glut-CACNA1H_LHX9_AZ255-gr01` by the DESeq2 disambiguation
step above, unaffected by any of this, because `disambiguate_with_markers()` did not exclude
placeholder genes at all. `AZ255-gr01` is not just an uncharacterized locus -- it is a mitochondrial
gene: this genome's chrM annotation (`config/paths.R`'s `GTF_LONSTR_BIOTYPE`, confirmed directly by
filtering `annotation_trans` to `seqnames == "chrM"`) gives every one of its 37 genes an anonymous
`nbis-#`/`AZ255-gr##`/`AZ255-gt##` placeholder ID rather than a real symbol, so the DESeq2 step had
picked a mitochondrial gene as `Glut-Arco-3`'s distinguishing "marker" -- exactly the kind of
contamination signal this whole line of investigation was trying to get out from under, not
independent confirmation that it was real. Fixed by passing `is_named_gene()` as `exclude_genes` into
`select_naming_genes()` inside `disambiguate_with_markers()` itself (see the `lookup` chunk above),
not only into this cross-check -- so a placeholder/mito/ribosomal gene can no longer become a
cluster's actual name via either route.

```{r}
#| label: specificity-score
source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "stats.R"))

## AverageExpression, not AggregateExpression -- this wants mean normalized expression per
## cluster (the scale calc_gene_specificity() expects), not the pseudobulk count sums DESeq2 needs.
##
## A detection-rate floor turned out not to be optional. The very first run of this (no floor)
## picked LOC116184446 for Glut-Arco-1 with specificity 0.698 -- checked directly against the
## object: that gene is detected in 1 of 1168 Arco-1 cells (mean expression 0.00086). The score has
## no term for how *many* cells support a gene's apparent specificity, so a gene hit in a single
## cell purely by chance and absent everywhere else scores at the theoretical maximum
## (log(n_clusters)), identical for every group of a given size regardless of which gene or
## cluster -- which is exactly the pattern that showed up (every 2-cluster group scoring 0.698,
## every 3-cluster group 1.102) and is what gave this away as noise rather than signal. pct_min
## mirrors naming_filters()'s own pct_in_min for the same reason it exists there.
specificity_candidates <- function(obj, cluster_col, clusters_in_group, assay = "SCT", top_n = 3,
                                   pct_min = 25) {
  cells <- colnames(obj)[obj@meta.data[[cluster_col]] %in% clusters_in_group]
  obj_grp <- subset(obj, cells = cells)
  obj_grp@meta.data[[cluster_col]] <- droplevels(factor(obj_grp@meta.data[[cluster_col]]))
  avg <- as.matrix(AverageExpression(obj_grp, assays = assay, group.by = cluster_col, layer = "data")[[assay]])
  avg <- avg[is_named_gene(rownames(avg)), , drop = FALSE]

  detect <- GetAssayData(obj_grp, assay = assay, layer = "counts") > 0
  groups_f <- obj_grp@meta.data[[cluster_col]]
  pct_in <- sapply(levels(groups_f), function(cl) Matrix::rowMeans(detect[, groups_f == cl, drop = FALSE])) * 100
  pct_in <- pct_in[rownames(avg), colnames(avg)]

  spec <- t(apply(avg, 1, calc_gene_specificity))
  colnames(spec) <- colnames(avg)
  spec[pct_in < pct_min] <- -Inf # excluded from ranking, not scored down -- avoids a false "tie"

  map_dfr(colnames(spec), function(cl) {
    ord <- order(spec[, cl], decreasing = TRUE)[seq_len(min(top_n, nrow(spec)))]
    tibble(cluster = cl, gene = rownames(spec)[ord], specificity = spec[ord, cl],
          pct_in_cluster = pct_in[ord, cl]) %>%
      dplyr::filter(is.finite(specificity)) # no candidate cleared pct_min for this cluster
  })
}

## Groups reused directly from de_tables (cluster_de()'s own `cluster` column enumerates exactly
## that group's members) rather than re-listed by hand, so this cannot silently drift from what
## disambiguate_with_markers() actually ran on.
disambig_groups <- map(de_tables, ~ unique(.x$cluster))

specificity_results <- imap_dfr(disambig_groups, function(clusters, grp) {
  specificity_candidates(obj, cluster_col, clusters) %>% mutate(group = grp)
})

top_specificity <- specificity_results %>%
  group_by(group, cluster) %>%
  slice_max(specificity, n = 1, with_ties = FALSE) %>%
  ungroup()

comparison <- bind_rows(glut_named, gaba_named) %>%
  filter(cluster %in% top_specificity$cluster) %>%
  transmute(cluster, deseq2_label = new_label, deseq2_gene = sub(".*_", "", new_label)) %>%
  left_join(top_specificity, by = "cluster")

kable(comparison, digits = 3, caption = paste(
  "DESeq2 pick vs. specificity-score top candidate, per disambiguated cluster. Agreement is a",
  "cross-check, not new evidence on its own; disagreement -- or a specificity candidate where the",
  "DESeq2 column is still a bare numeral -- is what is worth reading closely."
))
```

```{r}
#| label: apply
write_csv(lookup, lookup_fname)
message("Wrote ", lookup_fname)

obj$celltype_composite <- lookup$new_label[match(obj@meta.data[[cluster_col]], lookup$cluster)]

## Cells whose cluster was already excluded (Glut-Nido-3) get NA here, not a stale label --
## consistent with "already excluded", not silently re-included under a new name.
tibble(
  n_cells_total = ncol(obj),
  n_cells_labelled = sum(!is.na(obj$celltype_composite)),
  n_cells_excluded_artifact = sum(obj@meta.data[[cluster_col]] %in% excluded_clusters)
) %>%
  kable(caption = "Coverage of the new label")

qs_save(obj, obj_out_fname, nthreads = 8)
message("Wrote ", obj_out_fname)
```

## Dotplot: do the label genes actually look enriched where they name?

The point of this plot is to catch a labelling mistake, not to admire the labels: a gene pulled out
of `top_rank_agg` should show up as visibly higher in the row it names than everywhere else. Rows are
the *original* cluster identities (not the new composite label) so genes line up against a fixed,
familiar y-axis; columns are every distinct gene mentioned in any new label, ordered by which cluster
they name.

```{r}
#| label: label-genes
## Genes embedded in the new labels -- parsed the same way the labels themselves were built, not
## re-derived some other way, so this plot tests the actual rename rather than a re-interpretation
## of it. Manual overrides (Glut-GABA) and unchanged labels (song region, precursor, non-neuronal,
## the merge target's own pre-image) contribute no genes.
## Every gene-to-gene delimiter within a label is "_" (only the category prefix, e.g. "Glut-",
## keeps a hyphen) -- split on "_" alone, joining both multiple genes within one composite call
## (e.g. DACH2_MGAT4C) and the disambiguator (usually a real gene too, e.g. ...LHX9_NDNF, not the
## bare digit it would have been before the marker-analysis step). A label with no disambiguator
## collision has no trailing "_<token>" at all and is unaffected. A leftover pure-numeral token
## (only possible if a group's fallback stayed a bare number) is filtered out explicitly rather
## than warned about as a missing gene, since it never was one.
label_genes <- lookup %>%
  filter(cluster %in% c(glut_named$cluster, gaba_named$cluster)) %>%
  mutate(body = sub("^(Glut|GABA)-", "", new_label)) %>%
  separate_rows(body, sep = "_") %>%
  filter(!grepl("^[0-9]+$", body)) %>%
  distinct(cluster, gene = body)

genes_present <- intersect(unique(label_genes$gene), rownames(obj))
genes_missing <- setdiff(unique(label_genes$gene), rownames(obj))
if (length(genes_missing) > 0) {
  warning("Label genes not found in the object, dropped from the plot: ", paste(genes_missing, collapse = ", "),
          call. = FALSE)
}

## Row order: Glut clusters (song region first, then the composite-named ones grouped by which gene
## they were named for, then Nido/Pre), then GABA the same way -- so a gene's "home" row and the
## rows around it are visually adjacent rather than in whatever order factor() happened to produce.
glut_order <- c(song_region_clusters, precursor_clusters,
                glut_named %>% arrange(new_label) %>% pull(cluster),
                "Glut-Nido-1", "Glut-Nido-2")
gaba_order <- gaba_named %>% arrange(new_label) %>% pull(cluster)
plot_clusters <- intersect(c(glut_order, gaba_order), all_clusters) # excludes Glut-Nido-3

obj_plot <- subset(obj, cells = colnames(obj)[obj@meta.data[[cluster_col]] %in% plot_clusters])
obj_plot@meta.data[[cluster_col]] <- factor(obj_plot@meta.data[[cluster_col]], levels = plot_clusters)
```

```{r}
#| label: dotplot
#| fig-width: 16
#| fig-height: 10
DotPlot(obj_plot, features = genes_present, group.by = cluster_col, assay = "SCT") +
  RotatedAxis() +
  labs(title = "Label genes across Glut and GABA clusters (original cluster identity, pre-relabel)")
```

Same cells, same genes, same order -- just grouped by the label each cluster actually ends up with,
so the `Glut-NC-1`/`Glut-Nido-1` fold shows as one row (both were already adjacent above) and every
disambiguated pair reads by its own gene rather than its original cluster number.

```{r}
#| label: dotplot-renamed
#| fig-width: 16
#| fig-height: 10
new_label_order <- lookup$new_label[match(plot_clusters, lookup$cluster)]
new_label_order <- unique(new_label_order[!is.na(new_label_order)])
obj_plot$celltype_composite <- factor(obj_plot$celltype_composite, levels = new_label_order)

DotPlot(obj_plot, features = genes_present, group.by = "celltype_composite", assay = "SCT") +
  RotatedAxis() +
  labs(title = "Label genes across clusters, grouped by the new composite label")
```

## Question: does `Glut-Arco-4` show anomalous QC, explaining its low-confidence match?

`Glut-Arco-4` got the weakest composite call of anything actually relabelled (confidence 0.19,
"low" tier) -- a low match could mean either a genuinely novel population with no chicken
counterpart, or a technically poor cluster the integration correctly struggled to place. This checks
the latter using the same curation metrics built for `scripts/cluster_curation.R`, against its four
Arco siblings for reference.

This backup object has no `individual` column (it predates the souporcell reprocessing done earlier
this session), so `cluster_qc()`'s DE-dependent co-expression/class-panel scoring cannot run here --
that needs real pseudobulk DE, which needs real biological replicates, and faking one with e.g.
`sample_name` would pool unresolved birds together and produce a meaningless test (the same
limitation already documented in `glut_ex_style_naming.qmd`). It is not needed for this question
regardless: "does this cluster look anomalous" is answered by size, mito, depth, and doublet load,
none of which depend on DE, so those are computed directly from metadata instead.

```{r}
#| label: arco4-qc
arco_cells <- colnames(obj)[obj@meta.data[[cluster_col]] %in% paste0("Glut-Arco-", 1:5)]
qc_arco <- obj@meta.data[arco_cells, ] %>%
  as_tibble() %>%
  mutate(cluster = .data[[cluster_col]]) %>%
  group_by(cluster) %>%
  summarise(
    n_cells = n(),
    median_mito = median(perc_mito, na.rm = TRUE),
    median_nFeature = median(nFeature_RNA, na.rm = TRUE),
    frac_df_doublet = mean(DF_class == "Doublet", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(mito_mad = (median_mito - median(median_mito)) / stats::mad(median_mito))

kable(qc_arco, digits = 2, caption = "QC across the Arco family -- read Glut-Arco-4's row against its four siblings")
```

```{r}
#| label: arco4-verdict
verdict <- qc_arco %>% filter(cluster == "Glut-Arco-4")
tibble(
  n_cells_relative = verdict$n_cells / median(qc_arco$n_cells),
  mito_mad = verdict$mito_mad,
  frac_df_doublet_relative = verdict$frac_df_doublet / median(qc_arco$frac_df_doublet)
) %>%
  kable(digits = 2, caption = paste(
    "n_cells_relative: this cluster's size relative to the median Arco cluster --",
    "far below 1 means a small, thin cluster is the more likely explanation for a weak cross-",
    "species match than a genuinely unmatched population. mito_mad far from 0, or a doublet rate",
    "well above the sibling median, would point at a technical artefact instead."
  ))
```

## Question: does `Glut-Nido-2` form its own cluster against `GABA-1-1`/`GABA-1-2`?

The dual-transmitter hypothesis for `Glut-Nido-2` (the co-expression, depth, and cross-species
evidence gathered earlier all pointed the same way) makes a specific, checkable prediction: reclustered
together with genuine GABAergic cells, it should separate out as its own group rather than being
absorbed into either GABA population -- if it merges into GABA-1-1 or GABA-1-2, that is evidence it
*is* one of those populations rather than a distinct dual-identity type. `GABA-1-1`/`GABA-1-2`
specifically because that is what the composite integration's own runner-up call flagged
(`Inh_LGE_like_FOXP1_DGKH`, an LGE-like inhibitory type) as the closest inhibitory match.

```{r}
#| label: nido2-recluster
three_cells <- colnames(obj)[obj@meta.data[[cluster_col]] %in% c("Glut-Nido-2", "GABA-1-1", "GABA-1-2")]
obj_three <- subset(obj, cells = three_cells)
obj_three@meta.data[[cluster_col]] <- droplevels(factor(obj_three@meta.data[[cluster_col]]))

DefaultAssay(obj_three) <- "RNA"
obj_three <- obj_three %>%
  SCTransform(vars.to.regress = "perc_mito") %>%
  RunPCA(verbose = FALSE) %>%
  RunUMAP(reduction = "pca", dims = 1:15, reduction.name = "umap_three", verbose = FALSE) %>%
  FindNeighbors(reduction = "pca", dims = 1:15, verbose = FALSE) %>%
  FindClusters(resolution = 0.3, algorithm = 2, verbose = FALSE)

table(original = obj_three@meta.data[[cluster_col]], recluster = obj_three$seurat_clusters)
```

```{r}
#| label: nido2-plot
#| fig-width: 10
DimPlot(obj_three, group.by = cluster_col, reduction = "umap_three") + coord_equal() +
  labs(title = "Original identity")
DimPlot(obj_three, group.by = "seurat_clusters", reduction = "umap_three", label = TRUE) + coord_equal() +
  labs(title = "Reclustered together")
```

```{r}
#| label: nido2-verdict
## The check: does any recluster group hold Glut-Nido-2 cells with little admixture from either
## GABA population, or does Glut-Nido-2 split across groups that are otherwise dominated by
## GABA-1-1/GABA-1-2 cells?
tab <- table(original = obj_three@meta.data[[cluster_col]], recluster = obj_three$seurat_clusters)
comp <- as.data.frame(tab) %>%
  group_by(recluster) %>%
  mutate(frac = Freq / sum(Freq)) %>%
  ungroup() %>%
  filter(Freq > 0) %>%
  arrange(recluster, desc(frac))
kable(comp, digits = 2, caption = paste(
  "Composition of each recluster group. Glut-Nido-2 forming its own group (frac near 1 for one",
  "recluster label) supports a distinct population; Glut-Nido-2 spread across groups otherwise",
  "dominated by GABA-1-1/GABA-1-2 supports it being absorbed into one of them instead."
))
```

To pull this onto your own live object once the current preprocessing rerun finishes:

```r
lookup <- read_csv(here::here("snrna/naming/exploratory/apply_composite_labels", "composite_label_lookup.csv"))
obj$celltype_composite <- lookup$new_label[match(obj$cluster, lookup$cluster)]
```

## Session info

```{r}
#| label: session-info
sessionInfo()
```
