---
title: "Ex_-style aliases for the glutamatergic clusters"
format: html
editor: source
---

Applies the naming grammar validated in `zaremba_naming_validation.qmd` -- `Ex_<gene1>[_<gene2>]`,
from Zaremba et al. 2025 (*Science* 387:adp5182) -- to this project's own glutamatergic clusters, as
an alias next to the existing region-based names, not a replacement for them.

**What this does not do.** It does not recluster, does not rename `cluster`/`celltype`, and does not
claim that a generated `Ex_<gene>` alias denotes the same cell type Zaremba's identically-formatted
name does -- that is a cross-species homology claim, answered only by your own Xenium spatial
correspondences (supplied below as `xenium_overrides`) or by actually running the cross-species
integration in `snrna_zaremba-chk_integration_glut_sct_k10-f50-s30-d40.qmd`, not by string formatting.
Everywhere you have not supplied a confirmed correspondence, the alias is generated fresh from this
cluster's own top markers in *this* dataset -- informative, real, but not a Zaremba identity claim.

**Independent from the preprocessing/clustering/visualization notebook by design**, per your request:
this reads `obj_clustered.qs2` read-only and never sources or requires
`snrna_seurat_cellbender_preprocess.qmd`. It also does not write back into that file -- see
*Why not write onto `obj_clustered.qs2` directly* below.

**Nesting**, per the validation notebook's result: comparing a cluster only against its *region*
siblings (e.g. `Glut-HVC-1a` against `Glut-HVC-1`/`Glut-HVC-2` only, not against every other
glutamatergic cluster) found a real, uniquely-identifying marker for 69% of Zaremba's own supertypes
against 54% for a flat comparison across the whole population -- a large enough gap to make nesting
the default here rather than a coin flip. Read `zaremba_naming_validation.qmd` before trusting this
number in isolation: nesting was a wash on literally reproducing Zaremba's *specific* gene choices,
so it is adopted here for the "finds a real distinguishing marker" property, not because it recovers
their exact genes -- there is no ground truth for that on this dataset anyway, which is exactly why
that property is what matters here.

## Packages

```{r}
#| label: packages
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(knitr)

## Nothing here needs AnnotationHub/ensembldb -- no cross-species gene-ID mapping is required for
## this dataset's own gene symbols, so the dplyr-verb masking that bit zaremba_naming_validation.qmd
## does not apply. cluster_curation.R / cluster_naming.R still qualify every susceptible verb
## internally regardless, so this is robust either way.
source(here::here("scripts/cluster_curation.R"))
source(here::here("scripts/cluster_naming.R"))
```

## Parameters

```{r}
#| label: params
## EXPLORATORY -- not production. See ../README.md; the production filtering and naming live in
## ../hybrid_division_naming.qmd.
## The main clustered object, read-only, never written back to. It still has the old per-library
## `assignment` rather than `individual`/`batch`, and it still contains Glut-Nido-3, so
## drop_clusters below removes that artifact explicitly rather than silently naming it.
in_fname <- here::here("snrna/clustering/snrna_seurat_cellbender_preprocess",
                       "obj_clustered.qs2")
cluster_col <- "cluster" # set by snrna_seurat_cellbender_preprocess.qmd's `final_cluster`

## Excluded regardless of source object: confirmed artifact (see the manuscript-revision fix in
## snrna_seurat_cellbender_preprocess.qmd), a sink cluster with pathological SLC17A6/GAD1
## co-expression documented independently in xenium_label_transfer.qmd. Naming it would be naming
## something already slated for removal.
drop_clusters <- c("Glut-Nido-3")

out_dir <- here::here("snrna/naming/exploratory", "glut_ex_style_naming")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
alias_fname <- file.path(out_dir, "glut_ex_style_aliases.csv")
obj_out_fname <- file.path(out_dir, "obj_glut_ex_style.qs2")

filters <- naming_filters(max_genes = 2)

## Xenium-confirmed correspondences: fill these in from your spatial mapping session as you confirm
## them. Only clusters listed here get a Zaremba-attributed name; everything else gets a fresh,
## self-generated Ex_-style alias from its own markers (source = "generated from own markers").
## Example: xenium_overrides <- c("Glut-HVC-1" = "Ex_DACH2_SLIT2")
## character(), not c(): the latter evaluates to NULL, which zaremba_labels() now guards against,
## but writing the correctly-typed empty vector here doesn't rely on that guard.
xenium_overrides <- character()
override_source <- "xenium spatial mapping (fill in date/session)"
```

## Load

```{r}
#| label: load
stopifnot("input object not found at in_fname -- check the path above" = file.exists(in_fname))
obj <- qs_read(in_fname, nthreads = 8)
stopifnot(cluster_col %in% colnames(obj@meta.data))

## Glutamatergic, non-precursor cells only -- Glut-Pre-* are immature/undifferentiated, the same
## reason Ex_Pre_* was excluded from the Zaremba validation, and the same filter the existing
## snrna_zaremba-chk_integration...qmd notebook already applies to the bf side for the same reason.
## drop_clusters removes confirmed artifacts (Glut-Nido-3) regardless of what the source object
## still contains.
glut_cells <- colnames(obj)[grepl("^Glut", obj@meta.data[[cluster_col]]) &
                              !grepl("Pre", obj@meta.data[[cluster_col]]) &
                              !obj@meta.data[[cluster_col]] %in% drop_clusters]
obj_glut <- subset(obj, cells = glut_cells)
obj_glut@meta.data[[cluster_col]] <- droplevels(factor(obj_glut@meta.data[[cluster_col]]))

tibble(n_cells = ncol(obj_glut), n_clusters = n_distinct(obj_glut@meta.data[[cluster_col]])) %>%
  kable(caption = "Glutamatergic, non-precursor cells to be aliased")
```

## Region family: the nesting variable, parsed from your own labels

```{r}
#| label: region-family
## Raw region is parsed from the label itself (Glut-HVC-1a -> HVC, Glut-Nido-2 -> Nido,
## Glut-RA -> RA), not a separate metadata join. Reverted to nesting each region against itself --
## the two-group merge tried earlier (Arco+RA, HVC+NC+NR+Nido+LMAN) cost real naming quality (more
## unnamed clusters, real gene symbols replaced by uncharacterized loci and weaker multi-gene
## combinations) for a grouping that turned out not to be what was wanted region by region:
##
##   - NR, NC, Arco: back to separate, as originally.
##   - HVC, RA: separate, and should stay that way -- these are genuinely region-specific song
##     nuclei, so naming within-region is the right comparison, not a limitation to work around.
##   - Nido: separate. Also newly relevant here regardless of the region question: Nido-3 is
##     dropped (drop_clusters, above) as a confirmed artifact; Nido-2 looks like a real distinct
##     population (broadly distributed per-cell SLC17A6/GABA co-expression, not concentrated in one
##     library, unexplained by depth) rather than a Nido-3-style artifact or a stand-in for "Nido in
##     general" -- so it is named on its own here, not folded into a big cross-region comparison.
##   - LMAN: separate too, but treat its regional specificity with more caution than the others.
##     It is enriched in LMAN, but Xenium shows it more broadly distributed than that -- with the
##     caveat that the Xenium panel itself was designed from HVC/RA markers only, so it may simply
##     lack the resolution to detect true LMAN specificity, rather than LMAN specificity being
##     genuinely absent. Nested separately here because no alternative grouping was specified, not
##     because the region claim is as settled as HVC/RA's.
##
## region_group_map is kept as an explicit (currently identity) lookup, not removed, so a future
## merge is a one-line edit here rather than a rewrite -- and so this reasoning stays attached to
## the code that encodes it.
region_group_map <- c(
  Arco = "Arco", RA = "RA", HVC = "HVC", NC = "NC", NR = "NR", Nido = "Nido", LMAN = "LMAN"
)

raw_region <- function(cluster) {
  x <- sub("^Glut-", "", cluster)
  sub("-[0-9]+[a-z]?$", "", x)
}
region_family <- function(cluster) {
  raw <- raw_region(cluster)
  mapped <- region_group_map[raw]
  unname(ifelse(is.na(mapped), raw, mapped))
}

family_table <- tibble(cluster = levels(droplevels(factor(obj_glut@meta.data[[cluster_col]])))) %>%
  mutate(raw_region = raw_region(cluster), family = region_family(cluster)) %>%
  arrange(family, raw_region, cluster)

kable(family_table, caption = paste(
  "Parsed region and coarsened family per cluster -- check both columns before trusting the",
  "nesting below. A raw_region not in region_group_map (unexpected) would print as its own family."
))

family_table %>%
  count(family, name = "n_children") %>%
  kable(caption = "Clusters per family. A family of 1 has no sibling to nest against (see fallback below).")
```

## Generate aliases, nested by region family

```{r}
#| label: generate
sizes <- family_table %>% count(family, name = "n_children")
multi_families <- sizes %>% dplyr::filter(n_children > 1) %>% pull(family)
singleton_clusters <- family_table %>%
  dplyr::filter(family %in% (sizes %>% dplyr::filter(n_children == 1) %>% pull(family))) %>%
  pull(cluster)

## Nested: each cluster's markers chosen only against its region siblings.
nested_aliases <- map(multi_families, function(fam) {
  clusters_fam <- family_table %>% dplyr::filter(family == fam) %>% pull(cluster)
  cells_fam <- colnames(obj_glut)[obj_glut@meta.data[[cluster_col]] %in% clusters_fam]
  obj_fam <- subset(obj_glut, cells = cells_fam)
  zaremba_labels(obj_fam, cluster_col, overrides = xenium_overrides,
                override_source = override_source, filters = filters) %>%
    mutate(family = fam)
})

## Singleton families: nothing to nest against, so fall back to this cluster vs every other
## glutamatergic cluster in the dataset. Computed once, reused for every singleton rather than
## rerun per cluster -- it is the same flat DE regardless of which singleton is being named.
singleton_aliases <- if (length(singleton_clusters) > 0) {
  de_flat <- cluster_de(obj_glut, cluster_col)
  zaremba_labels(obj_glut, cluster_col, overrides = xenium_overrides,
                override_source = override_source, filters = filters, de = de_flat) %>%
    dplyr::filter(cluster %in% singleton_clusters) %>%
    left_join(family_table, by = "cluster")
} else {
  tibble()
}

aliases <- bind_rows(nested_aliases, singleton_aliases) %>%
  arrange(family, cluster)

kable(aliases, caption = paste(
  "celltype_ex_style per cluster. source distinguishes an override (your Xenium correspondence)",
  "from a name generated fresh from this cluster's own markers -- read it before treating any",
  "generated name as more than 'a real distinguishing gene', let alone a Zaremba identity claim."
))

tibble(
  n_clusters = nrow(aliases),
  n_overridden = sum(aliases$source == override_source),
  n_generated = sum(aliases$source == "generated from own markers"),
  n_unnamed = sum(aliases$source == "no marker cleared naming_filters()")
) %>%
  kable(caption = "Coverage")
```

## Unnamed clusters, if any

```{r}
#| label: unnamed
## A cluster with no marker clearing naming_filters() is not necessarily a bad cluster -- it may
## just be poorly separated from its region siblings at the current resolution. Worth reading beside
## scripts/cluster_curation.R's flags (no_unique_markers), which asks the same question for a
## different purpose (curation) using the same underlying select_naming_genes() call.
unnamed <- aliases %>% dplyr::filter(is.na(celltype_ex_style))
if (nrow(unnamed) > 0) {
  kable(unnamed %>% select(cluster, family, source), caption = "Clusters with no Ex_-style alias")
} else {
  message("Every cluster got an alias.")
}
```

## Write the lookup

```{r}
#| label: write-lookup
write_csv(aliases %>% select(cluster, family, celltype_ex_style, source, naming_genes,
                             n_naming_genes, unique_id),
          alias_fname)
message("Wrote ", alias_fname)
```

## Attach to the glutamatergic subset (a copy, not `obj_clustered.qs2`)

Why not write onto `obj_clustered.qs2` directly, given you asked for a metadata column on the
object: that file is owned by `snrna_seurat_cellbender_preprocess.qmd`, and this notebook is
independent of it by your request. If this notebook wrote back into that file, a later re-run of
the preprocessing notebook would silently overwrite the alias column with no record it had ever
been there, and "independent" would only be true in one direction. Instead this writes its own copy
of just the glutamatergic subset, with the alias attached, so nothing about the canonical object
changes as a side effect of running this one.

```{r}
#| label: attach
obj_glut <- AddMetaData(obj_glut, aliases %>% select(cluster, celltype_ex_style, source) %>%
                          column_to_rownames("cluster") %>%
                          .[obj_glut@meta.data[[cluster_col]], , drop = FALSE] %>%
                          `rownames<-`(colnames(obj_glut)))
qs_save(obj_glut, obj_out_fname, nthreads = 8)
message("Wrote ", obj_out_fname)
```

To pull the alias onto your own live object interactively instead:

```r
aliases <- read_csv(here::here("snrna/naming/exploratory/glut_ex_style_naming", "glut_ex_style_aliases.csv"))
obj$celltype_ex_style <- aliases$celltype_ex_style[match(obj$cluster, aliases$cluster)]
```

## Session info

```{r}
#| label: session-info
sessionInfo()
```
