---
title: "Zaremba Ex_-style naming: flat vs nested validation"
format: html
editor: source
---

**Exploratory — not production.** See `README.md`; the production cluster filtering and naming
live in `../hybrid_division_naming.qmd`.

Before adopting the `Ex_<gene1>[_<gene2>]` nomenclature from Zaremba et al. 2025 (*Science*
387:adp5182, "Developmental origins and evolution of pallial cell types and structures in birds")
for this project's own glutamatergic clusters, this notebook checks whether
`scripts/cluster_naming.R`'s DEG-based gene selection actually recovers *their* published gene
choices when run on *their own* object -- a ground-truth check with a known right answer, rather
than trusting the method on data where the right answer is unknown.

**Independent from the preprocessing/clustering/visualization notebook by design.** This reads only
the externally supplied Zaremba reference object
(`/hdd/sc_datasets/zaremba_gg_dev_2024/Gg_adult_snRNA_seq_srt.rds`) and this project's own
`scripts/cluster_curation.R` / `scripts/cluster_naming.R`. It does not read, source, or require
`snrna_seurat_cellbender_preprocess.qmd` or any of its outputs, and can be run at any point
regardless of where that notebook is in its own run. Applying whichever method wins here to this
project's actual glutamatergic clusters is a deliberate next step, not done in this notebook.

**The comparison, and why it's here at all.** Their metadata carries two annotation levels:
`anno_level_1` (7 excitatory subclasses, e.g. `Ex_DACH2_CALCR`) and `anno_level_3` (28 finer
supertypes, e.g. `Ex_DACH2_SLIT2`). A first pass (flat: every supertype's marker genes chosen by
comparing it against all 25 *other* supertypes at once, regardless of subclass) recovered their
literal gene choice for only 7-8 of 26 non-precursor supertypes, and 5 of the 6 total failures
(no gene cleared the naming filters at all) fell in the `DACH2` lineage specifically. The suspicion
this notebook tests: `DACH2_CALCR` and `DACH2_SV2C` are two *separate* six-child subclasses, and a
flat comparison drowns each supertype's distinguishing gene in comparisons against subclasses it was
never meant to be distinguished from. Their own naming more plausibly compared each supertype only
against its subclass siblings -- a **nested** comparison -- which this notebook runs directly and
scores the same way.

## Packages

```{r}
#| label: packages
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(AnnotationHub)
library(Matrix)
library(knitr)

## cluster_curation.R / cluster_naming.R qualify every dplyr verb susceptible to Bioconductor
## masking (filter/rename/select/desc) internally, so they are safe to source regardless of load
## order here -- AnnotationHub/ensembldb below attach exactly the generics (S4Vectors::rename,
## BiocGenerics::filter) that used to make an unqualified call inside a sourced file silently
## resolve to the wrong function. Source them after attaching everything else anyway, as a
## defensive habit, not because it is required now.
source(here::here("scripts/cluster_curation.R"))
source(here::here("scripts/cluster_naming.R"))
```

## Parameters

```{r}
#| label: params
zaremba_fname <- "/hdd/sc_datasets/zaremba_gg_dev_2024/Gg_adult_snRNA_seq_srt.rds"
stopifnot("Zaremba reference object not found" = file.exists(zaremba_fname))

out_dir <- here::here("snrna/naming/exploratory", "zaremba_naming_validation")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
obj_cache_fname <- file.path(out_dir, "obj_ex_symbols.qs2")

filters <- naming_filters(max_genes = 2)
redo <- list(load = FALSE)
```

## Load the reference and map to gene symbols

```{r}
#| label: load
## Excitatory, non-precursor cells only -- Ex_Pre_* are neuroblasts, not the mature types this
## nomenclature is being borrowed for.
if (!redo$load && file.exists(obj_cache_fname)) {
  obj_ex <- qs_read(obj_cache_fname, nthreads = 8)
} else {
  obj <- readRDS(zaremba_fname)
  obj$anno_level_1 <- as.character(obj$anno_level_1)
  obj$anno_level_3 <- as.character(obj$anno_level_3)
  ex_cells <- colnames(obj)[grepl("^Ex", obj$anno_level_1) & !grepl("Pre", obj$anno_level_1)]
  obj_ex <- subset(obj, cells = ex_cells)
  DefaultAssay(obj_ex) <- "RNA"

  ## Ensembl gene ID -> chicken gene symbol, same source as snrna_zaremba-chk_integration...qmd
  ## (AH53208, Ensembl 87 EnsDb for Gallus gallus). No finch ortholog step needed here: the
  ## comparison is against Zaremba's own chicken gene symbols, not a cross-species join.
  ah <- AnnotationHub()
  gg_db <- query(ah, c("Ensembl", "Gallus gallus", "EnsDb"))[["AH53208"]]
  ens_df <- AnnotationDbi::select(gg_db, keys = Features(obj_ex), c("SYMBOL", "GENENAME"), "GENEID")
  ens_df <- ens_df %>% dplyr::filter(!is.na(SYMBOL), SYMBOL != "")

  mat <- GetAssayData(obj_ex, assay = "RNA", layer = "counts")
  mat <- mat[ens_df$GENEID, ]
  ## Group-sum duplicate gene symbols via a sparse indicator matrix. Matrix.utils::aggregate.Matrix
  ## (used for this in the existing integration notebook) is archived on CRAN and unavailable
  ## through renv; this needs nothing beyond Matrix, already a hard Seurat dependency.
  grp <- factor(ens_df$SYMBOL)
  ind <- sparseMatrix(i = as.integer(grp), j = seq_along(grp), x = 1)
  mat <- ind %*% mat
  rownames(mat) <- levels(grp)

  obj_ex <- CreateSeuratObject(mat, meta.data = obj_ex@meta.data)
  obj_ex <- NormalizeData(obj_ex, verbose = FALSE)
  qs_save(obj_ex, obj_cache_fname, nthreads = 8)
}

tibble(n_cells = ncol(obj_ex), n_genes = nrow(obj_ex),
      n_supertypes = n_distinct(obj_ex$anno_level_3),
      n_subclasses = n_distinct(obj_ex$anno_level_1)) %>%
  kable(caption = "Excitatory, non-precursor cells and genes after symbol mapping")
```

## The nesting, read from their own metadata

```{r}
#| label: nesting
## Not assumed from string prefixes -- DACH2_CALCR and DACH2_SV2C are separate six-child subclasses
## whose children's names both start with "DACH2_", which is exactly what made the flat comparison
## fail on them. This table is the ground truth the nested loop below actually uses.
nesting <- obj_ex@meta.data %>%
  distinct(subclass = anno_level_1, supertype = anno_level_3) %>%
  arrange(subclass, supertype)

nesting %>%
  count(subclass, name = "n_children") %>%
  kable(caption = "Children per subclass. A subclass with 1 child has nothing to run DE against.")

kable(nesting, caption = "Full subclass -> supertype nesting")
```

## Scoring helper

```{r}
#| label: score-helper
## Case-insensitive gene-name comparison: the object's mapped gene symbols are not uniformly cased
## (e.g. one gene surfaced as "tac1" against the published "TAC1"), a mapping-table artefact, not a
## naming disagreement, so scoring must not let case decide a match.
score_against_published <- function(nm, published_col = "cluster") {
  nm %>%
    mutate(
      published = sub("^Ex_", "", .data[[published_col]]),
      published_genes = str_split(toupper(published), "_"),
      my_genes = str_split(toupper(coalesce(naming_genes, "")), "/"),
      overlap = map2_int(published_genes, my_genes, ~ length(intersect(.x, .y))),
      any_match = overlap > 0
    ) %>%
    dplyr::select(cluster = all_of(published_col), published, naming_genes, n_naming_genes,
           unique_id, overlap, any_match)
}
```

## Flat naming: every supertype against all 25 others

```{r}
#| label: flat
de_flat <- cluster_de(obj_ex, "anno_level_3", assay = "RNA")
nm_flat <- select_naming_genes(de_flat, filters = filters)
cmp_flat <- score_against_published(nm_flat) %>% mutate(method = "flat")

tibble(
  n_supertypes = nrow(cmp_flat),
  n_any_match = sum(cmp_flat$any_match),
  n_unique_id = sum(cmp_flat$unique_id),
  n_no_marker = sum(cmp_flat$n_naming_genes == 0)
) %>%
  kable(caption = "Flat comparison: every supertype scored against the entire excitatory population at once")
```

## Nested naming: each supertype against its subclass siblings only

```{r}
#| label: nested
## Subclasses with a single child cannot be tested by DE at all -- their "name" is definitionally
## the subclass itself, so score them as a trivial match rather than silently dropping them or
## running a meaningless one-group comparison.
subclass_sizes <- nesting %>% count(subclass, name = "n_children")
multi <- subclass_sizes %>% dplyr::filter(n_children > 1) %>% pull(subclass)
singleton <- subclass_sizes %>% dplyr::filter(n_children == 1) %>% pull(subclass)

nested_results <- map(multi, function(sc) {
  cells_sc <- colnames(obj_ex)[obj_ex$anno_level_1 == sc]
  obj_sc <- subset(obj_ex, cells = cells_sc)
  de_sc <- cluster_de(obj_sc, "anno_level_3", assay = "RNA")
  nm_sc <- select_naming_genes(de_sc, filters = filters)
  score_against_published(nm_sc) %>% mutate(method = "nested", subclass = sc)
})

nested_singleton <- map_dfr(singleton, function(sc) {
  st <- nesting %>% dplyr::filter(subclass == sc) %>% pull(supertype)
  tibble(cluster = st, published = sub("^Ex_", "", st), naming_genes = NA_character_,
        n_naming_genes = 0L, unique_id = TRUE, overlap = 1L, any_match = TRUE,
        method = "nested (trivial: only child of its subclass)", subclass = sc)
})

cmp_nested <- bind_rows(nested_results, nested_singleton)

tibble(
  n_supertypes = nrow(cmp_nested),
  n_any_match = sum(cmp_nested$any_match),
  n_unique_id = sum(cmp_nested$unique_id),
  n_no_marker = sum(cmp_nested$n_naming_genes == 0 & !grepl("trivial", cmp_nested$method))
) %>%
  kable(caption = "Nested comparison: every supertype scored only against its own subclass siblings")
```

## Head-to-head

```{r}
#| label: compare
side_by_side <- cmp_flat %>%
  dplyr::select(cluster, published, flat_genes = naming_genes, flat_match = any_match, flat_unique = unique_id) %>%
  left_join(
    cmp_nested %>% dplyr::select(cluster, subclass, nested_genes = naming_genes,
                          nested_match = any_match, nested_unique = unique_id),
    by = "cluster"
  ) %>%
  arrange(subclass, cluster)

kable(side_by_side, caption = paste(
  "Per-supertype comparison. flat_match / nested_match: did that method's chosen gene(s) include",
  "at least one of Zaremba's published genes for this supertype (case-insensitive)?"
))

tibble(
  method = c("flat", "nested"),
  n_any_match = c(sum(cmp_flat$any_match), sum(cmp_nested$any_match)),
  n_unique_id = c(sum(cmp_flat$unique_id), sum(cmp_nested$unique_id)),
  n_no_marker = c(sum(cmp_flat$n_naming_genes == 0),
                 sum(cmp_nested$n_naming_genes == 0 & !grepl("trivial", cmp_nested$method))),
  n_total = c(nrow(cmp_flat), nrow(cmp_nested))
) %>%
  kable(caption = "Headline comparison")

## The specific claim under test: does nesting fix the DACH2 lineage, where the flat method failed
## almost entirely?
side_by_side %>%
  dplyr::filter(grepl("DACH2", subclass)) %>%
  dplyr::select(subclass, cluster, flat_genes, flat_match, nested_genes, nested_match) %>%
  kable(caption = "DACH2_CALCR and DACH2_SV2C children specifically -- where the flat method broke")
```

## Conclusion

```{r}
#| label: conclusion
verdict <- tibble(
  flat_rate = sum(cmp_flat$any_match) / nrow(cmp_flat),
  nested_rate = sum(cmp_nested$any_match) / nrow(cmp_nested)
)
kable(verdict, digits = 3, caption = "Fraction of supertypes where the method's own top gene(s) matched a published gene")
```

Read the tables above before trusting either number in isolation: a higher match rate could still
be driven entirely by the DACH2 lineage (6+6 of 26 supertypes) rather than a general improvement,
and the `unique_id` columns matter independently of `any_match` -- a method can fail to name a
cluster the way Zaremba did while still finding *a* real, uniquely-identifying gene of its own, which
is a different (milder) kind of disagreement than finding no marker at all.

**What this notebook does not do.** It does not decide which method wins by fiat, and it does not
touch this project's own glutamatergic clusters. Whichever method scores better here is the one to
carry into a follow-up notebook that runs `zaremba_labels()` against the project's actual `obj_glut*`
objects -- additive, non-destructive, existing `celltype`/`cluster` columns untouched, exactly as
scoped for that step.

## Session info

```{r}
#| label: session-info
sessionInfo()
```
