---
title: "Untitled"
format: html
---


## Setup
```{r}
library(qs)
library(ArchR)
library(Seurat)
library(furrr)
library(tidyverse)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(scales)
library(ggrepel)
library(scCustomize)

source("~/nest/rstudio/utils/common_aesthetics.R")
pnames = names(position_colors)
names(position_colors) = case_when(pnames == "nido" ~ "nr",
                                     pnames == "ncl" ~ "nc",
                                     TRUE ~ pnames)

pnames2 = names(position_colors2)
names(position_colors2) = case_when(pnames2 == "nido" ~ "nr",
                                   pnames2 == "ncl" ~ "nc",
                                   TRUE ~ pnames2)
```


```{r}
proj_dir = "~/ssd/rstudio/multiome/motor-pathway/archr/motor-pathway_archr_processing_cluster-snrna-cr/"
analysis_prefix = "~/ssd/rstudio/multiome/motor-pathway/archr/motor-pathway_archr_processing_cluster-snrna-cr/"
out_name = "peaks"
out_dir = file.path(analysis_prefix, out_name)
dir.create(out_dir, recursive = T)

addArchRThreads(threads = 36) 
```

```{r}
proj = loadArchRProject(proj_dir)

```

## Add peaks
```{r}
#pathToMacs2 <- findMacs2()

if (! ("PeakMatrix" %in% getAvailableMatrices(proj))) {
  proj <- addReproduciblePeakSet(
    ArchRProj = proj, 
    threads = 1, 
    groupBy = "cluster",
    pathToMacs2 = '/home/brad/anaconda3/envs/scenicplus/bin/macs2',
    excludeChr = chrM,
    genomeSize = 1E9,
    plot=F,
    genomeAnnotation = genomeAnnotation,
    geneAnnotation = geneAnnotation
  )
  proj <- addPeakMatrix(proj)
  saveArchRProject(proj, load=F)
} 

```


### Export beds
```{r}
bed_dir = file.path(proj_dir, "beds")
dir.create(bed_dir)

bed_fname = file.path(bed_dir, "peak_set.bed")
gr = getPeakSet(proj)
rtracklayer::export(gr, bed_fname)
  
```

## Base composition of peaks
```{r}
library(BSgenome.LStriataDom.Brainard.lonStrDom2.ucsc)

gr_extend = trim(resize(gr, 2114, fix="center" ))
seqinfo_lsd = SeqinfoForBSGenome(BSgenome.LStriataDom.Brainard.lonStrDom2.ucsc)
gr_lsd = GRanges(seqnames = seqnames(seqinfo_lsd), ranges = IRanges(start=1, end=seqlengths(seqinfo_lsd)))
gr_extend = subsetByOverlaps(gr_extend, gr_lsd, type="within")
gr_seq = getSeq(BSgenome.LStriataDom.Brainard.lonStrDom2.ucsc, gr_extend)

gr_seq_alpha = alphabetFrequency(gr_seq)
sum(gr_seq_alpha[,"N"])

gr_seq_alpha_nind = gr_seq_alpha[,"N"]>0

sum(gr_seq_alpha_nind) / length(gr_seq_alpha_nind)

gr_filt = gr[!gr_seq_alpha_nind]

bed_fname = file.path(bed_dir, "peak_set_noN.bed")
rtracklayer::export(gr_filt, bed_fname)
```

## Differential peaks
```{r}
plan(multisession(workers = 4))
```


### Song-nonsong, Glut
```{r}
pair_list = list(ra= list("Glut-RA", "Glut-Arco-1"),
                 hvc_1 = list("Glut-HVC-1", "Glut-NC-1"),
                 hvc_2 = list("Glut-HVC-2", "Glut-NC-4")
                 )
```

```{r}
markers_peaks_sns_glut_se_fname = file.path(out_dir, "peaks_nonsong_glut_se.rds")
markers_peaks_sns_glut_fname = file.path(out_dir, "peaks_nonsong_glut.rds")

redo = F
if (redo) {
  
  marker_peaks_sns_glut = map(pair_list, \(pair_cur) {
      print(pair_cur)
      getMarkerFeatures(
        ArchRProj = proj, 
        useMatrix = "PeakMatrix", 
        groupBy = "cluster",
        useGroups = pair_cur[[1]],
        bgdGroups = pair_cur[[2]],
        bias = c("TSSEnrichment", "log10(nFrags)"),
        testMethod = "wilcoxon"
      )
  }) %>% set_names(names(pair_list))
  
  saveRDS(marker_peaks_sns_glut, markers_peaks_sns_glut_se_fname)
  
  marker_peaks_sns_glut_gr = map(marker_peaks_sns_glut, function(peaks_cur) {
    gr = getMarkers(peaks_cur, cutOff = "FDR <= 0.1", returnGR=T)
    gr
  })
  
  marker_peaks_sns_glut_df = map(marker_peaks_sns_glut_gr, as.data.frame) %>%
    bind_rows( .id="comparison")
  saveRDS(marker_peaks_sns_glut_df, markers_peaks_sns_glut_fname)
} else {
  marker_peaks_sns_glut = readRDS(markers_peaks_sns_glut_se_fname)
  marker_peaks_sns_glut_df = readRDS(markers_peaks_sns_glut_fname)
}

 marker_peaks_sns_glut_gr = map(marker_peaks_sns_glut, function(peaks_cur) {
    gr = getMarkers(peaks_cur, cutOff = "FDR <= 0.1", returnGR=T)
    gr
  })


```

```{r}
marker_peaks_sns_glut_gr_all = map(marker_peaks_sns_glut, function(peaks_cur) {
  gr = getMarkers(peaks_cur, cutOff = "FDR <= 1", returnGR=T)
  gr
})
marker_peaks_sns_glut_df_all = map(marker_peaks_sns_glut_gr_all, as.data.frame) %>%
  bind_rows( .id="comparison") %>%
  mutate(region = sprintf("%s:%s-%s", seqnames, start, end))
```

#### Compute average accessibility
```{r}
# Get summarized experiment of peak matrix aggregated by cluster
sePeak <- getGroupSE(
    ArchRProj = proj,
    useMatrix = "PeakMatrix",
    groupBy = "cluster"
)

rdata = rowData(sePeak) %>%
  as.data.frame() %>% 
  mutate(region = sprintf("%s:%s-%s", seqnames, start, end))
# Extract the average accessibility matrix (assays)
# This results in a matrix: Rows=Peaks, Columns=Clusters
averageAccessibility <- assay(sePeak)
rownames(averageAccessibility) = rdata$region
mat_avg = averageAccessibility
```

```{r}
# Estimate optimal k for k-means clustering of mat_sample_scale
# Using the same correlation distance as the heatmap
library(cluster)

set.seed(42)
idx = sample(nrow(mat_filter), min(nrow(mat_filter), 10000))
mat_sample = mat_filter[idx,]
mat_sample_scale = t(apply(mat_sample, 1, scale))
mat_test = mat_sample_scale
cor_dist_mat = as.dist(1 - cor(t(mat_test)))

k_range = 2:15
wss = numeric(length(k_range))
sil_avg = numeric(length(k_range))

set.seed(42)
for (i in seq_along(k_range)) {
  k = k_range[i]
  km = kmeans(mat_test, centers = k, nstart = 25, iter.max = 100)
  wss[i] = km$tot.withinss
  sil = silhouette(km$cluster, cor_dist_mat)
  sil_avg[i] = mean(sil[, "sil_width"])
}

par(mfrow = c(1, 2))
plot(k_range, wss, type = "b", pch = 19,
     xlab = "Number of clusters (k)", ylab = "Total within-cluster SS",
     main = "Elbow method")
plot(k_range, sil_avg, type = "b", pch = 19,
     xlab = "Number of clusters (k)", ylab = "Average silhouette width",
     main = "Silhouette method")
abline(v = k_range[which.max(sil_avg)], col = "red", lty = 2)

cat("Optimal k by silhouette:", k_range[which.max(sil_avg)], "\n")
```


#### Log2FC threshold
```{r}
mat_dar = marker_peaks_sns_glut_df_all %>%
  dplyr::select(comparison, Log2FC, region) %>%
  pivot_wider(names_from = "comparison", values_from = "Log2FC", values_fill = 0) %>%
  column_to_rownames("region") %>% as.matrix()
mat_dar = mat_dar[,c("hvc_1", "hvc_2", "ra")]

thresh = 2
mat_dar_keep = apply(mat_dar, 1, function(x) any(abs(x) >= thresh))
mat_dar_filter = mat_dar[mat_dar_keep,]

set.seed(42)
idx = sample(nrow(mat_dar_filter), min(nrow(mat_dar_filter), 1000))
mat_dar_sample = mat_dar_filter[idx,]

function(x) proxy::dist(x, method="correlation")
hm = Heatmap(mat_dar_sample,
             row_km = 6, 
             clustering_method_rows = "ward.D", 
             clustering_distance_rows = cor_dist, 
             cluster_columns = F,
             use_raster = TRUE)
hm
```

```{r}
ct_order = c("Glut-HVC-1", "Glut-NC-1", "Glut-HVC-2", "Glut-NC-4", "Glut-RA", "Glut-Arco-1")
mat_avg = mat_avg[,ct_order]

mat_avg_filter = mat_avg[names(mat_dar_keep)[mat_dar_keep],]

set.seed(42)
idx = sample(nrow(mat_avg_filter), min(nrow(mat_avg_filter), 1000))
mat_avg_sample = mat_avg_filter[idx,]

mat_avg_sample_scale = t(apply(mat_avg_sample, 1, scale))
colnames(mat_avg_sample_scale) = ct_order

nk = 5
km = kmeans(mat_avg_sample_scale, centers = nk)

function(x) proxy::dist(x, method="correlation")
hm = Heatmap(mat_avg_sample_scale,
             row_km = nk, 
             clustering_method_rows = "ward.D", 
             clustering_distance_rows = cor_dist, 
             cluster_columns = F,
             use_raster = TRUE)
hm
```


#### DAR
```{r fig.height = 10}
ct_order = c("Glut-HVC-1", "Glut-NC-1", "Glut-HVC-2", "Glut-NC-4", "Glut-RA", "Glut-Arco-1")
mat_avg = mat_avg[,ct_order]
#thresh = 1

fdr_thresh = 0.01
 marker_peaks_sns_glut_gr = map(marker_peaks_sns_glut, function(peaks_cur) {
    gr = getMarkers(peaks_cur, cutOff = sprintf("FDR <= %s", fdr_thresh), returnGR=T)
    gr
  })
  
marker_peaks_sns_glut_df = map(marker_peaks_sns_glut_gr, as.data.frame) %>%
  bind_rows( .id="comparison") %>%
  mutate(region = sprintf("%s:%s-%s", seqnames, start, end))

mat_filter = mat_avg[rownames(mat_avg) %in% marker_peaks_sns_glut_df$region,]

set.seed(42)
idx = sample(nrow(mat_filter), min(nrow(mat_filter), 1000))
mat_sample = mat_filter[idx,]

mat_sample_scale = t(apply(mat_sample, 1, scale))
colnames(mat_sample_scale) = ct_order
cor_dist = function(x) proxy::dist(x, method="correlation")


pal = brewer_pal(type="div", palette = 3 )(9)

min_val = min(mat_sample_scale)
max_val = max(mat_sample_scale)
cols = colorRamp2(c(min_val, 0, max_val), colors = c(pal[c(9,5,1)]))
width = 1
#height = width * nrow(mat) / ncol(mat)
height = 4

nk = 5

hm = Heatmap(mat_sample_scale,
             col = cols,
             row_km = nk, 
             clustering_method_rows = "ward.D", 
             clustering_distance_rows = cor_dist, 
             cluster_columns = F,
             column_names_gp = gpar(fontsize=6),
             show_row_names = F,
             width = unit(width, "in"),
             height = unit(height, "in"),
             use_raster = TRUE, raster_device_param = list(dpi=300))
hm

pdf(file.path(out_dir, sprintf("DAR_fdr%s_nk%s.pdf", fdr_thresh, nk)), height = height + 2, width = width + 2)
print(hm)
dev.off()
```

```{r}
mat_filter_scale = t(apply(mat_filter, 1, scale))
km = kmeans(mat_filter_scale, centers = nk)
```


### Calculate motif enrichment
```{r}

motifs_up_list = map(marker_peaks_sns_glut, function(peaks) {
    peakAnnoEnrichment(
      seMarker =  peaks,
      ArchRProj = proj,
      peakAnnotation = "MotifHS",
      cutOff = "FDR <= 0.01 & Log2FC >= .6"
    )
}) %>% set_names(names(marker_peaks_sns_glut))

motifs_up_df = map(motifs_up_list, function(x) {
  #map(peak_set, function(x) {
    df =  data.frame(TF = rownames(x), mlog10Padj = assay(x)[,1], enrichment = assay(x, "Enrichment")[,1])
    df = df[order(df$mlog10Padj, decreasing = TRUE),]
    df$rank <- seq_len(nrow(df))
    df
  }) %>% bind_rows(.id="pair")

```

```{r}
#plan(multisession(workers=length(markers_peaks_glut)))
motifs_down_list = map(marker_peaks_sns_glut, function(peaks) {
  #future_map(peaks, function(x) {
    peakAnnoEnrichment(
      seMarker =  peaks,
      ArchRProj = proj,
      peakAnnotation = "MotifHS",
      cutOff = "FDR <= 0.01 & Log2FC <= -0.6"
    )
  #}) 
}) %>% set_names(names(marker_peaks_sns_glut))

motifs_down_df = map(motifs_down_list, function(x) {
  #map(peak_set, function(x) {
    df =  data.frame(TF = rownames(x), mlog10Padj = assay(x)[,1], enrichment = assay(x, "Enrichment")[,1])
    df = df[order(df$mlog10Padj, decreasing = TRUE),]
    df$rank <- seq_len(nrow(df))
    df
  }) %>% bind_rows(.id="pair")
motifs_down_df = motifs_down_df %>%
  mutate(enrichment = -1 * enrichment)
```

```{r}
motifs_df = bind_rows(list(up = motifs_up_df, down = motifs_down_df), .id="direction")
motifs_df = motifs_df %>%
  mutate(Name = str_extract(TF, "[A-Z0-9]+")) %>%
  left_join(tfs)
```

### Plot motif enrichment

```{r fig.height = 10}
#motifs_df_cur = motifs_df %>% filter(pair == cur_set)

motifs_df_m = map(c("up", "down"), function(direction_cur) {
  motifs_df_cur = motifs_df %>% filter(direction == direction_cur)
  #browser()
  motifs_df_f = motifs_df_cur %>%
    group_by(pair) %>% 
    top_n(20, mlog10Padj)
  
  print(head(motifs_df_cur))
  motifs_df_m = motifs_df_cur %>%
    filter(TF %in% motifs_df_f$TF) %>% 
    select(pair, TF, enrichment) %>%
    pivot_wider(names_from = "pair", values_from = "enrichment", values_fill = 0) %>%
    column_to_rownames("TF") %>% as.matrix()
  #motifs_df_m = log2(motifs_df_m + 1)
  
  return(motifs_df_m)
}) %>% do.call(rbind, .)

pal = brewer_pal(type="div", palette = 3 )(9)
min_val = min(motifs_df_m)
max_val = max(motifs_df_m)
cols = colorRamp2(c(min_val, 0, max_val), colors = c(pal[c(9,5,1)]))

width = .6
height = width * nrow(motifs_df_m) / ncol(motifs_df_m)
height = 3
hm = ComplexHeatmap::Heatmap(motifs_df_m, col=cols, cluster_columns=F,
                             width = unit(width, "in"),
                             height = unit(height, "in")
                             )
print(hm)
pdf(file.path(out_dir, "song-glut_motifs_heatmap.pdf"))
print(hm)
dev.off()
```
##### Plot up and down separately
```{r}

top_motifs_n = 25
motifs_df_m = walk(c("up", "down"), function(direction_cur) {
  motifs_df_cur = motifs_df %>% filter(direction == direction_cur)
  #browser()
  motifs_df_f = motifs_df_cur %>%
    group_by(pair) %>% 
    top_n(top_motifs_n, mlog10Padj)
  
  print(head(motifs_df_cur))
  motifs_df_m = motifs_df_cur %>%
    filter(TF %in% motifs_df_f$TF) %>% 
    select(pair, TF, enrichment) %>%
    pivot_wider(names_from = "pair", values_from = "enrichment", values_fill = 0) %>%
    column_to_rownames("TF") %>% as.matrix()
  
  #pal = brewer_pal(type="div", palette = 3 )(9)
  min_val = min(motifs_df_m)
  max_val = max(motifs_df_m)
  
  if (direction_cur == "up") {
    pal = viridis_pal(option="C")(9)
  } else {
    pal = viridis_pal(option="E", direction = -1)(9)
  }
  cols = colorRamp2(c(min_val, max_val), colors = c(pal[c(1,9)]))
  
  width = .6
  height = width * nrow(motifs_df_m) / ncol(motifs_df_m)
  height = 3
  
  cor_dist = function(x) proxy::dist(x, method="correlation")
  
  hm = ComplexHeatmap::Heatmap(motifs_df_m, col=cols, 
                              # clustering_distance_rows = cor_dist,
                               cluster_columns=F,
                               row_names_gp = gpar(fontsize=6),
                               width = unit(width, "in"),
                               height = unit(height, "in")
  )
  print(hm)
  pdf(file.path(out_dir, str_interp("song-glut_motifs_${direction_cur}_top${top_motifs_n}_heatmap.pdf")), height = height + 2, width = width + 2)
  print(hm)
  dev.off()
})

```
#### Group by DBD
```{r}

top_motifs_n = 50
motifs_df_m = walk(c("up", "down"), function(direction_cur) {
  motifs_df_cur = motifs_df %>% filter(direction == direction_cur) 
  #browser()
  motifs_df_f = motifs_df_cur %>%
    group_by(pair) %>% 
    top_n(top_motifs_n, mlog10Padj)
  
  print(head(motifs_df_cur))
  motifs_df_m = motifs_df_cur %>%
    filter(TF %in% motifs_df_f$TF) %>% 
    filter(!is.na(DBD)) %>% 
    group_by(DBD, pair) %>% 
    summarize(enrichment = median(enrichment)) %>% 
    select(pair, DBD, enrichment) %>%
    pivot_wider(names_from = "pair", values_from = "enrichment", values_fill = 0) %>%
    column_to_rownames("DBD") %>% as.matrix()
  
  #pal = brewer_pal(type="div", palette = 3 )(9)
  min_val = min(motifs_df_m)
  max_val = max(motifs_df_m)
  
  if (direction_cur == "up") {
    pal = viridis_pal(option="C")(9)
  } else {
    pal = viridis_pal(option="E", direction = -1)(9)
  }
  cols = colorRamp2(c(min_val, max_val), colors = c(pal[c(1,9)]))
  
  width = .6
  height = width * nrow(motifs_df_m) / ncol(motifs_df_m)
  height = 3
  
  cor_dist = function(x) proxy::dist(x, method="correlation")
  
  hm = ComplexHeatmap::Heatmap(motifs_df_m, col=cols, 
                              # clustering_distance_rows = cor_dist,
                               cluster_columns=F,
                               row_names_gp = gpar(fontsize=6),
                               width = unit(width, "in"),
                               height = unit(height, "in")
  )
  print(hm)
  pdf(file.path(out_dir, str_interp("song-glut_motifs_dbd_${direction_cur}_top${top_motifs_n}_heatmap.pdf")), height = height + 2, width = width + 2)
  print(hm)
  dev.off()
})

```
```{r fig.height = 10}
#motifs_df_cur = motifs_df %>% filter(pair == cur_set)
motifs_up_df_f = motifs_up_df %>%
  group_by(pair) %>%
  top_n(20, mlog10Padj)

motifs_up_df_m = motifs_up_df %>% filter(TF %in% motifs_up_df_f$TF) %>% 
    select(pair, TF, enrichment) %>%
  pivot_wider(names_from = "pair", values_from = "enrichment", values_fill = 0) %>%
  # select(pair, TF, mlog10Padj) %>%
  # pivot_wider(names_from = "pair", values_from = "mlog10Padj", values_fill = 0) %>%
  column_to_rownames("TF") %>% as.matrix()
motifs_up_df_m = log2(motifs_up_df_m + 1)
#motifs_df_m[motifs_df_m==0] = NA
cols = viridisLite::viridis(9)

width = .6
height = width * nrow(motifs_up_df_m) / ncol(motifs_up_df_m)
hm = ComplexHeatmap::Heatmap(motifs_up_df_m, col=cols, cluster_columns=F,
                             width = unit(width, "in"),
                             height = unit(height, "in")
                             )
print(hm)
pdf(file.path(out_dir, "song-glut-enriched_motifs_heatmap.pdf"))
print(hm)
dev.off()
```

```{r fig.height=10, fig.width = 4}
motifs_df_f = motifs_df %>% filter(!grepl("FOS|JUN|SMARCC", TF)) %>%
  group_by(pair) %>%
  top_n(20, mlog10Padj)
motifs_df_m = motifs_df %>% 
  filter(TF %in% motifs_df_f$TF) %>% 
  select(pair, TF, enrichment) %>%
  pivot_wider(names_from = "TF", values_from = "enrichment", values_fill = 0) %>% column_to_rownames("pair") %>% as.matrix() %>% t()
cols = viridisLite::viridis(9)

width = .6
height = width * nrow(motifs_df_m) / ncol(motifs_df_m)
hm = ComplexHeatmap::Heatmap(motifs_df_m, col=cols,
                             width = unit(width, "in"),
                             height = unit(height, "in")
                             )
print(hm)
pdf(file.path(out_dir, "song-glut-enriched_motifs_no-fos-jun-smarcc_heatmap.pdf"))
print(hm)
dev.off()
```
```{r}
motifs_df_f = motifs_df %>% filter(!grepl("FOS|JUN|SMARCC", TF)) %>%
  group_by(pair) %>%
  filter(mlog10Padj>6) 

motifs_df_m = motifs_df %>% filter(TF %in% motifs_df_f$TF) %>% select(pair, TF, mlog10Padj) %>% pivot_wider(names_from = "TF", values_from = "mlog10Padj", values_fill = 0) %>% column_to_rownames("pair") %>% as.matrix() %>% t()
motifs_df_m = log2(motifs_df_m + 1)
cols = viridisLite::viridis(9)

width = .6
height = width * nrow(motifs_df_m) / ncol(motifs_df_m)
hm = ComplexHeatmap::Heatmap(motifs_df_m, col=cols, cluster_columns=F,
                             width = unit(width, "in"),
                             height = unit(height, "in")
                             )
print(hm)
pdf(file.path(out_dir, "song-glut-enriched_motifs_sig_heatmap.pdf"), height = height + 5, width = width + 5)
print(hm)
dev.off()
```

```{r}
proj_se_ge = getGroupSE(
  ArchRProj = proj,
  useMatrix = "GeneExpressionMatrix",
  groupBy = "cluster"
)
proj_df_ge = assay(proj_se_ge)
rownames(proj_df_ge) = getFeatures(proj, "GeneExpressionMatrix")
```




### HOMER motif enrichment (k-means clusters)
```{r}
# Build GRanges from km cluster assignments and export BED per cluster
km_df = data.frame(region = names(km$cluster), cluster = km$cluster)
km_gr = GRanges(km_df$region)
km_gr$cluster = km_df$cluster

homer_dir = file.path(out_dir, "homer_kmeans")
dir.create(homer_dir, recursive = TRUE)
genome_fa = "/mnt/nest/assembly/lonStrDom2/ncbi/GCF_005870125.1_lonStrDom2_genomic_ucsc_only.fna"

walk(seq_len(nk), function(k) {
  gr_k = km_gr[km_gr$cluster == k]
  bed_k = data.frame(seqnames(gr_k), start(gr_k), end(gr_k),
                     paste0(seqnames(gr_k), ":", start(gr_k), "-", end(gr_k)),
                     0, strand(gr_k))
  write_tsv(bed_k, file.path(homer_dir, sprintf("cluster_%d.bed", k)), col_names = FALSE)
})
```

```{r}
# Add HOMER to PATH
Sys.setenv(PATH = paste("/opt/homer/bin", Sys.getenv("PATH"), sep = ":"))

# Run HOMER findMotifsGenome.pl for each cluster
walk(seq_len(nk), function(k) {
  bed_k = file.path(homer_dir, sprintf("cluster_%d.bed", k))
  out_k = file.path(homer_dir, sprintf("cluster_%d", k))
  cmd = sprintf("findMotifsGenome.pl %s %s %s -size given -len 8,10,12 -p 24",
                bed_k, genome_fa, out_k)
  message(cmd)
  system(cmd)
})
```

```{r}
# Import HOMER knownResults.txt for each cluster
homer_motifs = map_dfr(seq_len(nk), function(k) {
  results_file = file.path(homer_dir, sprintf("cluster_%d", k), "knownResults.txt")
  if (file.exists(results_file)) {
    df = read_tsv(results_file, col_names = c("motif_name", "consensus", "pvalue",
                                               "log_pvalue", "qvalue",
                                               "n_target", "pct_target",
                                               "n_bg", "pct_bg"),
                  skip = 1, show_col_types = FALSE)
    df$cluster = k
    df
  }
})

homer_motifs = homer_motifs %>%
  mutate(pct_target = as.numeric(gsub("%", "", pct_target)),
         pct_bg = as.numeric(gsub("%", "", pct_bg)),
         tf = gsub("\\(.*", "", motif_name),
         enrichment = pct_target / pct_bg)
```

```{r}
# Top motifs per cluster
homer_top = homer_motifs %>%
  group_by(cluster) %>%
  top_n(10, -log_pvalue) %>%
  ungroup()

homer_top_mat = homer_motifs %>%
  filter(tf %in% homer_top$tf) %>%
  select(cluster, tf, enrichment) %>%
  pivot_wider(names_from = "cluster", values_from = "enrichment", values_fill = 0) %>%
  column_to_rownames("tf") %>% as.matrix()

hm_motifs = Heatmap(log2(homer_top_mat + 1),
                    name = "log2(enrichment+1)",
                    cluster_columns = FALSE,
                    column_labels = paste0("k", seq_len(ncol(homer_top_mat))))
hm_motifs
```

```{r}
# Import known motif matrices from knownResults/ directories
library(universalmotif)

homer_known_motifs = map_dfr(seq_len(nk), function(k) {
  motif_dir = file.path(homer_dir, sprintf("cluster_%d", k), "knownResults")
  motif_files = list.files(motif_dir, pattern = "^known\\d+\\.motif$", full.names = TRUE)
  map_dfr(motif_files, function(f) {
    lines = readLines(f)
    header = lines[1]
    mat_lines = lines[-1]
    mat = do.call(rbind, lapply(mat_lines, function(l) as.numeric(strsplit(trimws(l), "\\s+")[[1]])))
    colnames(mat) = c("A", "C", "G", "T")
    motif_rank = as.integer(gsub(".*known(\\d+)\\.motif", "\\1", f))
    motif_name_parsed = gsub("^>(\\S+).*", "\\1", header)
    tibble(cluster = k, rank = motif_rank, consensus = motif_name_parsed,
           motif_file = f, mat = list(mat))
  })
})

# Add rank to homer_motifs for joining
homer_motifs = homer_motifs %>%
  group_by(cluster) %>%
  mutate(rank = row_number()) %>%
  ungroup()
```

```{r}
# Filter known motifs by q-value threshold
known_qval_thresh = 1E-4

homer_known_motifs_filt = homer_known_motifs %>%
  inner_join(homer_motifs %>% select(cluster, rank, qvalue), by = c("cluster", "rank")) %>%
  filter(qvalue <= known_qval_thresh) %>%
  select(-qvalue)

# Convert filtered known motifs to universalmotif objects
umotifs_known = map(seq_len(nrow(homer_known_motifs_filt)), function(i) {
  row = homer_known_motifs_filt[i,]
  create_motif(t(row$mat[[1]]), type = "PPM",
               name = sprintf("k%d #%d: %s", row$cluster, row$rank, row$consensus))
})
```

```{r}
# Cluster known motifs by sequence similarity
sim_mat_known = compare_motifs(umotifs_known, method = "PCC")
sim_mat_known[!is.finite(sim_mat_known)] = 0  # treat non-comparable motifs as uncorrelated
motif_dist_known = as.dist(1 - sim_mat_known)
motif_hc_known = hclust(motif_dist_known, method = "average")
```

```{r}
plot(motif_hc_known, labels = FALSE)
```

```{r}
# Cut dendrogram to group similar known motifs
motif_group_known_h = 0.4  # PCC distance threshold; adjust as needed
motif_groups_known = cutree(motif_hc_known, h = motif_group_known_h)

motif_group_known_df = tibble(
  cluster = homer_known_motifs_filt$cluster,
  rank = homer_known_motifs_filt$rank,
  motif_group = motif_groups_known
)

# Remove old columns if re-running, then join
homer_motifs = homer_motifs %>%
  select(-any_of(c("motif_group", "motif_group_label"))) %>%
  left_join(motif_group_known_df, by = c("cluster", "rank"))

# Label each motif_group by the TF name of its most significant member
motif_group_known_labels = homer_motifs %>%
  filter(!is.na(motif_group)) %>%
  group_by(motif_group) %>%
  slice_min(log_pvalue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(motif_group, motif_group_label = tf)

homer_motifs = homer_motifs %>%
  select(-any_of("motif_group_label")) %>%
  left_join(motif_group_known_labels, by = "motif_group")
```

```{r fig.height=8, fig.width=6}
# Heatmap of known motif_group enrichment across clusters
motif_group_known_enrich = homer_motifs %>%
  filter(!is.na(motif_group), is.finite(enrichment)) %>%
  group_by(cluster, motif_group, motif_group_label) %>%
  summarize(enrichment = max(enrichment, na.rm = TRUE),
            log_pvalue = min(log_pvalue, na.rm = TRUE),
            .groups = "drop") %>%
  filter(is.finite(enrichment))

motif_group_known_enrich_mat = motif_group_known_enrich %>%
  mutate(label = sprintf("%d: %s", motif_group, motif_group_label)) %>%
  select(cluster, label, enrichment) %>%
  pivot_wider(names_from = "cluster", values_from = "enrichment", values_fill = 1) %>%
  column_to_rownames("label") %>% as.matrix()

Heatmap(log2(motif_group_known_enrich_mat),
        name = "log2(enrichment)",
        cluster_columns = TRUE,
        column_labels = paste0("k", seq_len(ncol(motif_group_known_enrich_mat))),
        row_names_gp = gpar(fontsize = 7))
```

```{r fig.height=8, fig.width=10}
# Similarity heatmap for known motifs
motif_names_known = map_chr(umotifs_known, function(m) m@name)
rownames(sim_mat_known) = motif_names_known
colnames(sim_mat_known) = motif_names_known

cluster_ids_known = homer_known_motifs_filt$cluster
names(cluster_ids_known) = motif_names_known
ha_known = HeatmapAnnotation(cluster = factor(cluster_ids_known),
                              col = list(cluster = setNames(
                                RColorBrewer::brewer.pal(nk, "Set1")[seq_len(nk)],
                                seq_len(nk))))

hm_known_sim = Heatmap(sim_mat_known,
        name = "PCC",
        cluster_rows = motif_hc_known,
        cluster_columns = motif_hc_known,
        top_annotation = ha_known,
        show_row_names = FALSE,
        show_column_names = FALSE)
hm_known_sim

pdf(file.path(out_dir, "known_motif_similarity_heatmap.pdf"), height = 10, width = 10)
draw(hm_known_sim)
dev.off()
```

```{r}
# Import HOMER homerResults.html (de novo) for each cluster
library(rvest)

homer_denovo = map_dfr(seq_len(nk), function(k) {
  html_file = file.path(homer_dir, sprintf("cluster_%d", k), "homerResults.html")
  if (file.exists(html_file)) {
    page = read_html(html_file)
    rows = page %>% html_nodes("tr") %>% tail(-1)  # skip header row
    map_dfr(seq_along(rows), function(i) {
      tds = rows[[i]] %>% html_nodes("td") %>% html_text(trim = TRUE)
      # Columns: Rank, Motif(SVG text), P-value, log P-value, % Targets, % Background, STD, Best Match, Motif File
      tibble(
        rank = as.integer(tds[1]),
        pvalue = tds[3],
        log_pvalue = as.numeric(tds[4]),
        pct_target = as.numeric(gsub("%", "", tds[5])),
        pct_bg = as.numeric(gsub("%", "", tds[6])),
        best_match = gsub("\\(.*", "", gsub("More Information.*", "", tds[8]))
      )
    }) %>% mutate(cluster = k)
  }
})

homer_denovo = homer_denovo %>%
  mutate(tf = trimws(best_match),
         enrichment = pct_target / pct_bg,
         motif_num = rank)
```

```{r}
# Top de novo motifs per cluster
homer_denovo_top = homer_denovo %>%
  group_by(cluster) %>%
  top_n(10, -log_pvalue) %>%
  ungroup()

homer_denovo_mat = homer_denovo %>%
  filter(tf %in% homer_denovo_top$tf) %>%
  group_by(cluster, tf) %>%
  slice_max(enrichment, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(cluster, tf, enrichment) %>%
  pivot_wider(names_from = "cluster", values_from = "enrichment", values_fill = 0) %>%
  column_to_rownames("tf") %>% as.matrix()

hm_denovo = Heatmap(log2(homer_denovo_mat + 1),
                    name = "log2(enrichment+1)",
                    cluster_columns = FALSE,
                    column_labels = paste0("k", seq_len(ncol(homer_denovo_mat))))
hm_denovo
```

```{r fig.height=3, fig.width=8}
# Import de novo motif matrices and display as sequence logos
library(universalmotif)

homer_denovo_motifs = map_dfr(seq_len(nk), function(k) {
  motif_dir = file.path(homer_dir, sprintf("cluster_%d", k), "homerResults")
  motif_files = list.files(motif_dir, pattern = "^motif\\d+\\.motif$", full.names = TRUE)
  map_dfr(motif_files, function(f) {
    lines = readLines(f)
    header = lines[1]
    mat_lines = lines[-1]
    mat = do.call(rbind, lapply(mat_lines, function(l) as.numeric(strsplit(trimws(l), "\\s+")[[1]])))
    colnames(mat) = c("A", "C", "G", "T")
    motif_num = as.integer(gsub(".*motif(\\d+)\\.motif", "\\1", f))
    consensus = gsub("^>(\\S+).*", "\\1", header)
    best_guess = gsub(".*BestGuess:", "", header)
    best_guess = gsub("\\s.*", "", best_guess)
    tibble(cluster = k, motif_num = motif_num, consensus = consensus,
           best_guess = best_guess, motif_file = f,
           mat = list(mat))
  })
})

# Add consensus and best_guess labels to homer_denovo
# homer_denovo = homer_denovo %>%
#   left_join(homer_denovo_motifs %>% select(cluster, motif_num, consensus, best_guess),
#             by = c("cluster", "motif_num"))

# Filter de novo motifs by p-value threshold
denovo_pval_thresh = 0.01

homer_denovo_motifs_filt = homer_denovo_motifs %>%
  inner_join(homer_denovo %>% select(cluster, motif_num, pvalue, log_pvalue),
             by = c("cluster", "motif_num")) %>%
  filter(as.numeric(pvalue) <= denovo_pval_thresh) %>%
  select(-pvalue, -log_pvalue)

# Display top motifs per cluster
homer_denovo_top_motifs = homer_denovo_motifs_filt %>%
  group_by(cluster) %>%
  filter(motif_num <= 10) %>%
  ungroup()

# Convert filtered motifs to universalmotif objects
umotifs_all = map(seq_len(nrow(homer_denovo_motifs_filt)), function(i) {
  row = homer_denovo_motifs_filt[i,]
  create_motif(t(row$mat[[1]]), type = "PPM",
               name = sprintf("k%d #%d: %s", row$cluster, row$motif_num, row$best_guess))
})
```

```{r fig.height=10, fig.width=12}
# Cluster all motifs by sequence similarity
sim_mat = compare_motifs(umotifs_all, method = "PCC")
sim_mat[!is.finite(sim_mat)] = 0
motif_dist = as.dist(1 - sim_mat)
motif_hc = hclust(motif_dist, method = "average")

# Plot motifs ordered by clustering
view_motifs(umotifs_all[motif_hc$order], dedup.names = TRUE)
```

```{r fig.height=8, fig.width=10}
# Similarity heatmap with dendrogram
motif_names = map_chr(umotifs_all, function(m) m@name)
rownames(sim_mat) = motif_names
colnames(sim_mat) = motif_names

# Color by cluster for annotation
cluster_ids = homer_denovo_motifs_filt$cluster
names(cluster_ids) = motif_names
ha = HeatmapAnnotation(cluster = factor(cluster_ids),
                        col = list(cluster = setNames(
                          RColorBrewer::brewer.pal(nk, "Set1")[seq_len(nk)],
                          seq_len(nk))))

Heatmap(sim_mat,
        name = "PCC",
        cluster_rows = motif_hc,
        cluster_columns = motif_hc,
        top_annotation = ha,
        column_names_gp = gpar(fontsize = 6),
        row_names_gp = gpar(fontsize = 6))
```

```{r}
plot(motif_hc, labels=F )
```

```{r}
# Cut dendrogram to group similar motifs
motif_group_h = 0.3  # PCC distance threshold; adjust as needed
motif_groups = cutree(motif_hc, h = motif_group_h)

# Build lookup from all motifs to their motif group
motif_group_df = tibble(
  cluster = homer_denovo_motifs_filt$cluster,
  motif_num = homer_denovo_motifs_filt$motif_num,
  motif_group = motif_groups
)

# Remove old motif_group columns if re-running
homer_denovo = homer_denovo %>%
  select(-any_of(c("motif_group", "motif_group_label")))

# Join motif_group into homer_denovo
homer_denovo = homer_denovo %>%
  left_join(motif_group_df, by = c("cluster", "motif_num"))
```

```{r}
# Label each motif_group by the best_guess of its most significant member
motif_group_labels = homer_denovo %>%
  filter(!is.na(motif_group)) %>%
  group_by(motif_group) %>%
  slice_min(log_pvalue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(motif_group, motif_group_label = best_match)

homer_denovo = homer_denovo %>%
  select(-any_of("motif_group_label")) %>%
  left_join(motif_group_labels, by = "motif_group")
```

```{r fig.height=8, fig.width=6}
# Heatmap of motif_group enrichment across clusters
# For each cluster x motif_group, take the max enrichment
motif_group_enrich = homer_denovo %>%
  filter(!is.na(motif_group), is.finite(enrichment)) %>%
  group_by(cluster, motif_group, motif_group_label) %>%
  summarize(enrichment = max(enrichment, na.rm = TRUE),
            log_pvalue = min(log_pvalue, na.rm = TRUE),
            .groups = "drop") %>%
  filter(is.finite(enrichment))

motif_group_enrich_mat = motif_group_enrich %>%
  mutate(label = sprintf("%d: %s", motif_group, motif_group_label)) %>%
  select(cluster, label, enrichment) %>%
  pivot_wider(names_from = "cluster", values_from = "enrichment", values_fill = 1) %>%
  column_to_rownames("label") %>% as.matrix()

Heatmap(log2(motif_group_enrich_mat),
        name = "log2(enrichment)",
        cluster_columns = TRUE,
        column_labels = paste0("k", seq_len(ncol(motif_group_enrich_mat))),
        row_names_gp = gpar(fontsize = 7))
```

#### Upset

```{r}
library(ComplexUpset)
```

```{r}
marker_peaks_sns_glut_df = marker_peaks_sns_glut_df %>%
  mutate(region = sprintf("%s:%s-%s", seqnames, start, end))
marker_peaks_sns_glut_w = marker_peaks_sns_glut_df %>%
  dplyr::select(-Log2FC, -group_name, -MeanDiff, -FDR) %>%
  pivot_wider(names_from = "comparison", values_from = "group", values_fill = 0)
```

```{r}
comparisons = unique(marker_peaks_sns_glut_df$comparison)
upset(marker_peaks_sns_glut_w, comparisons , name='', width_ratio=0.1)
```
```{r}
marker_peaks_sns_glut2_gr = map(marker_peaks_sns_glut, function(peaks_cur) {
  gr = getMarkers(peaks_cur, cutOff = "FDR < 0.01 & Log2FC >= 0.6", returnGR=T)
  gr
})

marker_peaks_sns_glut2_df = map(marker_peaks_sns_glut2_gr, as.data.frame) %>%
  bind_rows( .id="comparison")
```

```{r}
marker_peaks_sns_glut2_df = marker_peaks_sns_glut2_df %>%
  mutate(region = sprintf("%s:%s-%s", seqnames, start, end))
marker_peaks_sns_glut2_w = marker_peaks_sns_glut2_df %>%
  dplyr::select(-Log2FC, -group_name, -MeanDiff, -FDR) %>%
  pivot_wider(names_from = "comparison", values_from = "group", values_fill = 0)
```

```{r}
comparisons = unique(marker_peaks_sns_glut2_df$comparison)
upset(marker_peaks_sns_glut2_w, comparisons , name='', width_ratio=0.1, base_annotations=list(
  'Intersection size'=intersection_size(text_mapping=aes(label=!!upset_text_percentage()))
))
```


#### Union song enriched
```{r}
marker_peaks_glut_union = Reduce(GenomicRanges::union, marker_peaks_glut_gr)
```


#### Export beds
```{r}
bed_dir = file.path(out_dir, "beds")
dir.create(bed_dir)

walk(seq_along(marker_peaks_glut_df), function(i) {
  d = marker_peaks_glut_df[[i]]
  pair_cur = pairs_full[[i]]
  pair_cur_name = sprintf("%s_vs_%s", pair_cur[[1]], paste(pair_cur[[2]], collapse="_"))
  d_out = d %>% 
    arrange(seqnames, start) %>% 
    mutate(sign = sign(Log2FC)) %>%
    mutate(name = sprintf("%s:%s-%s", seqnames, start, end))
  
  signs = unique(d_out$sign)
  walk(unique(signs), function(sign_cur) {
    d_out_cur = d_out %>%
      filter(sign == sign_cur) %>% 
      select(seqnames, start, end, name, Log2FC, strand)
    fname = file.path(bed_dir, sprintf("%s_%s.bed", pair_cur_name, sign_cur))
    write_delim(d_out_cur, fname, delim="\t", col_names = F )
  })
  
  #cluster_cur = unique(d$cluster)
  

})

d = as.data.frame(marker_peaks_glut_inter)
d_out = d %>% 
    arrange(seqnames, start) %>% 
    mutate(name = sprintf("%s:%s-%s", seqnames, start, end)) %>%
    select(seqnames, start, end, name)
  #cluster_cur = unique(d$cluster)
  fname = file.path(bed_dir, sprintf("%s.bed", "Glut-RA_Glut-HVC-1_Glut-HVC-2_intersect"))
  write_delim(d_out, fname, delim="\t", col_names = F )
  
  
d = as.data.frame(marker_peaks_glut_union)
d_out = d %>% 
    arrange(seqnames, start) %>% 
    mutate(name = sprintf("%s:%s-%s", seqnames, start, end)) %>%
    select(seqnames, start, end, name)
  #cluster_cur = unique(d$cluster)
  fname = file.path(bed_dir, sprintf("%s.bed", "Glut-RA_Glut-HVC-1_Glut-HVC-2_union"))
  write_delim(d_out, fname, delim="\t", col_names = F )
  
```

### GABA
```{r}
proj$cluster = as.character(proj$cluster) ## Required for getMarkerFeatures or will error
  
```

```{r}
pair_list = list(gaba21_gaba11= list("GABA-2-1", "GABA-1-1"),
                 gaba41_gaba11 = list("GABA-4-1", "GABA-1-1")
                 )

#names(pair_list) = map_chr(pair_list, function(x) paste(x, collapse="_"))
```

```{r}
markers_peaks_gaba_se_fname = file.path(out_dir, "peaks_gaba_se.rds")
markers_peaks_gaba_fname = file.path(out_dir, "peaks_gaba.rds")

redo = F
if (redo) {
  
  marker_peaks_gaba = map(pair_list, \(pair_cur) {
      print(pair_cur)
      getMarkerFeatures(
        ArchRProj = proj, 
        useMatrix = "PeakMatrix", 
        groupBy = "cluster",
        useGroups = pair_cur[[1]],
        bgdGroups = pair_cur[[2]],
        bias = c("TSSEnrichment", "log10(nFrags)"),
        testMethod = "wilcoxon"
      )
  }) %>% set_names(names(pair_list))
  
  saveRDS(marker_peaks_gaba, markers_peaks_gaba_se_fname)
  
  marker_peaks_gaba_gr = map(marker_peaks_gaba, function(peaks_cur) {
    gr = getMarkers(peaks_cur, cutOff = "FDR <= 0.1", returnGR=T)
    gr
  })
  
  marker_peaks_gaba_df = map(marker_peaks_gaba_gr, as.data.frame) %>%
    bind_rows( .id="comparison")
  saveRDS(marker_peaks_gaba_df, markers_peaks_gaba_fname)
} else {
  marker_peaks_gaba = readRDS(markers_peaks_gaba_se_fname)
  marker_peaks_gaba_df = readRDS(markers_peaks_gaba_fname)
}

 marker_peaks_gaba_gr = map(marker_peaks_gaba, function(peaks_cur) {
    gr = getMarkers(peaks_cur, cutOff = "FDR <= 0.1", returnGR=T)
    gr
  })


```


### Astro
```{r}
pair_list = list(astro1_oligo = list("Astro-1", "Oligo"),
                 astro2_oligo = list("Astro-2", "Oligo")
                 )

#names(pair_list) = map_chr(pair_list, function(x) paste(x, collapse="_"))
```

```{r}
markers_peaks_astro_se_fname = file.path(out_dir, "peaks_astro_se.rds")
markers_peaks_astro_fname = file.path(out_dir, "peaks_astro.rds")

redo = F
if (redo) {
  
  marker_peaks_astro = map(pair_list, \(pair_cur) {
      print(pair_cur)
      getMarkerFeatures(
        ArchRProj = proj, 
        useMatrix = "PeakMatrix", 
        groupBy = "cluster",
        useGroups = pair_cur[[1]],
        bgdGroups = pair_cur[[2]],
        bias = c("TSSEnrichment", "log10(nFrags)"),
        testMethod = "wilcoxon"
      )
  }) %>% set_names(names(pair_list))
  
  saveRDS(marker_peaks_astro, markers_peaks_astro_se_fname)
  
  marker_peaks_astro_gr = map(marker_peaks_astro, function(peaks_cur) {
    gr = getMarkers(peaks_cur, cutOff = "FDR <= 0.1", returnGR=T)
    gr
  })
  
  marker_peaks_astro_df = map(marker_peaks_astro_gr, as.data.frame) %>%
    bind_rows( .id="comparison")
  saveRDS(marker_peaks_astro_df, markers_peaks_astro_fname)
} else {
  marker_peaks_astro = readRDS(markers_peaks_astro_se_fname)
  marker_peaks_astro_df = readRDS(markers_peaks_astro_fname)
}

 marker_peaks_astro_gr = map(marker_peaks_astro, function(peaks_cur) {
    gr = getMarkers(peaks_cur, cutOff = "FDR <= 0.1", returnGR=T)
    gr
  })


```

### Song-nonsong differences, same cluster
```{r}


proj$position_cluster = paste(proj$position, proj$cluster, sep="_")
proj$region_cluster = paste(proj$region, proj$cluster, sep="_")
proj$class = case_when(grepl("Glut", proj$cluster) ~ "Glut",
                       grepl("GABA", proj$cluster) ~ "GABA",
                       grepl("Pre", proj$cluster) ~ "Neurog",
                       TRUE ~ "Nonneural")
proj$position_class = paste(proj$position, proj$class, sep="_")
proj$region_class = paste(proj$region, proj$class, sep="_")

proj_md = proj@cellColData %>% as.data.frame() %>% distinct(position_cluster, cluster, position, region, region_cluster)
proj_md = proj_md %>%
  filter(!(region=="ARCO" & position=="nc"))  %>%
  filter(!(region=="ARCO" & position=="hvc"))

position_cluster_table = table(proj$position_cluster)
position_cluster_keep = names(position_cluster_table)[position_cluster_table>=20]
proj_md = proj_md %>% filter(position_cluster %in% position_cluster_keep)
position_clusters = split(proj_md$position_cluster, proj_md$region_cluster)
position_clusters = position_clusters[!grepl("Glut", names(position_clusters))]
position_clusters_lens = map_int(position_clusters, length)
position_clusters = position_clusters[position_clusters_lens==2]

position_clusters_unlist = unlist(position_clusters)
position_clusters_table = table(proj$position_cluster)
position_clusters_table = position_clusters_table[position_clusters_unlist]
position_clusters_ncells = min(position_clusters_table)

proj_md = proj@cellColData %>% as.data.frame() %>% distinct(position_class, region_class, region, position)
proj_md = proj_md %>%
  filter(!(region=="ARCO" & position=="nc")) %>%
  filter(!(region=="ARCO" & position=="hvc"))
position_class_table = table(proj$position_class)
position_class_keep = names(position_class_table)[position_class_table>20]
proj_md = proj_md %>% filter(position_class %in% position_class_keep)
position_class = split(proj_md$position_class, proj_md$region_class)
position_class_lens = map_int(position_class, length)
position_class = position_class[position_class_lens==2]

position_class_unlist = unlist(position_class)
position_class_table = table(proj$position_class)
position_class_table = position_class_table[position_class_unlist]
position_class_ncells = min(position_class_table)
to_compare = list(position_cluster=position_clusters, position_class=position_class)

```

```{r}
markers_peaks_sns_se_fname = file.path(out_dir, "peaks_nonsong_se.rds")
markers_peaks_sns_fname = file.path(out_dir, "peaks_nonsong.rds")

redo = F
if (redo) {
  
  marker_peaks_sns = imap(to_compare, \(list_cur, groupBy_cur) {
    future_map(list_cur, function(pair_cur) {
      print(pair_cur)
      getMarkerFeatures(
        ArchRProj = proj, 
        useMatrix = "PeakMatrix", 
        groupBy = groupBy_cur,
        useGroups = pair_cur[[1]],
        bgdGroups = pair_cur[[2]],
        bias = c("TSSEnrichment", "log10(nFrags)"),
        testMethod = "wilcoxon"
      )
    })
  })
  saveRDS(marker_peaks_sns, markers_peaks_sns_se_fname)
  
  marker_peaks_sns_gr = map(marker_peaks_sns, function(peaks_cur) {
    map(peaks_cur, function(x) {
      gr = getMarkers(x, cutOff = "FDR <= 0.1", returnGR=T)
      as.data.frame(gr)
    })
  })
  marker_peaks_sns_df = bind_rows(map(marker_peaks_sns_gr, function(x) bind_rows(x, .id="comparison")), .id="set")
  saveRDS(marker_peaks_sns_df, markers_peaks_sns_fname)
} else {
  marker_peaks_sns_df = readRDS(markers_peaks_sns_fname)
}




```

```{r}
marker_peaks_sns_n = marker_peaks_sns_df %>% group_by(comparison, set) %>%
  summarize(n=n())
```

```{r}
sets = unique(marker_peaks_sns_n$set)
marker_peaks_sns_split = split(marker_peaks_sns_n, marker_peaks_sns_n$set)
marker_peaks_sns_split[["position_class"]] = marker_peaks_sns_split[["position_class"]] %>%
  mutate(comparison = factor(comparison, levels=paste(c("ARCO", "NIDO"), rep(rev(c("Glut", "GABA", "Neurog", "Nonneural")), each=2), sep="_")))
```

```{r}

theme_common = theme(axis.text.y = element_text(size=6), axis.title = element_text(size=7))
iwalk(marker_peaks_sns_split, function(df_cur, set_cur) {
  gg = ggplot(df_cur, aes(comparison, n)) + 
    geom_bar(stat="identity") + 
    coord_flip() + 
    scale_y_continuous(labels=scales::scientific)
    theme_bw() + 
      theme(axis.text.x = element_text(size=6)) + 
    theme_common + 
    labs(y = "# of DARs", x="")
  print(gg)
  save_plot(file.path(out_dir, str_interp("DARs_song-vs-nonsong_${set_cur}_horiz.pdf")), gg, base_width=2, base_height=1)
  
  gg = ggplot(df_cur, aes(comparison, n)) + 
    geom_bar(stat="identity") + 
    scale_y_continuous(labels=scales::scientific) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5, size=6)) + 
    theme_common +
    labs(y = "# of DARs", x="")
  print(gg)
  save_plot(file.path(out_dir, str_interp("DARs_song-vs-nonsong_${set_cur}_vert.pdf")), gg, base_width=1.5, base_height=2)
})
```



### FSIs
```{r}
to_select = c("GABA-4-1_GABA-1-1", "GABA-2-1_GABA-1-1")
marker_peaks_gaba = marker_peaks_glut_df[to_select]

```

