---
title: "Multiome label transfer, hybrid snRNA labels"
format: html
editor: source
---

Repeats `multiome/seurat/seurat_preprocess.qmd`'s `## Import snRNA` curation strategy (its
`### Transfer labels` + `### Manually curate labels` sections) against the current hybrid-scheme
snRNA labels (`snrna/naming/hybrid_division_naming.qmd`, `celltype_hybrid`) instead of the snRNA
object's old raw `cluster` column that pipeline used.

**Two deliberate differences from the original, both requested directly:**

- **Separate script, separate object.** The original curation section ran in place inside the main
  preprocessing notebook and overwrote that notebook's own output object
  (`obj_clustered.qs`/`.h5seurat` under `MULTIOME_SEURAT_DIR`). This notebook reads that object
  read-only and writes its own copy (`obj_clustered_hybrid.qs2`, this directory) so the original
  multiome object is never touched.
- **Majority vote, not a hand-typed relabeling table.** The original's final step
  (`### Manually curate labels`) was a `case_when()` a human wrote by eyeballing the transfer
  concordance heatmap and a UMAP overlay -- a judgment call, not something derivable from the code
  or the data alone. Repeating the transfer mechanically here (same `FindTransferAnchors`/
  `TransferData` call, same reference/query roles) but automating the curation step as each
  multiome cluster's modal (most common) transferred label, printed in full below with its vote
  fraction so a low-confidence call is visible rather than silently accepted, and overridable via
  `cluster_hybrid_overrides` below without re-running the transfer.

**Votes are grouped by the multiome's own already-curated `cluster` column** (the pipeline's final,
hand-curated label), not a from-scratch reclustering. A from-scratch per-division reclustering was
tried first (matching `seurat_preprocess.qmd`'s exact resolutions/`idents_comb` tables) specifically
to avoid depending on that column's one known-artefactual entry (`Glut-Nido-3`, folded in by hand
from a separate raw cluster, `Glut-NC-3`) -- but the clustering pipeline itself has drifted since
this object was originally processed (package/algorithm updates), so a fresh reclustering no longer
reproduces the original's own clusters: no cells map to `HVCx`, `Arco` picks up an extra cluster, and
several GABA calls come out clearly wrong. Working from `cluster` avoids that drift; `Glut-Nido-3`
itself is confirmed artefactual on its own UMAP (see `## QC: Glut-Nido-3`) and excluded outright,
rather than folded into a vote.

The legacy `obj_clustered.qs` (this pipeline's format, from before this project standardized on
`qs2`) has already been converted to `obj_clustered.qs2` alongside it, by the same
`convert_qs_to_qs2.R` helper used elsewhere in `/ssd/brad/rstudio/multiome` -- a format-only,
read-only conversion (original `.qs` untouched); this notebook reads that `.qs2` copy.

```{r}
#| label: packages
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(ComplexHeatmap)
library(circlize)
library(here)

source(here::here("config/paths.R"))
source(here::here("config/figure_theme.R"))
```

## Parameters

```{r}
#| label: params
multiome_fname <- file.path(MULTIOME_SEURAT_DIR, "obj_clustered.qs2")
snrna_fname <- here::here("snrna/naming/hybrid_division_naming", "obj_hybrid_labels.qs2")

## Same ordering used throughout the hybrid-scheme figures (snrna/trees/celltypes_hclust_all_hybrid.qmd),
## repeated here (not sourced -- that notebook doesn't export it) so the concordance heatmap's columns
## read in the same division/family order as everywhere else this scheme is plotted.
ct_order <- c("Glut-DACH2-HVCra", "Glut-DACH2-HVCra-Int", "Glut-DACH2-HVCx",
             "Glut-DACH2-1", "Glut-DACH2-2", "Glut-DACH2-3", "Glut-DACH2-4",
             "Glut-DACH2-LMANsh", "Glut-DACH2-LMANco",
             "Glut-DACH2-5", "Glut-DACH2-6", "Glut-DACH2-7", "Glut-DACH2-8", "Glut-SATB2-1",
             "Glut-CACNA1H-RA", "Glut-CACNA1H-1", "Glut-CACNA1H-2", "Glut-CACNA1H-3", "Glut-CACNA1H-4",
             "Glut-GABA",
             "Glut-Im", "Glut-NB", "Glut-NSC",
             "GABA-LGE-1", "GABA-LGE-2",
             "GABA-MGE-SST-1", "GABA-MGE-SST-2", "GABA-MGE-SST-3",
             "GABA-MGE-PVALB-1", "GABA-MGE-PVALB-2", "GABA-MGE-PVALB-3",
             "GABA-MGE-LAMP5", "GABA-MGE-ST18", "GABA-MGE-LHX8",
             "GABA-CGE-1", "GABA-CGE-2",
             "GABA-Im",
             "Astro-1", "Astro-2", "Astro-3", "Epen",
             "Oligo-1", "Oligo-2", "Oligo-3", "OPC",
             "Micro", "Endo")

## Hand overrides for the majority-vote curation below, same escape hatch pattern as
## snrna/naming/hybrid_division_naming.qmd's *_overrides vectors: empty until a specific multiome
## cluster's vote is reviewed and judged wrong, then keyed here rather than by re-running the transfer.
cluster_hybrid_overrides <- setNames(c("Glut-CACNA1H-2", "Glut-DACH2-3", "Glut-DACH2-4", "GABA-Im", "GABA-CGE", "Oligo", "Astro", "Astro", "GABA-LGE-1", "GABA-LGE-2"),
                                     c("Glut-Arco-6", "Glut-NC-3", "Glut-NC-4", "GABA-Pre", "GABA-5", "Oligo", "Astro-1", "Astro-2", "GABA-1-1", "GABA-1-2")
                                     )

out_dir <- here::here("multiome/seurat", "label_transfer_hybrid")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
obj_out_fname <- file.path(out_dir, "obj_clustered_hybrid.qs2")
vote_fname <- file.path(out_dir, "cluster_hybrid_majority_vote.csv")
vote_breakdown_fname <- file.path(out_dir, "cluster_hybrid_vote_breakdown.csv")
```

## Load

Both read-only; neither is written back to.

```{r}
#| label: load
multiome <- qs_read(multiome_fname, nthreads = 8)
snrna <- qs_read(snrna_fname, nthreads = 8)
```

```{r}
#| label: prep-reference
## Same role as the original's `snrna = subset(snrna, subset=cluster!="Glut-Arco-4")` -- drop
## clusters this hybrid scheme excludes (celltype_hybrid is NA for exactly those, see
## snrna/naming/hybrid_division_naming.qmd's excluded_clusters) rather than naming just the one.
DefaultAssay(snrna) <- "SCT"
snrna <- subset(snrna, cells = colnames(snrna)[!is.na(snrna$celltype_hybrid)])
```

## QC: Glut-Nido-3

`Glut-Nido-3` (part of `cluster`) was flagged as artefactual. Confirmed on the multiome's own
existing UMAP: it doesn't occupy its own territory the way a real cluster does, it's scattered
across the `Glut-NC-1`/`Glut-NC-2`/`Glut-NC-4` region -- excluded below rather than voted on.

```{r fig.height=8, fig.width=8}
#| label: qc-glut-nido3
glut_cells <- colnames(multiome)[multiome$celltype == "Glut"]
DimPlot(subset(multiome, cells = glut_cells), group.by = "cluster", reduction = "umap_rna_int", label = TRUE) +
  ggtitle("Glutamatergic clusters (cluster, old post-curation naming)")
```

```{r}
#| label: exclude-glut-nido3
## Confirmed artefactual above (scattered across NC-1/NC-2/NC-4 rather than its own cluster) --
## dropped entirely so it can't pull a low-confidence vote onto any real label, and so it never
## reaches the saved object for downstream analysis (which only uses NC-1/NC-4 from this division
## anyway).
n_nido3 <- sum(multiome$cluster == "Glut-Nido-3")
message("Excluding ", n_nido3, " Glut-Nido-3 cell(s) (confirmed artefactual, not voted on).")
multiome <- subset(multiome, cells = colnames(multiome)[multiome$cluster != "Glut-Nido-3"])
```

## Transfer labels

Identical call to the original (`reference.reduction = "pca"`, `dims = 1:30`, SCT normalization on
both sides) -- only `refdata` changed, from the snRNA object's old `cluster` to `celltype_hybrid`.

```{r}
#| label: transfer
DefaultAssay(multiome) <- "SCT"
anchors <- FindTransferAnchors(reference = snrna,
                               normalization.method = "SCT",
                               query = multiome,
                               dims = 1:30,
                               reference.reduction = "pca")
predictions <- TransferData(anchorset = anchors, refdata = snrna$celltype_hybrid, dims = 1:30)
```

```{r}
#| label: apply-raw-transfer
multiome$cluster_hybrid_snrna <- predictions$predicted.id
```

## Concordance heatmap

Mean prediction score for each of the multiome's own existing clusters (`cluster` -- the label this
pipeline had already curated once, under the pre-hybrid naming scheme) against every hybrid snRNA
label, ordered by `ct_order`. Same construction as the original's QC heatmap, just against the new
reference labels.

```{r fig.height=10, fig.width=10}
#| label: concordance-heatmap
predictions$cluster <- multiome$cluster
predictions_stat <- predictions %>%
  select(-prediction.score.max) %>%
  select(c(one_of("cluster"), contains("prediction.score."))) %>%
  pivot_longer(contains("prediction.score."), names_to = "cluster_hybrid_snrna", values_to = "pred_score") %>%
  mutate(cluster_hybrid_snrna = sub("prediction.score.", "", cluster_hybrid_snrna)) %>%
  mutate(cluster_hybrid_snrna = gsub("\\.", "-", cluster_hybrid_snrna)) %>%
  group_by(cluster, cluster_hybrid_snrna) %>%
  summarize(pred_score_mean = mean(pred_score), .groups = "drop")

predictions_stat_mat <- predictions_stat %>%
  pivot_wider(names_from = "cluster_hybrid_snrna", values_from = "pred_score_mean") %>%
  column_to_rownames("cluster") %>%
  as.matrix()

predictions_stat_mat <- predictions_stat_mat[, intersect(ct_order, colnames(predictions_stat_mat))]

ncolors <- 9
cols <- circlize::colorRamp2(breaks = seq(0, 1, length.out = ncolors),
                             colors = scales::brewer_pal(palette = "Greys")(ncolors))

Heatmap(predictions_stat_mat, col = cols, cluster_columns = FALSE, rect_gp = gpar(lwd = 0.5),
       column_title = "Mean prediction score: multiome's own cluster (rows) x hybrid snRNA label (columns)")
```

## Curate: majority vote per multiome cluster

Each multiome cluster's final `cluster_hybrid` is its own cells' modal `cluster_hybrid_snrna` call.
Full per-cluster vote breakdown printed first -- not just the winner -- so a close or split vote is
visible before it is used.

```{r}
#| label: vote-breakdown
vote_breakdown <- multiome@meta.data %>%
  count(cluster, cluster_hybrid_snrna, name = "n_cells") %>%
  group_by(cluster) %>%
  mutate(vote_frac = n_cells / sum(n_cells)) %>%
  arrange(cluster, desc(vote_frac)) %>%
  ungroup()
write_csv(vote_breakdown, vote_breakdown_fname)

vote_majority <- vote_breakdown %>%
  group_by(cluster) %>%
  slice_max(n_cells, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(cluster_hybrid_snrna = if_else(cluster %in% names(cluster_hybrid_overrides),
                                        unname(cluster_hybrid_overrides[cluster]), cluster_hybrid_snrna)) %>%
  rename(cluster_hybrid = cluster_hybrid_snrna)
write_csv(vote_majority, vote_fname)

kable_low_conf <- vote_majority %>% filter(vote_frac < 0.5) %>% arrange(vote_frac)
if (nrow(kable_low_conf) > 0) {
  message(nrow(kable_low_conf), " multiome cluster(s) have NO majority (top vote < 50% of cells) ",
         "-- review these specifically, override via cluster_hybrid_overrides if the vote looks wrong:")
  print(kable_low_conf)
} else {
  message("Every multiome cluster's top vote already exceeds 50% of its own cells.")
}
knitr::kable(vote_majority, digits = 2,
            caption = "Final per-cluster curated label: modal transferred hybrid label and its vote fraction") %>%
  fig_kable_light()
```

```{r}
#| label: apply-curated
cluster_hybrid_map <- setNames(vote_majority$cluster_hybrid, vote_majority$cluster)
stopifnot("Every multiome cluster must resolve to a curated label" =
            all(unique(as.character(multiome$cluster)) %in% names(cluster_hybrid_map)))
multiome$cluster_hybrid <- unname(cluster_hybrid_map[as.character(multiome$cluster)])

kable_summary <- multiome@meta.data %>% count(cluster_hybrid, name = "n_cells") %>% arrange(desc(n_cells))
knitr::kable(kable_summary, caption = "Cell counts per final curated cluster_hybrid label") %>%
  fig_kable_light()
```

## QC: visual comparison

Same three-way overlay as the original (`cluster` / `cluster_hybrid_snrna` / `cluster_hybrid`,
plus `run` and `assignment` as batch/genotype checks) on the multiome's own existing UMAP -- this
notebook adds no new dimensionality reduction of its own.

```{r fig.height=10, fig.width=10}
#| label: qc-umap
reduction_cur <- "umap_rna_int"
DimPlot(multiome, group.by = "cluster", reduction = reduction_cur, label = TRUE) + ggtitle("cluster (pre-existing, old naming)")
DimPlot(multiome, group.by = "cluster_hybrid_snrna", reduction = reduction_cur, label = TRUE) + ggtitle("cluster_hybrid_snrna (raw transfer)")
DimPlot(multiome, group.by = "cluster_hybrid", reduction = reduction_cur, label = TRUE) + ggtitle("cluster_hybrid (majority-vote curated)")
DimPlot(multiome, group.by = "run", reduction = reduction_cur, label = TRUE)
DimPlot(multiome, group.by = "assignment", reduction = reduction_cur, label = FALSE)
```

## Save

Own copy, own filename -- `MULTIOME_SEURAT_DIR`'s `obj_clustered.qs`/`.qs2` are never written to by
this notebook.

```{r}
#| label: save
qs_save(multiome, obj_out_fname, nthreads = 1)
message("Wrote ", obj_out_fname)
```

```{r}
#| label: session-info
sessionInfo()
```
