---
title: "Cluster proportions / specificity heatmap, hybrid division-based labels"
format: html
editor: source
---

Reruns `celltypes_hclust_all.qmd`'s analysis and plotting (position-by-cluster proportion heatmap
with a tissue-specificity annotation bar) on the hybrid division-based labels from
`snrna/naming/hybrid_division_naming.qmd` (`celltype_hybrid`: major excitatory division kept
[`DACH2`/`SATB2`/`CACNA1H`], specific chicken-integration gene match dropped, song-region clusters
hand-labeled from anatomy/projections) instead of the raw `cluster` identity the original notebook
used. Independent of both source notebooks, by the same convention used throughout this project's
naming work: reads their outputs read-only, writes its own copy.

```{r}
#| label: packages
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)

library(pvclust)
library(future)
library(furrr)
library(ComplexHeatmap)
library(circlize)
library(scCustomize)
library(here)

source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "scRNA.R"))
source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "stats.R"))
```

## Parameters

```{r}
#| label: params
hclust_method <- "average"
dist_method <- "correlation"
```

## Directories

```{r}
#| label: dirs
script_name <- "celltypes_hclust_all_hybrid"
obj_fname <- here::here("snrna/naming/hybrid_division_naming", "obj_hybrid_labels.qs2")

out_dir <- here::here("snrna/trees", script_name)
out_sub_dir <- file.path(out_dir, str_interp("${dist_method}_${hclust_method}"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(out_sub_dir, showWarnings = FALSE)

## celltype_hybrid, not cluster -- the one substantive change from celltypes_hclust_all.qmd.
res_to_use <- "celltype_hybrid"

obj_out_fname <- file.path(out_dir, "obj_subset.qs2")

data_out_avg_fname <- file.path(out_dir, "data_avg.qs2")
data_out_pv_fname <- file.path(out_sub_dir, "pvclust.qs2") # path only, unused below -- matches
                                                            # the original, which never calls
                                                            # pvclust() in the code shown either

tfs <- read_delim(here::here("data/lambert_tfs/tfs.csv"), show_col_types = FALSE) %>%
  filter(DBD != "Unknown") # loaded but unused below, matching the original
```

## Load data

```{r}
#| label: load
redo <- TRUE
if (redo) {
  obj <- qs_read(obj_fname, nthreads = 8)
  idents <- FetchData(obj, res_to_use) %>%
    mutate_all(as.factor)
  Idents(obj) <- idents[, 1]

  ## Same doublet-assignment filter as the original (a no-op on this object specifically -- it was
  ## already filtered to both-caller singlets upstream, confirmed directly: 0 of 34295 cells have a
  ## "/" in `assignment` -- kept anyway so this notebook doesn't silently rely on that upstream state).
  ## New here: also drop cells whose cluster was excluded under the hybrid scheme (celltype_hybrid
  ## is NA for Glut-Arco-4 and Glut-Nido-3) -- carrying an excluded artifact cluster's cells into
  ## this analysis under a missing label would either error or silently create an "NA" column.
  cells <- Cells(obj)[!grepl("/", obj$assignment) & !is.na(obj@meta.data[[res_to_use]])]
  obj_sub <- subset(obj, cells = cells)

  qs_save(obj_sub, obj_out_fname, nthreads = 8)
} else {
  obj_sub <- qs_read(obj_out_fname, nthreads = 8)
}
```

```{r}
#| label: region
obj_sub$region <- case_when(obj_sub$position %in% c("ra", "arco") ~ "ARCO",
                            obj_sub$position %in% c("lman", "nr", "hvc", "nc") ~ "NIDO")
```

```{r}
#| label: cluster-table
table(obj_sub@meta.data[[res_to_use]])
```

### Manual order

`Glut-Nido-1` and `Glut-NC-1` are the same population under the hybrid scheme (confirmed in
`snrna/naming/exploratory/apply_composite_labels.qmd`: they independently won the same chicken match and were folded
together) and so share one label, `Glut-DACH2-1` -- listed once here, not twice, unlike the original
`ct_order`, which had them as separate rows.

```{r}
#| label: ct-order
ct_order <- c("Glut-DACH2-HVCra", "Glut-DACH2-HVCra-Pre", "Glut-DACH2-HVCx",
             "Glut-DACH2-1", "Glut-DACH2-2", "Glut-DACH2-3", "Glut-DACH2-4",
             "Glut-DACH2-LMANsh", "Glut-DACH2-LMANco",
             "Glut-DACH2-5", "Glut-DACH2-6", "Glut-DACH2-7", "Glut-DACH2-8", "Glut-SATB2-1",
             "Glut-CACNA1H-RA", "Glut-CACNA1H-1", "Glut-CACNA1H-2", "Glut-CACNA1H-3", "Glut-CACNA1H-4",
             "Glut-GABA",
             "Glut-Im", "Glut-NB", "Glut-NSC",
             "GABA-1-1", "GABA-1-2", "GABA-2-1", "GABA-2-2", "GABA-2-3",
             "GABA-3", "GABA-4-1", "GABA-4-2",
             "GABA-5-1", "GABA-5-2",
             "GABA-6", "GABA-7", "GABA-8",
             "GABA-Im",
             "Astro-1", "Astro-2", "Astro-3", "Epen",
             "Oligo-1", "Oligo-2", "Oligo-3", "OPC",
             "Micro", "Endo")

stopifnot(
  "ct_order does not exactly match the clusters present in obj_sub" =
    setequal(ct_order, unique(obj_sub@meta.data[[res_to_use]]))
)
```

## Average data

```{r}
#| label: aggregate
assay_to_use <- "SCT"
Idents(obj_sub) <- obj_sub@meta.data[[res_to_use]]
obj_int_avg <- AggregateExpression(obj_sub, assays = assay_to_use)
obj_int_avg1 <- log1p(obj_int_avg[[assay_to_use]])
colnames(obj_int_avg1) <- colnames(obj_int_avg[[assay_to_use]])
qs_save(obj_int_avg1, data_out_avg_fname, nthreads = 8)
```

```{r}
#| label: position-table
tab <- table(obj_sub$position, obj_sub@meta.data[[res_to_use]])
tab <- tab[c("hvc", "nc", "lman", "nr", "ra", "arco"), ct_order]
tab <- sweep(tab, 1, rowSums(tab), FUN = "/")
tab <- sweep(tab, 2, colSums(tab), FUN = "/")
tab_spec <- apply(tab, 2, calc_tissue_specificity)
```

```{r}
#| label: heatmap
scale_factor <- 0.08
height <- nrow(tab) * scale_factor
width <- ncol(tab) * scale_factor

tab <- tab[, ct_order]
tab_spec <- tab_spec[ct_order]

ncolors <- 9
cols <- colorRamp2(breaks = seq(0, 1, length.out = ncolors), colors = scales::brewer_pal(palette = "Greys")(ncolors))

## Cluster-type classification for the specificity-bar color, rewritten for the hybrid labels: the
## original matched "Glut-Pre"/"GABA-Pre" as literal substrings of the cluster name, which no longer
## works now that the precursor labels are Glut-NSC/Glut-NB/Glut-Im (no "Pre" substring at all) --
## classified here by an explicit name list instead of a substring that would silently stop matching.
## Glut-GABA (the Nido-2 dual-transmitter override) is deliberately caught by the "Glut" branch, not
## "GABA", matching how the original treated Glut-Nido-2 -- checked explicitly below rather than left
## to substring-match luck, since "Glut-GABA" also contains the literal substring "GABA".
glut_precursor <- c("Glut-NSC", "Glut-NB", "Glut-Im")
gaba_precursor <- c("GABA-Im")

bar_cols <- data.frame(cluster_type = c("Glut", "Glut-Pre", "GABA", "GABA-Pre", "Non-neuron"),
                       col = c("#008000", "#90ca90", "#800000", "#c37474", "#808000"))
ct_df <- data.frame(cluster = names(tab_spec)) %>%
  mutate(cluster_type = case_when(
    cluster %in% glut_precursor ~ "Glut-Pre",
    grepl("^Glut", cluster) ~ "Glut",
    cluster %in% gaba_precursor ~ "GABA-Pre",
    grepl("^GABA", cluster) ~ "GABA",
    TRUE ~ "Non-neuron")) %>%
  left_join(bar_cols, by = "cluster_type")
stopifnot("Glut-GABA classified as Glut, not GABA, as intended" =
           ct_df$cluster_type[ct_df$cluster == "Glut-GABA"] == "Glut")

bar_colors <- ct_df$col
ab <- anno_barplot(tab_spec, 0, which = "column", border = FALSE,
                   gp = gpar(fill = bar_colors))
hm_row <- HeatmapAnnotation(ab = ab, which = "column")

hm <- Heatmap(tab, cluster_columns = FALSE,
             cluster_rows = FALSE,
             col = cols,
             border = TRUE, border_gp = gpar(lwd = 0.5),
             height = unit(height, "in"),
             width = unit(width, "in"),
             column_names_gp = gpar(fontsize = 6),
             row_names_gp = gpar(fontsize = 6))
hm
hm2 <- hm_row %v% hm
hm2

pdf(file.path(out_sub_dir, "cluster_proportions_horiz_with-bar_ordered.pdf"), height = height + 4, width = width + 4)
print(hm2)
dev.off()

ab <- anno_barplot(tab_spec, 0, which = "row", border = TRUE,
                   gp = gpar(fill = bar_colors),
                   axis_param = list(gp = gpar(fontsize = 5)))
hm_row <- HeatmapAnnotation(ab = ab, which = "row")

hm <- Heatmap(t(tab), cluster_rows = FALSE, show_row_names = TRUE, row_names_side = "left",
             cluster_columns = FALSE,
             col = cols,
             border = TRUE, border_gp = gpar(lwd = 0.5),
             height = unit(width, "in"),
             width = unit(height, "in"),
             column_names_gp = gpar(fontsize = 6),
             row_names_gp = gpar(fontsize = 6))
hm
hm2 <- hm + hm_row
hm2
pdf(file.path(out_sub_dir, "cluster_proportions_vert_with-bar_ordered.pdf"), height = width + 4, width = height + 4)
print(hm2)
dev.off()
```

```{r}
#| label: session-info
sessionInfo()
```
