---
title: "Robustness of glutamatergic cell-type clustering to song-DEG removal"
format:
  html:
    toc: true
    df-print: paged
---

Reviewer-requested robustness analysis for the hierarchical clustering of
glutamatergic cell types in `celltypes_hclust_glut.qmd`. In that analysis, genes
differentially expressed in the song nuclei (Glut-RA within the arcopallial
dissections; Glut-HVC-* / Glut-LMAN-* within the nidopallial dissections) are
removed from the feature set before the cell types are clustered, so that the
tree reflects regional identity rather than song-nucleus identity.

This document addresses four points:

1. **Before removal.** The tree built with no song-DEG removal at all.
2. **Threshold sensitivity.** Trees built while defining the song DEGs at a
   range of adjusted-p thresholds (a looser threshold removes *more* genes).
3. **Bootstrap stability.** Two complementary bootstraps: `pvclust`
   multiscale bootstrap over genes (AU/BP support per branch) and a
   nucleus-level bootstrap that resamples cells within each cell type and
   measures how often each branch of the published tree is recovered.
4. **Splitting the dissections.** The published trees are built separately for
   the nidopallial (NIDO) and arcopallial (ARCO) dissections, which occupy
   distinct pallial domains. Every analysis below is therefore run a third
   time, labelled `ALL`, with all glutamatergic cell types clustered together
   in a single tree, so that the split can be compared against the pooled
   alternative. For `ALL`, the markers and the song DEGs are called on
   pseudobulk profiles with both dissections pooled, and the song-DEG set is
   the union of the Glut-RA, Glut-HVC-* and Glut-LMAN-* DEGs.

All parameters other than the song-DEG removal step are held at the values used
in `celltypes_hclust_glut.qmd` (`ward.D` on correlation distance, top 100
markers per cluster per direction at `p_adj < 0.1`, gene-specificity scaling).

```{r setup}
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(Matrix)
library(pvclust)
library(ComplexHeatmap)
library(circlize)
library(ggtree)
library(ape)
library(cowplot)
library(here)
set.seed(10)

source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "scRNA.R"))
source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "stats.R"))
source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "common_aesthetics.R"))

pnames = names(position_colors)
names(position_colors) = case_when(pnames == "nido" ~ "nr",
                                   pnames == "ncl" ~ "nc",
                                   TRUE ~ pnames)
```

# Parameters

```{r params}
hclust_method = "ward.D"
dist_method   = "correlation"
assay_to_use  = "SCT"
res_to_use    = "cluster"

## Number of top markers retained per cluster per direction (as published)
nsig = 100
## Significance threshold used to *select* the clustering features (as published)
feature_thresh = 0.1

## Song-DEG removal thresholds to sweep. NA = no removal (i.e. the clustering
## as it would be done before any song DEGs are taken out).
deg_thresholds = c(NA, 0.5, 0.25, 0.1, 0.05, 0.01, 0.001)
## The threshold used in the manuscript, for reference comparisons
ref_threshold = 0.1

## pvclust multiscale bootstrap replicates
nboot = 1000
n_cores = min(20L, max(1L, parallel::detectCores() - 4L))

## Cell-level bootstrap replicates
n_cellboot = 500

## Song-nucleus clusters, per analysis. ALL = every glutamatergic cell type
## clustered together, without splitting the nidopallial and arcopallial
## dissections (requested by a reviewer).
song_clusters = list(ARCO = "^Glut-RA$",
                     NIDO = "HVC|LMAN",
                     ALL  = "^Glut-RA$|HVC|LMAN")
## Which aggregated columns belong to each tree
region_cols  = list(ARCO = "RA|Arco",
                    NIDO = "HVC|NC|LMAN|NR|Nido",
                    ALL  = "Glut")
regions = c("ARCO", "NIDO", "ALL")

## Element-wise label for a threshold (format() on a vector pads to a common
## width, which would break the joins below, so map over elements).
thr_label = function(x) {
  vapply(x, function(v) if (is.na(v)) "none" else format(v, scientific = FALSE),
         character(1))
}
## Ordered threshold labels, most genes removed -> fewest
lev = c("none", thr_label(sort(deg_thresholds[!is.na(deg_thresholds)], decreasing = TRUE)))
ref_lab = thr_label(ref_threshold)
```

# Directories

```{r dirs}
script_name = "celltypes_hclust_glut_robustness"

obj_fname = here::here("snrna/clustering/snrna_seurat_cellbender_preprocess",
                       "obj_clustered.qs2")

out_dir     = here::here("snrna/trees", script_name)
out_sub_dir = file.path(out_dir, str_interp("${dist_method}_${hclust_method}"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(out_sub_dir, recursive = TRUE, showWarnings = FALSE)

obj_out_fname      = file.path(out_dir, "obj_clustered_glutamatergic.qs2")
data_out_avg_fname = file.path(out_dir, "data_avg.qs2")
markers_fname      = file.path(out_dir, sprintf("marker_genes_cluster_%s.rds", assay_to_use))

## Marker tables from the published run (identical cell subset); used as caches
## so that the pseudobulk DESeq2 tests are not recomputed unnecessarily.
## markers_cache holds the per-dissection tables, markers_cache_all the table
## computed with both dissections pooled.
markers_cache     = here::here("snrna/trees/celltypes_hclust_glut_no1a",
                               sprintf("marker_genes_cluster_%s.rds", assay_to_use))
markers_cache_all = here::here("snrna/trees/celltypes_hclust_glut_no1a",
                               sprintf("marker_genes_cluster_pos-together_%s.rds", assay_to_use))
```

# Load data

Same cell subset as the published analysis: glutamatergic clusters, excluding
the pre-motor/pre-song ("Pre") clusters, the Glut-HVC-1a cluster, and cells
whose dissection assignment spans two regions.

```{r load}
redo = !file.exists(obj_out_fname)
if (redo) {
  obj = qs_read(obj_fname)
  idents = FetchData(obj, res_to_use) %>% mutate_all(as.factor)
  Idents(obj) = idents[, 1]
  cells = Cells(obj)[grepl("Glut", Idents(obj)) &
                       !grepl("Pre", Idents(obj)) &
                       !grepl("1a", Idents(obj))]
  obj_sub = subset(obj, cells = cells)

  cells = Cells(obj_sub)[!grepl("/", obj_sub$assignment)]
  obj_sub = subset(obj_sub, cells = cells)

  qs_save(obj_sub, obj_out_fname)
  rm(obj); gc()
} else {
  obj_sub = qs_read(obj_out_fname)
}

obj_sub$region = case_when(obj_sub$position %in% c("ra", "arco") ~ "ARCO",
                           obj_sub$position %in% c("lman", "nr", "hvc", "nc") ~ "NIDO")
obj_sub$cluster = droplevels(as.factor(obj_sub@meta.data[[res_to_use]]))
table(obj_sub$cluster, obj_sub$region)
```

# Markers

Pseudobulk (cluster x animal) DESeq2 markers, computed separately within each
dissection region. These define both the song-DEG sets that get removed and the
feature sets used for clustering.

```{r markers}
compute_markers = function(obj_cur) {
  obj_agg = AggregateExpression(obj_cur, assay = assay_to_use,
                                group.by = c(res_to_use, "assignment"),
                                return.seurat = TRUE)
  Idents(obj_agg) = obj_agg$cluster
  FindAllMarkers(obj_agg, assay = assay_to_use, test.use = "DESeq2", only.pos = FALSE)
}

markers_agg = if (file.exists(markers_fname)) readRDS(markers_fname) else list()
markers_new = FALSE
for (reg in regions) {
  if (!is.null(markers_agg[[reg]])) next
  markers_new = TRUE
  markers_agg[[reg]] =
    if (reg == "ALL") {
      ## both dissections pooled
      if (file.exists(markers_cache_all)) readRDS(markers_cache_all) else compute_markers(obj_sub)
    } else if (file.exists(markers_cache)) {
      readRDS(markers_cache)[[reg]]
    } else {
      compute_markers(subset(obj_sub, region == reg))
    }
}
if (markers_new) saveRDS(markers_agg, markers_fname)
map(markers_agg[regions], ~ count(.x, cluster))
```

# Average expression

```{r avg}
if (file.exists(data_out_avg_fname)) {
  obj_int_avg1 = qs_read(data_out_avg_fname)
} else {
  Idents(obj_sub) = obj_sub$cluster
  obj_int_avg = AggregateExpression(obj_sub, assays = assay_to_use)
  obj_int_avg1 = log1p(obj_int_avg[[assay_to_use]])
  colnames(obj_int_avg1) = colnames(obj_int_avg[[assay_to_use]])
  qs_save(obj_int_avg1, data_out_avg_fname)
}
dim(obj_int_avg1)
```

# Helper functions

```{r helpers}
## Genes called as song DEGs at a given adjusted-p threshold.
## thresh = NA -> no genes removed (clustering "before" song-DEG removal).
song_degs = function(markers, song_pattern, thresh) {
  if (is.na(thresh)) return(character(0))
  markers %>%
    filter(grepl(song_pattern, cluster)) %>%
    filter(p_val_adj < thresh) %>%
    pull(gene) %>%
    unique()
}

## Feature set used for clustering, exactly as in celltypes_hclust_glut.qmd
## except that the removed-gene set is supplied.
select_features = function(markers, removed, nsig, feature_thresh) {
  markers %>%
    filter(!(gene %in% removed)) %>%
    mutate(sign = avg_log2FC > 0) %>%
    filter(p_val_adj < feature_thresh) %>%
    group_by(cluster, sign) %>%
    top_n(nsig, abs(avg_log2FC)) %>%
    ungroup() %>%
    distinct(gene, .keep_all = TRUE) %>%
    pull(gene)
}

## genes x cell-types matrix of scaled gene specificity, as published
specificity_matrix = function(avg, genes, col_pattern) {
  m = avg[genes, grepl(col_pattern, colnames(avg)), drop = FALSE]
  m = t(apply(m, 1, calc_gene_specificity, scale = TRUE))
  m[is.nan(m)] = 0
  as.matrix(m)
}

## Tip sets of every internal node of an hclust, in hclust merge order
## (i.e. the same order as pvclust's $edges rows).
hclust_clades = function(hc) {
  out = vector("list", nrow(hc$merge))
  for (i in seq_len(nrow(hc$merge))) {
    tips = character(0)
    for (j in hc$merge[i, ])
      tips = c(tips, if (j < 0) hc$labels[-j] else out[[j]])
    out[[i]] = sort(tips)
  }
  out
}
clade_key = function(tips) paste(sort(tips), collapse = "|")

## Tip sets for every node of a phylo object, indexed by node number.
phylo_tipsets = function(phy) {
  ntip = length(phy$tip.label)
  desc = vector("list", max(phy$edge))
  for (i in seq_len(ntip)) desc[[i]] = phy$tip.label[i]
  for (e in rev(seq_len(nrow(phy$edge)))) {
    p = phy$edge[e, 1]; ch = phy$edge[e, 2]
    desc[[p]] = c(desc[[p]], desc[[ch]])
  }
  desc
}

## Map hclust-ordered per-edge values (e.g. pvclust AU) onto ggtree node ids.
edge_values_to_nodes = function(hc, values) {
  phy = ape::as.phylo(hc)
  hc_keys = map_chr(hclust_clades(hc), clade_key)
  ts = phylo_tipsets(phy)
  node_keys = map_chr(ts, ~ if (is.null(.x)) NA_character_ else clade_key(.x))
  tibble(node = seq_along(node_keys), key = node_keys) %>%
    filter(!is.na(key)) %>%
    left_join(tibble(key = hc_keys, value = values), by = "key") %>%
    filter(!is.na(value))
}

## correlation-distance ward.D tree, matching pvclust's settings
build_tree = function(mat) {
  d = as.dist(1 - cor(mat))
  hclust(d, method = hclust_method)
}
```

# Threshold sweep

For each region and each song-DEG threshold, rebuild the feature set, recompute
the cell-type tree, and run the `pvclust` multiscale bootstrap.

```{r sweep}
#| message: false
sweep_fname = file.path(out_sub_dir, "pvclust_sweep.qs2")

## Cached per analysis x threshold, so that adding an analysis does not force
## the previously computed bootstraps to be rerun.
sweep = if (file.exists(sweep_fname)) qs_read(sweep_fname) else list()
sweep_new = FALSE
for (region_cur in regions) {
  for (thr in deg_thresholds) {
    thr_lab = thr_label(thr)
    key = paste(region_cur, thr_lab, sep = "_")
    if (!is.null(sweep[[key]])) next
    sweep_new = TRUE
    message("== ", region_cur, " threshold ", thr_lab)

    removed  = song_degs(markers_agg[[region_cur]], song_clusters[[region_cur]], thr)
    features = select_features(markers_agg[[region_cur]], removed, nsig, feature_thresh)
    mat      = specificity_matrix(obj_int_avg1, features, region_cols[[region_cur]])

    pv = pvclust(mat, method.dist = dist_method, method.hclust = hclust_method,
                 nboot = nboot, parallel = as.integer(n_cores))

    sweep[[key]] = list(region = region_cur, threshold = thr, threshold_label = thr_lab,
                        n_removed = length(removed), n_features = length(features),
                        features = features, removed = removed, pv = pv)
  }
}
if (sweep_new) qs_save(sweep, sweep_fname)
## keep only the analyses requested by `regions`, in that order
sweep = sweep[unlist(map(regions, function(r) paste(r, thr_label(deg_thresholds), sep = "_")))]

sweep_df = map_dfr(sweep, function(s) {
  au = s$pv$edges$au
  tibble(region = s$region, threshold_label = s$threshold_label,
         threshold = s$threshold, n_removed = s$n_removed,
         n_features = s$n_features,
         n_edges = length(au), mean_au = mean(au), median_au = median(au),
         n_au95 = sum(au >= 0.95), frac_au95 = mean(au >= 0.95))
}) %>%
  mutate(threshold_label = factor(threshold_label, levels = lev)) %>%
  arrange(region, threshold_label)

write_csv(sweep_df, file.path(out_sub_dir, "sweep_summary.csv"))
sweep_df
```

```{r sweep-plot}
#| fig-width: 9
#| fig-height: 6
p_counts = sweep_df %>%
  pivot_longer(c(n_removed, n_features, frac_au95), names_to = "metric") %>%
  mutate(metric = recode(metric,
                         n_removed  = "song DEGs removed",
                         n_features = "genes used for clustering",
                         frac_au95  = "fraction of branches AU >= 0.95")) %>%
  ggplot(aes(threshold_label, value, group = region, color = region)) +
  geom_line() + geom_point() +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "song-DEG adjusted-p threshold (none = no removal)", y = NULL) +
  theme_cowplot(font_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_counts)
save_plot(file.path(out_sub_dir, "sweep_summary.pdf"), p_counts,
          base_height = 3.5, base_width = 10)
```

# Trees across thresholds

Each tree is annotated with the `pvclust` approximately-unbiased (AU) support,
in percent, at every internal branch. Tips are coloured by dissection position.

```{r tree-fn}
## Display name for each analysis
region_name = c(ARCO = "ARCO", NIDO = "NIDO",
                ALL = "ALL (nido + arco combined)")

## Two lines, so that the longest title still fits the narrow page.
tree_title = function(region, threshold_label) {
  nm = if (region %in% names(region_name)) region_name[[region]] else region
  if (threshold_label == "none")
    sprintf("%s\nno song DEGs removed", nm)
  else
    sprintf("%s\nsong DEGs removed at p_adj < %s", nm, threshold_label)
}

tip_annotation = function(labels) {
  data.frame(labels = labels) |>
    mutate(position = str_extract(labels, "Glut-([A-Za-z]+)", group = 1)) |>
    mutate(position = if_else(position == "Nido", "NC", position),
           position = tolower(position)) |>
    mutate(col = position_colors[match(position, names(position_colors))])
}

## Tree geometry, matched to celltypes_hclust_glut.qmd: geom_tiplab at offset 0,
## hexpand(5), and a panel fixed by egg::set_panel_size so rows sit tight
## against one another. Font sizes are given in points and converted to the mm
## units that ggplot's `size` aesthetic expects; the per-tip row height tracks
## the tip font so that shrinking the text does not reopen gaps between labels.
tip_pt           = 6     # tip labels
support_pt       = 5     # branch-support labels (absent in the original)
tip_size         = tip_pt / .pt
support_size     = support_pt / .pt
tip_scale_factor = tip_pt * 1.45 / 72   # panel inches per tip (~1.45 line spacing)
panel_width      = 2                    # panel inches
## Branch line weight, given as an R lwd (lwd 1 = 1/96 in = 0.75 pt) and
## converted to the mm units ggplot's `linewidth` expects. ggtree's default is
## linewidth 0.5, i.e. lwd 1.42, which is heavier than this.
branch_lwd       = 0.5
branch_linewidth = branch_lwd / .pt

## hc:      hclust to draw
## values:  per-edge values in hclust merge order (e.g. pvclust AU), or NULL
## Returns a fixed-panel gtable; print()/plot() it, and size the device with
## tree_device_dim() so that there is no slack around the panel.
support_tree = function(hc, values, label_color = "grey30", title = NULL, subtitle = NULL) {
  hc_df = tip_annotation(hc$labels)
  g = ggtree(hc, branch.length = "none", linewidth = branch_linewidth) %<+% hc_df
  if (!is.null(values)) {
    node_df = edge_values_to_nodes(hc, values) %>%
      transmute(node, support_lab = as.character(round(value * 100)))
    g = g %<+% node_df
  }
  g = g +
    geom_tiplab(aes(color = col), angle = 0, hjust = 0, offset = 0,
                show.legend = FALSE, size = tip_size) +
    theme_dendrogram(plot.margin = margin(20, 20, 20, 20)) +
    scale_color_identity() +
    hexpand(5) +
    labs(title = title, subtitle = subtitle) +
    theme(plot.title = element_text(size = 9, face = "plain"),
          plot.subtitle = element_text(size = 7))
  if (!is.null(values))
    ## the extra left expansion keeps the root's label inside the panel
    g = g + geom_text2(aes(label = support_lab, subset = !isTip),
                       hjust = 1.15, vjust = -0.4, size = support_size,
                       color = label_color) +
      ggexpand(0.25, direction = -1)
  egg::set_panel_size(g,
                      width  = unit(panel_width, "in"),
                      height = unit(length(hc$labels) * tip_scale_factor, "in"))
}

## Device size that just contains the fixed panel plus margins and the title.
tree_device_dim = function(n_tips) {
  list(width  = panel_width + 2.0,
       height = n_tips * tip_scale_factor + 1.4)   # room for the title block
}

## Draw fixed-panel gtables one per page. plot() would call grid.newpage()
## before the first tree as well, leaving a blank leading page in the PDF.
draw_pages = function(plots) {
  for (i in seq_along(plots)) {
    if (i > 1) grid::grid.newpage()
    grid::grid.draw(plots[[i]])
  }
}

plot_tree = function(s, show_au = TRUE) {
  support_tree(s$pv$hclust,
               values = if (show_au) s$pv$edges$au else NULL,
               title = tree_title(s$region, s$threshold_label),
               subtitle = sprintf("%d genes removed, %d genes clustered\nbranch labels = pvclust AU (%%)",
                                  s$n_removed, s$n_features))
}
```

```{r trees}
#| fig-width: 5
#| fig-height: 7
#| results: hold
for (reg in regions) {
  keys = names(sweep)[map_chr(sweep, "region") == reg]
  n_tips = length(sweep[[keys[1]]]$pv$hclust$labels)
  dim = tree_device_dim(n_tips)
  walk(keys, ~ plot(plot_tree(sweep[[.x]])))
  pdf(file.path(out_sub_dir, sprintf("trees_%s_thresholds.pdf", tolower(reg))),
      height = dim$height, width = dim$width)
  draw_pages(map(keys, ~ plot_tree(sweep[[.x]])))
  dev.off()
}
```

# Topology agreement across thresholds

Two summaries of how much the tree changes with the threshold:

* **Cophenetic / Baker's gamma correlation** between the full dendrograms
  (Pearson and Spearman correlation of the cophenetic distances).
* **Adjusted Rand index** between the cell-type groupings obtained by cutting
  each tree at k = 2..8.

```{r topology}
coph = map(sweep, ~ as.vector(cophenetic(.x$pv$hclust)))

pairs_df = map_dfr(regions, function(reg) {
  keys = names(sweep)[map_chr(sweep, "region") == reg]
  expand_grid(a = keys, b = keys) %>%
    mutate(region = reg,
           lab_a = map_chr(a, ~ sweep[[.x]]$threshold_label),
           lab_b = map_chr(b, ~ sweep[[.x]]$threshold_label),
           cophenetic_r = map2_dbl(a, b, ~ cor(coph[[.x]], coph[[.y]])),
           baker_gamma  = map2_dbl(a, b, ~ cor(coph[[.x]], coph[[.y]], method = "spearman")),
           ari = map2_dbl(a, b, function(x, y) {
             ks = 2:min(8, length(sweep[[x]]$pv$hclust$labels) - 1)
             mean(map_dbl(ks, function(k)
               mclust::adjustedRandIndex(cutree(sweep[[x]]$pv$hclust, k)[sweep[[x]]$pv$hclust$labels],
                                         cutree(sweep[[y]]$pv$hclust, k)[sweep[[x]]$pv$hclust$labels])))
           }))
})
write_csv(select(pairs_df, -a, -b), file.path(out_sub_dir, "topology_agreement_pairwise.csv"))

## versus the published threshold
ref_df = pairs_df %>%
  filter(lab_b == ref_lab) %>%
  select(region, threshold_label = lab_a, cophenetic_r, baker_gamma, ari)
write_csv(ref_df, file.path(out_sub_dir, "topology_agreement_vs_published.csv"))
ref_df
```

```{r topology-plot}
#| fig-width: 9
#| fig-height: 4
p_top = ref_df %>%
  pivot_longer(c(cophenetic_r, baker_gamma, ari)) %>%
  mutate(threshold_label = factor(threshold_label, levels = lev),
         name = recode(name,
                       cophenetic_r = "cophenetic r",
                       baker_gamma  = "Baker's gamma",
                       ari          = "mean ARI (k = 2..min(8, n-1))")) %>%
  ggplot(aes(threshold_label, value, color = region, group = region)) +
  geom_line() + geom_point() +
  facet_wrap(~name) +
  ylim(0, 1) +
  labs(x = "song-DEG adjusted-p threshold", y = "agreement with published tree") +
  theme_cowplot(font_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_top)
save_plot(file.path(out_sub_dir, "topology_agreement.pdf"), p_top,
          base_height = 3.5, base_width = 10)
```

```{r topology-heatmap}
#| fig-width: 10
#| fig-height: 5
for (reg in regions) {
  m = pairs_df %>%
    filter(region == reg) %>%
    select(lab_a, lab_b, ari) %>%
    pivot_wider(names_from = lab_b, values_from = ari) %>%
    column_to_rownames("lab_a") %>%
    as.matrix()
  m = m[lev, lev]
  hm = Heatmap(m, name = "ARI", cluster_rows = FALSE, cluster_columns = FALSE,
               col = colorRamp2(c(0, 1), hcl_palette = "Blues", reverse = TRUE),
               column_title = sprintf("%s: pairwise ARI across thresholds", reg),
               cell_fun = function(j, i, x, y, w, h, fill)
                 grid.text(sprintf("%.2f", m[i, j]), x, y, gp = gpar(fontsize = 8)))
  print(hm)
  pdf(file.path(out_sub_dir, sprintf("topology_ari_%s.pdf", tolower(reg))),
      height = 5, width = 5.5)
  print(hm)
  dev.off()
}
```

# Branch-level bootstrap support

## Multiscale bootstrap over genes (pvclust)

AU support for every branch of the published (p_adj < 0.1) tree, and, for
comparison, for the same branches when they are present in the no-removal tree.

```{r au-table}
au_df = map_dfr(sweep, function(s) {
  hc = s$pv$hclust
  tibble(region = s$region, threshold_label = s$threshold_label,
         clade = map_chr(hclust_clades(hc), clade_key),
         n_tips = map_int(hclust_clades(hc), length),
         au = s$pv$edges$au, bp = s$pv$edges$bp)
})
write_csv(au_df, file.path(out_sub_dir, "branch_support_all.csv"))

ref_clades = au_df %>% filter(threshold_label == ref_lab)

## Is each published branch recovered at other thresholds, and with what support?
## Every published branch is crossed with every threshold, so branches that are
## absent at a threshold show up as NA rather than being dropped.
recovery = ref_clades %>%
  select(region, clade, n_tips, au_published = au) %>%
  crossing(threshold_label = lev) %>%
  left_join(au_df %>% select(region, threshold_label, clade, au_at_threshold = au),
            by = c("region", "clade", "threshold_label")) %>%
  mutate(threshold_label = factor(threshold_label, levels = lev),
         recovered = !is.na(au_at_threshold)) %>%
  arrange(region, desc(n_tips), threshold_label)
write_csv(recovery, file.path(out_sub_dir, "published_branch_recovery_per_branch.csv"))

recovery_summary = recovery %>%
  group_by(region, threshold_label) %>%
  summarise(n_published_branches = n(),
            n_recovered = sum(recovered),
            frac_recovered = mean(recovered),
            mean_au_when_recovered = mean(au_at_threshold, na.rm = TRUE), .groups = "drop") %>%
  arrange(region, threshold_label)
write_csv(recovery_summary, file.path(out_sub_dir, "published_branch_recovery.csv"))
recovery_summary
```

```{r au-plot}
#| fig-width: 9
#| fig-height: 4
p_au = au_df %>%
  mutate(threshold_label = factor(threshold_label, levels = lev)) %>%
  ggplot(aes(threshold_label, au)) +
  geom_hline(yintercept = 0.95, linetype = 2, color = "grey50") +
  geom_boxplot(outlier.shape = NA, fill = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6, aes(color = region)) +
  facet_wrap(~region) +
  labs(x = "song-DEG adjusted-p threshold", y = "pvclust AU support") +
  theme_cowplot(font_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
print(p_au)
save_plot(file.path(out_sub_dir, "branch_support_au.pdf"), p_au,
          base_height = 3.5, base_width = 8)
```

## Bootstrap over cells

`pvclust` resamples *genes*. To estimate stability with respect to the cells
(nuclei) that were sampled, we resample cells with replacement within each cell
type, re-aggregate the pseudobulk profiles, and rebuild the tree with the same
feature set and the same distance/linkage. The feature set is held fixed within
each bootstrap series (re-running the pseudobulk DESeq2 tests per replicate is
not tractable), so this measures sampling variability in the expression
profiles rather than in marker selection.

```{r cellboot}
#| message: false
cellboot_fname = file.path(out_sub_dir, "cellboot.qs2")

boot_tree = function(counts, cl_factor, features, keep_cols) {
  idx = unlist(lapply(split(seq_along(cl_factor), cl_factor),
                      function(ii) sample(ii, length(ii), replace = TRUE)),
               use.names = FALSE)
  cl_b = droplevels(cl_factor[idx])
  ind = sparse.model.matrix(~ 0 + cl_b)
  colnames(ind) = levels(cl_b)
  agg = as.matrix(counts[, idx, drop = FALSE] %*% ind)
  agg = log1p(agg)
  agg = agg[features, intersect(keep_cols, colnames(agg)), drop = FALSE]
  agg = t(apply(agg, 1, calc_gene_specificity, scale = TRUE))
  agg[is.nan(agg)] = 0
  build_tree(as.matrix(agg))
}

cellboot = if (file.exists(cellboot_fname)) qs_read(cellboot_fname) else list()
cellboot_todo = unlist(map(regions, ~ paste(.x, c("none", ref_lab), sep = "_")))
cellboot_todo = setdiff(cellboot_todo, names(cellboot))

if (length(cellboot_todo)) {
  counts_all = LayerData(obj_sub, assay = assay_to_use, layer = "counts")
  cl_all = droplevels(as.factor(obj_sub$cluster))

  for (reg in regions) {
    for (thr_lab in c("none", ref_lab)) {
      key = paste(reg, thr_lab, sep = "_")
      if (!(key %in% cellboot_todo)) next
      s = sweep[[key]]
      keep_cols = colnames(obj_int_avg1)[grepl(region_cols[[reg]], colnames(obj_int_avg1))]
      feats = intersect(s$features, rownames(counts_all))

      ## restrict the cells to those contributing to this region's columns
      cell_keep = which(as.character(cl_all) %in% keep_cols)
      counts_reg = counts_all[feats, cell_keep, drop = FALSE]
      cl_reg = droplevels(cl_all[cell_keep])

      trees = map(seq_len(n_cellboot), function(i) {
        if (i %% 50 == 0) message(key, " boot ", i)
        boot_tree(counts_reg, cl_reg, feats, keep_cols)
      })

      ref_hc = s$pv$hclust
      ref_keys = map_chr(hclust_clades(ref_hc), clade_key)
      boot_keys = map(trees, ~ map_chr(hclust_clades(.x), clade_key))
      support = map_dbl(ref_keys, function(k) mean(map_lgl(boot_keys, ~ k %in% .x)))

      cellboot[[key]] = list(region = reg, threshold_label = thr_lab,
                             clade = ref_keys,
                             n_tips = map_int(hclust_clades(ref_hc), length),
                             support = support, hclust = ref_hc)
    }
  }
  qs_save(cellboot, cellboot_fname)
  rm(counts_all); gc()
}
cellboot = cellboot[unlist(map(regions, ~ paste(.x, c("none", ref_lab), sep = "_")))]

cellboot_df = map_dfr(cellboot, function(x)
  tibble(region = x$region, threshold_label = x$threshold_label,
         clade = x$clade, n_tips = x$n_tips, cell_bootstrap = x$support))
write_csv(cellboot_df, file.path(out_sub_dir, "cell_bootstrap_support.csv"))

cellboot_df %>%
  group_by(region, threshold_label) %>%
  summarise(n_branches = n(),
            mean_support = mean(cell_bootstrap),
            n_ge_95 = sum(cell_bootstrap >= 0.95),
            n_ge_70 = sum(cell_bootstrap >= 0.70), .groups = "drop")
```

```{r cellboot-trees}
#| fig-width: 5
#| fig-height: 7
#| results: hold
plot_support_tree = function(x) {
  support_tree(x$hclust, values = x$support, label_color = "firebrick",
               title = tree_title(x$region, x$threshold_label),
               subtitle = sprintf("branch labels = %% of %d cell\nbootstraps recovering the branch",
                                  n_cellboot))
}
walk(cellboot, ~ plot(plot_support_tree(.x)))

## one page per analysis, sized to the tallest tree
cb_dim = tree_device_dim(max(map_int(cellboot, ~ length(.x$hclust$labels))))
pdf(file.path(out_sub_dir, "trees_cell_bootstrap.pdf"),
    height = cb_dim$height, width = cb_dim$width)
draw_pages(map(cellboot, plot_support_tree))
dev.off()
```

## Combined support table

```{r combined-support}
combined = ref_clades %>%
  select(region, clade, n_tips, au_gene_bootstrap = au, bp_gene_bootstrap = bp) %>%
  left_join(cellboot_df %>%
              filter(threshold_label == ref_lab) %>%
              select(region, clade, cell_bootstrap),
            by = c("region", "clade")) %>%
  arrange(region, desc(n_tips))
write_csv(combined, file.path(out_sub_dir, "branch_support_published_tree.csv"))
combined
```

# Where the song clusters sit

A compact summary of the question the removal step was meant to address: which
cell types the song-nucleus clusters group with, at each threshold.

```{r song-placement}
## Smallest clade containing a song cluster: its immediate sister group.
sister_of = function(hc, tip) {
  clades = hclust_clades(hc)
  containing = keep(clades, ~ tip %in% .x)
  if (!length(containing)) return(list(sister = character(0), clade = NA_character_))
  smallest = containing[[which.min(map_int(containing, length))]]
  list(sister = setdiff(smallest, tip), clade = clade_key(smallest))
}

song_tips = map(regions, function(reg)
  grep(song_clusters[[reg]], colnames(obj_int_avg1), value = TRUE) %>%
    intersect(colnames(obj_int_avg1)[grepl(region_cols[[reg]], colnames(obj_int_avg1))])) %>%
  set_names(regions)

## Any cluster from a song nucleus, for telling song/non-song pairs from
## song/song pairs
all_song_tips = unique(unlist(song_tips))

placement = map_dfr(sweep, function(s) {
  hc = s$pv$hclust
  au_by_clade = set_names(s$pv$edges$au, map_chr(hclust_clades(hc), clade_key))
  tips = intersect(song_tips[[s$region]], hc$labels)
  map_dfr(tips, function(tp) {
    r = sister_of(hc, tp)
    tibble(region = s$region, threshold_label = s$threshold_label,
           song_cluster = tp,
           sister = paste(r$sister, collapse = ", "),
           sister_list = list(r$sister),
           clade = r$clade,
           au = unname(au_by_clade[r$clade]),
           ## TRUE when the song cluster pairs with non-song cell types only
           song_nonsong = length(r$sister) > 0 && !any(r$sister %in% all_song_tips))
  })
}) %>%
  mutate(threshold_label = factor(threshold_label, levels = lev)) %>%
  arrange(region, song_cluster, threshold_label)

write_csv(select(placement, -sister_list),
          file.path(out_sub_dir, "song_cluster_placement.csv"))
select(placement, -sister_list)
```

## Agreement of the song / non-song pairings

The trees as a whole can rearrange without touching the relationships the
manuscript actually rests on, which are local: each song-nucleus cluster and the
non-song cell types it pairs with. This restricts the agreement measure to those
pairings. For every song cluster, its sister group is compared with the sister
group in the **published split trees at p_adj < `r ref_lab`** (Jaccard index of
the two sister sets: 1 = identical partners, 0 = no partner in common). The
same reference is used for the pooled `ALL` analysis, so that panel shows
whether pooling reproduces the published pairings.

```{r song-pair-agreement}
#| fig-width: 5
#| fig-height: 3.5
#| results: hold
jaccard = function(a, b) {
  if (!length(union(a, b))) return(1)
  length(intersect(a, b)) / length(union(a, b))
}

## reference = the published split trees (ARCO / NIDO) at the published threshold
ref_pairs = placement %>%
  filter(region %in% c("ARCO", "NIDO"), threshold_label == ref_lab) %>%
  select(song_cluster, ref_sister = sister_list)

pair_agree = placement %>%
  left_join(ref_pairs, by = "song_cluster") %>%
  mutate(agreement = map2_dbl(sister_list, ref_sister, jaccard),
         region = factor(region, levels = regions),
         song_cluster = factor(song_cluster, levels = rev(sort(unique(song_cluster)))))
write_csv(select(pair_agree, -sister_list, -ref_sister),
          file.path(out_sub_dir, "song_pair_agreement.csv"))

## sister names, one per line and without the shared "Glut-" prefix, so that
## they fit inside a tile
pair_agree = pair_agree %>%
  mutate(sister_lab = str_replace_all(str_replace_all(sister, "Glut-", ""), ", ", "\n"))

## One plot per analysis, with square tiles (coord_fixed on two discrete
## scales, each level one unit wide).
tile_in   = 0.4   # inches per tile
cell_pt   = 5     # in-tile text (the smallest type used anywhere here)
axis_pt   = 6     # axis text and titles

pair_plot = function(reg) {
  d = filter(pair_agree, region == reg) %>%
    mutate(song_cluster = factor(song_cluster, levels = rev(sort(unique(as.character(song_cluster))))))
  ggplot(d, aes(threshold_label, song_cluster)) +
    geom_tile(aes(fill = agreement), color = "grey70", linewidth = 0.2) +
    geom_text(aes(label = sister_lab, color = agreement > 0.5),
              size = cell_pt / .pt, lineheight = 0.9) +
    ## a song cluster pairing with another song cluster rather than with
    ## non-song cell types is marked in the corner of its tile. The nudge is in
    ## the discrete scale's own space.
    geom_point(data = filter(d, !song_nonsong),
               position = position_nudge(x = -0.42, y = 0.36),
               shape = 4, size = 1, stroke = 0.5, color = "firebrick") +
    coord_fixed(ratio = 1) +
    scale_fill_gradient(low = "white", high = "#2166AC", limits = c(0, 1),
                        name = "agreement with\npublished pairing") +
    scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "grey20"), guide = "none") +
    labs(x = "song-DEG adjusted-p threshold", y = NULL,
         title = region_name[[reg]],
         caption = "tile text = cell types the song cluster pairs with\nx = paired with another song cluster rather than with non-song cell types") +
    theme_cowplot(font_size = axis_pt) +
    theme(axis.text = element_text(size = axis_pt),
          axis.text.x = element_text(size = axis_pt, angle = 45, hjust = 1),
          axis.title = element_text(size = axis_pt),
          legend.title = element_text(size = axis_pt),
          legend.text = element_text(size = cell_pt),
          legend.key.size = unit(0.16, "in"),
          plot.title = element_text(size = 7),
          plot.caption = element_text(size = cell_pt, hjust = 0))
}

for (reg in regions) {
  p_pair = pair_plot(reg)
  print(p_pair)
  n_rows = n_distinct(filter(pair_agree, region == reg)$song_cluster)
  save_plot(file.path(out_sub_dir, sprintf("song_pair_agreement_%s.pdf", tolower(reg))),
            p_pair,
            base_width  = length(lev) * tile_in + 1.9,   # + y labels and legend
            base_height = n_rows * tile_in + 1.1)        # + x labels and caption
}
```

```{r song-pair-agreement-summary}
#| fig-width: 8
#| fig-height: 3.5
p_pair_mean = pair_agree %>%
  group_by(region, threshold_label) %>%
  summarise(mean_agreement = mean(agreement),
            frac_song_nonsong = mean(song_nonsong), .groups = "drop") %>%
  pivot_longer(c(mean_agreement, frac_song_nonsong)) %>%
  mutate(name = recode(name,
                       mean_agreement    = "mean agreement with published pairing",
                       frac_song_nonsong = "fraction pairing with non-song types")) %>%
  ggplot(aes(threshold_label, value, color = region, group = region)) +
  geom_line() + geom_point() +
  facet_wrap(~name) +
  ylim(0, 1) +
  labs(x = "song-DEG adjusted-p threshold", y = NULL) +
  theme_cowplot(font_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_pair_mean)
save_plot(file.path(out_sub_dir, "song_pair_agreement_summary.pdf"), p_pair_mean,
          base_height = 3.2, base_width = 9)
```

# Session info

```{r session}
sessionInfo()
```
