---
title: "Robustness of glutamatergic cell-type clustering to song-DEG removal (hybrid labels)"
format:
  html:
    toc: true
    df-print: paged
---

Repeats `celltypes_hclust_glut_robustness.qmd`'s analysis on the current hybrid division-based
glutamatergic labels (`snrna/naming/hybrid_division_naming.qmd`'s `celltype_hybrid`) instead of the
raw `cluster` identity. Independent of both source notebooks, by the same convention used throughout
this project's naming work: reads their outputs read-only, writes its own copy.

**What changes under the hybrid labels, and why -- checked directly, not assumed:**

- **Cluster set.** `Glut-Arco-4` and `Glut-Nido-3` (confirmed artifacts, `celltype_hybrid` is `NA`)
  are dropped, matching every other hybrid-labels notebook in this project. `Glut-Nido-1` is merged
  into `Glut-DACH2-1` together with `Glut-NC-1` (the same population, confirmed elsewhere) -- one
  fewer distinct cell type in the NIDO tree than the published analysis.
- **Cluster exclusion.** The published analysis excludes the "Pre" precursor clusters and
  `Glut-HVC-1a` by matching `"Pre"`/`"1a"` as literal substrings of the `cluster` label. Under the
  hybrid scheme this substring match is broken two different ways at once: the precursor labels
  (`Glut-NSC`/`Glut-NB`/`Glut-Im`) no longer contain `"Pre"` at all (so the old pattern would silently
  stop excluding them), and `Glut-HVC-1a` is now named `Glut-DACH2-HVCra-Int`, which does not contain
  `"Pre"` either (so the old pattern would silently stop excluding it too). An interim `-Pre` name for
  that cluster did match, but only by accident, not because the pattern still meant what it used to --
  which is exactly why the substring match is not trusted here. Replaced with an explicit name list -- see `glut_exclude`
  below -- so this can't silently drift the way `celltypes_hclust_all_hybrid.qmd` already found once
  with the `"Pre"`-substring GABA/Glut bar-color classification. `glut_exclude` also drops
  `Glut-SATB2-1` and `Glut-GABA` entirely (per direct instruction) -- neither is a `DACH2`/`CACNA1H`
  cluster, and the ARCO/NIDO split below is now defined by that division prefix, so keeping them in
  would leave them with no region to belong to.
- **`region_cols` (which pseudobulk columns belong to the ARCO vs. NIDO dissection).** The published
  analysis selects these by matching `"RA|Arco"` / region-prefixed substrings in the cluster label
  itself, which worked because those clusters were literally *named* after their dissection region.
  The hybrid scheme deliberately drops region-based naming for every non-song cluster (that's the
  point of it) but keeps the major excitatory division (`DACH2`/`CACNA1H`/`SATB2`) in every label, and
  that division tracks the dissection directly -- `DACH2` clusters came from the nidopallial
  dissection, `CACNA1H` from the arcopallial one (per direct instruction: subset by that marker
  prefix rather than by dissection metadata). Checked directly before trusting it (see the
  `region-membership` chunk): every `DACH2` cluster is >85% `NIDO` by actual dissection position and
  every `CACNA1H` cluster is 100% `ARCO`, so the prefix is a faithful, simpler stand-in for the
  published analysis's own region-prefix substring match -- not a new judgment call being introduced
  quietly.
- **Tip coloring by dissection position (`tip_annotation`).** Same underlying issue: the published
  version parses the *label string* (`Glut-([A-Za-z]+)` -> "Arco"/"NC"/"HVC"/...) to color tree tips.
  That parse would return nonsense on a hybrid label directly (e.g. "CACNA1H" is not a position).
  The fix is *not* to fall back on each cluster's real per-cell position majority, though -- a first
  attempt at that did, and it was wrong for two clusters: `Glut-NC-2` (-> `Glut-DACH2-2`) is a
  near-even hvc/nc split in the actual data (479 vs. 465 cells) and `Glut-NR-4` (-> `Glut-DACH2-8`) a
  near-even lman/nr split (350 vs. 320), so a real-data vote colors them by whichever side narrowly
  wins (hvc, lman) rather than by their *published* dissection identity (nc, nr) -- exactly the
  mismatch reported directly against the rendered trees. Fixed by keeping the raw, pre-relabeling
  cluster name (`orig_cluster`, preserved in the `load` chunk before `cluster` is overwritten with the
  hybrid label) and running the published notebook's *own* parsing rule against it instead of against
  the hybrid label -- so the tip color is still the cluster's nominal/published identity, just reached
  through the original name rather than the new one.
- **Pseudobulk replicate structure.** The published analysis pools cells by `(cluster, assignment)`
  for `AggregateExpression()`. `assignment` on this backup object is souporcell's per-library index,
  confirmed elsewhere in this project's work (`snrna/naming/exploratory/apply_composite_labels.qmd`) to NOT be comparable
  across libraries -- the same numeral means a different bird in a different library, so pooling on
  bare `assignment` risks silently merging two different birds' cells into one "replicate" and
  understating true biological variance. Fixed here the same way it was fixed there:
  `replicate_lib = paste(sample_name, assignment)`.
- **Marker cache.** The published analysis reuses `celltypes_hclust_glut_no1a`'s cached DESeq2 marker
  tables (`markers_cache`/`markers_cache_all`) as a shortcut. Those were computed on the *raw* cluster
  grouping, which is a genuinely different grouping now (NC-1/Nido-1 merged, Arco-4/Nido-3/HVCra-Int
  dropped) -- reusing them would silently mix marker calls computed against one cell/cluster
  partition into an analysis using another. Dropped; markers are recomputed fresh for this notebook
  and cached locally instead (`markers_fname`, as the published notebook itself already does for its
  own first run).
- **Everything else** (thresholds swept, `nsig`/`feature_thresh`, `ward.D` on correlation distance,
  `pvclust` and cell-level bootstrap replicate counts, tree/plot code) is unchanged from the published
  analysis.

```{r setup}
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(Matrix)
library(pvclust)
library(ComplexHeatmap)
library(circlize)
library(ggtree)
library(ape)
library(cowplot)
library(here)
set.seed(10)

source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "scRNA.R"))
source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "stats.R"))
source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "common_aesthetics.R"))

pnames = names(position_colors)
names(position_colors) = case_when(pnames == "nido" ~ "nr",
                                   pnames == "ncl" ~ "nc",
                                   TRUE ~ pnames)
```

# Parameters

```{r params}
hclust_method = "ward.D"
dist_method   = "correlation"
assay_to_use  = "SCT"
res_to_use    = "celltype_hybrid"

## Number of top markers retained per cluster per direction (as published)
nsig = 100
## Significance threshold used to *select* the clustering features (as published)
feature_thresh = 0.1

## Song-DEG removal thresholds to sweep. NA = no removal (i.e. the clustering
## as it would be done before any song DEGs are taken out).
deg_thresholds = c(NA, 0.5, 0.25, 0.1, 0.05, 0.01, 0.001)
## The threshold used in the manuscript, for reference comparisons
ref_threshold = 0.1

## pvclust multiscale bootstrap replicates
nboot = 1000
n_cores = min(20L, max(1L, parallel::detectCores() - 4L))

## Cell-level bootstrap replicates
n_cellboot = 500

## Song-nucleus clusters, per analysis, as EXACT hybrid labels/patterns (not
## region-prefix substrings -- see the note above). ALL = every glutamatergic
## cell type clustered together, without splitting the dissections.
song_clusters = list(ARCO = "^Glut-CACNA1H-RA$",
                     NIDO = "HVC|LMAN",
                     ALL  = "^Glut-CACNA1H-RA$|HVC|LMAN")
regions = c("ARCO", "NIDO", "ALL")

## Excluded from the Glut set entirely, matching the published analysis's own
## exclusion of the "Pre" precursor clusters and Glut-HVC-1a -- by explicit
## name now, not a substring that no longer means what it used to (see above).
glut_exclude = c("Glut-NSC", "Glut-NB", "Glut-Im", "Glut-DACH2-HVCra-Int",
                 "Glut-SATB2-1", "Glut-GABA")

## Element-wise label for a threshold (format() on a vector pads to a common
## width, which would break the joins below, so map over elements).
thr_label = function(x) {
  vapply(x, function(v) if (is.na(v)) "none" else format(v, scientific = FALSE),
         character(1))
}
## Ordered threshold labels, most genes removed -> fewest, with "none" (zero genes removed) placed
## at the end rather than the front -- it belongs at the low-removal extreme, not before it.
lev = c(thr_label(sort(deg_thresholds[!is.na(deg_thresholds)], decreasing = TRUE)), "none")
ref_lab = thr_label(ref_threshold)
```

# Directories

```{r dirs}
script_name = "celltypes_hclust_glut_robustness_hybrid"

obj_fname = here::here("snrna/naming/hybrid_division_naming", "obj_hybrid_labels.qs2")

out_dir     = here::here("snrna/trees", script_name)
out_sub_dir = file.path(out_dir, str_interp("${dist_method}_${hclust_method}"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(out_sub_dir, recursive = TRUE, showWarnings = FALSE)

obj_out_fname      = file.path(out_dir, "obj_clustered_glutamatergic.qs2")
data_out_avg_fname = file.path(out_dir, "data_avg.qs2")
markers_fname      = file.path(out_dir, sprintf("marker_genes_cluster_%s.rds", assay_to_use))
```

# Load data

Same cell subset logic as the published analysis, adapted for the hybrid labels: glutamatergic
clusters, excluding the developmental-stage clusters, the HVC-1a-equivalent cluster, and cells whose
dissection assignment spans two regions -- plus the confirmed-artifact clusters, which the published
analysis (built before those exclusions existed) did not need to drop explicitly.

```{r load}
redo = !file.exists(obj_out_fname)
if (redo) {
  obj = qs_read(obj_fname, nthreads = 8)
  cells = colnames(obj)[grepl("^Glut", obj@meta.data[[res_to_use]]) &
                         !is.na(obj@meta.data[[res_to_use]]) &
                         !(obj@meta.data[[res_to_use]] %in% glut_exclude)]
  obj_sub = subset(obj, cells = cells)

  ## No-op on this object (confirmed elsewhere: 0 of 34295 cells have a "/" in
  ## `assignment`, already filtered to both-caller singlets upstream) -- kept so this
  ## notebook doesn't silently rely on that upstream state, matching the published version.
  cells = colnames(obj_sub)[!grepl("/", obj_sub$assignment)]
  obj_sub = subset(obj_sub, cells = cells)

  qs_save(obj_sub, obj_out_fname, nthreads = 8)
  rm(obj); gc()
} else {
  obj_sub = qs_read(obj_out_fname, nthreads = 8)
}

obj_sub$region = case_when(obj_sub$position %in% c("ra", "arco") ~ "ARCO",
                           obj_sub$position %in% c("lman", "nr", "hvc", "nc") ~ "NIDO")
## The raw, published cluster label, kept under its own name before `cluster` is overwritten below --
## needed by hybrid_position_lookup, which colors tree tips by the cluster's PUBLISHED/nominal
## dissection identity (as celltypes_hclust_glut_robustness.qmd's own tip_annotation() did), not by
## a real per-cell position vote. See the region-membership chunk for why that distinction matters.
obj_sub$orig_cluster = droplevels(as.factor(obj_sub$cluster))
obj_sub$cluster = droplevels(as.factor(obj_sub@meta.data[[res_to_use]]))
## Cross-library-safe pseudoreplicate id -- see the note on the pseudobulk replicate structure above.
obj_sub$replicate_lib = paste(obj_sub$sample_name, obj_sub$assignment, sep = "_")
table(obj_sub$cluster, obj_sub$region)
```

## Region membership: subset by marker-prefix (division), validated against dissection position

Per direct instruction, `region_cols` is now defined by each cluster's division prefix --
`Glut-DACH2-*` -> NIDO, `Glut-CACNA1H-*` -> ARCO -- not by looking up dissection position at all.
That's a simplification of the data-driven majority-region check this notebook used before, so it is
kept here as a validation rather than dropped outright: every `DACH2` cluster really should be
overwhelmingly (checked: at least 85%) `NIDO` by actual dissection position, and every `CACNA1H`
cluster `ARCO`, or the prefix-based split would silently be wrong.

```{r region-membership}
region_majority = obj_sub@meta.data %>%
  count(cluster, region) %>%
  group_by(cluster) %>%
  mutate(frac = n / sum(n)) %>%
  slice_max(frac, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(division = case_when(grepl("^Glut-DACH2-", cluster) ~ "NIDO",
                              grepl("^Glut-CACNA1H-", cluster) ~ "ARCO",
                              TRUE ~ NA_character_))
stopifnot(
  "glut_exclude should have removed every cluster outside the DACH2/CACNA1H prefixes" =
    !anyNA(region_majority$division)
)
stopifnot(
  "a DACH2/CACNA1H cluster's dissection-position majority disagrees with its division prefix -- the prefix-based region_cols would be wrong" =
    all(region_majority$region == region_majority$division)
)
region_majority %>% arrange(frac) %>% select(cluster, division, region, frac, n)

## Column-selection regex per analysis, built directly from the division prefix (validated above),
## not from dissection metadata. Anchored so specificity_matrix()'s existing grepl-based column
## selection (built for the published analysis's region-prefix patterns) needs no further change.
region_cols = list(
  ARCO = "^Glut-CACNA1H-",
  NIDO = "^Glut-DACH2-",
  ALL  = "^Glut-(CACNA1H|DACH2)-"
)

## Lookup for tip coloring (see tip_annotation below): each hybrid cluster's PUBLISHED/nominal
## position, parsed from the raw cluster label exactly the way
## celltypes_hclust_glut_robustness.qmd's own tip_annotation() did (Glut-([A-Za-z]+) extract, with
## "Nido" -> "NC" as that notebook itself hardcodes) -- deliberately NOT a real per-cell majority
## vote. That distinction matters here: Glut-NC-2 (-> Glut-DACH2-2) is a near-even hvc/nc split in
## the actual data (479 vs. 465 cells) and Glut-NR-4 (-> Glut-DACH2-8) a near-even lman/nr split (350
## vs. 320) -- confirmed directly against the object -- so a raw-data vote colors them by whichever
## side narrowly wins (hvc, lman) rather than by their published dissection identity (nc, nr), which
## is what this notebook is trying to reproduce for comparison against the published tree.
hybrid_position_lookup = obj_sub@meta.data %>%
  distinct(cluster, orig_cluster) %>%
  mutate(orig_position = str_extract(as.character(orig_cluster), "Glut-([A-Za-z]+)", group = 1),
        orig_position = if_else(orig_position == "Nido", "NC", orig_position),
        position = tolower(orig_position)) %>%
  distinct(cluster, position)

## The one genuine merge (Glut-DACH2-1 = Glut-NC-1 + Glut-Nido-1) is nominally "nc" on both sides of
## the merge (via the Nido -> NC override above), so it collapses to a single row here rather than
## producing two conflicting positions for the same hybrid cluster -- checked, not assumed.
stopifnot(
  "a hybrid cluster maps to more than one nominal (published) dissection position" =
    !anyDuplicated(hybrid_position_lookup$cluster)
)
```

# Markers

Pseudobulk (cluster x library-qualified individual) DESeq2 markers, computed separately within each
dissection region. These define both the song-DEG sets that get removed and the feature sets used for
clustering. Recomputed fresh for this notebook -- see the note on the marker cache above.

```{r markers}
compute_markers = function(obj_cur) {
  obj_agg = AggregateExpression(obj_cur, assay = assay_to_use,
                                group.by = c("cluster", "replicate_lib"),
                                return.seurat = TRUE)
  Idents(obj_agg) = obj_agg$cluster
  FindAllMarkers(obj_agg, assay = assay_to_use, test.use = "DESeq2", only.pos = FALSE)
}

markers_agg = if (file.exists(markers_fname)) readRDS(markers_fname) else list()
markers_new = FALSE
for (reg in regions) {
  if (!is.null(markers_agg[[reg]])) next
  markers_new = TRUE
  markers_agg[[reg]] =
    if (reg == "ALL") {
      compute_markers(obj_sub)
    } else {
      compute_markers(subset(obj_sub, region == reg))
    }
}
if (markers_new) saveRDS(markers_agg, markers_fname)
map(markers_agg[regions], ~ count(.x, cluster))
```

# Average expression

```{r avg}
if (file.exists(data_out_avg_fname)) {
  obj_int_avg1 = qs_read(data_out_avg_fname, nthreads = 8)
} else {
  Idents(obj_sub) = obj_sub$cluster
  obj_int_avg = AggregateExpression(obj_sub, assays = assay_to_use)
  obj_int_avg1 = log1p(obj_int_avg[[assay_to_use]])
  colnames(obj_int_avg1) = colnames(obj_int_avg[[assay_to_use]])
  qs_save(obj_int_avg1, data_out_avg_fname, nthreads = 8)
}
dim(obj_int_avg1)
```

# Helper functions

```{r helpers}
## Genes called as song DEGs at a given adjusted-p threshold.
## thresh = NA -> no genes removed (clustering "before" song-DEG removal).
song_degs = function(markers, song_pattern, thresh) {
  if (is.na(thresh)) return(character(0))
  markers %>%
    filter(grepl(song_pattern, cluster)) %>%
    filter(p_val_adj < thresh) %>%
    pull(gene) %>%
    unique()
}

## Feature set used for clustering, exactly as in celltypes_hclust_glut.qmd
## except that the removed-gene set is supplied.
select_features = function(markers, removed, nsig, feature_thresh) {
  markers %>%
    filter(!(gene %in% removed)) %>%
    mutate(sign = avg_log2FC > 0) %>%
    filter(p_val_adj < feature_thresh) %>%
    group_by(cluster, sign) %>%
    top_n(nsig, abs(avg_log2FC)) %>%
    ungroup() %>%
    distinct(gene, .keep_all = TRUE) %>%
    pull(gene)
}

## genes x cell-types matrix of scaled gene specificity, as published
specificity_matrix = function(avg, genes, col_pattern) {
  m = avg[genes, grepl(col_pattern, colnames(avg)), drop = FALSE]
  m = t(apply(m, 1, calc_gene_specificity, scale = TRUE))
  m[is.nan(m)] = 0
  as.matrix(m)
}

## Tip sets of every internal node of an hclust, in hclust merge order
## (i.e. the same order as pvclust's $edges rows).
hclust_clades = function(hc) {
  out = vector("list", nrow(hc$merge))
  for (i in seq_len(nrow(hc$merge))) {
    tips = character(0)
    for (j in hc$merge[i, ])
      tips = c(tips, if (j < 0) hc$labels[-j] else out[[j]])
    out[[i]] = sort(tips)
  }
  out
}
clade_key = function(tips) paste(sort(tips), collapse = "|")

## Tip sets for every node of a phylo object, indexed by node number.
phylo_tipsets = function(phy) {
  ntip = length(phy$tip.label)
  desc = vector("list", max(phy$edge))
  for (i in seq_len(ntip)) desc[[i]] = phy$tip.label[i]
  for (e in rev(seq_len(nrow(phy$edge)))) {
    p = phy$edge[e, 1]; ch = phy$edge[e, 2]
    desc[[p]] = c(desc[[p]], desc[[ch]])
  }
  desc
}

## Map hclust-ordered per-edge values (e.g. pvclust AU) onto ggtree node ids.
edge_values_to_nodes = function(hc, values) {
  phy = ape::as.phylo(hc)
  hc_keys = map_chr(hclust_clades(hc), clade_key)
  ts = phylo_tipsets(phy)
  node_keys = map_chr(ts, ~ if (is.null(.x)) NA_character_ else clade_key(.x))
  tibble(node = seq_along(node_keys), key = node_keys) %>%
    filter(!is.na(key)) %>%
    left_join(tibble(key = hc_keys, value = values), by = "key") %>%
    filter(!is.na(value))
}

## correlation-distance ward.D tree, matching pvclust's settings
build_tree = function(mat) {
  d = as.dist(1 - cor(mat))
  hclust(d, method = hclust_method)
}
```

# Threshold sweep

For each region and each song-DEG threshold, rebuild the feature set, recompute
the cell-type tree, and run the `pvclust` multiscale bootstrap.

```{r sweep}
#| message: false
sweep_fname = file.path(out_sub_dir, "pvclust_sweep.qs2")

## Cached per analysis x threshold, so that adding an analysis does not force
## the previously computed bootstraps to be rerun.
sweep = if (file.exists(sweep_fname)) qs_read(sweep_fname, nthreads = 8) else list()
sweep_new = FALSE
for (region_cur in regions) {
  for (thr in deg_thresholds) {
    thr_lab = thr_label(thr)
    key = paste(region_cur, thr_lab, sep = "_")
    if (!is.null(sweep[[key]])) next
    sweep_new = TRUE
    message("== ", region_cur, " threshold ", thr_lab)

    removed  = song_degs(markers_agg[[region_cur]], song_clusters[[region_cur]], thr)
    features = select_features(markers_agg[[region_cur]], removed, nsig, feature_thresh)
    mat      = specificity_matrix(obj_int_avg1, features, region_cols[[region_cur]])

    pv = pvclust(mat, method.dist = dist_method, method.hclust = hclust_method,
                 nboot = nboot, parallel = as.integer(n_cores))

    sweep[[key]] = list(region = region_cur, threshold = thr, threshold_label = thr_lab,
                        n_removed = length(removed), n_features = length(features),
                        features = features, removed = removed, pv = pv)
  }
}
if (sweep_new) qs_save(sweep, sweep_fname, nthreads = 8)
## keep only the analyses requested by `regions`, in that order
sweep = sweep[unlist(map(regions, function(r) paste(r, thr_label(deg_thresholds), sep = "_")))]

## ALL dropped from this table/plot specifically (per direct instruction) -- it stays in `sweep`
## itself and in every other section below (trees, topology, bootstrap, song placement), which are
## separate analyses, not "the sweep analysis" this instruction was scoped to.
sweep_df = map_dfr(sweep, function(s) {
  au = s$pv$edges$au
  tibble(region = s$region, threshold_label = s$threshold_label,
         threshold = s$threshold, n_removed = s$n_removed,
         n_features = s$n_features,
         n_edges = length(au), mean_au = mean(au), median_au = median(au),
         n_au95 = sum(au >= 0.95), frac_au95 = mean(au >= 0.95))
}) %>%
  filter(region != "ALL") %>%
  mutate(threshold_label = factor(threshold_label, levels = lev)) %>%
  arrange(region, threshold_label)

write_csv(sweep_df, file.path(out_sub_dir, "sweep_summary.csv"))
sweep_df
```

```{r sweep-plot}
#| fig-width: 9
#| fig-height: 6
p_counts = sweep_df %>%
  pivot_longer(c(n_removed, n_features, frac_au95), names_to = "metric") %>%
  mutate(metric = recode(metric,
                         n_removed  = "song DEGs removed",
                         n_features = "genes used for clustering",
                         frac_au95  = "fraction of branches AU >= 0.95")) %>%
  ggplot(aes(threshold_label, value, group = region, color = region)) +
  geom_line() + geom_point() +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "song-DEG adjusted-p threshold (none = no removal)", y = NULL) +
  theme_cowplot(font_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_counts)
save_plot(file.path(out_sub_dir, "sweep_summary.pdf"), p_counts,
          base_height = 3.5, base_width = 10)
```

# Trees across thresholds

Each tree is annotated with the `pvclust` approximately-unbiased (AU) support,
in percent, at every internal branch. Tips are coloured by dissection position.

```{r tree-fn}
## Display name for each analysis
region_name = c(ARCO = "ARCO", NIDO = "NIDO",
                ALL = "ALL (nido + arco combined)")

## Two lines, so that the longest title still fits the narrow page.
tree_title = function(region, threshold_label) {
  nm = if (region %in% names(region_name)) region_name[[region]] else region
  if (threshold_label == "none")
    sprintf("%s\nno song DEGs removed", nm)
  else
    sprintf("%s\nsong DEGs removed at p_adj < %s", nm, threshold_label)
}

## Position looked up directly from the data (hybrid_position_lookup, built in the
## region-membership chunk) rather than parsed from the label string -- see the note at the top of
## this notebook on why the old Glut-([A-Za-z]+) parse doesn't work for hybrid labels.
tip_annotation = function(labels) {
  data.frame(labels = labels) %>%
    left_join(hybrid_position_lookup, by = c("labels" = "cluster")) %>%
    mutate(col = position_colors[match(position, names(position_colors))])
}

## Tree geometry, matched to celltypes_hclust_glut.qmd: geom_tiplab at offset 0,
## hexpand(5), and a panel fixed by egg::set_panel_size so rows sit tight
## against one another. Font sizes are given in points and converted to the mm
## units that ggplot's `size` aesthetic expects; the per-tip row height tracks
## the tip font so that shrinking the text does not reopen gaps between labels.
tip_pt           = 6     # tip labels
support_pt       = 5     # branch-support labels (absent in the original)
tip_size         = tip_pt / .pt
support_size     = support_pt / .pt
tip_scale_factor = tip_pt * 1.45 / 72   # panel inches per tip (~1.45 line spacing)
panel_width      = 2                    # panel inches
## Branch line weight, given as an R lwd (lwd 1 = 1/96 in = 0.75 pt) and
## converted to the mm units ggplot's `linewidth` expects. ggtree's default is
## linewidth 0.5, i.e. lwd 1.42, which is heavier than this.
branch_lwd       = 0.5
branch_linewidth = branch_lwd / .pt

## hc:      hclust to draw
## values:  per-edge values in hclust merge order (e.g. pvclust AU), or NULL
## Returns a fixed-panel gtable; print()/plot() it, and size the device with
## tree_device_dim() so that there is no slack around the panel.
support_tree = function(hc, values, label_color = "grey30", title = NULL, subtitle = NULL) {
  hc_df = tip_annotation(hc$labels)
  g = ggtree(hc, branch.length = "none", linewidth = branch_linewidth) %<+% hc_df
  if (!is.null(values)) {
    node_df = edge_values_to_nodes(hc, values) %>%
      transmute(node, support_lab = as.character(round(value * 100)))
    g = g %<+% node_df
  }
  g = g +
    geom_tiplab(aes(color = col), angle = 0, hjust = 0, offset = 0,
                show.legend = FALSE, size = tip_size) +
    theme_dendrogram(plot.margin = margin(20, 20, 20, 20)) +
    scale_color_identity() +
    hexpand(5) +
    labs(title = title, subtitle = subtitle) +
    theme(plot.title = element_text(size = 9, face = "plain"),
          plot.subtitle = element_text(size = 7))
  if (!is.null(values))
    ## the extra left expansion keeps the root's label inside the panel
    g = g + geom_text2(aes(label = support_lab, subset = !isTip),
                       hjust = 1.15, vjust = -0.4, size = support_size,
                       color = label_color) +
      ggexpand(0.25, direction = -1)
  egg::set_panel_size(g,
                      width  = unit(panel_width, "in"),
                      height = unit(length(hc$labels) * tip_scale_factor, "in"))
}

## Device size that just contains the fixed panel plus margins and the title.
tree_device_dim = function(n_tips) {
  list(width  = panel_width + 2.0,
       height = n_tips * tip_scale_factor + 1.4)   # room for the title block
}

## Draw fixed-panel gtables one per page. plot() would call grid.newpage()
## before the first tree as well, leaving a blank leading page in the PDF.
draw_pages = function(plots) {
  for (i in seq_along(plots)) {
    if (i > 1) grid::grid.newpage()
    grid::grid.draw(plots[[i]])
  }
}

plot_tree = function(s, show_au = TRUE) {
  support_tree(s$pv$hclust,
               values = if (show_au) s$pv$edges$au else NULL,
               title = tree_title(s$region, s$threshold_label),
               subtitle = sprintf("%d genes removed, %d genes clustered\nbranch labels = pvclust AU (%%)",
                                  s$n_removed, s$n_features))
}
```

```{r trees}
#| fig-width: 5
#| fig-height: 7
#| results: hold
for (reg in regions) {
  keys = names(sweep)[map_chr(sweep, "region") == reg]
  n_tips = length(sweep[[keys[1]]]$pv$hclust$labels)
  dim = tree_device_dim(n_tips)
  walk(keys, ~ plot(plot_tree(sweep[[.x]])))
  pdf(file.path(out_sub_dir, sprintf("trees_%s_thresholds.pdf", tolower(reg))),
      height = dim$height, width = dim$width)
  draw_pages(map(keys, ~ plot_tree(sweep[[.x]])))
  dev.off()
}
```

# Topology agreement across thresholds

Two summaries of how much the tree changes with the threshold:

* **Cophenetic / Baker's gamma correlation** between the full dendrograms
  (Pearson and Spearman correlation of the cophenetic distances).
* **Adjusted Rand index** between the cell-type groupings obtained by cutting
  each tree at k = 2..8.

```{r topology}
coph = map(sweep, ~ as.vector(cophenetic(.x$pv$hclust)))

pairs_df = map_dfr(regions, function(reg) {
  keys = names(sweep)[map_chr(sweep, "region") == reg]
  expand_grid(a = keys, b = keys) %>%
    mutate(region = reg,
           lab_a = map_chr(a, ~ sweep[[.x]]$threshold_label),
           lab_b = map_chr(b, ~ sweep[[.x]]$threshold_label),
           cophenetic_r = map2_dbl(a, b, ~ cor(coph[[.x]], coph[[.y]])),
           baker_gamma  = map2_dbl(a, b, ~ cor(coph[[.x]], coph[[.y]], method = "spearman")),
           ari = map2_dbl(a, b, function(x, y) {
             ks = 2:min(8, length(sweep[[x]]$pv$hclust$labels) - 1)
             mean(map_dbl(ks, function(k)
               mclust::adjustedRandIndex(cutree(sweep[[x]]$pv$hclust, k)[sweep[[x]]$pv$hclust$labels],
                                         cutree(sweep[[y]]$pv$hclust, k)[sweep[[x]]$pv$hclust$labels])))
           }))
})
write_csv(select(pairs_df, -a, -b), file.path(out_sub_dir, "topology_agreement_pairwise.csv"))

## versus the published threshold
ref_df = pairs_df %>%
  filter(lab_b == ref_lab) %>%
  select(region, threshold_label = lab_a, cophenetic_r, baker_gamma, ari)
write_csv(ref_df, file.path(out_sub_dir, "topology_agreement_vs_published.csv"))
ref_df
```

```{r topology-plot}
#| fig-width: 9
#| fig-height: 4
## ALL dropped from this plot specifically (per direct instruction, same as sweep_summary and
## song_pair_agreement_summary above) -- it stays in ref_df itself (topology_agreement_vs_published.csv)
## and in the per-region topology_ari_<region>.pdf heatmaps below.
p_top = ref_df %>%
  filter(region != "ALL") %>%
  pivot_longer(c(cophenetic_r, baker_gamma, ari)) %>%
  mutate(threshold_label = factor(threshold_label, levels = lev),
         name = recode(name,
                       cophenetic_r = "cophenetic r",
                       baker_gamma  = "Baker's gamma",
                       ari          = "mean ARI (k = 2..min(8, n-1))")) %>%
  ggplot(aes(threshold_label, value, color = region, group = region)) +
  geom_line() + geom_point() +
  facet_wrap(~name) +
  ylim(0, 1) +
  labs(x = "song-DEG adjusted-p threshold", y = "agreement with published tree") +
  theme_cowplot(font_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_top)
save_plot(file.path(out_sub_dir, "topology_agreement.pdf"), p_top,
          base_height = 3.5, base_width = 10)
```

```{r topology-heatmap}
#| fig-width: 10
#| fig-height: 5
for (reg in regions) {
  m = pairs_df %>%
    filter(region == reg) %>%
    select(lab_a, lab_b, ari) %>%
    pivot_wider(names_from = lab_b, values_from = ari) %>%
    column_to_rownames("lab_a") %>%
    as.matrix()
  m = m[lev, lev]
  hm = Heatmap(m, name = "ARI", cluster_rows = FALSE, cluster_columns = FALSE,
               col = colorRamp2(c(0, 1), hcl_palette = "Blues", reverse = TRUE),
               column_title = sprintf("%s: pairwise ARI across thresholds", reg),
               cell_fun = function(j, i, x, y, w, h, fill)
                 grid.text(sprintf("%.2f", m[i, j]), x, y, gp = gpar(fontsize = 8)))
  print(hm)
  pdf(file.path(out_sub_dir, sprintf("topology_ari_%s.pdf", tolower(reg))),
      height = 5, width = 5.5)
  print(hm)
  dev.off()
}
```

# Branch-level bootstrap support

## Multiscale bootstrap over genes (pvclust)

AU support for every branch of the published (p_adj < 0.1) tree, and, for
comparison, for the same branches when they are present in the no-removal tree.

```{r au-table}
au_df = map_dfr(sweep, function(s) {
  hc = s$pv$hclust
  tibble(region = s$region, threshold_label = s$threshold_label,
         clade = map_chr(hclust_clades(hc), clade_key),
         n_tips = map_int(hclust_clades(hc), length),
         au = s$pv$edges$au, bp = s$pv$edges$bp)
})
write_csv(au_df, file.path(out_sub_dir, "branch_support_all.csv"))

ref_clades = au_df %>% filter(threshold_label == ref_lab)

## Is each published branch recovered at other thresholds, and with what support?
## Every published branch is crossed with every threshold, so branches that are
## absent at a threshold show up as NA rather than being dropped.
recovery = ref_clades %>%
  select(region, clade, n_tips, au_published = au) %>%
  crossing(threshold_label = lev) %>%
  left_join(au_df %>% select(region, threshold_label, clade, au_at_threshold = au),
            by = c("region", "clade", "threshold_label")) %>%
  mutate(threshold_label = factor(threshold_label, levels = lev),
         recovered = !is.na(au_at_threshold)) %>%
  arrange(region, desc(n_tips), threshold_label)
write_csv(recovery, file.path(out_sub_dir, "published_branch_recovery_per_branch.csv"))

recovery_summary = recovery %>%
  group_by(region, threshold_label) %>%
  summarise(n_published_branches = n(),
            n_recovered = sum(recovered),
            frac_recovered = mean(recovered),
            mean_au_when_recovered = mean(au_at_threshold, na.rm = TRUE), .groups = "drop") %>%
  arrange(region, threshold_label)
write_csv(recovery_summary, file.path(out_sub_dir, "published_branch_recovery.csv"))
recovery_summary
```

```{r au-plot}
#| fig-width: 9
#| fig-height: 4
p_au = au_df %>%
  mutate(threshold_label = factor(threshold_label, levels = lev)) %>%
  ggplot(aes(threshold_label, au)) +
  geom_hline(yintercept = 0.95, linetype = 2, color = "grey50") +
  geom_boxplot(outlier.shape = NA, fill = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6, aes(color = region)) +
  facet_wrap(~region) +
  labs(x = "song-DEG adjusted-p threshold", y = "pvclust AU support") +
  theme_cowplot(font_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
print(p_au)
save_plot(file.path(out_sub_dir, "branch_support_au.pdf"), p_au,
          base_height = 3.5, base_width = 8)
```

## Bootstrap over cells

`pvclust` resamples *genes*. To estimate stability with respect to the cells
(nuclei) that were sampled, we resample cells with replacement within each cell
type, re-aggregate the pseudobulk profiles, and rebuild the tree with the same
feature set and the same distance/linkage. The feature set is held fixed within
each bootstrap series (re-running the pseudobulk DESeq2 tests per replicate is
not tractable), so this measures sampling variability in the expression
profiles rather than in marker selection.

```{r cellboot}
#| message: false
cellboot_fname = file.path(out_sub_dir, "cellboot.qs2")

boot_tree = function(counts, cl_factor, features, keep_cols) {
  idx = unlist(lapply(split(seq_along(cl_factor), cl_factor),
                      function(ii) sample(ii, length(ii), replace = TRUE)),
               use.names = FALSE)
  cl_b = droplevels(cl_factor[idx])
  ind = sparse.model.matrix(~ 0 + cl_b)
  colnames(ind) = levels(cl_b)
  agg = as.matrix(counts[, idx, drop = FALSE] %*% ind)
  agg = log1p(agg)
  agg = agg[features, intersect(keep_cols, colnames(agg)), drop = FALSE]
  agg = t(apply(agg, 1, calc_gene_specificity, scale = TRUE))
  agg[is.nan(agg)] = 0
  build_tree(as.matrix(agg))
}

cellboot = if (file.exists(cellboot_fname)) qs_read(cellboot_fname, nthreads = 8) else list()
cellboot_todo = unlist(map(regions, ~ paste(.x, c("none", ref_lab), sep = "_")))
cellboot_todo = setdiff(cellboot_todo, names(cellboot))

if (length(cellboot_todo)) {
  counts_all = LayerData(obj_sub, assay = assay_to_use, layer = "counts")
  cl_all = droplevels(as.factor(obj_sub$cluster))

  for (reg in regions) {
    for (thr_lab in c("none", ref_lab)) {
      key = paste(reg, thr_lab, sep = "_")
      if (!(key %in% cellboot_todo)) next
      s = sweep[[key]]
      keep_cols = colnames(obj_int_avg1)[grepl(region_cols[[reg]], colnames(obj_int_avg1))]
      feats = intersect(s$features, rownames(counts_all))

      ## restrict the cells to those contributing to this region's columns
      cell_keep = which(as.character(cl_all) %in% keep_cols)
      counts_reg = counts_all[feats, cell_keep, drop = FALSE]
      cl_reg = droplevels(cl_all[cell_keep])

      trees = map(seq_len(n_cellboot), function(i) {
        if (i %% 50 == 0) message(key, " boot ", i)
        boot_tree(counts_reg, cl_reg, feats, keep_cols)
      })

      ref_hc = s$pv$hclust
      ref_keys = map_chr(hclust_clades(ref_hc), clade_key)
      boot_keys = map(trees, ~ map_chr(hclust_clades(.x), clade_key))
      support = map_dbl(ref_keys, function(k) mean(map_lgl(boot_keys, ~ k %in% .x)))

      cellboot[[key]] = list(region = reg, threshold_label = thr_lab,
                             clade = ref_keys,
                             n_tips = map_int(hclust_clades(ref_hc), length),
                             support = support, hclust = ref_hc)
    }
  }
  qs_save(cellboot, cellboot_fname, nthreads = 8)
  rm(counts_all); gc()
}
cellboot = cellboot[unlist(map(regions, ~ paste(.x, c("none", ref_lab), sep = "_")))]

cellboot_df = map_dfr(cellboot, function(x)
  tibble(region = x$region, threshold_label = x$threshold_label,
         clade = x$clade, n_tips = x$n_tips, cell_bootstrap = x$support))
write_csv(cellboot_df, file.path(out_sub_dir, "cell_bootstrap_support.csv"))

cellboot_df %>%
  group_by(region, threshold_label) %>%
  summarise(n_branches = n(),
            mean_support = mean(cell_bootstrap),
            n_ge_95 = sum(cell_bootstrap >= 0.95),
            n_ge_70 = sum(cell_bootstrap >= 0.70), .groups = "drop")
```

```{r cellboot-trees}
#| fig-width: 5
#| fig-height: 7
#| results: hold
plot_support_tree = function(x) {
  support_tree(x$hclust, values = x$support, label_color = "firebrick",
               title = tree_title(x$region, x$threshold_label),
               subtitle = sprintf("branch labels = %% of %d cell\nbootstraps recovering the branch",
                                  n_cellboot))
}
walk(cellboot, ~ plot(plot_support_tree(.x)))

## one page per analysis, sized to the tallest tree
cb_dim = tree_device_dim(max(map_int(cellboot, ~ length(.x$hclust$labels))))
pdf(file.path(out_sub_dir, "trees_cell_bootstrap.pdf"),
    height = cb_dim$height, width = cb_dim$width)
draw_pages(map(cellboot, plot_support_tree))
dev.off()
```

## Combined support table

```{r combined-support}
combined = ref_clades %>%
  select(region, clade, n_tips, au_gene_bootstrap = au, bp_gene_bootstrap = bp) %>%
  left_join(cellboot_df %>%
              filter(threshold_label == ref_lab) %>%
              select(region, clade, cell_bootstrap),
            by = c("region", "clade")) %>%
  arrange(region, desc(n_tips))
write_csv(combined, file.path(out_sub_dir, "branch_support_published_tree.csv"))
combined
```

# Where the song clusters sit

A compact summary of the question the removal step was meant to address: which
cell types the song-nucleus clusters group with, at each threshold.

```{r song-placement}
## Smallest clade containing a song cluster: its immediate sister group.
sister_of = function(hc, tip) {
  clades = hclust_clades(hc)
  containing = keep(clades, ~ tip %in% .x)
  if (!length(containing)) return(list(sister = character(0), clade = NA_character_))
  smallest = containing[[which.min(map_int(containing, length))]]
  list(sister = setdiff(smallest, tip), clade = clade_key(smallest))
}

song_tips = map(regions, function(reg)
  grep(song_clusters[[reg]], colnames(obj_int_avg1), value = TRUE) %>%
    intersect(colnames(obj_int_avg1)[grepl(region_cols[[reg]], colnames(obj_int_avg1))])) %>%
  set_names(regions)

## Any cluster from a song nucleus, for telling song/non-song pairs from
## song/song pairs
all_song_tips = unique(unlist(song_tips))

placement = map_dfr(sweep, function(s) {
  hc = s$pv$hclust
  au_by_clade = set_names(s$pv$edges$au, map_chr(hclust_clades(hc), clade_key))
  tips = intersect(song_tips[[s$region]], hc$labels)
  map_dfr(tips, function(tp) {
    r = sister_of(hc, tp)
    tibble(region = s$region, threshold_label = s$threshold_label,
           song_cluster = tp,
           sister = paste(r$sister, collapse = ", "),
           sister_list = list(r$sister),
           clade = r$clade,
           au = unname(au_by_clade[r$clade]),
           ## TRUE when the song cluster pairs with non-song cell types only
           song_nonsong = length(r$sister) > 0 && !any(r$sister %in% all_song_tips))
  })
}) %>%
  mutate(threshold_label = factor(threshold_label, levels = lev)) %>%
  arrange(region, song_cluster, threshold_label)

write_csv(select(placement, -sister_list),
          file.path(out_sub_dir, "song_cluster_placement.csv"))
select(placement, -sister_list)
```

## Bootstrap support for the song / non-song pairings

The trees as a whole can rearrange without touching the relationships the
manuscript rests on, which are local: each song-nucleus cluster and the non-song
cell types it pairs with. These panels show, for every song cluster and every
song-DEG threshold, which cell types it pairs with (tile text) and how well
supported that pairing is (tile fill). Support is the `pvclust` approximately
unbiased (AU) value for the smallest clade containing the song cluster, i.e. the
branch that joins it to the cell types named in the tile, from multiscale
bootstrap resampling of genes (`r nboot` replicates at each of 10 scaling
levels). The cell-level bootstrap support is added to the accompanying CSV for
the two thresholds where it was run.

```{r song-pair-support}
#| fig-width: 5
#| fig-height: 3.5
#| results: hold
## Jaccard overlap of a pairing with the same song cluster's pairing at the
## reference threshold *within the same analysis*; carried in the CSV as a
## convenience, not shown in the plot.
jaccard = function(a, b) {
  if (!length(union(a, b))) return(1)
  length(intersect(a, b)) / length(union(a, b))
}
ref_pairs = placement %>%
  filter(threshold_label == ref_lab) %>%
  select(region, song_cluster, ref_sister = sister_list)

pair_support = placement %>%
  left_join(ref_pairs, by = c("region", "song_cluster")) %>%
  ## cell-level bootstrap support, where that bootstrap was run for this clade
  left_join(cellboot_df %>%
              mutate(threshold_label = factor(threshold_label, levels = lev)) %>%
              select(region, threshold_label, clade, cell_bootstrap),
            by = c("region", "threshold_label", "clade")) %>%
  mutate(au_pct = 100 * au,
         jaccard_vs_ref_threshold = map2_dbl(sister_list, ref_sister, jaccard),
         region = factor(region, levels = regions),
         song_cluster = factor(song_cluster, levels = rev(sort(unique(song_cluster)))))
write_csv(select(pair_support, -sister_list, -ref_sister),
          file.path(out_sub_dir, "song_pair_support.csv"))

## sister names, one per line and without the shared "Glut-" prefix, so that
## they fit inside a tile
pair_support = pair_support %>%
  mutate(sister_lab = str_replace_all(str_replace_all(sister, "Glut-", ""), ", ", "\n"))

## One plot per analysis. Tiles are wider than they are tall: the column count
## is fixed at the threshold series, while the row pitch is kept tight so the
## row labels sit close together. No coord_fixed, so the tile aspect follows the
## saved device size directly.
tile_w_in = 0.55  # inches per column
tile_h_in = 0.22  # inches per row (the row pitch)
cell_pt   = 5     # in-tile text (the smallest type used anywhere here)
axis_pt   = 6     # axis text and titles

## Standard position colours for the song nuclei (HVC / LMAN / RA), the same
## ones used for the tree tips.
song_position = function(x) case_when(grepl("LMAN", x) ~ "lman",
                                      grepl("HVC", x)  ~ "hvc",
                                      grepl("RA", x)   ~ "ra",
                                      TRUE             ~ NA_character_)

pair_plot = function(d, ttl) {
  ## rows grouped by song nucleus; the discrete y scale runs bottom to top, so
  ## the level order is the reverse of the reading order
  lv = unique(as.character(d$song_cluster))
  lv = rev(lv[order(match(song_position(lv), c("hvc", "lman", "ra")), lv)])
  d = mutate(d, song_cluster = factor(as.character(song_cluster), levels = lv))
  axis_cols = unname(position_colors[song_position(lv)])

  ggplot(d, aes(threshold_label, song_cluster)) +
    geom_tile(aes(fill = au_pct), color = "grey70", linewidth = 0.2) +
    geom_text(aes(label = sister_lab, color = au_pct > 60),
              size = cell_pt / .pt, lineheight = 0.9) +
    ## a song cluster pairing with another song cluster rather than with
    ## non-song cell types is marked in the corner of its tile. The nudge is in
    ## the discrete scale's own space.
    geom_point(data = filter(d, !song_nonsong),
               position = position_nudge(x = -0.42, y = 0.36),
               shape = 4, size = 1, stroke = 0.5, color = "firebrick") +
    scale_fill_gradient(low = "white", high = "#2166AC", limits = c(0, 100),
                        name = "AU support\nfor the pairing (%)") +
    scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "grey20"), guide = "none") +
    labs(x = "song-DEG adjusted-p threshold", y = NULL,
         title = ttl,
         caption = "tile text = cell types the song cluster pairs with\nx = paired with another song cluster rather than with non-song cell types") +
    theme_cowplot(font_size = axis_pt) +
    theme(axis.text = element_text(size = axis_pt),
          axis.text.x = element_text(size = axis_pt, angle = 45, hjust = 1),
          axis.text.y = element_text(size = axis_pt, color = axis_cols),
          axis.title = element_text(size = axis_pt),
          legend.title = element_text(size = axis_pt),
          legend.text = element_text(size = cell_pt),
          legend.key.size = unit(0.16, "in"),
          plot.title = element_text(size = 7),
          plot.caption = element_text(size = cell_pt, hjust = 0))
}

## The ARCO and NIDO song clusters go into ONE heatmap -- their pairings still
## come from the two separate clusterings, this only puts them on a single set
## of axes, which sidesteps having to match tile geometry across files. The
## pooled ALL analysis stays in its own panel, since it is a different tree.
##
## set_panel_size() pins the panel to ncol x nrow tiles of exactly
## tile_w_in x tile_h_in; sizing the page instead (panel = page minus
## title/axis/caption) lets whatever those elements do not use be absorbed by
## the panel and split across its rows, which made a one-row panel's tiles
## taller than a four-row panel's.
gt_width_in  = function(g) grid::convertWidth(sum(g$widths), "in", valueOnly = TRUE)
gt_height_in = function(g) grid::convertHeight(sum(g$heights), "in", valueOnly = TRUE)

save_pair_plot = function(d, ttl, fname) {
  n_rows = n_distinct(d$song_cluster)
  gt = egg::set_panel_size(pair_plot(d, ttl),
                           width  = unit(length(lev) * tile_w_in, "in"),
                           height = unit(n_rows * tile_h_in, "in"))
  plot(gt)
  save_plot(file.path(out_sub_dir, fname), gt,
            base_width  = gt_width_in(gt),
            base_height = gt_height_in(gt))
}

## single heatmap: every song cluster, each from its own (ARCO or NIDO) tree
save_pair_plot(filter(pair_support, region %in% c("ARCO", "NIDO")),
               "song / non-song pairings, separate ARCO and NIDO clusterings",
               "song_pair_support_split.pdf")

## the pooled analysis, for comparison
save_pair_plot(filter(pair_support, region == "ALL"),
               region_name[["ALL"]],
               "song_pair_support_all.pdf")
```

```{r song-pair-support-summary}
#| fig-width: 8
#| fig-height: 3.5
## ALL dropped from this summary plot specifically (per direct instruction, same as sweep_summary
## above) -- it stays in pair_support itself and in the per-region song_pair_support_<region>.pdf
## tile plots above.
p_pair_mean = pair_support %>%
  filter(region != "ALL") %>%
  group_by(region, threshold_label) %>%
  summarise(mean_au = mean(au, na.rm = TRUE),
            frac_song_nonsong = mean(song_nonsong), .groups = "drop") %>%
  pivot_longer(c(mean_au, frac_song_nonsong)) %>%
  mutate(name = recode(name,
                       mean_au           = "mean AU support for the pairing",
                       frac_song_nonsong = "fraction pairing with non-song types")) %>%
  ggplot(aes(threshold_label, value, color = region, group = region)) +
  geom_line() + geom_point() +
  facet_wrap(~name) +
  ylim(0, 1) +
  labs(x = "song-DEG adjusted-p threshold", y = NULL) +
  theme_cowplot(font_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_pair_mean)
save_plot(file.path(out_sub_dir, "song_pair_support_summary.pdf"), p_pair_mean,
          base_height = 3.2, base_width = 9)
```

# Session info

```{r session}
sessionInfo()
```
