---
title: "Hybrid division-based naming for glutamatergic clusters"
format: html
editor: source
---

`exploratory/apply_composite_labels.qmd` names every glutamatergic cluster after its specific chicken-integration
match (`Ex_DACH2_MGAT4C` -> `Glut-DACH2_MGAT4C`, etc.). That implies a specific cross-species
correspondence for every cluster, which is more than the data actually supports for most of them
(moderate confidence, one winner out of four methods that often disagree) -- and is flatly wrong for
the song nuclei, which are evolutionarily novel structures with no real chicken counterpart to begin
with. This notebook is a deliberately weaker claim: keep only the three excitatory major divisions
that hold up across birds (`DACH2`, `SATB2`, `CACNA1H`), hand-label the song-region clusters by their
own anatomy/projection evidence rather than any chicken match, and leave every other cluster a bare
numbered member of its division rather than implying a specific chicken identity it doesn't have good
support for.

**This is the production cluster-filtering and naming notebook for the glutamatergic clusters.**
It is the single place where the cluster exclusions (`excluded_clusters`) and the published cluster
labels (`celltype_hybrid`) are defined; everything downstream -- `snrna/trees/celltypes_hclust_*_hybrid.qmd`,
`snrna/reduction_viz/combined_all_umap_hybrid.R`, `snrna/deg/dotplot_manual_markers_hybrid.qmd`,
`snrna/integration/colquitt2021_label_transfer.qmd` -- reads its `obj_hybrid_labels.qs2`. The other
notebooks under `snrna/naming/exploratory/` are the exploratory work that led here and are **not**
production: nothing reads their outputs, and they are kept only as a record of the reasoning.

Independent of the preprocessing notebook, by the same convention: reads the main clustered object
(`snrna/clustering/snrna_seurat_cellbender_preprocess/obj_clustered.qs2`) read-only and writes its
own copy, never writing back to the canonical object.

## The rules, as given

- **Major division is kept, specific chicken gene match is dropped.** `DACH2`, `SATB2`, `CACNA1H`
  are the three excitatory subclass markers reported to hold across birds; everything past that
  (which specific chicken population a cluster's `top_rank_agg` happened to win) is dropped.
- **Song-region clusters (HVC, RA, LMAN) are hand-labeled from known anatomy/projections, not from
  the chicken match**, since these are the clusters where the chicken correspondence is weakest to
  begin with (see the QC section below -- confirmed directly, not assumed):
  - `Glut-HVC-1` -> `Glut-DACH2-HVCra` (RA-projecting)
  - `Glut-HVC-1a` -> `Glut-DACH2-HVCra-Pre` (a distinct sub-label for this HVC-1 subtype, not merged
    into HVCra, per direct instruction)
  - `Glut-HVC-2` -> `Glut-DACH2-HVCx` (Area X-projecting)
  - `Glut-RA` -> `Glut-CACNA1H-RA`
  - `Glut-LMAN-1` -> `Glut-DACH2-LMANsh` (shelf)
  - `Glut-LMAN-2` -> `Glut-DACH2-LMANco` (core)
- **Every other glutamatergic cluster gets `Glut-<division>-<n>`**, `n` a plain sequential index
  within its division (order: original cluster label, ascending -- a bookkeeping choice, not a
  biological claim; documented so it can be re-ordered if that turns out to matter).
- **`Glut-Nido-1` keeps folding into `Glut-NC-1`'s own result** (the same population, confirmed in
  `exploratory/apply_composite_labels.qmd`), so it takes whatever number `Glut-NC-1` lands on rather than being
  numbered independently.
- **`GABA-Pre` -> `GABA-Im`**, matching the immature-interneuron developmental-stage relabel already
  applied to its GABAergic-side counterpart in `exploratory/apply_composite_labels.qmd`, and to `Glut-Im` here.
- **Every other GABAergic cluster is named from its own cross-species match**, anchored on
  `snrna/integration/composite_scoring/results/gg_adult_hybrid/composite_calls_summary.csv` -- the
  chicken (Zaremba) integration re-run directly against this notebook's own hybrid-labeled object
  (`obj_hybrid_labels.h5ad`), confirmed to key on the same `cluster` values used here, unlike the
  older `composite_gg_adult` table `exploratory/apply_composite_labels.qmd` used (which predates the
  hybrid relabel and does not line up with `Glut-*` clusters post-rename -- fine for GABA, whose names
  this scheme leaves alone either way, but not safely reusable for anything Glut-side). `top_rank_agg`
  (`Inh_<LGE|MGE|CGE>_like_<gene(s)>`) gives both a developmental-origin division (ganglionic eminence
  of origin -- LGE/MGE/CGE, the GABAergic analogue of `DACH2`/`SATB2`/`CACNA1H`) and, for most
  clusters, a specific marker gene, unlike the Glut side where the specific chicken match was judged
  too weak to keep. Kept here because the confidence tiers are markedly better (mostly "high",
  `>=0.6`) and the calls line up with known interneuron biology (SST/PVALB = MGE, VIP/LAMP5-adjacent
  = CGE) -- an independent check the Glut song-region hand-labels also had, from anatomy rather than a
  second species.
  - **`GABA-4-2`'s own `top_rank_agg` (`Inh_MGE_like_ST18_PVALB`) is overridden to its own
    rank-aggregation runner-up (`Inh_MGE_like_SST_PVALB`)**, the same way the Glut song-region
    hand-labels override a chicken match the data itself contradicts: the rank-agg winner only barely
    beat the runner-up (margin 0.08, overall confidence just "moderate", the four underlying methods
    split 2-2 with `saturn`'s own vote nonsensical), and **confirmed directly against this object's
    own expression** -- not assumed -- `ST18` is detected in 0.6% of `GABA-4-2`'s own cells versus
    42.9% in `GABA-7` (the cluster that actually earns the `ST18` name), while `SST`/`PVALB` detection
    (60%/18%) sits much closer to `GABA-3`'s and `GABA-4-1`'s own SST_PVALB profile (89%/48%,
    88%/65%) than to any ST18 population. This folds `GABA-4-2` into the SST_PVALB group below.
  - **A chicken match being real is not the same claim as its gene being a real discriminating
    marker in this species, and this is not hypothetical**: `SLC24A2` and `ABI3BP` (both embedded in
    composite calls above) are not the top-expressing GABA cluster for their own gene in this
    object -- `SLC24A2` is highest in the three SST_PVALB clusters, not the SST_SLC24A2 clusters it
    was meant to name, and `ABI3BP` shows no real signal anywhere (flat ~0.05-0.13 across all 13
    clusters), most likely a genuine chicken/finch species difference. So every candidate gene is
    **verified directly against this object's own expression** before being used: its own cluster
    must be the single highest-expressing GABA cluster for that gene, by at least a 1.5-fold margin
    over the runner-up (see the verification section below for the full pass/fail table). A gene
    absent from this genome's own annotation altogether (`CACNA1B`, confirmed against this object's
    `rownames` and the other seven `CACNA1<A-I,S>` paralogs, all present) fails this check
    automatically -- unconfirmed is treated the same as disconfirmed, not as a pass by default.
  - **A cluster whose candidate gene passes verification, and is not shared with another cluster's
    own chicken match, keeps it as its name**: `GABA-CGE-MAN1A1` failed (MAN1A1 is actually highest
    in three unrelated MGE clusters) and fell back below, but `GABA-MGE-ANO1`, `GABA-MGE-ST18`, and
    `GABA-MGE-LHX8` all passed cleanly (the `LHX8` example this scheme was asked for by name, and
    independently corroborated by the mouse/Yao integration in the same `composite_scoring` output:
    `GABA-8`'s own top mouse match is an `Lhx8`-labeled striatal/pallidal interneuron type).
  - **`GABA-6` (`GABA-MGE-ANO1` by the rule above) is overridden by hand to `GABA-MGE-LAMP5`**:
    LAMP5 marks a recognized interneuron class in its own right, and the mouse/Yao integration in the
    same `composite_scoring` output independently calls this cluster an `Lamp5`-labeled type (high
    confidence 0.52) -- a real second line of cross-species evidence `ANO1` alone doesn't have.
    Flagged as worth checking by that mouse call and **confirmed directly against this object's own
    expression** -- not assumed -- `LAMP5` is detected in 64.1% of this cluster's own cells versus
    11.9% in the next-highest GABA cluster (a 6.3-fold margin), at least as clean a discriminator as
    `ANO1` itself.
  - **Every remaining collision group (`GABA-1-1`/`GABA-1-2` on `LGE_FOXP2`,
    `GABA-2-1`/`GABA-2-2`/`GABA-2-3` on `SST_SLC24A2`/`SST_ABI3BP`, `GABA-3`/`GABA-4-1`/`GABA-4-2` on
    `SST_PVALB` after the override above, and `GABA-5-1`/`GABA-5-2` on `CGE_MAN1A1`/`CGE_CACNA1B`)
    gets a real distinguishing gene from this object's own within-group differential expression**
    (the same `cluster_de()`/`select_naming_genes()` machinery `exploratory/apply_composite_labels.qmd`
    uses), **not the chicken match's own suggested second gene**, which already failed verification
    for exactly this reason for three of the four groups (`SLC24A2`, `ABI3BP`, `MAN1A1`/`CACNA1B`):
    `GABA-1-1` -> `MAML2`, `GABA-1-2` -> `MDGA1`, `GABA-2-1` -> `SLC12A5`, `GABA-2-2` -> `SHISA9`,
    `GABA-2-3` -> `DIAPH3`, `GABA-3` -> `UNC5C`, `GABA-4-1` -> `CHRM2`, `GABA-4-2` -> `IL1RAP`,
    `GABA-5-1` -> `PTPRK`, `GABA-5-2` -> `STK32C` (all real DESeq2 pseudobulk hits, padj <= 1e-4, and
    confirmed directly by detection rate against their own group siblings, not merely a
    significance-test artifact).
  - **The two MGE groups additionally keep a real class marker for the whole group as a prefix on
    top of that within-group pick** (`SST`, `PVALB` -- both real per the QC section above, unlike the
    chicken match's own more specific second gene), giving `GABA-MGE-SST-SLC12A5` etc. **LGE and CGE
    use a single gene with no class-marker prefix instead**, per direct instruction (a class marker
    was tried for both -- `MEIS2` for LGE, `DPF3` for CGE, each found the same "ask the data, not the
    chicken match" way, a group-vs-rest differential expression -- but was dropped again for looking
    inconsistent alongside a single-gene LGE/CGE name): `GABA-LGE-MAML2`, `GABA-LGE-MDGA1`,
    `GABA-CGE-PTPRK` (before the `GABA-5-1` override just below), `GABA-CGE-STK32C`.
  - **`GABA-5-1`'s own within-group DE pick (`PTPRK`) is overridden by hand to `PENK`**: requested by
    name and **confirmed directly against this object's own expression** -- not assumed -- `PENK` is
    detected in 99.1% of `GABA-5-1`'s own cells (mean SCT 2.79) versus 53.6% in `GABA-5-2` (its only
    group sibling) and 50.3% in the next-highest GABA cluster overall (`GABA-7`), a dramatically
    cleaner, more specific signal than `PTPRK`'s own within-group numbers (93%/22%). `GABA-5-2` keeps
    its own algorithmic pick (`STK32C`), unaffected by this and independently well verified (63%
    vs. 22% within the group).
- **Untouched**: `Glut-Nido-2` (`Glut-GABA`, the dual-transmitter override -- not a single division to
  begin with), the three Glut precursor clusters (`Glut-NSC`/`Glut-NB`/`Glut-Im`, developmental-stage
  labels independent of this scheme), and all non-neuronal clusters.

```{r}
#| label: packages
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(knitr)

source(here::here("scripts/cluster_curation.R"))
source(here::here("scripts/cluster_naming.R"))

## Excludes annotation placeholders (`^LOC[0-9]` NCBI-style loci, `^nbis-`/`-gr[0-9]+$` NBIS-pipeline
## loci -- every chrM gene in this genome's annotation gets one of these) and ribosomal protein genes
## (`^RP[SL][0-9]`/`^MRP[SL][0-9]`, sparing the `RPS6K*` kinase family) from candidacy as a GABA
## within-group naming gene below -- same definition `exploratory/apply_composite_labels.qmd` uses,
## so a housekeeping or contamination-associated gene cannot become a cluster's name here either.
is_named_gene <- function(gene) {
  !grepl("^LOC[0-9]", gene) & !grepl("^nbis-", gene, ignore.case = TRUE) & !grepl("-gr[0-9]+$", gene) &
    !grepl("^RP[SL][0-9]+[A-Z]?[0-9]?$", gene) & !grepl("^MRP[SL][0-9]+[A-Z]?[0-9]?$", gene)
}
```

## Parameters

```{r}
#| label: params
in_fname <- here::here("snrna/clustering/snrna_seurat_cellbender_preprocess",
                       "obj_clustered.qs2")
composite_fname <- here::here("snrna/integration/composite_gg_adult", "composite_calls_summary.csv")
## Chicken integration re-run directly against this notebook's own hybrid-labeled output -- its
## `query` values already match this object's GABA cluster names 1:1 (confirmed below), unlike
## `composite_fname` above, which predates the hybrid relabel and is kept only for the Glut division
## read (where the pre-relabel `Glut-*` names are exactly what is wanted for that lookup).
gaba_composite_fname <- here::here("snrna/integration/composite_scoring/results/gg_adult_hybrid",
                                   "composite_calls_summary.csv")
cluster_col <- "cluster"

out_dir <- here::here("snrna/naming", "hybrid_division_naming")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
lookup_fname <- file.path(out_dir, "hybrid_label_lookup.csv")
obj_out_fname <- file.path(out_dir, "obj_hybrid_labels.qs2")

## Confirmed artifacts, dropped upstream/elsewhere of any naming exercise. Glut-Arco-1 was excluded
## here briefly by mistake (meant to be Glut-Arco-4, already covered) and has been brought back into
## the generic CACNA1H pool below.
excluded_clusters <- c("Glut-Nido-3", "Glut-Arco-4")

## Hand-labeled from anatomy/projection evidence, not from the chicken match -- see the QC section
## for the confirmed mismatch between these and each cluster's own top_rank_agg.
song_region_overrides <- c(
  "Glut-HVC-1"  = "Glut-DACH2-HVCra",
  "Glut-HVC-1a" = "Glut-DACH2-HVCra-Pre",
  "Glut-HVC-2"  = "Glut-DACH2-HVCx",
  "Glut-RA"     = "Glut-CACNA1H-RA",
  "Glut-LMAN-1" = "Glut-DACH2-LMANsh",
  "Glut-LMAN-2" = "Glut-DACH2-LMANco"
)

## Developmental-stage labels: the Glut side already established in exploratory/apply_composite_labels.qmd,
## carried over unchanged; GABA-Pre -> GABA-Im added here to match, on the GABAergic side.
precursor_overrides <- c("Glut-Pre-1" = "Glut-NSC", "Glut-Pre-2" = "Glut-NB", "Glut-Pre-3" = "Glut-Im",
                         "GABA-Pre" = "GABA-Im")

## The dual-transmitter override, also carried over unchanged.
nido2_override <- c("Glut-Nido-2" = "Glut-GABA")

## Nido-1 is the same population as NC-1 (confirmed in exploratory/apply_composite_labels.qmd via their shared,
## independently-arrived-at top_rank_agg winner) -- it takes NC-1's own generic number rather than
## being numbered on its own.
fold_targets <- c("Glut-Nido-1" = "Glut-NC-1")
```

```{r}
#| label: load
obj <- qs_read(in_fname, nthreads = 8)
composite <- read_csv(composite_fname, show_col_types = FALSE)
stopifnot(setequal(composite$query, levels(factor(obj@meta.data[[cluster_col]]))))

gaba_composite <- read_csv(gaba_composite_fname, show_col_types = FALSE)
## Confirmed 1:1 against this object's own GABA cluster names, restricted to the clusters this
## table is actually used for below (GABA-Pre is excluded from this composite match -- see the
## precursor override -- so its absence/rename to GABA-Im on the composite side is expected, not
## checked here).
gaba_clusters_raw <- levels(factor(obj@meta.data[[cluster_col]]))[
  grepl("^GABA", levels(factor(obj@meta.data[[cluster_col]])))
]
stopifnot(
  "gg_adult_hybrid composite$query does not cover this object's own (non-precursor) GABA clusters" =
    all(setdiff(gaba_clusters_raw, "GABA-Pre") %in% gaba_composite$query)
)

## Library-qualified souporcell assignment for the GABA within-group naming DE below -- this backup
## object predates the souporcell reprocessing (no `individual`/`batch` columns), so raw `assignment`
## alone is a *per-library* index and would silently pool distinct birds across libraries if used
## directly. Qualifying it by library first is what makes it valid: within one library, distinct
## `assignment` values are genuinely distinct known birds (that library's own souporcell run). Same
## construction as exploratory/apply_composite_labels.qmd.
obj$replicate_lib <- paste(obj$sample_name, obj$assignment, sep = "_")
```

## Major division per cluster

```{r}
#| label: division
divisions <- c("DACH2", "SATB2", "CACNA1H")

## Division read directly off each cluster's own top_rank_agg winner (Ex_<division>_<gene> ->
## <division>) -- the same field apply_composite_labels.qmd used for the full composite name, just
## truncated to its first token. Not re-derived some other way, so this can't silently drift from
## what that notebook's own eligible-cluster computation saw.
glut_division <- composite %>%
  filter(grepl("^Glut", query)) %>%
  transmute(cluster = query, division = sub("_.*", "", sub("^Ex_", "", top_rank_agg)))

kable(glut_division %>% filter(!division %in% divisions),
      caption = paste("Glut clusters whose own top_rank_agg winner is NOT one of DACH2/SATB2/CACNA1H",
                      "-- expected here only for clusters excluded, overridden, or precursor-staged",
                      "below (checked explicitly in the next chunk), not silently dropped."))
```

```{r}
#| label: division-check
non_division <- glut_division %>% filter(!division %in% divisions) %>% pull(cluster)
unexplained <- setdiff(non_division, c(excluded_clusters, names(song_region_overrides),
                                       names(precursor_overrides), names(nido2_override),
                                       names(fold_targets)))
if (length(unexplained) > 0) {
  stop("Glut cluster(s) with a non-DACH2/SATB2/CACNA1H top_rank_agg and no override to explain it: ",
       paste(unexplained, collapse = ", "))
}
cat("All", length(non_division), "off-division clusters are accounted for by an exclusion, hand-label,",
   "precursor stage, or fold target above -- none silently miscategorized.\n")
```

**Confirmed, not assumed**: the chicken match itself disagrees with the hand-labels below for two of
the six song-region clusters -- `Glut-HVC-1`'s own `top_rank_agg` is `Ex_SATB2_SOX6` (not DACH2), and
`Glut-RA`'s is `Ex_KIAA1217` (not CACNA1H at all, moderate confidence). That is exactly the kind of
disagreement this notebook exists to route around for the song system specifically: these are
evolutionarily novel structures, so a weak or wrong single-species match is expected, not a reason to
second-guess the anatomy/projection evidence used to hand-label them instead.

## GABA division and informative naming

Unlike the Glut side, the GABA composite calls are mostly high confidence and line up with known
interneuron biology, so the specific chicken match is kept rather than truncated to a bare division
number -- see the rules above for why.

```{r}
#| label: gaba-division
gaba_divisions <- c("LGE", "MGE", "CGE")

## Shared logic for extracting a division/body from a composite call: strip the chicken-side prefix,
## drop "_like_" for readability (Inh_LGE_like_FOXP2 -> LGE-FOXP2 -- a formatting choice, not a change
## in which chicken population a cluster was matched to). No naming decision yet -- that is deferred
## to the verification step below, since a chicken match being real is not the same question as
## whether it is a real *discriminating marker in this species*.
assign_gaba_names <- function(composite, eligible_clusters) {
  composite %>%
    dplyr::filter(query %in% eligible_clusters) %>%
    dplyr::mutate(division_gene = sub("_like_", "-", sub("^Inh_", "", top_rank_agg))) %>%
    dplyr::transmute(cluster = query,
                     division = sub("-.*", "", division_gene),
                     body = division_gene,
                     confidence, confidence_tier)
}

## GABA-4-2's own top_rank_agg (Inh_MGE_like_ST18_PVALB) only barely beat the SST_PVALB runner-up in
## the rank-aggregation step (margin 0.08 of 1, overall confidence just "moderate" at 0.32, and
## agg_methods_concordant = FALSE): individually, gsi and samap each independently call SST_PVALB
## (gsi at 0.73, its highest per-method confidence of the four), only cca calls ST18_PVALB, and
## saturn's own top call is a nonsensical excitatory type (Ex_CACNA1H_CPA6) -- not a real second
## vote either way. **Confirmed directly against this object's own expression**, not assumed: ST18 is
## detected in just 0.6% of GABA-4-2's own cells (vs. 42.9% in GABA-7, the cluster that actually earns
## the ST18 name), while SST/PVALB are detected in 60%/18% of its cells -- much closer to GABA-3's and
## GABA-4-1's own SST_PVALB profile (89%/48% and 88%/65%) than to any ST18 population. Overridden to
## the runner-up call here, the same way the Glut song-region hand-labels override a chicken match
## the data itself contradicts.
gaba_body_overrides <- c("GABA-4-2" = "Inh_MGE_like_SST_PVALB")

gaba_eligible <- setdiff(gaba_clusters_raw, "GABA-Pre")
gaba_composite_overridden <- gaba_composite %>%
  mutate(top_rank_agg = if_else(query %in% names(gaba_body_overrides),
                                gaba_body_overrides[query], top_rank_agg))
gaba_named <- assign_gaba_names(gaba_composite_overridden, gaba_eligible)

kable(gaba_named %>% filter(!division %in% gaba_divisions),
      caption = paste("GABA clusters whose own top_rank_agg winner is NOT LGE/MGE/CGE -- expected",
                      "to be empty; a non-empty result here means a chicken cluster this table",
                      "matched to is not one of the three ganglionic-eminence divisions."))
stopifnot(
  "A GABA cluster's top_rank_agg winner is not LGE/MGE/CGE-derived -- check gaba_composite directly" =
    all(gaba_named$division %in% gaba_divisions)
)
```

### Verify each candidate marker gene against this object's own expression

A chicken cross-species match being a real, confident correspondence does not make its named gene a
real *discriminating marker in this species* -- SLC24A2 and ABI3BP (both embedded in composite calls
above) turned out not to be, most likely a genuine species difference between what discriminates a
population in chicken and what discriminates the same broad class in finch. Rather than trust the
composite call's gene at face value, each candidate's own single-cell expression is checked directly:
its own cluster must be the single highest-expressing GABA cluster for that gene (a strict top-1, not
merely "elevated"), by a margin over the runner-up, or the specific gene is dropped in favor of the
raw cluster's own numeric suffix -- the same fallback already used for a collision. A gene absent from this
object's own annotation altogether (`CACNA1B`, see the rules above) fails this check automatically:
unconfirmed is treated the same as disconfirmed here, not as an exception.

```{r}
#| label: gaba-verify-markers
## The gene actually being asked to carry the name: the last "_"-delimited token of the body once
## the division prefix is stripped (MGE-SST_PVALB -> PVALB, CGE-MAN1A1 -> MAN1A1, LGE-FOXP2 ->
## FOXP2). For a compound body this is the part beyond the shared class marker (SST, itself real and
## broadly shared across several MGE clusters, is not what is being verified here).
gaba_named <- gaba_named %>%
  mutate(distinguishing_gene = sub(".*_", "", sub("^[A-Z]+-", "", body)))

DefaultAssay(obj) <- "SCT"
gaba_expr_genes <- intersect(unique(gaba_named$distinguishing_gene), rownames(obj))
gaba_cells <- colnames(obj)[obj@meta.data[[cluster_col]] %in% gaba_eligible]
gaba_expr_mat <- GetAssayData(obj, assay = "SCT", layer = "data")[gaba_expr_genes, gaba_cells, drop = FALSE]
gaba_expr_cluster <- droplevels(factor(obj@meta.data[gaba_cells, cluster_col]))
gaba_avg <- sapply(levels(gaba_expr_cluster),
                   function(cl) Matrix::rowMeans(gaba_expr_mat[, gaba_expr_cluster == cl, drop = FALSE]))

## own cluster must be the single top-ranked GABA cluster for its distinguishing gene, and clear the
## runner-up by at least margin_min-fold -- a real discriminating marker, not merely "detectable".
verify_specific <- function(gene, own_cluster, avg, margin_min = 1.5) {
  if (!gene %in% rownames(avg)) return(FALSE)
  ranked <- sort(avg[gene, ], decreasing = TRUE)
  ranked[1] > 0 && names(ranked)[1] == own_cluster && (length(ranked) < 2 || ranked[1] >= margin_min * ranked[2])
}

gaba_named <- gaba_named %>%
  rowwise() %>%
  mutate(verified = verify_specific(distinguishing_gene, cluster, gaba_avg)) %>%
  ungroup()

dup_bodies <- gaba_named %>% dplyr::count(body) %>% dplyr::filter(n > 1) %>% dplyr::pull(body)
gaba_named <- gaba_named %>% mutate(use_specific = verified & !(body %in% dup_bodies))

kable(gaba_named %>% select(cluster, body, distinguishing_gene, verified, use_specific),
      caption = paste("Every GABA cluster's candidate distinguishing gene and whether it passed",
                      "direct verification against this object's own expression. use_specific = FALSE",
                      "for any reason (failed verification, gene absent from this object, or shared",
                      "with another cluster's own chicken match) falls back to the raw cluster's own",
                      "numeric suffix, division-prefixed."))
```

```{r}
#| label: gaba-final-labels
## Two pools now. (1) A cluster whose composite-derived candidate gene is confirmed and exclusive to
## it keeps that name as-is. (2) Every other group (both MGE collisions, and now LGE/CGE too) gets a
## real distinguishing gene from this object's own within-group differential expression, prefixed by
## a real class marker for the whole group -- see the two chunks below. A bare division-numeral
## (tried first here for LGE/CGE, since neither's composite class marker held up under verification)
## was rejected as looking inconsistent next to the MGE groups' gene-based names, once real
## discriminating genes turned out to exist for LGE/CGE too (see class-marker discovery below).
gaba_specific <- gaba_named %>% filter(use_specific) %>% mutate(new_label = paste0("GABA-", body)) %>%
  select(cluster, new_label)

## Real distinguishing gene per cluster within a collision group, from this object's own differential
## expression -- the same cluster_de()/select_naming_genes() machinery
## exploratory/apply_composite_labels.qmd uses -- not the chicken match's own suggested second gene,
## which already failed verification above for exactly this reason (SLC24A2, ABI3BP, MAN1A1/CACNA1B:
## real, confident chicken matches that are simply not real discriminating markers in this species).
## class_label is an optional extra prefix on top of the group's own within-group pick, kept only for
## the two MGE groups below (SST, PVALB): both are real, broadly-shared markers across their whole
## group (see the QC section above), unlike LGE/CGE, where no group-wide marker held up well enough
## to be worth the extra length -- LGE and CGE get a single gene each instead.
within_group_gene_family <- function(obj, cluster_col, clusters, division, class_label = NULL,
                                     filters = naming_filters(max_genes = 1)) {
  cells <- colnames(obj)[obj@meta.data[[cluster_col]] %in% clusters]
  og <- subset(obj, cells = cells)
  og@meta.data[[cluster_col]] <- droplevels(factor(og@meta.data[[cluster_col]]))
  DefaultAssay(og) <- "SCT"
  de <- cluster_de(og, cluster_col, replicate_col = "replicate_lib")
  excl <- unique(de$feature)[!is_named_gene(unique(de$feature))]
  nm <- select_naming_genes(de, filters = filters, exclude_genes = excl)
  stopifnot(
    "No naming gene cleared naming_filters() for at least one cluster in this group" =
      !anyNA(nm$naming_genes)
  )
  prefix <- if (is.null(class_label)) division else paste0(division, "-", class_label)
  tibble(cluster = nm$cluster, new_label = paste0("GABA-", prefix, "-", nm$naming_genes))
}

mge_sst_family <- within_group_gene_family(obj, cluster_col, c("GABA-2-1", "GABA-2-2", "GABA-2-3"), "MGE", "SST")
mge_pvalb_family <- within_group_gene_family(obj, cluster_col, c("GABA-3", "GABA-4-1", "GABA-4-2"), "MGE", "PVALB")
lge_family <- within_group_gene_family(obj, cluster_col, c("GABA-1-1", "GABA-1-2"), "LGE")
cge_family <- within_group_gene_family(obj, cluster_col, c("GABA-5-1", "GABA-5-2"), "CGE")

## GABA-5-1's own within-group DE pick (PTPRK) is overridden to PENK here: checked directly, PENK is
## detected in 99.1% of GABA-5-1's own cells (mean SCT 2.79) versus 53.6% in GABA-5-2 (its only group
## sibling) and 50.3% in the next-highest GABA cluster overall (GABA-7) -- a dramatically cleaner,
## more specific signal than PTPRK's own within-group numbers (93%/22%), and requested by name.
## GABA-5-2 keeps its own algorithmic pick (STK32C), unaffected by this -- still well verified
## (63% vs. 22% within the group) independent of what GABA-5-1 ends up called.
cge_family <- cge_family %>%
  mutate(new_label = if_else(cluster == "GABA-5-1", "GABA-CGE-PENK", new_label))

kable(bind_rows(mge_sst_family, mge_pvalb_family, lge_family, cge_family),
      caption = paste("Every collision group's own within-group differential-expression naming gene,",
                      "replacing the chicken match's own second gene (already shown not to",
                      "discriminate in this species above)."))

gaba_named <- bind_rows(
  gaba_specific,
  mge_sst_family,
  mge_pvalb_family,
  lge_family,
  cge_family
) %>%
  left_join(gaba_named %>% select(cluster, division, confidence_tier), by = "cluster")
stopifnot(setequal(gaba_named$cluster, gaba_eligible), !anyDuplicated(gaba_named$cluster))

## GABA-6 (GABA-MGE-ANO1 by the rule above) is overridden to GABA-MGE-LAMP5 -- still an MGE-division
## label, but named for LAMP5 rather than the chicken top_rank_agg's own ANO1, because a second,
## independent line of cross-species evidence points at LAMP5 specifically: the mouse/Yao integration
## in the same `composite_scoring` output independently calls this cluster an `Lamp5`-labeled type
## (high confidence 0.52), which `ANO1` alone doesn't have. `LAMP5` was not the chicken `top_rank_agg`
## call and so never entered the verification step above on its own; overridden by hand here because
## the mouse integration flagged it as a candidate worth checking directly, and **confirmed directly
## against this object's own expression** -- not assumed -- `LAMP5` is detected in 64.1% of this
## cluster's own cells versus 11.9% in the next-highest GABA cluster (a 6.3-fold margin), at least as
## clean a discriminator as `ANO1` itself.
gaba_label_overrides <- c("GABA-6" = "GABA-MGE-LAMP5")
gaba_named <- gaba_named %>%
  mutate(new_label = dplyr::if_else(cluster %in% names(gaba_label_overrides),
                                    unname(gaba_label_overrides[cluster]), new_label))

kable(gaba_named %>% arrange(new_label) %>% select(cluster, new_label, confidence_tier),
      caption = "Final GABA labels: a verified composite gene, or a within-group DE gene.")
```

**Confirmed, not assumed**: the `GABA-5-1` -> `PENK` override above, checked directly against its
algorithmic replacement (`PTPRK`) and its only group sibling, `GABA-5-2`.

```{r}
#| label: gaba-penk-qc
penk_ptprk_mat <- GetAssayData(obj, assay = "SCT", layer = "data")[c("PENK", "PTPRK"), gaba_cells, drop = FALSE]
tibble(cluster = as.character(obj@meta.data[gaba_cells, cluster_col]),
      PENK = penk_ptprk_mat["PENK", ], PTPRK = penk_ptprk_mat["PTPRK", ]) %>%
  group_by(cluster) %>%
  summarise(n = n(), pct_PENK = mean(PENK > 0) * 100, pct_PTPRK = mean(PTPRK > 0) * 100, .groups = "drop") %>%
  arrange(desc(pct_PENK)) %>%
  kable(digits = 1, caption = paste(
    "PENK and PTPRK detection rate across all eligible GABA clusters -- PENK should read as more",
    "specific to GABA-5-1 than PTPRK was, which is why it replaced PTPRK as GABA-5-1's own name."
  ))
```

**Confirmed, not assumed**: the eight within-group naming genes above (two MGE groups of three, one
LGE pair, one CGE pair), checked directly against detection rate within their own group -- each own
cluster should read as the clear top, not merely "detected".

```{r}
#| label: gaba-family-qc
## Built from the family tibbles themselves (not a separately hardcoded gene list), so this can't
## silently drift from what gaba-final-labels actually picked. The naming gene is always the last
## "-"-delimited token of new_label (division and class-marker tokens never contain a literal "-").
family_groups <- list(
  list(label = "MGE-SST", clusters = c("GABA-2-1", "GABA-2-2", "GABA-2-3"), df = mge_sst_family),
  list(label = "MGE-PVALB", clusters = c("GABA-3", "GABA-4-1", "GABA-4-2"), df = mge_pvalb_family),
  list(label = "LGE", clusters = c("GABA-1-1", "GABA-1-2"), df = lge_family),
  list(label = "CGE", clusters = c("GABA-5-1", "GABA-5-2"), df = cge_family)
)
map_dfr(family_groups, function(g) {
  genes <- setNames(sub(".*-", "", g$df$new_label), g$df$cluster)
  cells <- colnames(obj)[obj@meta.data[[cluster_col]] %in% g$clusters]
  mat <- GetAssayData(obj, assay = "SCT", layer = "data")[unique(genes), cells, drop = FALSE]
  ids <- obj@meta.data[cells, cluster_col]
  map_dfr(names(genes), function(cl) {
    gene <- genes[[cl]]
    tibble(group = g$label, own_cluster = cl, gene = gene,
          pct_in_own = mean(mat[gene, ids == cl] > 0) * 100,
          pct_in_others = mean(mat[gene, ids != cl] > 0) * 100)
  })
}) %>%
  kable(digits = 1, caption = paste(
    "Each within-group naming gene's detection rate in the cluster it names versus every other",
    "cluster in its own group."
  ))
```

**Confirmed, not assumed**: the direct evidence behind the `GABA-6` -> `GABA-LAMP5` override above --
`LAMP5` detection rate against every other GABA cluster.

```{r}
#| label: gaba-lamp5-qc
lamp5_mat <- GetAssayData(obj, assay = "SCT", layer = "data")["LAMP5", gaba_cells, drop = FALSE]
tibble(cluster = as.character(obj@meta.data[gaba_cells, cluster_col]), LAMP5 = lamp5_mat[1, ]) %>%
  group_by(cluster) %>%
  summarise(n = n(), pct_LAMP5 = mean(LAMP5 > 0) * 100, mean_LAMP5 = mean(LAMP5), .groups = "drop") %>%
  arrange(desc(pct_LAMP5)) %>%
  kable(digits = 2, caption = "LAMP5 detection rate and mean expression across all eligible GABA clusters.")
```

**Confirmed, not assumed**: the direct evidence behind the `GABA-4-2` override above -- detection rate
(fraction of a cluster's own cells with nonzero SCT expression) for `ST18`/`SST`/`PVALB`, against the
cluster that actually earns the `ST18` name and its own new SST_PVALB collision-mates.

```{r}
#| label: gaba-4-2-qc
qc_clusters <- c("GABA-4-2", "GABA-7", "GABA-3", "GABA-4-1")
qc_mat <- GetAssayData(subset(obj, cells = colnames(obj)[obj@meta.data[[cluster_col]] %in% qc_clusters]),
                       assay = "SCT", layer = "data")[c("ST18", "SST", "PVALB"), ]
qc_cluster_f <- obj@meta.data[colnames(qc_mat), cluster_col]
tibble(cluster = qc_cluster_f, ST18 = qc_mat["ST18", ], SST = qc_mat["SST", ], PVALB = qc_mat["PVALB", ]) %>%
  group_by(cluster) %>%
  summarise(n = n(), pct_ST18 = mean(ST18 > 0) * 100, pct_SST = mean(SST > 0) * 100,
           pct_PVALB = mean(PVALB > 0) * 100, .groups = "drop") %>%
  arrange(factor(cluster, levels = qc_clusters)) %>%
  kable(digits = 1, caption = paste(
    "GABA-4-2 (its own top_rank_agg says ST18_PVALB, overridden above to SST_PVALB): its own ST18",
    "detection rate reads like the SST_PVALB clusters (GABA-3/GABA-4-1), not like GABA-7, the",
    "cluster whose own ST18 detection rate actually earns that name."
  ))
```

## Build the hybrid labels

```{r}
#| label: build-labels
all_clusters <- levels(factor(obj@meta.data[[cluster_col]]))
glut_clusters <- all_clusters[grepl("^Glut", all_clusters)]

hand_labeled <- c(song_region_overrides, precursor_overrides, nido2_override)

## Generic pool: every Glut cluster not excluded, not hand-labeled, not itself a fold source (Nido-1
## is assigned after, from its fold target's own result).
generic_pool <- setdiff(glut_clusters, c(excluded_clusters, names(hand_labeled), names(fold_targets)))

generic <- glut_division %>%
  filter(cluster %in% generic_pool) %>%
  ## Every member of generic_pool must resolve to one of the three real divisions -- the
  ## division-check chunk above already guarantees this for anything NOT excluded/overridden, and
  ## everything excluded/overridden was just removed from generic_pool, so this should never fire;
  ## kept as a hard stop rather than a silent NA if that invariant is ever broken by a future edit.
  { stopifnot(all(.$division %in% divisions)); . } %>%
  arrange(division, cluster) %>%
  group_by(division) %>%
  mutate(new_label = paste0("Glut-", division, "-", row_number())) %>%
  ungroup() %>%
  select(cluster, new_label)

fold_labels <- tibble(
  cluster = names(fold_targets),
  new_label = generic$new_label[match(fold_targets, generic$cluster)]
)
stopifnot(!anyNA(fold_labels$new_label))

hybrid_lookup <- bind_rows(
  generic,
  fold_labels,
  tibble(cluster = names(hand_labeled), new_label = unname(hand_labeled)),
  tibble(cluster = excluded_clusters, new_label = NA_character_),
  gaba_named %>% select(cluster, new_label)
) %>%
  ## Non-neuronal clusters are untouched by this scheme -- carried through as-is rather than
  ## silently dropped from the lookup, so every cluster in the object has exactly one row here.
  ## GABA-Pre is excluded from this bucket: it already got its own row above via
  ## hand_labeled/precursor_overrides (-> GABA-Im), and every other GABAergic cluster already got
  ## its own row via gaba_named just above, so both must be excluded here or they would each get a
  ## second, conflicting row for the same cluster.
  bind_rows(tibble(cluster = setdiff(all_clusters, c(glut_clusters, names(hand_labeled), gaba_named$cluster)),
                   new_label = setdiff(all_clusters, c(glut_clusters, names(hand_labeled), gaba_named$cluster))))

stopifnot(setequal(hybrid_lookup$cluster, all_clusters), !anyDuplicated(hybrid_lookup$cluster))

kable(hybrid_lookup %>% arrange(cluster), caption = "Full hybrid-scheme lookup, one row per cluster")
```

```{r}
#| label: apply
write_csv(hybrid_lookup, lookup_fname)
message("Wrote ", lookup_fname)

obj$celltype_hybrid <- hybrid_lookup$new_label[match(obj@meta.data[[cluster_col]], hybrid_lookup$cluster)]
qs_save(obj, obj_out_fname, nthreads = 8)
message("Wrote ", obj_out_fname)
```

## Sanity check: do the three divisions actually separate cleanly?

```{r}
#| label: division-dotplot
Idents(obj) <- "celltype_hybrid"
glut_idents <- sort(na.omit(unique(hybrid_lookup$new_label[hybrid_lookup$cluster %in% glut_clusters])))
DotPlot(subset(obj, idents = glut_idents), features = divisions, assay = "SCT") +
  RotatedAxis() +
  labs(title = "DACH2 / SATB2 / CACNA1H expression across hybrid-scheme Glut labels")
```

The point of this one is to catch a labelling mistake, not to admire the labels: a gene pulled out of
a GABA cluster's own composite match should show up as visibly higher in the row it names than the
other GABA clusters. Clusters left on the raw-suffix fallback (see above) contribute no gene of their
own here.

```{r}
#| label: gaba-label-genes-dotplot
#| fig-width: 12
#| fig-height: 8
## Genes embedded in each GABA cluster's own new label, parsed the same way the labels themselves
## were built ("GABA-" and its division token dropped; the division token is optional in the strip
## below only so a future standalone override with no division prefix wouldn't silently break it --
## remaining "-"/"_"-delimited tokens are genes) -- not re-derived some other way, so this plot tests
## the actual rename rather than a re-interpretation of it.
gaba_label_genes <- gaba_named %>%
  mutate(body = sub(paste0("^GABA-((", paste(gaba_divisions, collapse = "|"), ")-)?"), "", new_label)) %>%
  separate_rows(body, sep = "[-_]") %>%
  filter(!grepl("^[0-9]+$", body)) %>%
  distinct(cluster, gene = body)

genes_present <- intersect(unique(gaba_label_genes$gene), rownames(obj))
genes_missing <- setdiff(unique(gaba_label_genes$gene), rownames(obj))
if (length(genes_missing) > 0) {
  warning("GABA label genes not found in the object, dropped from the plot: ",
          paste(genes_missing, collapse = ", "), call. = FALSE)
}

gaba_idents <- sort(na.omit(unique(hybrid_lookup$new_label[hybrid_lookup$cluster %in% gaba_named$cluster])))
DotPlot(subset(obj, idents = gaba_idents), features = genes_present, assay = "SCT") +
  RotatedAxis() +
  labs(title = "Label genes across hybrid-scheme GABA labels")
```

`CACNA1B` (named in `GABA-CGE-CACNA1B`, i.e. `GABA-5-2`) is expected to be missing from this plot and
the `genes_missing` warning above -- this genome's own annotation has no `CACNA1B`-symbol gene at all
(confirmed against `rownames(obj)` and the other seven `CACNA1<A,C,D,E,G,H,I,S>` paralogs, all
present), so that specific label cannot be sanity-checked against expression the way every other one
here can; see the rules section above.

```{r}
#| label: session-info
sessionInfo()
```
