---
title: "Manual-marker dotplot, hybrid division-based labels"
format: html
editor: source
---

Brings over the curated (not top-DEG) manual-marker dotplot from
`snrna_seurat_cellbender.0.01_preprocess/deg/all_deg.qmd`'s `dotplot_manual-markers` chunk, on the
current hybrid division-based cluster labels (`snrna/naming/hybrid_division_naming.qmd`) instead of
the raw `cluster` identity that notebook used. The cluster set and order are harmonized with
`celltypes_hclust_all_hybrid.qmd`'s tree analysis -- same `ct_order`, same cluster omissions
(`Glut-Arco-4`, `Glut-Nido-3` dropped, not just unlabeled) -- and the cluster axis is oriented so this
dotplot can be read directly alongside that notebook's
`cluster_proportions_vert_with-bar_ordered.pdf` (clusters top-to-bottom in the same order).

Independent of the source notebooks, by the same convention used throughout this project's naming
work: reads their outputs read-only, writes its own copy.

```{r}
#| label: packages
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(scCustomize)
library(cowplot)
library(egg)
theme_set(theme_cowplot())
```

## Parameters

```{r}
#| label: params
in_fname <- here::here("snrna/naming/hybrid_division_naming", "obj_hybrid_labels.qs2")
res_to_use <- "celltype_hybrid"
assay_to_use <- "SCT"

out_dir <- here::here("snrna/deg", "dotplot_manual_markers_hybrid")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

## Identical to celltypes_hclust_all_hybrid.qmd's ct_order -- not re-derived, so this can't
## silently drift from the cluster set/order that notebook's vertical heatmap uses.
ct_order <- c("Glut-DACH2-HVCra", "Glut-DACH2-HVCra-Pre", "Glut-DACH2-HVCx",
             "Glut-DACH2-1", "Glut-DACH2-2", "Glut-DACH2-3", "Glut-DACH2-4",
             "Glut-DACH2-LMANsh", "Glut-DACH2-LMANco",
             "Glut-DACH2-5", "Glut-DACH2-6", "Glut-DACH2-7", "Glut-DACH2-8", "Glut-SATB2-1",
             "Glut-CACNA1H-RA", "Glut-CACNA1H-1", "Glut-CACNA1H-2", "Glut-CACNA1H-3", "Glut-CACNA1H-4",
             "Glut-GABA",
             "Glut-Im", "Glut-NB", "Glut-NSC",
             "GABA-LGE-1", "GABA-LGE-2",
             "GABA-MGE-SST-1", "GABA-MGE-SST-2", "GABA-MGE-SST-3",
             "GABA-MGE-PVALB-1", "GABA-MGE-PVALB-2", "GABA-MGE-PVALB-3",
             "GABA-MGE-LAMP5", "GABA-MGE-ST18", "GABA-MGE-LHX8",
             "GABA-CGE-1", "GABA-CGE-2",
             "GABA-Im",
             "Astro-1", "Astro-2", "Astro-3", "Epen",
             "Oligo-1", "Oligo-2", "Oligo-3", "OPC",
             "Micro", "Endo")

## The curated marker list from all_deg.qmd's dotplot_manual-markers chunk, unchanged -- this
## request is about harmonizing the CLUSTER axis, not re-picking the genes.
manual_markers <- c("SLC17A6", "SYT1", "DACH2", "SATB2", "LHX9", "CACNA1H",
                    "NECTIN3", "DCX", "SOX4", "NR2E1",
                    "GAD1", "FOXP2", "NPY", "CEMIP", "NXPH1", "TTLL5", "PENK", "CRH", "LAMP5", "CALB2", "LHX8",
                    "SLC1A3", "SPEF2", "PLP1", "PDGFRA", "CSF1R", "FLI1")
```

## Load, apply the same cluster omissions as the tree analysis

```{r}
#| label: load
obj <- qs_read(in_fname, nthreads = 8)
DefaultAssay(obj) <- assay_to_use

## Same omission as celltypes_hclust_all_hybrid.qmd: celltype_hybrid is NA for exactly the excluded
## artifact clusters (Glut-Arco-4, Glut-Nido-3) -- dropped, not left as an unlabeled row.
obj <- subset(obj, cells = colnames(obj)[!is.na(obj@meta.data[[res_to_use]])])

stopifnot(
  "ct_order does not exactly match the clusters present in obj" =
    setequal(ct_order, unique(obj@meta.data[[res_to_use]]))
)

## rev(ct_order): ggplot's discrete axis draws the FIRST factor level at the axis origin, so
## reversing here is what puts ct_order's first entry (Glut-DACH2-HVCra) at the TOP of the plot --
## matching cluster_proportions_vert_with-bar_ordered.pdf's top-to-bottom row order, not bottom-to-top.
obj@meta.data[[res_to_use]] <- factor(obj@meta.data[[res_to_use]], levels = rev(ct_order))
```

## Dotplot

```{r}
#| label: dotplot-manual-markers
#| fig-height: 10
#| fig-width: 10
markers_present <- intersect(manual_markers, rownames(obj))
missing <- setdiff(manual_markers, markers_present)
if (length(missing) > 0) {
  warning("Manual markers not present in this object's ", assay_to_use, " assay, dropped: ",
         paste(missing, collapse = ", "))
}

scale_factor <- 0.08
height <- length(ct_order) * scale_factor
width <- length(markers_present) * scale_factor

gg <- DotPlot_scCustom(obj, features = markers_present, group.by = res_to_use, flip_axes = FALSE,
                      dot.scale = 2, dot.min = 0.05, col.min = 0)
gg <- gg +
  theme(axis.text.x = element_text(size = 6, angle = 90, hjust = 1, vjust = 0.5),
       panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.25),
       axis.text.y = element_text(size = 6),
       axis.line = element_blank(),
       axis.ticks = element_blank())

gg <- egg::set_panel_size(p = gg, width = unit(width, "in"), height = unit(height, "in"))
plot(gg)
save_plot(file.path(out_dir, "dotplot_manual-markers.pdf"), gg, base_height = height + 5, base_width = width + 5)
```

```{r}
#| label: session-info
sessionInfo()
```
