---
title: "Glut-DACH2-HVCra-Int against Lovell et al. 2008's HVC-vs-Shelf gene lists"
format:
  html:
    toc: true
    df-print: paged
---

`snrna/naming/hybrid_division_naming.qmd` renamed `Glut-HVC-1a` to `Glut-DACH2-HVCra-Int`
(*intermediate*, replacing an earlier `-Pre`/precursor name that the data did not support): a
population sitting transcriptionally between `Glut-DACH2-HVCra` and the surrounding nidopallial
`Glut-DACH2-1`/`Glut-DACH2-2` clusters. In the Xenium mapping this cluster forms a border around HVC.

The classical anatomical name for the tissue immediately around HVC is the **shelf**, and it has a
published transcriptional description: Lovell PV, Clayton DF, Replogle KL, Mello CV (2008),
"Birdsong 'transcriptomics': neurochemical specializations of the oscine song system", *PLoS ONE*
3(10):e3440, which used laser-capture microdissection plus microarray on six adult male zebra finches
to compare HVC against the adjacent shelf. This notebook asks whether `Glut-DACH2-HVCra-Int`'s
position relative to `Glut-DACH2-HVCra` lines up with that published HVC/shelf axis.

Independent of the naming notebook, by the same convention used throughout this project: reads its
output (`obj_hybrid_labels.qs2`) read-only and writes only its own results.

**Gene lists.** `lovell2008_hvc_shelf/lovell2008_tableS1_genes.csv` is the paper's Table S1 (the
primary FDR < 0.05 list), parsed from the journal's own supplementary file: 255 unique annotated
genes, 119 with `+` regulation (enriched in HVC) and 136 with `-` (enriched in the shelf, i.e. lower
in HVC). The secondary/ND supplementary lists (Tables S2/S5) are deliberately not used -- the primary
list is the one with a stated FDR.

**Two caveats, up front.** (1) The paper's contrast is a bulk tissue contrast, so its "shelf" list is
whatever is higher in the surrounding tissue than in HVC; it is not a marker set for one cell type,
and generic nidopallial clusters score high on it too. So the informative statistic is where
`Glut-DACH2-HVCra-Int` falls *between* HVCra and the surrounding nidopallium, not its shelf score in
isolation. (2) The shelf proper is the ventral auditory region below HVC, while the Xenium border is
a ring; agreement with this gene list is evidence about the HVC/surround axis, not proof that this
cluster is the auditory shelf specifically.

```{r}
#| label: packages
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(knitr)
```

## Parameters

```{r}
#| label: params
obj_fname <- here::here("snrna/naming/hybrid_division_naming", "obj_hybrid_labels.qs2")
genes_fname <- here::here("snrna/deg/lovell2008_hvc_shelf", "lovell2008_tableS1_genes.csv")

out_dir <- here::here("snrna/deg", "lovell2008_hvc_shelf")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

target <- "Glut-DACH2-HVCra-Int"
reference <- "Glut-DACH2-HVCra"
```

```{r}
#| label: load
obj <- qs_read(obj_fname, nthreads = 8)
DefaultAssay(obj) <- "SCT"
obj <- obj[, !is.na(obj$celltype_hybrid)]
Idents(obj) <- obj$celltype_hybrid

lovell <- read_csv(genes_fname, show_col_types = FALSE)
stopifnot(setequal(lovell$enriched_in, c("HVC", "Shelf")), !anyDuplicated(lovell$gene))

hvc_genes <- intersect(lovell$gene[lovell$enriched_in == "HVC"], rownames(obj))
shelf_genes <- intersect(lovell$gene[lovell$enriched_in == "Shelf"], rownames(obj))
tibble(list = c("HVC-enriched (+)", "Shelf-enriched (-)"),
       n_published = c(sum(lovell$enriched_in == "HVC"), sum(lovell$enriched_in == "Shelf")),
       n_in_this_object = c(length(hvc_genes), length(shelf_genes))) %>%
  kable(caption = "Lovell 2008 Table S1 genes recoverable in this object's own annotation")
```

Genes are matched on symbol. Some of the paper's genes have no symbol-matched entry in this genome's
annotation and are simply dropped -- the loss is reported above rather than assumed away, and it
applies to both lists roughly evenly, so it does not bias the direction of the test below.

## Is the HVCra-Int vs. HVCra difference the published HVC/shelf difference?

```{r}
#| label: de
de <- FindMarkers(obj, ident.1 = target, ident.2 = reference,
                  min.pct = 0.15, logfc.threshold = 0.4) %>%
  rownames_to_column("gene") %>%
  filter(p_val_adj < 0.05)
de_up <- de$gene[de$avg_log2FC > 0]    # higher in HVCra-Int than in HVCra
de_dn <- de$gene[de$avg_log2FC < 0]    # lower in HVCra-Int than in HVCra
write_csv(de, file.path(out_dir, "de_hvcra_int_vs_hvcra.csv"))
```

```{r}
#| label: enrichment
## One-sided Fisher tests against the whole detected transcriptome. All four combinations are run,
## not just the two that are expected to be positive -- the two "wrong-direction" tests are the
## control, and are what would catch a result driven by list size or by generic gene properties
## rather than by direction.
universe <- rownames(obj)
enrich <- function(gene_set, de_set, set_label, de_label) {
  a <- length(intersect(gene_set, de_set))
  tab <- matrix(c(a, length(setdiff(de_set, gene_set)),
                  length(setdiff(gene_set, de_set)),
                  length(universe) - a - length(setdiff(de_set, gene_set)) -
                    length(setdiff(gene_set, de_set))), nrow = 2)
  ft <- fisher.test(tab, alternative = "greater")
  tibble(published_list = set_label, de_direction = de_label, overlap = a,
         odds_ratio = round(unname(ft$estimate), 2), p_value = signif(ft$p.value, 3))
}
bind_rows(
  enrich(shelf_genes, de_up, "Shelf-enriched", paste("up in", target)),
  enrich(hvc_genes, de_dn, "HVC-enriched", paste("down in", target)),
  enrich(hvc_genes, de_up, "HVC-enriched", paste("up in", target)),
  enrich(shelf_genes, de_dn, "Shelf-enriched", paste("down in", target))
) %>%
  kable(caption = paste0("Lovell 2008 lists vs. ", target, " differential expression against ",
                         reference, " (one-sided Fisher, universe = all detected genes)"))
```

The first two rows are the test; the last two are the control and should be unremarkable. A cluster
that differs from HVCra along the published HVC/shelf axis gives strong enrichment in the first two
and none in the last two.

```{r}
#| label: overlap-genes
tibble(gene = sort(intersect(shelf_genes, de_up)), published = "Shelf-enriched",
       direction = paste("up in", target)) %>%
  bind_rows(tibble(gene = sort(intersect(hvc_genes, de_dn)), published = "HVC-enriched",
                   direction = paste("down in", target))) %>%
  write_csv(file.path(out_dir, "lovell_overlap_genes.csv")) %>%
  kable(caption = "The individual genes driving the two concordant overlaps")
```

## Where each cluster sits on the published axis

```{r}
#| label: module-scores
## Score every cell on both published lists, then take the difference. This puts the target cluster
## on the same axis as HVCra and as the surrounding nidopallium, which is the comparison that
## matters -- an absolute shelf score on its own is not interpretable (see the caveats above).
scored <- AddModuleScore(obj, features = list(hvc_genes, shelf_genes),
                         name = "lovell", assay = "SCT", seed = 1)
score_sel <- c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx", "Glut-CACNA1H-RA", "Glut-DACH2-LMANco",
               "Glut-DACH2-HVCra-Int", "Glut-DACH2-LMANsh", "Glut-DACH2-1", "Glut-DACH2-2",
               "Glut-DACH2-3", "Glut-DACH2-4", "Glut-DACH2-5", "Glut-DACH2-6", "Glut-DACH2-7",
               "Glut-DACH2-8", "Glut-SATB2-1", "Glut-Im", "Glut-NB", "Glut-NSC")
scores <- tibble(celltype = scored$celltype_hybrid,
                 hvc_score = scored$lovell1, shelf_score = scored$lovell2) %>%
  filter(celltype %in% score_sel) %>%
  group_by(celltype) %>%
  summarise(n_cells = n(),
            hvc_score = round(median(hvc_score), 3),
            shelf_score = round(median(shelf_score), 3),
            hvc_minus_shelf = round(median(hvc_score - shelf_score), 3)) %>%
  arrange(desc(hvc_minus_shelf))
write_csv(scores, file.path(out_dir, "lovell_module_scores.csv"))
kable(scores, caption = "Median per-cluster Lovell 2008 HVC and Shelf module scores")
```

```{r}
#| label: module-score-plot
#| fig-width: 7
#| fig-height: 5
scores %>%
  mutate(celltype = fct_reorder(celltype, hvc_minus_shelf),
         highlight = case_when(celltype == target ~ "HVCra-Int",
                               celltype %in% c("Glut-DACH2-HVCra", "Glut-DACH2-1",
                                               "Glut-DACH2-2") ~ "anchor",
                               TRUE ~ "other")) %>%
  ggplot(aes(hvc_minus_shelf, celltype, fill = highlight)) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  scale_fill_manual(values = c(`HVCra-Int` = "#d1495b", anchor = "#30638e", other = "grey70")) +
  labs(x = "Lovell 2008 HVC score - Shelf score (median per cluster)", y = NULL,
       title = "Position on the published HVC/shelf axis") +
  theme_bw()
ggsave(file.path(out_dir, "lovell_axis_barplot.pdf"), width = 7, height = 5)
```

`Glut-DACH2-HVCra-Int` is expected to land between `Glut-DACH2-HVCra` (most HVC-like) and
`Glut-DACH2-1`/`Glut-DACH2-2` (the surrounding nidopallium, most shelf-like) -- the same intermediate
position the naming notebook's own DE-based axis puts it in, reached here from an entirely
independent, published, cross-study gene set.

```{r}
#| label: session-info
sessionInfo()
```
