---
title: "Song vs surround DEG, excitatory (hybrid labels)"
format: html
editor: source
---

Parallels `song-surround_deg_glut.qmd`, but on the current (hybrid division-based) labels from
`snrna/naming/hybrid_division_naming.qmd` instead of the raw `cluster` identity.

Three things change beyond the names themselves, and each is noted where it happens:

- the two clusters that notebook excludes (`Glut-Arco-4`, `Glut-Nido-3`) are gone from the object;
- `Glut-NC-1` and `Glut-Nido-1` are **one** cluster under the hybrid scheme, `Glut-DACH2-1`, so the
  surround term of the HVCra comparison is that merged cluster, not `Glut-NC-1` alone;
- the pseudobulk replicate is the bird (`individual` from souporcell), not `assignment`.

```{r}
library(tidyverse)
library(Seurat)
library(scCustomize)
library(qs2)
library(UpSetR)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ComplexHeatmap)
library(cowplot)
library(ggpubr)
library(here)
select = dplyr::select

source(here::here("config/paths.R"))
```

```{r}
script_name = "song-surround_deg_glut_hybrid"
data_obj = here::here("snrna/naming/hybrid_division_naming", "obj_hybrid_labels.qs2")
out_dir = here::here("snrna/deg", script_name)
dir.create(out_dir, recursive = T)
res_to_use = "celltype_hybrid"
```

```{r}
obj = qs_read(data_obj, nthreads = 1)

## Souporcell doublets. The preprocessing notebook already dropped them, so this is a no-op on the
## current object -- kept because it is the guard that made the old version correct and costs
## nothing if an upstream re-run ever stops filtering.
cells = Cells(obj)[!grepl("/", obj$assignment)]
obj = subset(obj, cells=cells)

## Excluded clusters: celltype_hybrid is NA for exactly Glut-Arco-4 and Glut-Nido-3.
obj = subset(obj, cells = Cells(obj)[!is.na(obj@meta.data[[res_to_use]])])
obj@meta.data[[res_to_use]] = droplevels(factor(obj@meta.data[[res_to_use]]))
```

## Individual identity

The pseudobulk replicate has to be the bird. `assignment` is souporcell's *per-library* cluster
index (0/1/2), recycled independently in each of the six libraries, so it names three arbitrary
labels six times over rather than three birds. The old version of this notebook grouped on it,
which was survivable only as long as every cluster sat inside a single library. Under the hybrid
scheme that is no longer true -- `Glut-DACH2-1` merges `Glut-NC-1` with `Glut-Nido-1` -- so a
pseudobulk column keyed on `assignment` would pool cells from different birds.

`snrna/clustering/snrna_souporcell_clustering.qmd` resolves the indices into six birds by matching
clusters across libraries on genotype; its per-cell table is read here rather than re-derived, and
joined on cell name, which is exact (asserted below).

```{r}
soup = read_csv(here::here("snrna/clustering/snrna_souporcell_clustering",
                           "souporcell_assignments.csv"),
                show_col_types = FALSE)

stopifnot(
  "souporcell assignments do not cover every cell -- check the cell-name suffix convention" =
    all(Cells(obj) %in% soup$cell)
)

md = tibble(cell = Cells(obj)) %>%
  left_join(soup %>% select(cell, sample_id, individual, batch), by = "cell") %>%
  mutate(individual = factor(individual, levels = sprintf("bird%s", 1:6)),
         soup_batch = factor(sprintf("batch%s", batch)),
         soup_library = sample_id) %>%
  select(cell, individual, soup_batch, soup_library) %>%
  column_to_rownames("cell")

## A silent NA here would become a seventh pseudobulk "bird", so assert rather than assume.
stopifnot("unassigned individuals after the join" = !any(is.na(md$individual)))
obj = AddMetaData(obj, md)
```

```{r}
table(obj@meta.data[[res_to_use]], obj$position)
```

```{r}
table(obj@meta.data[[res_to_use]], obj$individual)
```

## Song / surround pairs

Element 1 of each pair is the song-region cluster, the rest its nidopallial or arcopallial
surround. Renamed one-for-one from the `cluster`-based version:

| old song | old surround | hybrid song | hybrid surround |
|---|---|---|---|
| `Glut-RA` | `Glut-Arco-1` | `Glut-CACNA1H-RA` | `Glut-CACNA1H-1` |
| `Glut-HVC-1` | `Glut-NC-1` | `Glut-DACH2-HVCra` | `Glut-DACH2-1` |
| `Glut-HVC-2` | `Glut-NC-4` | `Glut-DACH2-HVCx` | `Glut-DACH2-4` |
| `Glut-LMAN-2` | `Glut-NR-4`, `Glut-NR-2` | `Glut-DACH2-LMANco` | `Glut-DACH2-8`, `Glut-DACH2-6` |
| `Glut-LMAN-1` | `Glut-NR-1`, `Glut-NR-3` | `Glut-DACH2-LMANsh` | `Glut-DACH2-5`, `Glut-DACH2-7` |

`Glut-DACH2-1` is `Glut-NC-1` **and** `Glut-Nido-1` merged, so the HVCra surround is a larger,
slightly different set of cells than it was -- the one place where the comparison itself, not just
its name, has moved.

```{r}
pairs = list(`Glut-CACNA1H-RA`   = c("Glut-CACNA1H-RA", "Glut-CACNA1H-1"),
             `Glut-DACH2-HVCra`  = c("Glut-DACH2-HVCra", "Glut-DACH2-1"),
             `Glut-DACH2-HVCx`   = c("Glut-DACH2-HVCx", "Glut-DACH2-4"),
             `Glut-DACH2-LMANco` = c("Glut-DACH2-LMANco", c("Glut-DACH2-8", "Glut-DACH2-6")),
             `Glut-DACH2-LMANsh` = c("Glut-DACH2-LMANsh", c("Glut-DACH2-5", "Glut-DACH2-7")))

stopifnot("a pair names a label that is not in celltype_hybrid" =
            all(unlist(pairs) %in% levels(obj@meta.data[[res_to_use]])))

## The song clusters, in the order every figure below uses.
song_cts = names(pairs)[c(2, 3, 4, 5, 1)]  # HVCra, HVCx, LMANco, LMANsh, RA
```

# Markers

```{r}
assay_to_use = "SCT"
markers_fname = file.path(out_dir, sprintf("marker_genes_%s_%s.rds", res_to_use, assay_to_use))

obj_agg = AggregateExpression(obj, assay=assay_to_use,
                              group.by=c(res_to_use, "individual"), return.seurat = T)
stopifnot("AggregateExpression did not carry the grouping variable through" =
            res_to_use %in% colnames(obj_agg@meta.data))
Idents(obj_agg) = obj_agg@meta.data[[res_to_use]]

## Cached on the same principle as the embeddings in combined_all_umap_glut_hybrid.R: the DESeq2
## pseudobulk fits are the expensive step here and everything below them is plotting. Set TRUE by
## hand (or delete the rds) after changing the pairs, the object, the assay or the pseudobulk
## grouping -- nothing detects that for you.
redo_markers = !file.exists(markers_fname)
cat(if (redo_markers) "no cached markers -- fitting\n" else
    sprintf("reusing cached markers: %s\n", markers_fname))

if (redo_markers) {

  ## As in the `cluster`-based version, `ident.2` takes the first surround cluster only -- the pair
  ## vectors carry the others for the dotplots and violins further down, which do use all of them.
  markers_agg = map(pairs, function(pair_cur) {
    print(pair_cur)
    res = FindMarkers(obj_agg, assay=assay_to_use,
                      ident.1=pair_cur[1], ident.2=pair_cur[2],
                      test.use="DESeq2", only.pos = F)
    res = res %>% rownames_to_column(var="gene")
    res
  }) %>% set_names(names(pairs)) %>% bind_rows(.id="compare")

  saveRDS(markers_agg, markers_fname)
} else {
  markers_agg = readRDS(markers_fname)
}
```

## Scatters

```{r}
markers_agg_df = markers_agg %>%
  filter(p_val_adj < .1) %>%
  select(avg_log2FC, compare, gene) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC")
markers_agg_mat = markers_agg_df %>% column_to_rownames("gene") %>%
  as.matrix()

cor(markers_agg_mat, use = "complete")
```

```{r}

ct_pairs = combn(names(pairs), m = 2, simplify = F)

ngenes = 50

markers_agg_df_cur = markers_agg_df %>%
    filter(!grepl("^LOC", gene))

walk(ct_pairs, function(x) {
  ct1 = x[1]
  ct2 = x[2]

markers1 = markers_agg_df_cur %>%
  top_n(ngenes, !!as.symbol(ct1)) %>%
  select(gene) %>% unlist()

markers2 = markers_agg_df_cur %>%
  top_n(ngenes, !!as.symbol(ct2)) %>%
  select(gene) %>% unlist()

top_markers = intersect(markers1, markers2)

markers3 = markers_agg_df_cur %>%
  top_n(-1 * ngenes, !!as.symbol(ct1)) %>%
  select(gene) %>% unlist()

markers4 = markers_agg_df_cur %>%
  top_n(-1 *ngenes, !!as.symbol(ct2)) %>%
  select(gene) %>% unlist()

bottom_markers = intersect(markers3, markers4)
to_plot_markers = c(bottom_markers, top_markers)


markers_agg_df_cur = markers_agg_df %>% select(gene, !!as.symbol(ct1), !!as.symbol(ct2)) %>%
  mutate(to_plot = if_else(gene %in% to_plot_markers, gene, NA))
gg = ggplot(markers_agg_df_cur, aes(!!as.symbol(ct1), !!as.symbol(ct2))) +
  geom_vline(xintercept=0, linetype = 2) +
  geom_hline(yintercept=0, linetype = 2) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label=to_plot), na.rm = T, force_pull = 2) +
  theme_bw() +
  coord_equal()
print(gg)
save_plot(file.path(out_dir, str_interp("scatter_${ct1}_${ct2}.pdf")), gg)
})
```
## Correlation
Spearman correlation of the song-vs-surround log2 fold changes between every pair of song clusters:
does a gene that moves in HVCra move the same way in RA? The gene set is everything significant in
**at least one** pair, and each gene's fold change is then read from all five comparisons whether or
not it was significant there.

That last point is the whole design and the version of this chunk inherited from
`song-surround_deg_glut.qmd` had it backwards, alongside two other faults -- it correlated with
Pearson rather than Spearman, passed the gene matrix rather than the correlation matrix to
`Heatmap()`, and assigned the result without ever printing or saving it, so nothing was produced at
all.

The gene-set point: filtering to `p_val_adj < .1` *before* pivoting wide leaves NA wherever a gene
failed significance, and `cor(use = "complete")` then silently keeps only the genes significant in
**all five** pairs. Genes that clear the threshold everywhere are by construction the consistent
ones, so that set inflates the correlations it is used to measure. Taking the union and then reading
the unfiltered fold changes keeps NA meaning one thing only -- `FindMarkers` did not return the gene
for that comparison, having filtered it for low expression -- and those few rows are dropped.

Spearman rather than Pearson because the question is whether the pairs *rank* genes alike; the
log2FC distributions differ in spread between comparisons (HVCra and RA return far more significant
genes than the other three), which a Pearson coefficient would partly be measuring.

```{r}
sig_genes = markers_agg %>% filter(p_val_adj < .1) %>% distinct(gene) %>% pull(gene)

cor_input = markers_agg %>%
  filter(gene %in% sig_genes) %>%
  select(gene, compare, avg_log2FC) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC") %>%
  column_to_rownames("gene") %>%
  as.matrix()
cor_input = cor_input[, song_cts]

n_union = nrow(cor_input)
cor_input = na.omit(cor_input)
cat(sprintf("genes significant in >=1 pair: %s; returned by all five comparisons: %s\n",
            n_union, nrow(cor_input)))

markers_agg_cor = cor(cor_input, method = "spearman")
print(round(markers_agg_cor, 3))
write.csv(round(markers_agg_cor, 4), file.path(out_dir, "correlation_spearman.csv"))
```

The full matrix, with the diagonal blanked. The diagonal is 1 by construction and, on a scale that
has to reach 1 to hold it, every off-diagonal coefficient is pushed into the bottom of the ramp;
blanking it lets the scale be spent on the comparisons that carry information. Both triangles are
drawn, so each coefficient appears twice, symmetric about the empty diagonal.

```{r}
## NA, not 0 -- na_col paints these white, and a real zero would also be white, so they must not be
## made the same value. The diagonal is blank because it is uninformative, not because it is small.
cor_tri = markers_agg_cor
diag(cor_tri) = NA

## 0-1 greyscale. Anything at or below zero therefore reads as white and is not distinguishable
## from a coefficient of zero -- or, here, from the blanked diagonal. The printed matrix above is
## the check on that, and the range is reported here so a negative value cannot pass unnoticed.
off_diag = markers_agg_cor[row(markers_agg_cor) != col(markers_agg_cor)]
cat(sprintf("off-diagonal Spearman rho: min %.3f, max %.3f\n", min(off_diag), max(off_diag)))

cor_cols = circlize::colorRamp2(breaks = c(0, 1), colors = c("white", "black"))

cell_size = 0.2
hm = Heatmap(cor_tri,
             name = "Spearman rho",
             col = cor_cols,
             na_col = "white",
             cluster_rows = FALSE, cluster_columns = FALSE,
             row_names_gp = gpar(fontsize = 6),
             column_names_gp = gpar(fontsize = 6),
             rect_gp = gpar(col = "grey80", lwd = 0.5),
             width = unit(cell_size * ncol(cor_tri), "inch"),
             height = unit(cell_size * nrow(cor_tri), "inch"))

pdf(file.path(out_dir, "correlation_spearman.pdf"),
    width = cell_size * ncol(cor_tri) + 2.6,
    height = cell_size * nrow(cor_tri) + 1.8)
print(hm)
dev.off()
print(hm)
```

## Heatmap

The heatmaps below are drawn gene-per-row: a tall, half-inch-wide strip, which is the right shape
for reading down a gene list and the wrong shape for a figure that has to sit across a column. Each
one therefore also gets a horizontal counterpart -- the same matrix transposed, so the comparisons
are rows and the genes are columns, the gene dendrogram moves to the top and the height and width
budgets swap with it. Same data, same clustering, same colour scale; only the orientation changes.

In the transposed version the comparison names become row labels, read horizontally instead of
rotated, and the gene marks (where there are any) hang off the bottom rather than the right.

```{r}
save_heatmap_horizontal = function(mat, hc, fname, col,
                                   height = 15, width = 0.5,
                                   mark_genes = NULL,
                                   show_gene_names = FALSE,
                                   name = "avg_log2FC") {
  mat_t = t(mat)

  ## `height`/`width` are the *vertical* figure's budget, passed through unchanged so the two
  ## orientations stay tied to one set of numbers: the gene axis keeps `height` inches and the
  ## comparison axis keeps `width`, they simply change places.
  ##
  ## The marks are taken as gene *names* and resolved against the matrix here, so the positions and
  ## the labels are derived from one lookup and cannot fall out of step -- a gene that is not in the
  ## matrix drops from both. (The vertical versions above pass a na.omit()'d index alongside the
  ## unfiltered label vector, which is only correct while every gene matches.)
  ha = NULL
  if (length(mark_genes) > 0) {
    idx = match(mark_genes, colnames(mat_t))
    keep = !is.na(idx)
    if (any(keep)) {
      ha = HeatmapAnnotation(
        foo = anno_mark(at = idx[keep], labels = mark_genes[keep], side = "bottom",
                        labels_gp = gpar(fontsize = 5), lines_gp = gpar(lwd = .5)),
        which = "column"
      )
    }
  }

  ## Named, so the legend reads `avg_log2FC` rather than the auto-generated `matrix_N` the
  ## unnamed vertical versions above fall back to.
  hm = Heatmap(mat_t, col = col, name = name,
               cluster_rows = FALSE,
               cluster_columns = hc,
               show_row_dend = FALSE,
               show_column_dend = TRUE,
               column_dend_gp = gpar(lwd = .5),
               show_column_names = show_gene_names,
               row_names_gp = gpar(fontsize = 6),
               bottom_annotation = ha,
               height = unit(width, "inch"),
               width = unit(height, "inch"))

  pdf(fname, width = height + 4, height = width + 4)
  print(hm)
  dev.off()
  invisible(hm)
}
```

```{r}
ngenes = 50
markers_agg_select = markers_agg %>%
  group_by(compare) %>%
  top_n(ngenes, abs(avg_log2FC))


markers_agg_mat = markers_agg %>%
  filter(gene %in% markers_agg_select$gene) %>%
  select(avg_log2FC, compare, gene) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC") %>%
  column_to_rownames("gene") %>%
  as.matrix()

## A gene in the top 50 of one comparison need not have been returned by FindMarkers for all five,
## and dist() propagates the resulting NA into every distance, which hclust rejects outright. The
## `heatmap_sig` chunk below already drops incomplete rows for the same reason; do it here too
## rather than leave the chunk dependent on the gene sets happening to line up. 17 of 200 rows go.
markers_agg_mat = na.omit(markers_agg_mat)

height = 15
width = 0.5

max_val = 4
cols_cur = circlize::colorRamp2(breaks=c(-1*max_val, 0, max_val), colors = c(scales::muted("blue"), "white", scales::muted("red")))

hc = hclust(dist(markers_agg_mat), method = "ward.D")
hm = Heatmap(markers_agg_mat, col = cols_cur,
             cluster_rows = hc,
             show_column_dend = F,
             height = unit(height, "inch"),
             width = unit(width, "inch"))
pdf(file.path(out_dir, "heatmap.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

height = 2
width = 0.5

hm = Heatmap(markers_agg_mat, col = cols_cur,
             cluster_rows = hc,
             show_column_dend = F,
             show_row_names = F,
             height = unit(height, "inch"),
             width = unit(width, "inch"))
pdf(file.path(out_dir, "heatmap_no_genes.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

## Literal 15/0.5 rather than `height`/`width`, which were just reassigned for the no-genes variant.
save_heatmap_horizontal(markers_agg_mat, hc, file.path(out_dir, "heatmap_horizontal.pdf"),
                        col = cols_cur, height = 15, width = 0.5)
```
```{r}

markers_agg_select = markers_agg %>%
  filter(p_val_adj < 0.1) %>%
  filter(abs(avg_log2FC) >= 1)


markers_agg_mat = markers_agg %>%
  filter(gene %in% markers_agg_select$gene) %>%
  select(avg_log2FC, compare, gene) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC") %>%
  column_to_rownames("gene") %>%
  as.matrix()

markers_agg_mat = markers_agg_mat[, song_cts]
markers_agg_mat = na.omit(markers_agg_mat)

hc = hclust(dist(markers_agg_mat), method = "ward.D")

height = 15
width = 0.5

max_val = 4
cols_cur = circlize::colorRamp2(breaks=c(-1*max_val, 0, max_val), colors = c(scales::muted("blue"), "white", scales::muted("red")))

hc = hclust(dist(markers_agg_mat), method = "ward.D")
hm = Heatmap(markers_agg_mat, col = cols_cur,
             cluster_rows = hc,
             cluster_columns = F,
             show_column_dend = F,
             height = unit(height, "inch"),
             width = unit(width, "inch"))
pdf(file.path(out_dir, "heatmap_sig.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()



height = 2
width = 0.5

hm = Heatmap(markers_agg_mat, col = cols_cur,
             cluster_rows = hc,
             cluster_columns = F,
             show_column_dend = F,
             show_row_names = F,
             height = unit(height, "inch"),
             width = unit(width, "inch"))
pdf(file.path(out_dir, "heatmap_sig_no_genes.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

save_heatmap_horizontal(markers_agg_mat, hc, file.path(out_dir, "heatmap_sig_horizontal.pdf"),
                        col = cols_cur, height = 15, width = 0.5)


dist_mat = as.matrix(dist(markers_agg_mat))
hc_cut = dynamicTreeCut::cutreeDynamic(hc, cutHeight = 100,  method="hybrid", distM = dist_mat, minClusterSize = 20)
genes_split = split(hc$labels, hc_cut)
table(hc_cut)
```

```{r}

markers_agg = markers_agg %>%
  mutate(avg_log2FC_sign = sign(avg_log2FC))

markers_agg_select = markers_agg %>%
  filter(p_val_adj < 0.1) %>%
  filter(abs(avg_log2FC) >= 1)

markers_agg_select_label = markers_agg_select %>%
  group_by(compare, avg_log2FC_sign ) %>%
  filter(!grepl("LOC|nbis", gene)) %>%
  top_n(5, abs(avg_log2FC)) %>%
  ungroup() %>%
  distinct(gene)

markers_agg_mat = markers_agg %>%
  filter(gene %in% markers_agg_select$gene) %>%
  select(avg_log2FC, compare, gene) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC", values_fill = 0) %>%
  column_to_rownames("gene") %>%
  as.matrix()

markers_agg_mat = markers_agg_mat[, song_cts]

hc = hclust(dist(markers_agg_mat), method = "ward.D2")

height = 3
width = 0.5

max_val = 4
cols_cur = circlize::colorRamp2(breaks=c(-1*max_val, 0, max_val), colors = c(scales::muted("blue"), "white", scales::muted("red")))

markers_agg_select_label_ind = na.omit(match(markers_agg_select_label$gene,
                                             rownames(markers_agg_mat)))

ha = rowAnnotation(foo = anno_mark(at = markers_agg_select_label_ind,
                                   labels = markers_agg_select_label$gene,
                                   labels_gp = gpar(fontsize=5),
                                   lines_gp=gpar(lwd=.5)) )

hm = Heatmap(markers_agg_mat, col = cols_cur,

             show_row_names = F,
             cluster_rows = hc,
             cluster_columns = F,
             right_annotation = ha,
             show_column_dend = F,
             row_dend_gp = gpar(lwd=.5),
             height = unit(height, "inch"),
             width = unit(width, "inch"))
print(hm)
pdf(file.path(out_dir, "heatmap_sig_label.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

save_heatmap_horizontal(markers_agg_mat, hc, file.path(out_dir, "heatmap_sig_label_horizontal.pdf"),
                        col = cols_cur, height = 3, width = 0.5,
                        mark_genes = markers_agg_select_label$gene)



dist_mat = as.matrix(dist(markers_agg_mat))
hc_cut = dynamicTreeCut::cutreeDynamic(hc, method="hybrid", distM = dist_mat, minClusterSize = 20)
table(hc_cut)

hc_cut_df = data.frame(gene = rownames(markers_agg_mat), hc_cut = hc_cut)
markers_agg_select_label = markers_agg_select %>%
  left_join(hc_cut_df) %>%
  group_by(hc_cut) %>%
  filter(!grepl("LOC|nbis", gene)) %>%
  top_n(4, abs(avg_log2FC)) %>%
  ungroup() %>%
  distinct(gene)

height = 3
width = 0.5

max_val = 4
cols_cur = circlize::colorRamp2(breaks=c(-1*max_val, 0, max_val), colors = c(scales::muted("blue"), "white", scales::muted("red")))

markers_agg_select_label_ind = na.omit(match(markers_agg_select_label$gene,
                                             rownames(markers_agg_mat)))

ha = rowAnnotation(foo = anno_mark(at = markers_agg_select_label_ind,
                                   labels = markers_agg_select_label$gene,
                                   labels_gp = gpar(fontsize=5),
                                   lines_gp=gpar(lwd=.5)) )

hm = Heatmap(markers_agg_mat, col = cols_cur,
             show_row_names = F,
             cluster_rows = hc,
             cluster_columns = F,
             right_annotation = ha,
             show_column_dend = F,
             row_dend_gp = gpar(lwd=.5),
             height = unit(height, "inch"),
             width = unit(width, "inch"))
print(hm)
pdf(file.path(out_dir, "heatmap_sig_label_cluster.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

save_heatmap_horizontal(markers_agg_mat, hc,
                        file.path(out_dir, "heatmap_sig_label_cluster_horizontal.pdf"),
                        col = cols_cur, height = 3, width = 0.5,
                        mark_genes = markers_agg_select_label$gene)
```

### UMAPs
```{r}
umap_dir = file.path(out_dir, "umaps")
dir.create(umap_dir)
```

```{r}
obj_umap_fname = here::here("snrna/reduction_viz/combined_all_umap_glut_hybrid", "obj_clustered.qs2")
obj_umap = qs_read(obj_umap_fname, nthreads = 8)
```

```{r}
## Same parameter names as the `cluster`-based version -- the sweep in
## combined_all_umap_glut_hybrid.R is the same grid -- but these were originally picked by eye from
## the old embeddings. The hybrid embeddings are fit on a different cell set, so the panels in
## snrna/reduction_viz/combined_all_umap_glut_hybrid are worth a look before these are trusted as
## the best of the nine.
reductions_to_use = c(arco="dims10nn30mindist0.4", nido="dims20nn40mindist0.5")
```

```{r}
genes_to_plot = markers_agg_select_label$gene
## MAGIC_SCT is not on obj_hybrid_labels.qs2, so in practice only SCT is plotted; the intersect
## keeps the loop correct either way.
assays_to_use = c("SCT", "MAGIC_SCT")

## Panel size and point size. The old values -- 3/4 in panels at pt.size 3/2 -- drew points several
## times wider than the spacing between cells, so 15,000 nidopallial cells merged into a field of
## overlapping discs and nothing below the level of a whole cluster could be read. `pt.size` is in
## ggplot size units here, not pixels, so the rasteriser was never the cause and raising
## raster.dpi alone would not have helped.
##
## These were set by rendering the same gene at a range of sizes and looking at the output: the
## panels are enlarged and the points are brought down to roughly the inter-cell spacing, which is
## the value at which individual cells resolve and the expression gradient is still readable as a
## surface. The two regions differ because the arcopallial embedding holds a quarter the cells in a
## comparable area, so it can carry a larger point on a smaller panel.
##
## Drawn as vector, not rasterised. `FeaturePlot_scCustom(raster = TRUE)` ignores `raster.dpi`
## outright -- the embedded bitmap comes out 389 x 342 px at 72 ppi whether that argument says 1024,
## 2048 or 4096 (checked with `pdfimages -list`) -- so the point layer was being baked at 72 ppi and
## no setting of it would have helped. Vector costs about 900 KB for a nidopallial panel and 225 KB
## for an arcopallial one, which against the 2 GB of cached objects in this tree is not a reason to
## ship a soft figure.
plot_dims = list(arco = c(height = 5, width = 5), nido = c(height = 6, width = 6))
point_sizes = list(arco = 1.0, nido = 0.6)
walk(names(obj_umap), function(region_cur) {
  obj_cur = obj_umap[[region_cur]]
  reduction_to_use = reductions_to_use[region_cur]
  plot_dim = plot_dims[[region_cur]]
  point_size = point_sizes[[region_cur]]

  walk(intersect(assays_to_use, Assays(obj_cur)), function(assay_to_use) {
    DefaultAssay(obj_cur) = assay_to_use
    walk(genes_to_plot, function(gene_to_plot) {
      gg = FeaturePlot_scCustom(obj_cur,
                                features=gene_to_plot,
                                reduction=reduction_to_use,
                                raster = FALSE,
                                pt.size = point_size) +
        coord_equal()  +
        theme_void()
      save_plot(file.path(umap_dir,
                          str_interp("${region_cur}_${gene_to_plot}_${assay_to_use}.pdf")),
                gg, base_height = plot_dim["height"], base_width = plot_dim["width"])
    })
  })
})
```

### GO, DEG clusters

```{r}
obj_agg_cluster = AggregateExpression(obj, assay=assay_to_use, group.by=res_to_use, return.seurat = T)
mat = GetAssayData(obj_agg_cluster, assay = assay_to_use, layer = "data")

to_keep = apply(mat, 1, function(x) sum(x>0) >= (ncol(mat) * 1))
mat_expr = mat[to_keep,]
```

```{r}
go_res_fname = file.path(out_dir, "go_res.rds")
redo = F
if (redo) {
  entrezid = bitr(Features(obj), fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

  go_res = imap(genes_split, function(gene_set, gene_set_name) {

    gene_set_cur = entrezid$ENTREZID[match(gene_set, entrezid$SYMBOL)]
    gene_set_cur = gene_set_cur[!is.na(gene_set_cur)]

    universe = rownames(mat_expr)
    universe = entrezid$ENTREZID[match(universe, entrezid$SYMBOL)]
    universe = universe[!is.na(universe)]

    go_res = enrichGO(gene_set_cur, universe = universe, OrgDb = org.Hs.eg.db, ont = "ALL", maxGSSize = 200 )
    as.data.frame(go_res)
  }) %>% set_names(names(genes_split)) %>% bind_rows(.id="gene_set")

  saveRDS(go_res, go_res_fname)
} else {
  go_res = readRDS(go_res_fname)
}
```


## Dotplots

```{r}
cts = unlist(pairs)

## Song cluster next to its own surround, in the order the figure reads. As in the `cluster`-based
## version this lists one surround per song cluster, so the second surround of each LMAN pair falls
## outside the factor levels and is dropped on the next line.
ct_levels = c("Glut-DACH2-HVCra", "Glut-DACH2-1",
              "Glut-DACH2-HVCx", "Glut-DACH2-4",
              "Glut-DACH2-LMANco", "Glut-DACH2-8",
              "Glut-DACH2-LMANsh", "Glut-DACH2-5",
              "Glut-CACNA1H-RA", "Glut-CACNA1H-1")

obj_sub = subset(obj, cells = Cells(obj)[obj@meta.data[[res_to_use]] %in% cts])
obj_sub@meta.data[[res_to_use]] = factor(obj_sub@meta.data[[res_to_use]], levels = ct_levels)
obj_sub = subset(obj_sub, cells = Cells(obj_sub)[!is.na(obj_sub@meta.data[[res_to_use]])])
Idents(obj_sub) = obj_sub@meta.data[[res_to_use]]
DefaultAssay(obj_sub) = assay_to_use

obj_agg_cluster = AggregateExpression(obj, assay=assay_to_use, group.by=c(res_to_use), return.seurat = T)

```

```{r}

ngenes = 5

## Canvas = panel + the room the axis labels need, with the two budgeted separately. Folding the
## label allowance into one per-column factor (the old `length(cts) * .2`) meant the panel geometry
## moved whenever the label allowance was retuned, and it over-counted the columns: `cts` is the 12
## entries of `pairs`, but `ct_levels` keeps 10 and the other two are dropped before plotting, so a
## fifth of the width was being reserved for columns that are never drawn.
##
## The x labels are rotated to vertical, so a column costs only the width of its dot -- the label
## contributes nothing horizontally. At the old size each column had twice the width that each gene
## row had height, which is the gap this closes: the columns now sit at roughly the row pitch.
## 0.075 is close to the floor: `dot.scale = 2` puts the largest dot at about 0.078 in across, so
## at this column pitch a full-size dot fills its column edge to edge. Going narrower again means
## bringing dot.scale down with it or the dots start to collide.
dot_col_width   = 0.08  # panel inches per cluster column
dot_row_height  = 0.078  # panel inches per gene row
dot_ylab_width  = 0.72   # room for the gene names
dot_xlab_height = 1.0    # room for the rotated cluster names

n_cols = length(ct_levels)
width = n_cols * dot_col_width + dot_ylab_width

markers_agg_select_up = markers_agg %>%
  filter(!grepl("^LOC", gene)) %>%
  group_by(compare) %>%
  top_n(ngenes, avg_log2FC) %>%
  ungroup() %>%
  select(gene) %>% unlist() %>% unique()

markers_agg_mat = GetAssayData(obj_agg_cluster, assay = "SCT", layer = "data")
markers_agg_mat = markers_agg_mat[markers_agg_select_up, cts]
height = nrow(markers_agg_mat) * dot_row_height + dot_xlab_height

hc = hclust(dist(markers_agg_mat), method = "ward.D")

markers_agg_select = hc$labels[hc$order]
gg = DotPlot_scCustom(obj_sub, features = markers_agg_select, group.by = res_to_use, flip_axes = T, dot.scale=2, dot.min = .1)
gg = gg +
  theme(axis.text.x = element_text(size=6, angle = 90, hjust = 1, vjust = .5),
        axis.text.y = element_text(size=6),
        axis.line = element_line(linewidth=.5))
gg


save_plot(file.path(out_dir, str_interp("dotplot_up_${ngenes}_legend.pdf")), gg, base_height=height, base_width=width)
gg = gg + theme(legend.position="none")
save_plot(file.path(out_dir, str_interp("dotplot_up_${ngenes}.pdf")), gg, base_height=height, base_width=width)



markers_agg_select_down = markers_agg %>%
  filter(!grepl("^LOC", gene)) %>%
  group_by(compare) %>%
  top_n(-1*ngenes, avg_log2FC) %>%
  ungroup() %>%
  select(gene) %>% unlist() %>% unique()

markers_agg_mat = GetAssayData(obj_agg_cluster, assay = "SCT", layer = "data")
markers_agg_mat = markers_agg_mat[markers_agg_select_down, cts]
height = nrow(markers_agg_mat) * dot_row_height + dot_xlab_height

hc = hclust(dist(markers_agg_mat), method = "ward.D")

markers_agg_select = hc$labels[hc$order]

gg = DotPlot_scCustom(obj_sub, features = markers_agg_select, group.by = res_to_use, flip_axes = T, dot.scale=2, dot.min = .1)
gg = gg +
  theme(axis.text.x = element_text(size=6, angle = 90, hjust = 1, vjust = .5),
        axis.text.y = element_text(size=6),
        axis.line = element_line(linewidth=.5))
gg



gg_leg = get_legend(gg, position = "right")
as_ggplot(gg_leg)

gg = gg + theme(legend.position="none")
save_plot(file.path(out_dir, str_interp("dotplot_down_${ngenes}.pdf")), gg, base_height=height, base_width=width)

save_plot(file.path(out_dir, str_interp("dotplot_down_${ngenes}_legend.pdf")), gg_leg, base_height=height, base_width=width)

```
## Volcano

```{r}
markers_agg = markers_agg %>%
  mutate(sig = p_val_adj < .1,
         avg_log2FC_sign = sign(avg_log2FC),
         p_val_adj_mlog10 = -1 * log10(p_val_adj))

markers_agg_to_plot = markers_agg %>%
  group_by(avg_log2FC_sign, compare) %>%
  filter(!grepl("LOC", gene)) %>%
  top_n(5, p_val_adj_mlog10) %>%
  mutate(gene_to_label = gene) %>% select(compare, gene_to_label, gene)

markers_agg_tmp = markers_agg %>% left_join(markers_agg_to_plot)
gg = ggplot(markers_agg_tmp, aes(avg_log2FC, p_val_adj_mlog10)) +
  geom_point(size=.1) +
  ggrepel::geom_text_repel(aes(label=gene_to_label), size=2.1) +
  facet_grid(.~compare, scales="free") +
  labs(x="", y="") +
  theme_bw() +
  theme(legend.position="none",
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth=.25),
        axis.line = element_line(linewidth=.25),
        axis.ticks = element_line(linewidth=.25),
        axis.text = element_text(size=6),
        strip.text = element_text(size=7),
        strip.background = element_blank())
gg = egg::set_panel_size(p=gg, width = unit(.6, "in"), height=unit(1, "in"))
plot(gg)
save_plot(file.path(out_dir, "deg-celltype-pairs_volcano.pdf"), gg, base_height = 2.5, base_width=1, ncol=5)
```

## Violins

```{r}
obj_sub_vln = subset(obj, cells = Cells(obj)[obj@meta.data[[res_to_use]] %in% cts])
obj_sub_vln@meta.data[[res_to_use]] = factor(obj_sub_vln@meta.data[[res_to_use]], levels = ct_levels)
```

```{r}
VlnPlot_scCustom(obj_sub_vln, features = c("DPP6", "ADARB1"), group.by=res_to_use)
```

## Overlap of DEG

```{r}
markers_agg_split = split(markers_agg, markers_agg$compare)
markers_agg_sig = map(markers_agg_split, function(x) {
  x %>%
    dplyr::filter(p_val_adj < 0.1) %>%
    dplyr::select(gene) %>%
    unlist()
})
set_order = rev(c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx", "Glut-CACNA1H-RA",
                  "Glut-DACH2-LMANco", "Glut-DACH2-LMANsh"))
markers_agg_sig_upset = fromList(markers_agg_sig)
gg = upset(fromList(markers_agg_sig), group.by = "sets", order.by="freq", keep.order = T, sets = set_order, empty.intersections = "on", nintersects = 100)
gg
```

```{r}

intersect_names = list(SMP = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx", "Glut-CACNA1H-RA"),
                       NIDO = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx", "Glut-DACH2-LMANsh", "Glut-DACH2-LMANco"),
                       SONG = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx", "Glut-DACH2-LMANsh", "Glut-DACH2-LMANco", "Glut-CACNA1H-RA"),
                       HVC = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx"),
                       LMAN  = c("Glut-DACH2-LMANsh", "Glut-DACH2-LMANco"),
                       PP = c("Glut-DACH2-HVCra", "Glut-DACH2-LMANsh"),
                       PS = c("Glut-DACH2-HVCx", "Glut-DACH2-LMANco"))

intersects = map(intersect_names, function(inter_cur) {
  sort(Reduce(intersect, markers_agg_sig[inter_cur]))
}) %>% set_names(names(intersect_names))
```

## GO Terms

```{r}
obj_agg_cluster = AggregateExpression(obj, assay=assay_to_use, group.by=res_to_use, return.seurat = T)
mat = GetAssayData(obj_agg_cluster, assay = assay_to_use, layer = "data")

thresh = .05
plot(density(mat[,"Glut-DACH2-HVCra"]))
mat_expr = map(1:ncol(mat), function(i) rownames(mat)[mat[,i] >= thresh]) %>%
  set_names(colnames(mat))
```

```{r}

entrezid = bitr(markers_agg_split[[1]]$gene, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

go_res = map(names(intersect_names), function(compare) {
  cts = intersect_names[[compare]]

  gene_set = entrezid$ENTREZID[match(intersects[[compare]], entrezid$SYMBOL)]
  gene_set = gene_set[!is.na(gene_set)]

  universe = Reduce(intersect, mat_expr[cts])
  universe = entrezid$ENTREZID[match(universe, entrezid$SYMBOL)]
  universe = universe[!is.na(universe)]

  go_res = enrichGO(gene_set, universe = universe, OrgDb = org.Hs.eg.db, ont = "ALL", maxGSSize = 200 )
  as.data.frame(go_res)
}) %>% set_names(names(intersect_names)) %>% bind_rows(.id="compare")
```

## GSEA

```{r}
library(fgsea)
gmt_files = GMT_C5_V7  # bundled in data/gene_lists/

gmt_pathway = gmtPathways(gmt_files)
```

```{r}
gse_res_list = map(markers_agg_split, function(compare) {
  compare = compare %>%
    arrange(-avg_log2FC)
  gene_set = compare %>% select(avg_log2FC) %>% unlist()
  names(gene_set) = compare$gene
  gene_set = gene_set[!is.na(gene_set)]
  gene_set = gene_set[!is.na(names(gene_set))]
  gsea_res = fgseaMultilevel(pathways = gmt_pathway,
                             stats = gene_set,
                             minSize=20,
                             maxSize=200)

  gsea_res_filt = gsea_res %>% filter(padj<.2) %>% arrange(pval)
  collapsedPathways = collapsePathways(gsea_res_filt, pathways = gmt_pathway, stats = gene_set, pval.threshold = .05)
  gsea_res_ind = gsea_res %>% filter(pathway %in% collapsedPathways$mainPathways) %>%
    arrange(-NES)
  return(list(gsea_res, gsea_res_ind))
})
gse_res = map(gse_res_list, 1) %>% set_names(names(markers_agg_split)) %>% bind_rows(.id="compare")
gse_res_ind = map(gse_res_list, 2) %>% set_names(names(markers_agg_split)) %>% bind_rows(.id="compare")
```

```{r}
set_order = song_cts
  gse_res_ind_select = gse_res_ind %>%
    mutate(p.adjust_sign = -1 * log10(padj)) %>%
    mutate(NES_sign = sign(NES)) %>%
    filter(padj<.05) %>%
    select(pathway, compare) %>%
    mutate(is_sig = TRUE)

  gse_res_ind_filt = gse_res %>%
    filter(pathway %in% gse_res_ind_select$pathway) %>%
    distinct(compare, pathway, .keep_all=T) %>%
    left_join(gse_res_ind_select) %>%
    mutate(is_sig = if_else(is.na(is_sig), FALSE, is_sig)) %>%
    mutate(pathway = sub("^GO", "", pathway ),
           pathway = gsub("_", " ", pathway),
           pathway = tolower(pathway)) %>%
    mutate(pathway = case_when(
      grepl("establishment of protein", pathway) ~ "protein localization to ER",
      grepl("anaphase promoting complex", pathway) ~ "APC-dependent catabolic process",
       ## GO_CELL_CELL_ADHESION_VIA_PLASMA_MEMBRANE_ADHESION_MOLECULES and its homophilic child
       ## GO_HOMOPHILIC_CELL_ADHESION_VIA_PLASMA_MEMBRANE_ADHESION_MOLECULES both match the second
       ## pattern. Both are significant here, so collapsing them to one label put two rows on the
       ## same (compare, pathway) cell and pivot_wider turned NES into a list column. Matched most
       ## specific first, and kept as two rows -- they are two terms.
       grepl("homophilic cell adhesion via plasma", pathway) ~ "homophilic cell adhesion",
       grepl("cell adhesion via plasma", pathway) ~ "cell-cell adhesion",
       grepl("nuclear transcribed", pathway) ~ "nonsense mediated decay",

       grepl("cellular component assembly involved in morphogenesis", pathway) ~ "cellular component assembly in morphogenesis",
                               TRUE ~ pathway))

  go_res_mat = gse_res_ind_filt %>%
    select(compare, pathway, NES) %>%
    pivot_wider(names_from = "compare", values_from = "NES", values_fill = 0) %>%
    column_to_rownames("pathway") %>% as.matrix()

  hc_row = hclust(dist(go_res_mat), method="ward.D")
  hc_row_order = hc_row$labels[hc_row$order]


  gse_res_ind_filt = gse_res_ind_filt %>%
    mutate(leadingEdge_len = map_int(leadingEdge, length))

  gse_res_ind_filt = gse_res_ind_filt %>%
    mutate(pathway = factor(pathway, levels=hc_row_order)) %>%
    mutate(compare = factor(compare, levels=set_order)) %>%
    complete(pathway, compare, fill=list(NES=NA, size=NA, leadingEdge_len = NA))

  size_limits = seq(round(min(gse_res_ind_filt$leadingEdge_len, na.rm=T), -1), round(max(gse_res_ind_filt$leadingEdge_len, na.rm=T), -1), by=20)
  gg = ggplot(gse_res_ind_filt, aes(compare, pathway)) +
    geom_point(shape=21, aes(size=leadingEdge_len, fill=NES, color=is_sig)) +
    scale_color_manual(values=c(`TRUE`="black", `FALSE`=NA)) +
    scale_fill_gradient2(low=scales::muted("blue"), high=scales::muted("red")) +
    scale_size_continuous(range=c(.5, 3), breaks=size_limits) +
    labs(x = "", y ="") +
    theme_bw() +
    theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1, size=6),
          axis.text.y = element_text(hjust=1, size=6),
          axis.ticks = element_line(size=.25),
          panel.border = element_blank()) +
    coord_fixed()

  ncols = ncol(go_res_mat)
  nrows = nrow(go_res_mat)

  scale_factor = .1
  width = ncols * scale_factor
  height = nrows * scale_factor
  gg = egg::set_panel_size(gg, width = unit(width, "in"), height = unit(height, "in"))
  plot(gg)
  save_plot(file.path(out_dir, sprintf("gse_terms_by_compare.pdf")), gg, base_height = height + 4, base_width = width + 4)
```
