---
title: "Song vs surround DEG, excitatory (hybrid labels)"
format: html
editor: source
---

Parallels `song-surround_deg_glut.qmd`, but on the current (hybrid division-based) labels from
`snrna/naming/hybrid_division_naming.qmd` instead of the raw `cluster` identity.

Three things change beyond the names themselves, and each is noted where it happens:

- the two clusters that notebook excludes (`Glut-Arco-4`, `Glut-Nido-3`) are gone from the object;
- `Glut-NC-1` and `Glut-Nido-1` are **one** cluster under the hybrid scheme, `Glut-DACH2-1`, so the
  surround term of the HVCra comparison is that merged cluster, not `Glut-NC-1` alone;
- the pseudobulk replicate is the bird (`individual` from souporcell), not `assignment`.

```{r}
library(tidyverse)
library(Seurat)
library(scCustomize)
library(qs2)
library(UpSetR)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ComplexHeatmap)
library(cowplot)
library(ggpubr)
library(here)
suppressPackageStartupMessages(library(DESeq2))

## DESeq2 drags in SummarizedExperiment and matrixStats, which mask things this notebook depends on:
## SummarizedExperiment::Assays returns an S4 object rather than Seurat's character vector, and
## matrixStats::count shadows dplyr's. Loaded here at the top, once, rather than part-way down the
## document, so the masking is the same for every chunk instead of changing halfway -- and the three
## that matter are pinned back explicitly. `Assays` is qualified at its call site instead, since it
## is Seurat's generic rather than a plain function.
select = dplyr::select
count  = dplyr::count
filter = dplyr::filter

source(here::here("config/paths.R"))
```

```{r}
script_name = "song-surround_deg_glut_hybrid"
data_obj = here::here("snrna/naming/hybrid_division_naming", "obj_hybrid_labels.qs2")
out_dir = here::here("snrna/deg", script_name)
dir.create(out_dir, recursive = T)
res_to_use = "celltype_hybrid"
```

```{r}
obj = qs_read(data_obj, nthreads = 1)

## Souporcell doublets. The preprocessing notebook already dropped them, so this is a no-op on the
## current object -- kept because it is the guard that made the old version correct and costs
## nothing if an upstream re-run ever stops filtering.
cells = Cells(obj)[!grepl("/", obj$assignment)]
obj = subset(obj, cells=cells)

## Excluded clusters: celltype_hybrid is NA for exactly Glut-Arco-4 and Glut-Nido-3.
obj = subset(obj, cells = Cells(obj)[!is.na(obj@meta.data[[res_to_use]])])
obj@meta.data[[res_to_use]] = droplevels(factor(obj@meta.data[[res_to_use]]))
```

## Individual identity

The pseudobulk replicate has to be the bird. `assignment` is souporcell's *per-library* cluster
index (0/1/2), recycled independently in each of the six libraries, so it names three arbitrary
labels six times over rather than three birds. The old version of this notebook grouped on it,
which was survivable only as long as every cluster sat inside a single library. Under the hybrid
scheme that is no longer true -- `Glut-DACH2-1` merges `Glut-NC-1` with `Glut-Nido-1` -- so a
pseudobulk column keyed on `assignment` would pool cells from different birds.

`snrna/clustering/snrna_souporcell_clustering.qmd` resolves the indices into six birds by matching
clusters across libraries on genotype; its per-cell table is read here rather than re-derived, and
joined on cell name, which is exact (asserted below).

```{r}
soup = read_csv(here::here("snrna/clustering/snrna_souporcell_clustering",
                           "souporcell_assignments.csv"),
                show_col_types = FALSE)

stopifnot(
  "souporcell assignments do not cover every cell -- check the cell-name suffix convention" =
    all(Cells(obj) %in% soup$cell)
)

md = tibble(cell = Cells(obj)) %>%
  left_join(soup %>% select(cell, sample_id, individual, batch), by = "cell") %>%
  mutate(individual = factor(individual, levels = sprintf("bird%s", 1:6)),
         soup_batch = factor(sprintf("batch%s", batch)),
         soup_library = sample_id) %>%
  select(cell, individual, soup_batch, soup_library) %>%
  column_to_rownames("cell")

## A silent NA here would become a seventh pseudobulk "bird", so assert rather than assume.
stopifnot("unassigned individuals after the join" = !any(is.na(md$individual)))
obj = AddMetaData(obj, md)
```

```{r}
table(obj@meta.data[[res_to_use]], obj$position)
```

```{r}
table(obj@meta.data[[res_to_use]], obj$individual)
```

## Song / surround pairs

Element 1 of each pair is the song-region cluster, the rest its nidopallial or arcopallial
surround. Renamed one-for-one from the `cluster`-based version:

| old song | old surround | hybrid song | hybrid surround |
|---|---|---|---|
| `Glut-RA` | `Glut-Arco-1` | `Glut-CACNA1H-RA` | `Glut-CACNA1H-1` |
| `Glut-HVC-1` | `Glut-NC-1` | `Glut-DACH2-HVCra` | `Glut-DACH2-1` |
| `Glut-HVC-2` | `Glut-NC-4` | `Glut-DACH2-HVCx` | `Glut-DACH2-4` |
| `Glut-LMAN-2` | `Glut-NR-4`, `Glut-NR-2` | `Glut-DACH2-LMANco` | `Glut-DACH2-8`, `Glut-DACH2-6` |
| `Glut-LMAN-1` | `Glut-NR-1`, `Glut-NR-3` | `Glut-DACH2-LMANsh` | `Glut-DACH2-5`, `Glut-DACH2-7` |

`Glut-DACH2-1` is `Glut-NC-1` **and** `Glut-Nido-1` merged, so the HVCra surround is a larger,
slightly different set of cells than it was -- the one place where the comparison itself, not just
its name, has moved.

```{r}
pairs = list(`Glut-CACNA1H-RA`   = c("Glut-CACNA1H-RA", "Glut-CACNA1H-1"),
             `Glut-DACH2-HVCra`  = c("Glut-DACH2-HVCra", "Glut-DACH2-1"),
             `Glut-DACH2-HVCx`   = c("Glut-DACH2-HVCx", "Glut-DACH2-4"),
             `Glut-DACH2-LMANco` = c("Glut-DACH2-LMANco", c("Glut-DACH2-8", "Glut-DACH2-6")),
             `Glut-DACH2-LMANsh` = c("Glut-DACH2-LMANsh", c("Glut-DACH2-5", "Glut-DACH2-7")))

stopifnot("a pair names a label that is not in celltype_hybrid" =
            all(unlist(pairs) %in% levels(obj@meta.data[[res_to_use]])))

## The song clusters, in the order every figure below uses.
song_cts = names(pairs)[c(2, 3, 4, 5, 1)]  # HVCra, HVCx, LMANco, LMANsh, RA
```

## Genes eligible to be displayed

Every panel below that picks genes by effect size -- the scatter labels, the heatmap marks, the
volcano labels, the dotplots -- draws from the same eligibility rule, defined once here.

Excluded are the unnamed NCBI gene models (`LOC*`) and the mitochondrial features, which in this
annotation are `AZ255-gr01`, `AZ255-gr02` and the 45 `nbis-*` entries. There are no standard `MT-`,
`ND*`, `COX*` or `CYTB` symbols in the object -- those names are what the mitochondrial genome is
called here. Mitochondrial transcripts are abundant and move together, so left in they crowd out
named genes whenever a panel is chosen by fold change.

This governs only what is eligible to be *labelled or plotted as a candidate*. Nothing is removed
from the DE results, the correlation matrix or any of the heatmap matrices.

```{r}
## Anchored. The previous filters were written three different ways -- "^LOC", "LOC|nbis" and a bare
## "LOC" -- and the two unanchored ones also dropped BLOC1S2, BLOC1S4, BLOC1S5, BLOC1S6, NPLOC4,
## ELOC and CLOCK, none of which is an unnamed model. Anchoring to the start fixes that and makes
## the three sites agree.
display_exclude = "^LOC|^AZ255|^nbis"
is_displayable = function(gene) !grepl(display_exclude, gene)

## Significance threshold, shared by the song/non-song contrast and the core-set logic below so the
## two cannot drift apart. Same 0.1 the rest of the notebook uses.
sig_thresh = 0.1
```

# Markers

```{r}
assay_to_use = "SCT"
markers_fname = file.path(out_dir, sprintf("marker_genes_%s_%s.rds", res_to_use, assay_to_use))

obj_agg = AggregateExpression(obj, assay=assay_to_use,
                              group.by=c(res_to_use, "individual"), return.seurat = T)
stopifnot("AggregateExpression did not carry the grouping variable through" =
            res_to_use %in% colnames(obj_agg@meta.data))
Idents(obj_agg) = obj_agg@meta.data[[res_to_use]]

## Cached on the same principle as the embeddings in combined_all_umap_glut_hybrid.R: the DESeq2
## pseudobulk fits are the expensive step here and everything below them is plotting. Set TRUE by
## hand (or delete the rds) after changing the pairs, the object, the assay or the pseudobulk
## grouping -- nothing detects that for you.
redo_markers = !file.exists(markers_fname)
cat(if (redo_markers) "no cached markers -- fitting\n" else
    sprintf("reusing cached markers: %s\n", markers_fname))

if (redo_markers) {

  ## As in the `cluster`-based version, `ident.2` takes the first surround cluster only -- the pair
  ## vectors carry the others for the dotplots and violins further down, which do use all of them.
  markers_agg = map(pairs, function(pair_cur) {
    print(pair_cur)
    res = FindMarkers(obj_agg, assay=assay_to_use,
                      ident.1=pair_cur[1], ident.2=pair_cur[2],
                      test.use="DESeq2", only.pos = F)
    res = res %>% rownames_to_column(var="gene")
    res
  }) %>% set_names(names(pairs)) %>% bind_rows(.id="compare")

  saveRDS(markers_agg, markers_fname)
} else {
  markers_agg = readRDS(markers_fname)
}

markers_unpaired = markers_agg
```

## The five contrasts, paired

The fits above treat the pseudobulk columns as independent samples. They are not: each bird
contributes both a song and a surround column, so bird identity is a nuisance term that can be
modelled instead of left in the residual. Same argument as the pooled song/non-song contrast below,
applied per pair -- `~ individual + grp`, raw `RNA` counts, DESeq2 directly.

Each pair reduces to 6 samples, 3 birds a side: HVC and RA draw on birds 1-3, LMAN on birds 4-6. A
bird needs at least `min_cells_pb` cells on *both* sides to be usable, since one side alone cannot
be paired; none are lost at 20, but the rule is enforced rather than assumed.

Each song type is tested against its **one** corresponding non-song type -- five contrasts, five
surrounds -- which is what `ident.2 = pair_cur[2]` does in the unpaired fits above and what the
paired fits do here. The extra clusters in the `pairs` vectors (`Glut-DACH2-6`, `Glut-DACH2-7`) are
not part of any contrast: they are there for `cts = unlist(pairs)`, which drives the dotplot and
violin columns further down, where the neighbouring non-song clusters are worth showing alongside
the tested pair.

**One thing changes relative to the unpaired fits, beyond the pairing itself**: raw counts rather
than SCT, for the reason given in the pooled contrast -- DESeq2's model and size factors expect
summed UMIs.

```{r}

paired_pairs_fname = file.path(out_dir, "song_surround_paired_deseq2.rds")
min_cells_pb = 20

paired_contrast = function(song_ct, surround_cts) {
  labs = c(song_ct, surround_cts)
  o = subset(obj, cells = Cells(obj)[obj@meta.data[[res_to_use]] %in% labs])
  o$grp = if_else(o@meta.data[[res_to_use]] == song_ct, "song", "surround")

  ## BPCells-backed on disk; pulled in as dgCMatrix so the indicator multiply below is ordinary
  ## Matrix arithmetic. Built by hand rather than via AggregateExpression so the sample names
  ## round-trip exactly into the design matrix.
  cts = as(GetAssayData(o, assay = "RNA", layer = "counts"), "dgCMatrix")
  g = factor(paste(o$grp, o$individual, sep = "__"))
  n_cells = as.integer(table(g))
  pb = as.matrix(cts %*% Matrix::sparse.model.matrix(~ 0 + g))
  colnames(pb) = levels(g)

  cd = tibble(sample = colnames(pb), n_cells = n_cells) %>%
    separate(sample, into = c("grp", "individual"), sep = "__", remove = FALSE)

  ## A bird is usable only if both of its sides clear the cell floor.
  ok = cd %>% filter(n_cells >= min_cells_pb) %>%
    dplyr::count(individual) %>% filter(n == 2) %>% pull(individual)
  dropped = setdiff(unique(cd$individual), ok)
  if (length(dropped)) cat(sprintf("  %s: dropping %s (under %s cells on one side)\n",
                                   song_ct, paste(dropped, collapse = ", "), min_cells_pb))

  cd = cd %>% filter(individual %in% ok) %>%
    mutate(grp = factor(grp, levels = c("surround", "song")), individual = factor(individual))
  pb = pb[, cd$sample, drop = FALSE]
  cd = cd %>% column_to_rownames("sample")

  stopifnot("not enough paired birds for this contrast" =
              nrow(cd) >= 4 && n_distinct(cd$individual) >= 2)

  dds = DESeqDataSetFromMatrix(round(pb), colData = cd, design = ~ individual + grp)
  keep = rowSums(counts(dds) >= 10) >= (ncol(dds) / 2)
  dds = dds[keep, ]
  dds = DESeq(dds, quiet = TRUE)

  r = results(dds, name = "grp_song_vs_surround") %>% as.data.frame() %>%
    rownames_to_column("gene") %>% as_tibble()
  cat(sprintf("  %-20s birds=%s samples=%s tested=%s sig=%s\n", song_ct,
              n_distinct(cd$individual), nrow(cd), nrow(r),
              sum(!is.na(r$padj) & r$padj < sig_thresh)))
  r
}

redo_paired_pairs = !file.exists(paired_pairs_fname)
cat(if (redo_paired_pairs) "no cached paired per-pair fits -- fitting\n" else
    sprintf("reusing cached paired per-pair fits: %s\n", paired_pairs_fname))

if (redo_paired_pairs) {
  markers_paired = imap(pairs, function(p, nm) paired_contrast(p[1], p[2])) %>%
    bind_rows(.id = "compare")
  saveRDS(markers_paired, paired_pairs_fname)
} else {
  markers_paired = readRDS(paired_pairs_fname)
}

## Mapped into the column names the rest of the notebook expects, so every figure downstream reads
## the paired fits without further change.
markers_paired_std = markers_paired %>%
  transmute(compare, gene,
            p_val = pvalue,
            avg_log2FC = log2FoldChange,
            p_val_adj = padj,
            baseMean, lfcSE)
```

```{r}
## Paired against unpaired, per contrast. The pairing should buy power without moving the direction
## of the effect; where the fold-change correlation is weak, the two fits are not saying the same
## thing and that is worth knowing before either is trusted.
cmp_pairs = markers_paired_std %>%
  select(compare, gene, lfc_p = avg_log2FC, padj_p = p_val_adj) %>%
  inner_join(markers_unpaired %>% select(compare, gene, lfc_u = avg_log2FC, padj_u = p_val_adj),
             by = c("compare", "gene"))

cmp_pairs %>%
  group_by(compare) %>%
  summarise(tested_both = n(),
            sig_paired = sum(!is.na(padj_p) & padj_p < sig_thresh),
            sig_unpaired = sum(!is.na(padj_u) & padj_u < sig_thresh),
            sig_both = sum(!is.na(padj_p) & padj_p < sig_thresh &
                             !is.na(padj_u) & padj_u < sig_thresh),
            rho_log2FC = cor(lfc_p, lfc_u, method = "spearman", use = "complete"),
            .groups = "drop") %>%
  as.data.frame() %>% print()

write_csv(markers_paired_std %>% arrange(compare, p_val_adj),
          file.path(out_dir, "song_surround_paired_deg.csv"))
```

```{r}
## The switch. TRUE makes the paired fits the result set every figure below is built from; FALSE
## restores the unpaired FindMarkers fits. Nothing else in the notebook needs editing either way --
## the schemas match -- so this is the one line to change to see the whole document both ways.
use_paired_pairs = TRUE

markers_agg = if (use_paired_pairs) markers_paired_std else markers_unpaired
cat(sprintf("downstream figures use the %s fits: %s significant of %s rows\n",
            if (use_paired_pairs) "PAIRED" else "unpaired",
            sum(!is.na(markers_agg$p_val_adj) & markers_agg$p_val_adj < sig_thresh),
            nrow(markers_agg)))

## What the display-eligibility rule removes from whichever result set is in play.
local({
  g = unique(markers_agg$gene)
  cat(sprintf("display candidates: %s of %s genes; excluded %s LOC, %s AZ255, %s nbis\n",
              sum(is_displayable(g)), length(g),
              sum(grepl("^LOC", g)), sum(grepl("^AZ255", g)), sum(grepl("^nbis", g))))
})
```

## Scatters

```{r}
markers_agg_df = markers_agg %>%
  filter(p_val_adj < .1) %>%
  select(avg_log2FC, compare, gene) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC")
markers_agg_mat = markers_agg_df %>% column_to_rownames("gene") %>%
  as.matrix()

cor(markers_agg_mat, use = "complete")
```

```{r}

ct_pairs = combn(names(pairs), m = 2, simplify = F)

ngenes = 50

markers_agg_df_cur = markers_agg_df %>%
    filter(is_displayable(gene))

walk(ct_pairs, function(x) {
  ct1 = x[1]
  ct2 = x[2]

markers1 = markers_agg_df_cur %>%
  top_n(ngenes, !!as.symbol(ct1)) %>%
  select(gene) %>% unlist()

markers2 = markers_agg_df_cur %>%
  top_n(ngenes, !!as.symbol(ct2)) %>%
  select(gene) %>% unlist()

top_markers = intersect(markers1, markers2)

markers3 = markers_agg_df_cur %>%
  top_n(-1 * ngenes, !!as.symbol(ct1)) %>%
  select(gene) %>% unlist()

markers4 = markers_agg_df_cur %>%
  top_n(-1 *ngenes, !!as.symbol(ct2)) %>%
  select(gene) %>% unlist()

bottom_markers = intersect(markers3, markers4)
to_plot_markers = c(bottom_markers, top_markers)


markers_agg_df_cur = markers_agg_df %>% select(gene, !!as.symbol(ct1), !!as.symbol(ct2)) %>%
  mutate(to_plot = if_else(gene %in% to_plot_markers, gene, NA))
gg = ggplot(markers_agg_df_cur, aes(!!as.symbol(ct1), !!as.symbol(ct2))) +
  geom_vline(xintercept=0, linetype = 2) +
  geom_hline(yintercept=0, linetype = 2) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label=to_plot), na.rm = T, force_pull = 2) +
  theme_bw() +
  coord_equal()
print(gg)
save_plot(file.path(out_dir, str_interp("scatter_${ct1}_${ct2}.pdf")), gg)
})
```
## Correlation
Spearman correlation of the song-vs-surround log2 fold changes between every pair of song clusters:
does a gene that moves in HVCra move the same way in RA? The gene set is everything significant in
**at least one** pair, and each gene's fold change is then read from all five comparisons whether or
not it was significant there.

That last point is the whole design and the version of this chunk inherited from
`song-surround_deg_glut.qmd` had it backwards, alongside two other faults -- it correlated with
Pearson rather than Spearman, passed the gene matrix rather than the correlation matrix to
`Heatmap()`, and assigned the result without ever printing or saving it, so nothing was produced at
all.

The gene-set point: filtering to `p_val_adj < .1` *before* pivoting wide leaves NA wherever a gene
failed significance, and `cor(use = "complete")` then silently keeps only the genes significant in
**all five** pairs. Genes that clear the threshold everywhere are by construction the consistent
ones, so that set inflates the correlations it is used to measure. Taking the union and then reading
the unfiltered fold changes keeps NA meaning one thing only -- `FindMarkers` did not return the gene
for that comparison, having filtered it for low expression -- and those few rows are dropped.

Spearman rather than Pearson because the question is whether the pairs *rank* genes alike; the
log2FC distributions differ in spread between comparisons (HVCra and RA return far more significant
genes than the other three), which a Pearson coefficient would partly be measuring.

```{r}
sig_genes = markers_agg %>% filter(p_val_adj < .1) %>% distinct(gene) %>% pull(gene)

cor_input = markers_agg %>%
  filter(gene %in% sig_genes) %>%
  select(gene, compare, avg_log2FC) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC") %>%
  column_to_rownames("gene") %>%
  as.matrix()
cor_input = cor_input[, song_cts]

n_union = nrow(cor_input)
cor_input = na.omit(cor_input)
cat(sprintf("genes significant in >=1 pair: %s; returned by all five comparisons: %s\n",
            n_union, nrow(cor_input)))

markers_agg_cor = cor(cor_input, method = "spearman")
print(round(markers_agg_cor, 3))
write.csv(round(markers_agg_cor, 4), file.path(out_dir, "correlation_spearman.csv"))
```

The full matrix, with the diagonal blanked. The diagonal is 1 by construction and, on a scale that
has to reach 1 to hold it, every off-diagonal coefficient is pushed into the bottom of the ramp;
blanking it lets the scale be spent on the comparisons that carry information. Both triangles are
drawn, so each coefficient appears twice, symmetric about the empty diagonal.

```{r}
## NA, not 0 -- na_col paints these white, and a real zero would also be white, so they must not be
## made the same value. The diagonal is blank because it is uninformative, not because it is small.
cor_tri = markers_agg_cor
diag(cor_tri) = NA

## 0-1 greyscale. Anything at or below zero therefore reads as white and is not distinguishable
## from a coefficient of zero -- or, here, from the blanked diagonal. The printed matrix above is
## the check on that, and the range is reported here so a negative value cannot pass unnoticed.
off_diag = markers_agg_cor[row(markers_agg_cor) != col(markers_agg_cor)]
cat(sprintf("off-diagonal Spearman rho: min %.3f, max %.3f\n", min(off_diag), max(off_diag)))

cor_cols = circlize::colorRamp2(breaks = c(0, 1), colors = c("white", "black"))

cell_size = 0.2
hm = Heatmap(cor_tri,
             name = "Spearman rho",
             col = cor_cols,
             na_col = "white",
             cluster_rows = FALSE, cluster_columns = FALSE,
             row_names_gp = gpar(fontsize = 6),
             column_names_gp = gpar(fontsize = 6),
             rect_gp = gpar(col = "grey80", lwd = 0.5),
             width = unit(cell_size * ncol(cor_tri), "inch"),
             height = unit(cell_size * nrow(cor_tri), "inch"))

pdf(file.path(out_dir, "correlation_spearman.pdf"),
    width = cell_size * ncol(cor_tri) + 2.6,
    height = cell_size * nrow(cor_tri) + 1.8)
print(hm)
dev.off()
print(hm)
```

## Heatmap

The heatmaps below are drawn gene-per-row: a tall, half-inch-wide strip, which is the right shape
for reading down a gene list and the wrong shape for a figure that has to sit across a column. Each
one therefore also gets a horizontal counterpart -- the same matrix transposed, so the comparisons
are rows and the genes are columns, the gene dendrogram moves to the top and the height and width
budgets swap with it. Same data, same clustering, same colour scale; only the orientation changes.

In the transposed version the comparison names become row labels, read horizontally instead of
rotated, and the gene marks (where there are any) hang off the bottom rather than the right.

```{r}
save_heatmap_horizontal = function(mat, hc, fname, col,
                                   height = 15, width = 0.5,
                                   mark_genes = NULL,
                                   show_gene_names = FALSE,
                                   name = "avg_log2FC") {
  mat_t = t(mat)

  ## `height`/`width` are the *vertical* figure's budget, passed through unchanged so the two
  ## orientations stay tied to one set of numbers: the gene axis keeps `height` inches and the
  ## comparison axis keeps `width`, they simply change places.
  ##
  ## The marks are taken as gene *names* and resolved against the matrix here, so the positions and
  ## the labels are derived from one lookup and cannot fall out of step -- a gene that is not in the
  ## matrix drops from both. (The vertical versions above pass a na.omit()'d index alongside the
  ## unfiltered label vector, which is only correct while every gene matches.)
  ha = NULL
  if (length(mark_genes) > 0) {
    idx = match(mark_genes, colnames(mat_t))
    keep = !is.na(idx)
    if (any(keep)) {
      ha = HeatmapAnnotation(
        foo = anno_mark(at = idx[keep], labels = mark_genes[keep], side = "bottom",
                        labels_gp = gpar(fontsize = 5), lines_gp = gpar(lwd = .5)),
        which = "column"
      )
    }
  }

  ## Named, so the legend reads `avg_log2FC` rather than the auto-generated `matrix_N` the
  ## unnamed vertical versions above fall back to.
  hm = Heatmap(mat_t, col = col, name = name,
               cluster_rows = FALSE,
               cluster_columns = hc,
               show_row_dend = FALSE,
               show_column_dend = TRUE,
               column_dend_gp = gpar(lwd = .5),
               show_column_names = show_gene_names,
               row_names_gp = gpar(fontsize = 6),
               bottom_annotation = ha,
               height = unit(width, "inch"),
               width = unit(height, "inch"))

  pdf(fname, width = height + 4, height = width + 4)
  print(hm)
  dev.off()
  invisible(hm)
}
```

```{r}
ngenes = 50
markers_agg_select = markers_agg %>%
  group_by(compare) %>%
  top_n(ngenes, abs(avg_log2FC))


markers_agg_mat = markers_agg %>%
  filter(gene %in% markers_agg_select$gene) %>%
  select(avg_log2FC, compare, gene) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC") %>%
  column_to_rownames("gene") %>%
  as.matrix()

## A gene in the top 50 of one comparison need not have been returned by FindMarkers for all five,
## and dist() propagates the resulting NA into every distance, which hclust rejects outright. The
## `heatmap_sig` chunk below already drops incomplete rows for the same reason; do it here too
## rather than leave the chunk dependent on the gene sets happening to line up. 17 of 200 rows go.
markers_agg_mat = na.omit(markers_agg_mat)

height = 15
width = 0.5

max_val = 4
cols_cur = circlize::colorRamp2(breaks=c(-1*max_val, 0, max_val), colors = c(scales::muted("blue"), "white", scales::muted("red")))

hc = hclust(dist(markers_agg_mat), method = "ward.D")
hm = Heatmap(markers_agg_mat, col = cols_cur,
             cluster_rows = hc,
             show_column_dend = F,
             height = unit(height, "inch"),
             width = unit(width, "inch"))
pdf(file.path(out_dir, "heatmap.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

height = 2
width = 0.5

hm = Heatmap(markers_agg_mat, col = cols_cur,
             cluster_rows = hc,
             show_column_dend = F,
             show_row_names = F,
             height = unit(height, "inch"),
             width = unit(width, "inch"))
pdf(file.path(out_dir, "heatmap_no_genes.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

## Literal 15/0.5 rather than `height`/`width`, which were just reassigned for the no-genes variant.
save_heatmap_horizontal(markers_agg_mat, hc, file.path(out_dir, "heatmap_horizontal.pdf"),
                        col = cols_cur, height = 15, width = 0.5)
```
```{r}

markers_agg_select = markers_agg %>%
  filter(p_val_adj < 0.1) %>%
  filter(abs(avg_log2FC) >= 1)


markers_agg_mat = markers_agg %>%
  filter(gene %in% markers_agg_select$gene) %>%
  select(avg_log2FC, compare, gene) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC") %>%
  column_to_rownames("gene") %>%
  as.matrix()

markers_agg_mat = markers_agg_mat[, song_cts]
markers_agg_mat = na.omit(markers_agg_mat)

hc = hclust(dist(markers_agg_mat), method = "ward.D")

height = 15
width = 0.5

max_val = 4
cols_cur = circlize::colorRamp2(breaks=c(-1*max_val, 0, max_val), colors = c(scales::muted("blue"), "white", scales::muted("red")))

hc = hclust(dist(markers_agg_mat), method = "ward.D")
hm = Heatmap(markers_agg_mat, col = cols_cur,
             cluster_rows = hc,
             cluster_columns = F,
             show_column_dend = F,
             height = unit(height, "inch"),
             width = unit(width, "inch"))
pdf(file.path(out_dir, "heatmap_sig.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()



height = 2
width = 0.5

hm = Heatmap(markers_agg_mat, col = cols_cur,
             cluster_rows = hc,
             cluster_columns = F,
             show_column_dend = F,
             show_row_names = F,
             height = unit(height, "inch"),
             width = unit(width, "inch"))
pdf(file.path(out_dir, "heatmap_sig_no_genes.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

save_heatmap_horizontal(markers_agg_mat, hc, file.path(out_dir, "heatmap_sig_horizontal.pdf"),
                        col = cols_cur, height = 15, width = 0.5)


dist_mat = as.matrix(dist(markers_agg_mat))
hc_cut = dynamicTreeCut::cutreeDynamic(hc, cutHeight = 100,  method="hybrid", distM = dist_mat, minClusterSize = 20)
genes_split = split(hc$labels, hc_cut)
table(hc_cut)
```

```{r}

markers_agg = markers_agg %>%
  mutate(avg_log2FC_sign = sign(avg_log2FC))

markers_agg_select = markers_agg %>%
  filter(p_val_adj < 0.1) %>%
  filter(abs(avg_log2FC) >= 1)

markers_agg_select_label = markers_agg_select %>%
  group_by(compare, avg_log2FC_sign ) %>%
  filter(is_displayable(gene)) %>%
  top_n(5, abs(avg_log2FC)) %>%
  ungroup() %>%
  distinct(gene)

markers_agg_mat = markers_agg %>%
  filter(gene %in% markers_agg_select$gene) %>%
  select(avg_log2FC, compare, gene) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC", values_fill = 0) %>%
  column_to_rownames("gene") %>%
  as.matrix()

markers_agg_mat = markers_agg_mat[, song_cts]

hc = hclust(dist(markers_agg_mat), method = "ward.D2")

height = 3
width = 0.5

max_val = 4
cols_cur = circlize::colorRamp2(breaks=c(-1*max_val, 0, max_val), colors = c(scales::muted("blue"), "white", scales::muted("red")))

markers_agg_select_label_ind = na.omit(match(markers_agg_select_label$gene,
                                             rownames(markers_agg_mat)))

ha = rowAnnotation(foo = anno_mark(at = markers_agg_select_label_ind,
                                   labels = markers_agg_select_label$gene,
                                   labels_gp = gpar(fontsize=5),
                                   lines_gp=gpar(lwd=.5)) )

hm = Heatmap(markers_agg_mat, col = cols_cur,

             show_row_names = F,
             cluster_rows = hc,
             cluster_columns = F,
             right_annotation = ha,
             show_column_dend = F,
             row_dend_gp = gpar(lwd=.5),
             height = unit(height, "inch"),
             width = unit(width, "inch"))
print(hm)
pdf(file.path(out_dir, "heatmap_sig_label.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

save_heatmap_horizontal(markers_agg_mat, hc, file.path(out_dir, "heatmap_sig_label_horizontal.pdf"),
                        col = cols_cur, height = 3, width = 0.5,
                        mark_genes = markers_agg_select_label$gene)



dist_mat = as.matrix(dist(markers_agg_mat))
hc_cut = dynamicTreeCut::cutreeDynamic(hc, method="hybrid", distM = dist_mat, minClusterSize = 20)
table(hc_cut)

hc_cut_df = data.frame(gene = rownames(markers_agg_mat), hc_cut = hc_cut)
markers_agg_select_label = markers_agg_select %>%
  left_join(hc_cut_df) %>%
  group_by(hc_cut) %>%
  filter(is_displayable(gene)) %>%
  top_n(4, abs(avg_log2FC)) %>%
  ungroup() %>%
  distinct(gene)

height = 3
width = 0.5

max_val = 4
cols_cur = circlize::colorRamp2(breaks=c(-1*max_val, 0, max_val), colors = c(scales::muted("blue"), "white", scales::muted("red")))

markers_agg_select_label_ind = na.omit(match(markers_agg_select_label$gene,
                                             rownames(markers_agg_mat)))

ha = rowAnnotation(foo = anno_mark(at = markers_agg_select_label_ind,
                                   labels = markers_agg_select_label$gene,
                                   labels_gp = gpar(fontsize=5),
                                   lines_gp=gpar(lwd=.5)) )

hm = Heatmap(markers_agg_mat, col = cols_cur,
             show_row_names = F,
             cluster_rows = hc,
             cluster_columns = F,
             right_annotation = ha,
             show_column_dend = F,
             row_dend_gp = gpar(lwd=.5),
             height = unit(height, "inch"),
             width = unit(width, "inch"))
print(hm)
pdf(file.path(out_dir, "heatmap_sig_label_cluster.pdf"), width = width + 4, height = height + 4)
print(hm)
dev.off()

save_heatmap_horizontal(markers_agg_mat, hc,
                        file.path(out_dir, "heatmap_sig_label_cluster_horizontal.pdf"),
                        col = cols_cur, height = 3, width = 0.5,
                        mark_genes = markers_agg_select_label$gene)
```

### Transcription factors only

The same significant-gene heatmap restricted to transcription factors. In a heatmap of a thousand
genes the TFs are there but unfindable, and they are the ones that carry a claim about regulatory
identity rather than about effector genes -- so they get their own panel, short enough to label
every row.

The TF definition is the bundled Lambert et al. list, `data/lambert_tfs/tfs.csv`, filtered to
`DBD != "Unknown"` exactly as `snrna/trees/celltypes_hclust_glut.qmd` and the ArchR notebooks do.
That drops the ~1000 entries with no assigned DNA-binding domain, which are the least certain calls
in that table. Matching those notebooks matters more than the particular cut: a TF here should mean
the same thing it means elsewhere in the project.

Selection is the same as `heatmap_sig` -- significant at `p_val_adj < sig_thresh` and
`|avg_log2FC| >= 1` in at least one contrast.

Where it differs is what happens to a TF that was not tested in every contrast. `heatmap_sig` drops
those rows; here they are kept and the untested cells drawn as zero, following `heatmap_sig_label`.
Dropping them would lose about half the DE TFs, which defeats a figure whose whole purpose is to
show the TFs. The cost is that a filled cell reads as "no change" when it really means "too lowly
expressed in that pair to test" -- the two are not the same, the proportion filled is reported
below, and the CSV keeps the real values with their NAs intact.

```{r}
tfs = read_delim(here::here("data/lambert_tfs/tfs.csv"), show_col_types = FALSE) %>%
  filter(DBD != "Unknown")

tf_de_genes = markers_agg %>%
  filter(!is.na(p_val_adj), p_val_adj < sig_thresh, abs(avg_log2FC) >= 1, gene %in% tfs$Name) %>%
  distinct(gene) %>% pull(gene)

## values_fill = 0 rather than dropping incomplete rows. Half the DE TFs are missing a value in at
## least one contrast -- DESeq2's count prefilter removed them there, which for a TF usually means
## it is too lowly expressed in that pair to test -- and `na.omit` would have silently halved the
## panel, which defeats the point of a TF-specific figure.
##
## The filled cells are drawn as zero, i.e. as "no change", which is an assumption and not a
## measurement: a gene too lowly expressed to test is not the same as a gene that did not move. The
## count is reported below and the accompanying CSV keeps the real values, NAs included, so nothing
## downstream inherits the imputation.
tf_mat = markers_agg %>%
  filter(gene %in% tf_de_genes) %>%
  select(gene, compare, avg_log2FC) %>%
  pivot_wider(names_from = "compare", values_from = "avg_log2FC", values_fill = 0) %>%
  column_to_rownames("gene") %>%
  as.matrix()
tf_mat = tf_mat[, song_cts, drop = FALSE]

n_measured = markers_agg %>%
  filter(gene %in% tf_de_genes, compare %in% song_cts, !is.na(avg_log2FC)) %>% nrow()
n_complete = markers_agg %>%
  filter(gene %in% tf_de_genes, compare %in% song_cts, !is.na(avg_log2FC)) %>%
  dplyr::count(gene) %>% filter(n == length(song_cts)) %>% nrow()

cat(sprintf("TF list: %s with a known DBD; DE TFs: %s (%s complete across all five)\n",
            nrow(tfs), length(tf_de_genes), n_complete))
cat(sprintf("heatmap cells: %s measured, %s filled at 0 (%.1f%%)\n",
            n_measured, length(tf_de_genes) * length(song_cts) - n_measured,
            100 * (length(tf_de_genes) * length(song_cts) - n_measured) /
              (length(tf_de_genes) * length(song_cts))))

## Written out with the fold changes, since the point of this panel is usually to pick TFs to follow
## up rather than to read the colours.
markers_agg %>%
  filter(gene %in% rownames(tf_mat)) %>%
  left_join(tfs %>% select(gene = Name, DBD), by = "gene") %>%
  arrange(gene, compare) %>%
  write_csv(file.path(out_dir, "heatmap_tf_de.csv"))
```

```{r}
if (nrow(tf_mat) > 1) {
  max_val = 4
  cols_cur = circlize::colorRamp2(breaks = c(-max_val, 0, max_val),
                                  colors = c(scales::muted("blue"), "white", scales::muted("red")))

  hc_tf = hclust(dist(tf_mat), method = "ward.D")

  ## Square cells: one size for both dimensions, so the panel is sized from the row and column
  ## counts rather than to a fixed height. 0.085 in is about 6 pt, which is the floor for 5 pt row
  ## labels not to collide -- every row is labelled here, which is the reason this variant exists.
  cell_size = 0.085
  height = cell_size * nrow(tf_mat)
  width  = cell_size * ncol(tf_mat)

  hm = Heatmap(tf_mat, name = "avg_log2FC", col = cols_cur,
               cluster_rows = hc_tf,
               cluster_columns = FALSE,
               show_row_names = TRUE,
               row_names_gp = gpar(fontsize = 5),
               column_names_gp = gpar(fontsize = 6),
               row_dend_gp = gpar(lwd = .5),
               show_column_dend = FALSE,
               height = unit(height, "inch"),
               width = unit(width, "inch"))

  pdf(file.path(out_dir, "heatmap_tf.pdf"), width = width + 3, height = height + 2)
  print(hm)
  dev.off()
  print(hm)

  save_heatmap_horizontal(tf_mat, hc_tf, file.path(out_dir, "heatmap_tf_horizontal.pdf"),
                          col = cols_cur, height = height, width = width,
                          show_gene_names = TRUE)
} else {
  cat("fewer than two DE TFs with complete rows -- no heatmap drawn\n")
}
```

### UMAPs
```{r}
umap_dir = file.path(out_dir, "umaps")
dir.create(umap_dir, showWarnings = FALSE, recursive = TRUE)

## Cleared first. The gene list here is whatever the label selection returned this run, so a panel
## for a gene that has since dropped out would otherwise sit in the directory indefinitely, dated
## from an old run and indistinguishable from current output. (This bit me: two AZ255 panels from an
## interrupted render survived the run that excluded them, and looked like the exclusion had failed.)
unlink(list.files(umap_dir, pattern = "\\.pdf$", full.names = TRUE))
```

```{r}
obj_umap_fname = here::here("snrna/reduction_viz/combined_all_umap_glut_hybrid", "obj_clustered.qs2")
obj_umap = qs_read(obj_umap_fname, nthreads = 8)
```

```{r}
## Same parameter names as the `cluster`-based version -- the sweep in
## combined_all_umap_glut_hybrid.R is the same grid -- but these were originally picked by eye from
## the old embeddings. The hybrid embeddings are fit on a different cell set, so the panels in
## snrna/reduction_viz/combined_all_umap_glut_hybrid are worth a look before these are trusted as
## the best of the nine.
reductions_to_use = c(arco="dims10nn30mindist0.4", nido="dims20nn40mindist0.5")
```

```{r}
genes_to_plot = markers_agg_select_label$gene
## MAGIC_SCT is not on obj_hybrid_labels.qs2, so in practice only SCT is plotted; the intersect
## keeps the loop correct either way.
assays_to_use = c("SCT", "MAGIC_SCT")

## Panel size and point size. The old values -- 3/4 in panels at pt.size 3/2 -- drew points several
## times wider than the spacing between cells, so 15,000 nidopallial cells merged into a field of
## overlapping discs and nothing below the level of a whole cluster could be read. `pt.size` is in
## ggplot size units here, not pixels, so the rasteriser was never the cause and raising
## raster.dpi alone would not have helped.
##
## These were set by rendering the same gene at a range of sizes and looking at the output: the
## panels are enlarged and the points are brought down to roughly the inter-cell spacing, which is
## the value at which individual cells resolve and the expression gradient is still readable as a
## surface. The two regions differ because the arcopallial embedding holds a quarter the cells in a
## comparable area, so it can carry a larger point on a smaller panel.
##
## Points rasterised, everything else vector -- but through ggrastr directly, not through
## `FeaturePlot_scCustom(raster = TRUE)`, which ignores `raster.dpi` outright: the embedded bitmap
## comes out 389 x 342 px at 72 ppi whether that argument says 1024, 2048 or 4096 (checked with
## `pdfimages -list`). That 72 ppi is the soft resolution, and no value of raster.dpi would have
## changed it.
##
## Building the plot with `raster = FALSE` and rasterising the Point layer afterwards puts the
## points at a dpi that is actually honoured -- 3230 x 2842 px at 600 ppi for a nidopallial panel --
## while the title, colourbar and its labels stay as real text. 600 dpi to match fig_save() in
## config/figure_theme.R. About 1.5 MB a nidopallial panel, 720 KB an arcopallial one.
plot_dims = list(arco = c(height = 5, width = 5), nido = c(height = 6, width = 6))
point_sizes = list(arco = 1.0, nido = 0.6)
umap_raster_dpi = 600
walk(names(obj_umap), function(region_cur) {
  obj_cur = obj_umap[[region_cur]]
  reduction_to_use = reductions_to_use[region_cur]
  plot_dim = plot_dims[[region_cur]]
  point_size = point_sizes[[region_cur]]

  walk(intersect(assays_to_use, SeuratObject::Assays(obj_cur)), function(assay_to_use) {
    DefaultAssay(obj_cur) = assay_to_use

    ## Only the genes this object actually carries. The embedding objects were re-SCTransformed per
    ## region, so their SCT assay holds a reduced feature set, while the paired fits name genes from
    ## the full RNA assay -- a gene can be a legitimate marker and still have no row here. Before
    ## the switch to paired fits the two sets came from the same SCT matrix and this never bit.
    genes_present = intersect(genes_to_plot, rownames(obj_cur))
    missing_genes = setdiff(genes_to_plot, genes_present)
    if (length(missing_genes))
      cat(sprintf("%s/%s: %s of %s label genes absent from the %s assay (%s)\n",
                  region_cur, assay_to_use, length(missing_genes), length(genes_to_plot),
                  assay_to_use, paste(missing_genes, collapse = ", ")))

    walk(genes_present, function(gene_to_plot) {
      gg = FeaturePlot_scCustom(obj_cur,
                                features=gene_to_plot,
                                reduction=reduction_to_use,
                                raster = FALSE,
                                pt.size = point_size) +
        coord_equal()  +
        theme_void()
      gg = ggrastr::rasterise(gg, layers = "Point", dpi = umap_raster_dpi)
      save_plot(file.path(umap_dir,
                          str_interp("${region_cur}_${gene_to_plot}_${assay_to_use}.pdf")),
                gg, base_height = plot_dim["height"], base_width = plot_dim["width"])
    })
  })
})
```

### GO, DEG clusters

```{r}
obj_agg_cluster = AggregateExpression(obj, assay=assay_to_use, group.by=res_to_use, return.seurat = T)
mat = GetAssayData(obj_agg_cluster, assay = assay_to_use, layer = "data")

to_keep = apply(mat, 1, function(x) sum(x>0) >= (ncol(mat) * 1))
mat_expr = mat[to_keep,]
```

```{r}
go_res_fname = file.path(out_dir, "go_res.rds")
redo = F
if (redo) {
  entrezid = bitr(Features(obj), fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

  go_res = imap(genes_split, function(gene_set, gene_set_name) {

    gene_set_cur = entrezid$ENTREZID[match(gene_set, entrezid$SYMBOL)]
    gene_set_cur = gene_set_cur[!is.na(gene_set_cur)]

    universe = rownames(mat_expr)
    universe = entrezid$ENTREZID[match(universe, entrezid$SYMBOL)]
    universe = universe[!is.na(universe)]

    go_res = enrichGO(gene_set_cur, universe = universe, OrgDb = org.Hs.eg.db, ont = "ALL", maxGSSize = 200 )
    as.data.frame(go_res)
  }) %>% set_names(names(genes_split)) %>% bind_rows(.id="gene_set")

  saveRDS(go_res, go_res_fname)
} else {
  go_res = readRDS(go_res_fname)
}
```


## Dotplots

```{r}
cts = unlist(pairs)

## Song cluster next to its own surround, in the order the figure reads. As in the `cluster`-based
## version this lists one surround per song cluster, so the second surround of each LMAN pair falls
## outside the factor levels and is dropped on the next line.
ct_levels = c("Glut-DACH2-HVCra", "Glut-DACH2-1",
              "Glut-DACH2-HVCx", "Glut-DACH2-4",
              "Glut-DACH2-LMANco", "Glut-DACH2-8",
              "Glut-DACH2-LMANsh", "Glut-DACH2-5",
              "Glut-CACNA1H-RA", "Glut-CACNA1H-1")

obj_sub = subset(obj, cells = Cells(obj)[obj@meta.data[[res_to_use]] %in% cts])
obj_sub@meta.data[[res_to_use]] = factor(obj_sub@meta.data[[res_to_use]], levels = ct_levels)
obj_sub = subset(obj_sub, cells = Cells(obj_sub)[!is.na(obj_sub@meta.data[[res_to_use]])])
Idents(obj_sub) = obj_sub@meta.data[[res_to_use]]
DefaultAssay(obj_sub) = assay_to_use

obj_agg_cluster = AggregateExpression(obj, assay=assay_to_use, group.by=c(res_to_use), return.seurat = T)

```

```{r}

ngenes = 5

## Canvas = panel + the room the axis labels need, with the two budgeted separately. Folding the
## label allowance into one per-column factor (the old `length(cts) * .2`) meant the panel geometry
## moved whenever the label allowance was retuned, and it over-counted the columns: `cts` is the 12
## entries of `pairs`, but `ct_levels` keeps 10 and the other two are dropped before plotting, so a
## fifth of the width was being reserved for columns that are never drawn.
##
## The x labels are rotated to vertical, so a column costs only the width of its dot -- the label
## contributes nothing horizontally. At the old size each column had twice the width that each gene
## row had height, which is the gap this closes: the columns now sit at roughly the row pitch.
## 0.075 is close to the floor: `dot.scale = 2` puts the largest dot at about 0.078 in across, so
## at this column pitch a full-size dot fills its column edge to edge. Going narrower again means
## bringing dot.scale down with it or the dots start to collide.
dot_col_width   = 0.08  # panel inches per cluster column
dot_row_height  = 0.078  # panel inches per gene row
dot_ylab_width  = 0.72   # room for the gene names
dot_xlab_height = 1.0    # room for the rotated cluster names

n_cols = length(ct_levels)
width = n_cols * dot_col_width + dot_ylab_width

markers_agg_select_up = markers_agg %>%
  filter(is_displayable(gene)) %>%
  group_by(compare) %>%
  top_n(ngenes, avg_log2FC) %>%
  ungroup() %>%
  select(gene) %>% unlist() %>% unique()

markers_agg_mat = GetAssayData(obj_agg_cluster, assay = "SCT", layer = "data")
markers_agg_mat = markers_agg_mat[markers_agg_select_up, cts]
height = nrow(markers_agg_mat) * dot_row_height + dot_xlab_height

hc = hclust(dist(markers_agg_mat), method = "ward.D")

markers_agg_select = hc$labels[hc$order]
gg = DotPlot_scCustom(obj_sub, features = markers_agg_select, group.by = res_to_use, flip_axes = T, dot.scale=2, dot.min = .1)
gg = gg +
  theme(axis.text.x = element_text(size=6, angle = 90, hjust = 1, vjust = .5),
        axis.text.y = element_text(size=6),
        axis.line = element_line(linewidth=.5))
gg


save_plot(file.path(out_dir, str_interp("dotplot_up_${ngenes}_legend.pdf")), gg, base_height=height, base_width=width)
gg = gg + theme(legend.position="none")
save_plot(file.path(out_dir, str_interp("dotplot_up_${ngenes}.pdf")), gg, base_height=height, base_width=width)



markers_agg_select_down = markers_agg %>%
  filter(is_displayable(gene)) %>%
  group_by(compare) %>%
  top_n(-1*ngenes, avg_log2FC) %>%
  ungroup() %>%
  select(gene) %>% unlist() %>% unique()

markers_agg_mat = GetAssayData(obj_agg_cluster, assay = "SCT", layer = "data")
markers_agg_mat = markers_agg_mat[markers_agg_select_down, cts]
height = nrow(markers_agg_mat) * dot_row_height + dot_xlab_height

hc = hclust(dist(markers_agg_mat), method = "ward.D")

markers_agg_select = hc$labels[hc$order]

gg = DotPlot_scCustom(obj_sub, features = markers_agg_select, group.by = res_to_use, flip_axes = T, dot.scale=2, dot.min = .1)
gg = gg +
  theme(axis.text.x = element_text(size=6, angle = 90, hjust = 1, vjust = .5),
        axis.text.y = element_text(size=6),
        axis.line = element_line(linewidth=.5))
gg



gg_leg = get_legend(gg, position = "right")
as_ggplot(gg_leg)

gg = gg + theme(legend.position="none")
save_plot(file.path(out_dir, str_interp("dotplot_down_${ngenes}.pdf")), gg, base_height=height, base_width=width)

save_plot(file.path(out_dir, str_interp("dotplot_down_${ngenes}_legend.pdf")), gg_leg, base_height=height, base_width=width)

```
## Volcano

```{r}
markers_agg = markers_agg %>%
  mutate(sig = p_val_adj < .1,
         avg_log2FC_sign = sign(avg_log2FC),
         p_val_adj_mlog10 = -1 * log10(p_val_adj))

markers_agg_to_plot = markers_agg %>%
  group_by(avg_log2FC_sign, compare) %>%
  filter(is_displayable(gene)) %>%
  top_n(5, p_val_adj_mlog10) %>%
  mutate(gene_to_label = gene) %>% select(compare, gene_to_label, gene)

markers_agg_tmp = markers_agg %>% left_join(markers_agg_to_plot)
gg = ggplot(markers_agg_tmp, aes(avg_log2FC, p_val_adj_mlog10)) +
  geom_point(size=.1) +
  ggrepel::geom_text_repel(aes(label=gene_to_label), size=2.1) +
  facet_grid(.~compare, scales="free") +
  labs(x="", y="") +
  theme_bw() +
  theme(legend.position="none",
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth=.25),
        axis.line = element_line(linewidth=.25),
        axis.ticks = element_line(linewidth=.25),
        axis.text = element_text(size=6),
        strip.text = element_text(size=7),
        strip.background = element_blank())
gg = egg::set_panel_size(p=gg, width = unit(.6, "in"), height=unit(1, "in"))
plot(gg)
save_plot(file.path(out_dir, "deg-celltype-pairs_volcano.pdf"), gg, base_height = 2.5, base_width=1, ncol=5)
```

## Violins

```{r}
obj_sub_vln = subset(obj, cells = Cells(obj)[obj@meta.data[[res_to_use]] %in% cts])
obj_sub_vln@meta.data[[res_to_use]] = factor(obj_sub_vln@meta.data[[res_to_use]], levels = ct_levels)
```

```{r}
VlnPlot_scCustom(obj_sub_vln, features = c("DPP6", "ADARB1"), group.by=res_to_use)
```

## Song vs non-song, as one contrast

The five pairs asked five separate questions. This asks one: pooling every song cluster against
every one of their surrounds, what separates song neurons from non-song neurons?

Two design points.

**What counts as non-song.** The five surrounds already paired with a song nucleus -- one each:
`Glut-CACNA1H-1` for RA, `Glut-DACH2-1` and `-4` for the two HVC clusters, `Glut-DACH2-8` and `-5`
for the two LMAN clusters -- not every other excitatory cluster, and not the additional neighbours
carried in the `pairs` vectors for the dotplot. Those five are the anatomically matched controls,
and both sides of the contrast then draw on the same regions in the same proportions.
Pooling against all excitatory neurons instead would put arcopallial against nidopallial identity
inside the contrast and confound region with song.

**What counts as a replicate.** The bird, not the cluster. Aggregating to one pseudobulk per
song/non-song per bird gives 6 against 6 rather than treating each cluster x bird column as an
independent sample, which would be pseudoreplication -- the same trap `assignment` sets further up.
Each bird contributes to both sides, so the design is naturally paired, and `FindMarkers` cannot
model that. Ignoring the pairing inflates the within-group dispersion and so costs power: the test
is conservative here, not anti-conservative. If a paired fit is wanted, the same pseudobulk matrix
run through DESeq2 directly with `~ individual + songness` is the one-step change.

```{r}
songness_fname = file.path(out_dir, sprintf("song_vs_nonsong_%s.rds", assay_to_use))

## The five corresponding non-song types, one per song type -- `pairs[[i]][2]`, the same surround
## each per-pair contrast uses. Not `setdiff(unlist(pairs), song_cts)`, which would also sweep in
## Glut-DACH2-6 and -7; those are in the pairs vectors for the dotplot columns, not as contrast
## terms.
song_labels    = song_cts
nonsong_labels = unname(map_chr(pairs[song_cts], 2))
stopifnot("song and non-song overlap" = length(intersect(song_labels, nonsong_labels)) == 0)
cat("song:    ", paste(song_labels, collapse = ", "), "\n")
cat("non-song:", paste(nonsong_labels, collapse = ", "), "\n")

obj_songness = subset(obj, cells = Cells(obj)[obj@meta.data[[res_to_use]] %in%
                                                c(song_labels, nonsong_labels)])
obj_songness$songness = if_else(obj_songness@meta.data[[res_to_use]] %in% song_labels,
                                "song", "nonsong")
print(table(obj_songness$songness, obj_songness$individual))

redo_songness = !file.exists(songness_fname)
cat(if (redo_songness) "no cached song/non-song contrast -- fitting\n" else
    sprintf("reusing cached song/non-song contrast: %s\n", songness_fname))

if (redo_songness) {
  obj_agg_songness = AggregateExpression(obj_songness, assay = assay_to_use,
                                         group.by = c("songness", "individual"),
                                         return.seurat = TRUE)
  Idents(obj_agg_songness) = obj_agg_songness$songness

  songness_markers = FindMarkers(obj_agg_songness, assay = assay_to_use,
                                 ident.1 = "song", ident.2 = "nonsong",
                                 test.use = "DESeq2", only.pos = FALSE) %>%
    rownames_to_column(var = "gene")
  saveRDS(songness_markers, songness_fname)
} else {
  songness_markers = readRDS(songness_fname)
}

songness_markers = songness_markers %>%
  mutate(sig = !is.na(p_val_adj) & p_val_adj < sig_thresh,
         displayable = is_displayable(gene))

cat(sprintf("song vs non-song: %s significant of %s tested (%s up, %s down)\n",
            sum(songness_markers$sig), nrow(songness_markers),
            sum(songness_markers$sig & songness_markers$avg_log2FC > 0),
            sum(songness_markers$sig & songness_markers$avg_log2FC < 0)))

write_csv(songness_markers %>% arrange(p_val_adj),
          file.path(out_dir, "song_vs_nonsong_deg.csv"))
print(songness_markers %>% filter(sig, displayable) %>% arrange(-abs(avg_log2FC)) %>% head(30))
```

```{r}
## Volcano, same construction as the per-pair one below so the two read alike.
## Labels chosen by significance within each direction, with a fold-change floor -- the same rule as
## the per-pair volcano below, which ranks on `p_val_adj_mlog10`.
##
## Ranking on |avg_log2FC| alone (what this did first) labels the wrong points. These are unshrunk
## MLE fold changes, so the largest ones belong disproportionately to low-expression genes with wide
## standard errors: big effect, weak p. On a volcano those sit far out on x but low on y, down in
## the band near the significance floor, so the labels land on visually unremarkable points while
## the genes at the top of the plot -- the ones the eye goes to -- go unlabelled.
##
## The |log2FC| >= 1 floor keeps the other failure mode out: a very highly expressed gene can reach
## a tiny p on a fold change too small to care about.
sn_lab = songness_markers %>%
  filter(sig, displayable, abs(avg_log2FC) >= 1) %>%
  group_by(direction = sign(avg_log2FC)) %>%
  slice_min(p_val_adj, n = 12, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(gene, gene_to_label = gene)

sn_plot = songness_markers %>%
  mutate(p_mlog10 = -log10(p_val_adj)) %>%
  left_join(sn_lab, by = "gene")

gg = ggplot(sn_plot, aes(avg_log2FC, p_mlog10)) +
  geom_point(aes(colour = sig), size = .3, stroke = 0) +
  scale_colour_manual(values = c(`TRUE` = "black", `FALSE` = "grey75"), guide = "none") +
  ggrepel::geom_text_repel(aes(label = gene_to_label), size = 2.1, max.overlaps = 30, na.rm = TRUE) +
  labs(x = "avg_log2FC (song / non-song)", y = "-log10 adjusted p") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = .25),
        axis.text = element_text(size = 6),
        axis.title = element_text(size = 7))
gg = egg::set_panel_size(gg, width = unit(2.2, "in"), height = unit(2.2, "in"))
plot(gg)
save_plot(file.path(out_dir, "song_vs_nonsong_volcano.pdf"), gg,
          base_height = 2.2 + 1.2, base_width = 2.2 + 1.2)
```

### Paired, in DESeq2 directly

The contrast above leaves the pairing on the table. Every bird contributes both a song and a
non-song sample, so bird-to-bird differences are a nuisance term that can be modelled rather than
absorbed into the residual. `~ individual + songness` does that; `FindMarkers` has no way to express
it.

Three departures from the chunk above, each deliberate:

- **Raw counts, from the `RNA` assay.** DESeq2's negative binomial model and its size factors expect
  summed UMI counts; SCT's corrected values are already variance-stabilised and are not what the
  model is written for. The chunk above uses SCT because that is what the five per-pair fits use and
  comparability mattered there. Here correctness does, so the pseudobulk is built from raw counts.
- **The pseudobulk is built by hand**, as a cell-by-group indicator multiply, rather than through
  `AggregateExpression`. The sample names then round-trip exactly, which matters because the design
  matrix is keyed on them; Seurat rewrites separators in group names and the two would have to be
  reconciled by string surgery.
- **No shrinkage on the fold changes.** `apeglm` and `ashr` are not in this library, so `lfcShrink`
  has nothing good to offer; the reported log2FC is the MLE. A light count prefilter keeps the worst
  of the low-count noise out, but the fold change for a marginal gene is still noisier than the
  shrunken estimate would be. Rank by p value, not by log2FC, for anything marginal.

What the pairing does *not* fix: birds 1-3 carry the HVC and RA nuclei, birds 4-6 the LMAN nuclei,
so "song" means different nuclei in different birds. The paired term asks, within each bird, whether
its song neurons differ from its own surround -- which is the right question, but it is not the same
as asking it of one nucleus.

```{r}

paired_fname = file.path(out_dir, "song_vs_nonsong_paired_deseq2.rds")

redo_paired = !file.exists(paired_fname)
cat(if (redo_paired) "no cached paired fit -- fitting\n" else
    sprintf("reusing cached paired fit: %s\n", paired_fname))

if (redo_paired) {
  ## The RNA counts layer is BPCells-backed (on disk), not an in-memory sparse matrix, so it has no
  ## @x slot and does not take a matrix product directly. Pulled into memory as dgCMatrix first --
  ## 14k cells is nothing, and everything downstream is then ordinary Matrix arithmetic.
  cts_cells = as(GetAssayData(obj_songness, assay = "RNA", layer = "counts"), "dgCMatrix")

  ## Raw counts, asserted rather than assumed -- a non-integer matrix here would mean the RNA layer
  ## is not what this model wants, and DESeq2 would round it silently.
  stopifnot("RNA counts layer is not integer-valued" =
              all(abs(cts_cells@x - round(cts_cells@x)) < 1e-8))

  grp = factor(paste(obj_songness$songness, obj_songness$individual, sep = "__"))
  pb = cts_cells %*% Matrix::sparse.model.matrix(~ 0 + grp)
  colnames(pb) = levels(grp)
  pb = as.matrix(pb)

  coldata = tibble(sample = colnames(pb)) %>%
    separate(sample, into = c("songness", "individual"), sep = "__", remove = FALSE) %>%
    mutate(songness = factor(songness, levels = c("nonsong", "song")),
           individual = factor(individual)) %>%
    column_to_rownames("sample")

  ## Every bird must appear on both sides or the paired term is not estimable.
  stopifnot("design is not paired -- a bird is missing one side" =
              all(table(coldata$individual, coldata$songness) == 1))
  print(table(coldata$individual, coldata$songness))

  dds = DESeqDataSetFromMatrix(round(pb), colData = coldata,
                              design = ~ individual + songness)

  ## Prefilter: expressed in at least half the samples. Keeps DESeq2's dispersion fit off the long
  ## tail of genes seen in one or two pseudobulk columns.
  keep = rowSums(counts(dds) >= 10) >= (ncol(dds) / 2)
  cat(sprintf("prefilter keeps %s of %s genes\n", sum(keep), length(keep)))
  dds = dds[keep, ]

  dds = DESeq(dds)
  print(resultsNames(dds))

  paired_res = results(dds, name = "songness_song_vs_nonsong") %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    as_tibble()
  saveRDS(paired_res, paired_fname)
} else {
  paired_res = readRDS(paired_fname)
}

paired_res = paired_res %>%
  mutate(sig = !is.na(padj) & padj < sig_thresh,
         displayable = is_displayable(gene))

cat(sprintf("paired song vs non-song: %s significant of %s tested (%s up, %s down)\n",
            sum(paired_res$sig), nrow(paired_res),
            sum(paired_res$sig & paired_res$log2FoldChange > 0),
            sum(paired_res$sig & paired_res$log2FoldChange < 0)))

## About a third of tested genes clear p_adj < 0.1 here, which is a property of the design rather
## than a licence to call them all interesting: six deep pseudobulk columns a side, with the bird
## effect modelled out, leave DESeq2 with very little dispersion, and a tiny consistent shift then
## reaches significance. The fold change is what separates a real difference from a detectable one.
cat(sprintf("of those, |log2FC| >= 1: %s; >= 2: %s\n",
            sum(paired_res$sig & abs(paired_res$log2FoldChange) >= 1),
            sum(paired_res$sig & abs(paired_res$log2FoldChange) >= 2)))

write_csv(paired_res %>% arrange(padj),
          file.path(out_dir, "song_vs_nonsong_paired_deg.csv"))
print(paired_res %>% filter(sig, displayable) %>% arrange(padj) %>% head(30))
```

```{r}
## Paired against unpaired, on the genes both tested. The pairing should buy power without moving
## the effect sizes much -- if the fold changes disagree, the two are not measuring the same thing
## and the discrepancy matters more than either result.
cmp = paired_res %>%
  select(gene, lfc_paired = log2FoldChange, sig_paired = sig) %>%
  inner_join(songness_markers %>% select(gene, lfc_unpaired = avg_log2FC, sig_unpaired = sig),
             by = "gene")

cat(sprintf("genes in both: %s\n", nrow(cmp)))
cat(sprintf("significant: paired %s, unpaired %s, both %s\n",
            sum(cmp$sig_paired), sum(cmp$sig_unpaired), sum(cmp$sig_paired & cmp$sig_unpaired)))
cat(sprintf("log2FC Spearman: %.3f; Pearson: %.3f\n",
            cor(cmp$lfc_paired, cmp$lfc_unpaired, method = "spearman"),
            cor(cmp$lfc_paired, cmp$lfc_unpaired, method = "pearson")))

gg = ggplot(cmp, aes(lfc_unpaired, lfc_paired)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, linewidth = .25, colour = "grey50") +
  geom_point(size = .3, stroke = 0, alpha = .4) +
  labs(x = "avg_log2FC, unpaired (SCT, FindMarkers)",
       y = "log2FoldChange, paired (RNA, ~ individual + songness)") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = .25),
        axis.text = element_text(size = 6),
        axis.title = element_text(size = 6))
gg = egg::set_panel_size(gg, width = unit(2, "in"), height = unit(2, "in"))
plot(gg)
save_plot(file.path(out_dir, "song_vs_nonsong_paired_vs_unpaired.pdf"), gg,
          base_height = 2 + 1.4, base_width = 2 + 1.4)
```

## Core and shared gene sets

Which genes separate a song nucleus from its surround in the *same direction* everywhere, and which
do so only within HVC or only within LMAN.

Direction is the point. The UpSet and `intersects` below take the intersection of significant gene
lists without regard to sign, so a gene up in HVCra and down in RA counts as shared there; here it
does not. A gene qualifies for a set only if it is significant in every comparison of that set *and*
its fold change has the same sign in all of them.

Three sets, plus their exclusive forms:

- **core** -- same direction in all five song/surround pairs, significant in at least three of them.
- **HVC-common** -- both HVC pairs (HVCra, HVCx), significant in both.
- **LMAN-common** -- both LMAN pairs (LMANco, LMANsh), significant in both.

"Exclusive" means the gene additionally fails to reach significance in *any* of the pairs outside
that set -- so HVC-exclusive genes are HVC's own, not core genes seen through an HVC window.

Because core needs significance in only three of five, it is **not** a subset of HVC-common or of
LMAN-common: a core gene can make its three elsewhere and miss in HVCx, say. The counts printed
below report that overlap rather than assume it. Requiring all five instead gave a core of 5 genes,
which was largely a statement about the weakest comparison's power.

The threshold is `p_val_adj < 0.1`, the same one the rest of the notebook uses, and the
per-comparison fold changes and adjusted p values are written out alongside so a near-miss can be
told from a real divergence.

```{r}

hvc_cts  = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx")
lman_cts = c("Glut-DACH2-LMANco", "Glut-DACH2-LMANsh")

de_long = markers_agg %>%
  transmute(gene, compare,
            avg_log2FC,
            sig = !is.na(p_val_adj) & p_val_adj < sig_thresh,
            dir = sign(avg_log2FC))

## Same-signed across *every* comparison in `cts`, and significant in at least `k` of them.
##
## The two conditions do different work. Direction concordance is required everywhere, including
## where the gene misses significance, so a gene that moves the other way in one nucleus is out
## regardless of how strong it is in the rest. Significance is required only k times, which is what
## keeps the least-powered pair from setting the size of the set: HVCra and RA return 1422 and 1252
## significant genes, HVCx, LMANco and LMANsh only 194, 160 and 256, so demanding all five made the
## core a function of the weakest comparison's power more than of shared biology.
##
## The gene must also have been returned for all of `cts` -- otherwise its direction is undefined
## somewhere and "same direction everywhere" cannot be checked.
concordant_set = function(cts, k = length(cts)) {
  stopifnot(k <= length(cts))
  de_long %>%
    filter(compare %in% cts) %>%
    group_by(gene) %>%
    summarise(n_present = n_distinct(compare),
              n_sig = sum(sig),
              n_dir = n_distinct(dir[dir != 0]),
              direction = if (n_distinct(dir[dir != 0]) == 1) dir[dir != 0][1] else NA_real_,
              .groups = "drop") %>%
    filter(n_present == length(cts), n_sig >= k, n_dir == 1) %>%
    transmute(gene, direction = if_else(direction > 0, "up", "down"))
}

## k = 3 of 5 for the core set. The two-comparison sets keep k = 2, their full size -- there is no
## room to relax a set of two without it ceasing to mean "shared".
core_k = 3
core_set = concordant_set(song_cts, k = core_k)
hvc_set  = concordant_set(hvc_cts)
lman_set = concordant_set(lman_cts)

## Exclusive: not significant in any comparison outside the set, in either direction.
sig_outside = function(cts) {
  de_long %>% filter(!compare %in% cts, sig) %>% distinct(gene) %>% pull(gene)
}
hvc_excl  = hvc_set  %>% filter(!gene %in% sig_outside(hvc_cts))
lman_excl = lman_set %>% filter(!gene %in% sig_outside(lman_cts))

cat(sprintf(paste0("core (same direction in all 5, significant in >=", core_k, "): %s  (%s up, %s down)\n",
                   "HVC-common: %s   HVC-exclusive: %s\n",
                   "LMAN-common: %s  LMAN-exclusive: %s\n",
                   "core genes also in HVC-common: %s; in LMAN-common: %s\n"),
            nrow(core_set), sum(core_set$direction == "up"), sum(core_set$direction == "down"),
            nrow(hvc_set), nrow(hvc_excl), nrow(lman_set), nrow(lman_excl),
            sum(core_set$gene %in% hvc_set$gene), sum(core_set$gene %in% lman_set$gene)))
```

```{r}
## One table, one row per gene per set, carrying the five fold changes and the five adjusted p
## values so a membership can be checked against the numbers behind it. `displayable` flags the
## unnamed models and mitochondrial features -- they are kept in the analysis and only excluded
## from figure labels, so the flag is how they stay visible here.
wide_fc = markers_agg %>%
  select(gene, compare, avg_log2FC) %>%
  pivot_wider(names_from = compare, values_from = avg_log2FC,
              names_glue = "log2FC_{compare}")
wide_p = markers_agg %>%
  select(gene, compare, p_val_adj) %>%
  pivot_wider(names_from = compare, values_from = p_val_adj,
              names_glue = "padj_{compare}")

core_gene_sets = bind_rows(
  core_set %>% mutate(set = "core"),
  hvc_set  %>% mutate(set = "HVC-common"),
  hvc_excl %>% mutate(set = "HVC-exclusive"),
  lman_set %>% mutate(set = "LMAN-common"),
  lman_excl %>% mutate(set = "LMAN-exclusive")
) %>%
  mutate(set = factor(set, levels = c("core", "HVC-common", "HVC-exclusive",
                                      "LMAN-common", "LMAN-exclusive")),
         displayable = is_displayable(gene)) %>%
  left_join(wide_fc, by = "gene") %>%
  left_join(wide_p, by = "gene") %>%
  arrange(set, direction, gene)

write_csv(core_gene_sets, file.path(out_dir, "core_gene_sets.csv"))
print(core_gene_sets %>% count(set, direction))
```

```{r}
## The core genes as a heatmap over the five pairs. Rows ordered by mean fold change rather than
## clustered: the set is defined by having the same sign everywhere, so a dendrogram would mostly
## re-say that, where the ordering shows how strongly each gene separates and in which direction.
core_mat = core_gene_sets %>%
  filter(set == "core") %>%
  distinct(gene, .keep_all = TRUE) %>%
  select(gene, all_of(paste0("log2FC_", song_cts))) %>%
  column_to_rownames("gene") %>%
  as.matrix()
colnames(core_mat) = song_cts
core_mat = core_mat[order(rowMeans(core_mat)), , drop = FALSE]

if (nrow(core_mat) > 0) {
  max_val = 4
  cols_cur = circlize::colorRamp2(breaks = c(-max_val, 0, max_val),
                                  colors = c(scales::muted("blue"), "white", scales::muted("red")))

  ## Labelled only where the name means something; the flag above keeps the rest in the table.
  core_label = rownames(core_mat)[is_displayable(rownames(core_mat))]
  core_label_ind = match(core_label, rownames(core_mat))

  cell_h = 0.10
  hm = Heatmap(core_mat, name = "avg_log2FC", col = cols_cur,
               cluster_rows = FALSE, cluster_columns = FALSE,
               show_row_names = FALSE,
               column_names_gp = gpar(fontsize = 6),
               right_annotation = rowAnnotation(
                 foo = anno_mark(at = core_label_ind, labels = core_label,
                                 labels_gp = gpar(fontsize = 5), lines_gp = gpar(lwd = .5))),
               width = unit(0.12 * ncol(core_mat), "inch"),
               height = unit(cell_h * nrow(core_mat), "inch"))

  pdf(file.path(out_dir, "core_genes_heatmap.pdf"),
      width = 0.12 * ncol(core_mat) + 3, height = cell_h * nrow(core_mat) + 2)
  print(hm)
  dev.off()
  print(hm)
}
```

## Overlap of DEG

```{r}
markers_agg_split = split(markers_agg, markers_agg$compare)
markers_agg_sig = map(markers_agg_split, function(x) {
  x %>%
    dplyr::filter(p_val_adj < 0.1) %>%
    dplyr::select(gene) %>%
    unlist()
})
set_order = rev(c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx", "Glut-CACNA1H-RA",
                  "Glut-DACH2-LMANco", "Glut-DACH2-LMANsh"))
markers_agg_sig_upset = fromList(markers_agg_sig)
gg = upset(fromList(markers_agg_sig), group.by = "sets", order.by="freq", keep.order = T, sets = set_order, empty.intersections = "on", nintersects = 100)
gg
```

```{r}

intersect_names = list(SMP = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx", "Glut-CACNA1H-RA"),
                       NIDO = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx", "Glut-DACH2-LMANsh", "Glut-DACH2-LMANco"),
                       SONG = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx", "Glut-DACH2-LMANsh", "Glut-DACH2-LMANco", "Glut-CACNA1H-RA"),
                       HVC = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCx"),
                       LMAN  = c("Glut-DACH2-LMANsh", "Glut-DACH2-LMANco"),
                       PP = c("Glut-DACH2-HVCra", "Glut-DACH2-LMANsh"),
                       PS = c("Glut-DACH2-HVCx", "Glut-DACH2-LMANco"))

intersects = map(intersect_names, function(inter_cur) {
  sort(Reduce(intersect, markers_agg_sig[inter_cur]))
}) %>% set_names(names(intersect_names))
```

## GO Terms

```{r}
obj_agg_cluster = AggregateExpression(obj, assay=assay_to_use, group.by=res_to_use, return.seurat = T)
mat = GetAssayData(obj_agg_cluster, assay = assay_to_use, layer = "data")

thresh = .05
plot(density(mat[,"Glut-DACH2-HVCra"]))
mat_expr = map(1:ncol(mat), function(i) rownames(mat)[mat[,i] >= thresh]) %>%
  set_names(colnames(mat))
```

```{r}

entrezid = bitr(markers_agg_split[[1]]$gene, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

go_res = map(names(intersect_names), function(compare) {
  cts = intersect_names[[compare]]

  gene_set = entrezid$ENTREZID[match(intersects[[compare]], entrezid$SYMBOL)]
  gene_set = gene_set[!is.na(gene_set)]

  universe = Reduce(intersect, mat_expr[cts])
  universe = entrezid$ENTREZID[match(universe, entrezid$SYMBOL)]
  universe = universe[!is.na(universe)]

  go_res = enrichGO(gene_set, universe = universe, OrgDb = org.Hs.eg.db, ont = "ALL", maxGSSize = 200 )
  as.data.frame(go_res)
}) %>% set_names(names(intersect_names)) %>% bind_rows(.id="compare")
```

## GSEA

```{r}
library(fgsea)
gmt_files = GMT_C5_V7  # bundled in data/gene_lists/

gmt_pathway = gmtPathways(gmt_files)
```

```{r}
gse_res_list = map(markers_agg_split, function(compare) {
  compare = compare %>%
    arrange(-avg_log2FC)
  gene_set = compare %>% select(avg_log2FC) %>% unlist()
  names(gene_set) = compare$gene
  gene_set = gene_set[!is.na(gene_set)]
  gene_set = gene_set[!is.na(names(gene_set))]
  gsea_res = fgseaMultilevel(pathways = gmt_pathway,
                             stats = gene_set,
                             minSize=20,
                             maxSize=200)

  gsea_res_filt = gsea_res %>% filter(padj<.2) %>% arrange(pval)
  collapsedPathways = collapsePathways(gsea_res_filt, pathways = gmt_pathway, stats = gene_set, pval.threshold = .05)
  gsea_res_ind = gsea_res %>% filter(pathway %in% collapsedPathways$mainPathways) %>%
    arrange(-NES)
  return(list(gsea_res, gsea_res_ind))
})
gse_res = map(gse_res_list, 1) %>% set_names(names(markers_agg_split)) %>% bind_rows(.id="compare")
gse_res_ind = map(gse_res_list, 2) %>% set_names(names(markers_agg_split)) %>% bind_rows(.id="compare")
```

```{r}
set_order = song_cts
  gse_res_ind_select = gse_res_ind %>%
    mutate(p.adjust_sign = -1 * log10(padj)) %>%
    mutate(NES_sign = sign(NES)) %>%
    filter(padj<.05) %>%
    select(pathway, compare) %>%
    mutate(is_sig = TRUE)

  gse_res_ind_filt = gse_res %>%
    filter(pathway %in% gse_res_ind_select$pathway) %>%
    distinct(compare, pathway, .keep_all=T) %>%
    left_join(gse_res_ind_select) %>%
    mutate(is_sig = if_else(is.na(is_sig), FALSE, is_sig)) %>%
    mutate(pathway = sub("^GO", "", pathway ),
           pathway = gsub("_", " ", pathway),
           pathway = tolower(pathway)) %>%
    mutate(pathway = case_when(
      grepl("establishment of protein", pathway) ~ "protein localization to ER",
      grepl("anaphase promoting complex", pathway) ~ "APC-dependent catabolic process",
       ## GO_CELL_CELL_ADHESION_VIA_PLASMA_MEMBRANE_ADHESION_MOLECULES and its homophilic child
       ## GO_HOMOPHILIC_CELL_ADHESION_VIA_PLASMA_MEMBRANE_ADHESION_MOLECULES both match the second
       ## pattern. Both are significant here, so collapsing them to one label put two rows on the
       ## same (compare, pathway) cell and pivot_wider turned NES into a list column. Matched most
       ## specific first, and kept as two rows -- they are two terms.
       grepl("homophilic cell adhesion via plasma", pathway) ~ "homophilic cell adhesion",
       grepl("cell adhesion via plasma", pathway) ~ "cell-cell adhesion",
       grepl("nuclear transcribed", pathway) ~ "nonsense mediated decay",

       grepl("cellular component assembly involved in morphogenesis", pathway) ~ "cellular component assembly in morphogenesis",
                               TRUE ~ pathway))

  go_res_mat = gse_res_ind_filt %>%
    select(compare, pathway, NES) %>%
    pivot_wider(names_from = "compare", values_from = "NES", values_fill = 0) %>%
    column_to_rownames("pathway") %>% as.matrix()

  hc_row = hclust(dist(go_res_mat), method="ward.D")
  hc_row_order = hc_row$labels[hc_row$order]


  gse_res_ind_filt = gse_res_ind_filt %>%
    mutate(leadingEdge_len = map_int(leadingEdge, length))

  gse_res_ind_filt = gse_res_ind_filt %>%
    mutate(pathway = factor(pathway, levels=hc_row_order)) %>%
    mutate(compare = factor(compare, levels=set_order)) %>%
    complete(pathway, compare, fill=list(NES=NA, size=NA, leadingEdge_len = NA))

  size_limits = seq(round(min(gse_res_ind_filt$leadingEdge_len, na.rm=T), -1), round(max(gse_res_ind_filt$leadingEdge_len, na.rm=T), -1), by=20)
  gg = ggplot(gse_res_ind_filt, aes(compare, pathway)) +
    geom_point(shape=21, aes(size=leadingEdge_len, fill=NES, color=is_sig)) +
    scale_color_manual(values=c(`TRUE`="black", `FALSE`=NA)) +
    scale_fill_gradient2(low=scales::muted("blue"), high=scales::muted("red")) +
    scale_size_continuous(range=c(.5, 3), breaks=size_limits) +
    labs(x = "", y ="") +
    theme_bw() +
    theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1, size=6),
          axis.text.y = element_text(hjust=1, size=6),
          axis.ticks = element_line(size=.25),
          panel.border = element_blank()) +
    coord_fixed()

  ncols = ncol(go_res_mat)
  nrows = nrow(go_res_mat)

  scale_factor = .1
  width = ncols * scale_factor
  height = nrows * scale_factor
  gg = egg::set_panel_size(gg, width = unit(width, "in"), height = unit(height, "in"))
  plot(gg)
  save_plot(file.path(out_dir, sprintf("gse_terms_by_compare.pdf")), gg, base_height = height + 4, base_width = width + 4)
```
