---
title: "Curation and naming: Glut-Nido"
format: html
editor: source
---

One curation round for the Glut-Nido subset, as the trial run for the scheme in
`scripts/cluster_curation.R` and `scripts/cluster_naming.R`. Nido is the right subset to start with
because it has already been curated by hand *twice* (`obj_glutamatergic_nido` ->
`obj_glutamatergic_nido_filt` in `snrna_seurat_cellbender_preprocess.qmd`), so the automated flags
can be scored against the calls you actually made rather than against nothing.

**What this replaces.** The hand loop is: look at the subclusters, drop the ones with high mito or
co-expressed markers, recluster the survivors. This notebook proposes the same drops from measured
quantities, records them per *cell barcode*, and reruns the clustering from the ledger. Two
properties follow, and both are the point:

- **Decisions survive reclustering.** The ledger keys on barcodes, not cluster ids, so renumbering
  cannot silently repoint a decision at different cells -- the same failure mode as souporcell's
  per-library `assignment`.
- **Cells are marked, not deleted.** `curation_status` / `exclude_reason` travel with every cell.
  Since this dataset is the reference others take names from, downstream projects need to be able to
  reconcile counts and see what was removed and why.

Nothing here is applied automatically: the flags are a proposal, you confirm them, and the confirmed
set is what the ledger records.

```{r}
#| label: packages
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(knitr)
library(clustree) # attached, not just loaded -- see the preprocess notebook for why

source(file.path(Sys.getenv("COLQUITTLAB_UTILS", "/opt/colquittlab/utils"), "R", "scRNA.R"))
source(here::here("scripts/cluster_curation.R"))
source(here::here("scripts/cluster_naming.R"))

select <- dplyr::select
filter <- dplyr::filter

## sctransform parallelises its per-gene regression through future, and future refuses to ship more
## than 500 MiB of globals into a future by default. SCTransform's own chunked data clears that on
## an object this size (~730 MiB for 16k cells), and the failure surfaces as
## `getGlobalsAndPackages ... The total size of the 19 globals exported for future expression
## ('FUN()')` -- which names neither future nor SCTransform, so it is worth knowing on sight.
##
## The option is the actual fix: the size check runs when the future's globals are collected, which
## happens under a sequential plan too, so plan(sequential) alone does not avoid it. Setting the plan
## as well keeps SCTransform from forking a copy of a large object for no gain.
## snrna_seurat_cellbender_preprocess.qmd sets the same two lines (107-108) for the same reason.
library(future)
options(future.globals.maxSize = Inf)
plan(sequential)
```

```{r}
#| label: params
subset_name <- "glut_nido"
in_dir <- here::here("snrna/clustering", "snrna_seurat_cellbender_preprocess")
obj_fname <- file.path(in_dir, "obj_glutamatergic_nido.qs2")

out_dir <- here::here("snrna/clustering", "cluster_curation_glut_nido")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

## Ledgers are small, hand-confirmed, and reviewable in a diff: source, not output. They belong in
## git next to the code, not in the output tree.
ledger_dir <- here::here("data/curation")
dir.create(ledger_dir, recursive = TRUE, showWarnings = FALSE)
ledger_fname <- file.path(ledger_dir, sprintf("%s_ledger.csv", subset_name))

rna_dims <- 10
resolutions <- seq(.3, 1, .1)
cluster_prefix_cur <- "cluster_glut_nido"
round_n <- 1L

## Recorded rather than defaulted, so the run says what it used. `doublet_frac` is left at its NA
## default -- disabled -- because the preprocessing notebook has already kept only cells that both
## DoubletFinder and souporcell call singlets, so every cluster here is at exactly zero and a rule
## on it could never fire. Residual doublets surface as marker co-expression instead.
thresholds <- curation_thresholds(mito_mad = 3, min_cells = 20)
filters <- naming_filters()
panels <- class_panels()

seed <- 10
set.seed(seed)
```

## The panel

```{r}
#| label: panel
## Derived from ct_markers in gene_lists.R, with three deliberate deviations -- see class_panels().
## Read this table before trusting any class call below: every class assignment is only as good as
## the row that defines it.
tibble(
  class = names(panels$panels),
  markers = map_chr(panels$panels, paste, collapse = ", ")
) %>%
  kable(caption = "Cell-class panels (specificity scoring ignores the pan-neuronal genes below)")

tibble(
  set = names(panels$exclusive),
  members = map_chr(panels$exclusive, paste, collapse = ", ")
) %>%
  kable(caption = "Mutually exclusive sets: two present members in one cluster is a co-expression flag")

tibble(
  state = names(panels$states),
  markers = map_chr(panels$states, paste, collapse = ", ")
) %>%
  kable(caption = paste(
    "States, scored but never counted as violations -- a cluster may legitimately be Glut *and*",
    "cycling. Note `Cycling` is ct_markers$Mito, which is mitotic (TOP2A/PCNA/...), not",
    "mitochondrial; perc_mito is a separate metric entirely."
  ))

tibble(non_specific = panels$non_specific) %>%
  kable(caption = "In a panel but excluded from specificity scoring (pan-neuronal)")
```

## Load

```{r}
#| label: load
stopifnot("run snrna_seurat_cellbender_preprocess.qmd first" = file.exists(obj_fname))
obj <- qs_read(obj_fname, nthreads = 8)

## Re-apply any decisions from previous rounds before doing anything else: a round always starts
## from the surviving cells, and this is what makes re-running the notebook idempotent.
ledger <- read_ledger(ledger_fname)
obj <- apply_ledger(obj, ledger, subset_name)

tibble(
  n_cells_total = ncol(obj),
  n_kept = sum(obj$curation_status == "keep"),
  n_excluded_previous_rounds = sum(obj$curation_status == "exclude")
) %>%
  kable(caption = "Starting point")
```

## Cluster the surviving cells

```{r}
#| label: cluster
reduction_cur <- umap_name(rna_dims)
res_to_use <- cluster_name(cluster_prefix_cur, rna_dims, 0.7)

obj_cur <- subset(obj, cells = kept_cells(obj))

## Recluster only when there is something new to react to. On round 1 the object arrives straight
## from the preprocessing notebook with this exact clustering already on it, and re-running
## SCTransform + process_rna would spend an hour reproducing it -- not bit-identically, either,
## which would make the flags below refer to slightly different clusters than the ones you looked
## at upstream. Once cells have been excluded, the clustering genuinely has to be redone, because
## avoiding the distortion those cells cause is the entire point of the loop.
needs_recluster <- !all(c(res_to_use %in% colnames(obj_cur@meta.data),
                          reduction_cur %in% Reductions(obj_cur))) ||
  any(obj$curation_status == "exclude")

if (needs_recluster) {
  message("Reclustering ", ncol(obj_cur), " surviving cells")
  obj_cur <- obj_cur %>% SCTransform(vars.to.regress = "perc_mito")
  obj_cur <- process_rna(obj_cur,
                         cluster_prefix = cluster_prefix_cur,
                         rna_dims_list = rna_dims,
                         resolutions = resolutions,
                         algorithm = 2,
                         run_sct = TRUE,
                         run_parallel = FALSE)
} else {
  message("Scoring the clustering already on the object (no exclusions yet)")
}

stopifnot("chosen resolution not in the object" = res_to_use %in% colnames(obj_cur@meta.data))
```

```{r}
#| label: umap
#| fig-width: 10
DimPlot(obj_cur, group.by = res_to_use, label = TRUE, reduction = reduction_cur) + coord_equal()
DimPlot(obj_cur, group.by = "individual", reduction = reduction_cur) + coord_equal()
DimPlot(obj_cur, group.by = "position", reduction = reduction_cur) + coord_equal()
```

## Score the clusters

```{r}
#| label: score
## One presto pass feeds both the QC metrics and the naming genes -- it is the expensive step.
de <- cluster_de(obj_cur, res_to_use)
qc <- cluster_qc(obj_cur, res_to_use, de = de, panels = panels)
naming <- select_naming_genes(de, filters = filters)
flagged <- flag_clusters(qc, thresholds = thresholds, naming = naming)

flagged %>%
  select(cluster, n_cells, median_mito, mito_mad, median_nFeature,
         classes_present, states_present, coexpr_violation, flag_reason) %>%
  kable(digits = 2, caption = paste(
    "Per-cluster QC. mito_mad is MADs above the subset median, not an absolute cut -- the global",
    "perc_mito filter already ran upstream, so what is left to detect is enrichment within Nido.",
    "median_nFeature is reported, not flagged: elevated complexity is doublet-like but also true of",
    "genuinely large neurons, so it is context for the co-expression call rather than a rule."
  ))

## Verification, not curation: both callers' doublets were removed upstream, so these must be zero.
## A non-zero value means this object did not go through that filter and the co-expression rule is
## then carrying more weight than it was meant to.
doublet_check <- flagged %>%
  summarise(max_df = max(coalesce(frac_df_doublet, 0)), max_soc = max(coalesce(frac_soc_doublet, 0)))
if (doublet_check$max_df > 0 || doublet_check$max_soc > 0) {
  warning("Doublets present in this object -- expected none after preprocessing. ",
          "Set curation_thresholds(doublet_frac = ...) to flag on them.", call. = FALSE)
}
kable(doublet_check, caption = "Upstream doublet removal check (both must be 0)")
```

```{r}
#| label: naming
naming %>%
  left_join(qc %>% select(cluster, top_class, class_margin), by = "cluster") %>%
  kable(digits = 3, caption = paste(
    "Naming genes: the smallest set that is present in this cluster and in no sibling.",
    "A two- or three-gene name means no single gene distinguished the cluster -- the length is a",
    "measurement, not a style choice. n_naming_genes == 0 means no marker of its own: merge or drop."
  ))
```

```{r}
#| label: anatomy
## Region is an enrichment claim about sample composition, not a marker claim, so it is tested with
## birds as the replicate unit. `confounded` marks clusters living in a single batch, where a region
## effect cannot be told apart from a batch effect -- for Nido that is the check that matters, since
## lman/nr are batch 2 and share no birds with the other four libraries.
anatomy <- region_enrichment(obj_cur, res_to_use)
kable(anatomy, digits = 2, caption = "Region enrichment (region = NA where unsupported or confounded)")

labels <- make_labels(naming, qc = qc, anatomy = anatomy, prefix = "Glut-Nido")
kable(labels %>% select(cluster, class, region, naming_genes, unique_id, label),
      caption = "Proposed labels: <class>-<region>-<genes>, fields dropped where unsupported")
```

## Confirm the proposal

```{r}
#| label: propose
## Everything above is a proposal. Edit `overrides` to disagree with it -- clusters listed in
## `keep_anyway` are retained despite their flags, clusters in `drop_anyway` are dropped with the
## reason you give. Both are recorded in the ledger, so a hand override is as auditable as an
## automatic flag.
overrides <- list(
  keep_anyway = character(),
  drop_anyway = c() # e.g. c("7" = "residual doublets not caught by either caller")
)

decided <- flagged %>%
  mutate(
    flagged = case_when(
      cluster %in% overrides$keep_anyway ~ FALSE,
      cluster %in% names(overrides$drop_anyway) ~ TRUE,
      TRUE ~ flagged
    ),
    flag_reason = case_when(
      cluster %in% overrides$keep_anyway ~ "",
      cluster %in% names(overrides$drop_anyway) ~ paste0("manual: ", overrides$drop_anyway[cluster]),
      TRUE ~ flag_reason
    )
  )

proposal <- propose_exclusions(obj_cur, res_to_use, decided, subset_name, round = round_n)

proposal %>%
  count(cluster_at_decision, reason, name = "n_cells") %>%
  kable(caption = "Cells this round proposes to exclude")

tibble(
  n_cells = ncol(obj_cur),
  n_proposed_excluded = nrow(proposal),
  frac_excluded = round(nrow(proposal) / ncol(obj_cur), 3)
) %>%
  kable(caption = paste(
    "Drop fraction. A few percent is artefact removal; a large fraction is a different operation",
    "wearing its clothes, and means the thresholds are wrong rather than the cells."
  ))
```

```{r}
#| label: write-ledger
## Set to TRUE once the tables above look right. Kept behind a flag so rendering the notebook to
## look at it cannot silently commit a curation decision.
commit_decisions <- FALSE

if (commit_decisions) {
  ledger <- write_ledger(ledger, proposal, ledger_fname)
  message("Wrote ", nrow(proposal), " decisions to ", ledger_fname)
} else {
  message("commit_decisions = FALSE; ledger not modified")
}
```

## Did the curation delete biology?

```{r}
#| label: audit
## The check that matters for this dataset. lman/nr are batch 2 -- different birds, higher ambient,
## different doublet rates -- so a quality-driven rule can preferentially delete batch-2 cells, and
## the result would read as "LMAN lacks cell type X". Enrichment near 1 means a level is losing
## cells at the average rate; a level well above 1 is a finding to report, not a filter to apply.
obj_audit <- apply_ledger(obj_cur, bind_rows(ledger, proposal), subset_name)
excluded_composition(obj_audit) %>%
  kable(digits = 3, caption = "Composition of the excluded cells against the design")
```

```{r}
#| label: save
qs_save(list(qc = flagged, naming = naming, anatomy = anatomy, labels = labels,
             thresholds = thresholds, filters = filters, round = round_n),
        file.path(out_dir, sprintf("%s_round%s_evidence.qs2", subset_name, round_n)), nthreads = 8)
```

Next round: set `round_n <- 2`, re-render, and the `load` chunk will start from the survivors. Stop
when nothing is flagged -- and if that has not happened by round 3, the thresholds are wrong rather
than the cells.

## Session info

```{r}
#| label: session-info
sessionInfo()
```
