---
title: "snRNA souporcell reprocessing: genotype-based individual assignment"
format: html
editor: source
---

Reprocesses the souporcell runs for the six 210607 snRNA libraries using the approach developed in
`~/ssd/rstudio/multiome/dev_full1/lonStrDom3/dev-full1_multiome_souporcell_clustering.qmd` (itself
derived from `~/ssd/rstudio/stable`): read each run's `cluster_genotypes.vcf` as a clusters x variants
genotype matrix **plus a matching read-depth matrix**, mask calls that rest on too few reads, then
compare clusters *across* libraries by genotype to recover which souporcell cluster in which library
is the same bird.

What this replaces: `snrna_seurat_cellbender_preprocess.qmd` currently reads each run's `clusters.tsv`
and keeps `assignment` as-is. That number is a **per-library** cluster index -- `ra_run1` cluster 0 and
`hvc_run1` cluster 0 are unrelated birds -- so every use of `assignment` downstream (the
`DimPlot(group.by = "assignment")` panels, the `table(cluster, assignment)` contingency tables) is
really plotting three arbitrary labels recycled six times, and there is currently no individual
identity in the object at all. This notebook derives that identity from the genotypes.

## The headline result, and a correction to the assumed design

This notebook was written expecting a complete block: three birds dissected into all six regions,
3 x 6 = 18 slots, every library holding the same individuals. **The genotypes contradict that.** The
libraries fall into two batches that share no birds at all:

| batch | libraries                       | birds |
|-------|---------------------------------|-------|
| 1     | `ra`, `arco`, `hvc`, `nc`       | 3     |
| 2     | `lman`, `nr`                    | 3     |

Within batch 1 the four libraries match each other cleanly -- same-bird distances of 0.03-0.08
against a different-bird background of 0.55-0.70, the same well-separated picture stable measured.
Within batch 2, `lman` and `nr` match each other at 0.04-0.13. But every `ra`/`arco`/`hvc`/`nc` x
`lman`/`nr` distance sits at 0.55-0.65, i.e. at the different-bird background, with no exceptions:
those are different animals, not the same animals poorly genotyped.

So there are **six birds**, each in four or two libraries, and `individual` is only comparable
*within* a batch. Anything downstream that treats bird as a random effect or pairs regions within an
animal can only do so inside a batch; `lman`/`nr` cannot be compared to the other four at the
individual level at all. The `batch-structure` chunk below derives this from the genotypes rather
than restating it, so a reprocessing run will re-report it rather than inherit it.

`lman` and `nr` also show an intermediate band at 0.29-0.40 -- above same-bird, well below
different-bird -- between the *other* combinations of their clusters. That is what relatedness looks
like; those three birds are plausibly siblings. It matters here only because a loose threshold chains
through that band, which is what the calibration below is built to reject.

**Two further differences from the dev-full1 notebook.**

1. *There is no hand-filled truth sheet for this dataset.* dev-full1 could calibrate `min_depth` and
   the same-bird threshold against 90 labelled slots and score the result with agreement/ARI. Here
   both parameters are chosen from label-free criteria instead:
   - `min_depth` from the ratio of matched to all-pairs genotype distance (stable's criterion:
     a depth filter is paying its way while that ratio keeps falling), floored by how many
     comparable sites survive;
   - the same-bird threshold from a structural criterion that assumes no design at all -- no
     component may contain two clusters from the same library, because a bird is one individual and
     cannot be two of a library's clusters. An earlier version of this notebook used the design
     instead (largest threshold giving 3 components of 6, bounded by size <= 6); that criterion was
     unreachable, and its fallback -- largest threshold within the size bound -- chose 0.40, deep
     inside the different-bird range, chaining two `lman` and two `nr` clusters into one "bird".
     Library-disjointness rejects exactly that, and needs to know nothing about the dissection.

2. *1:1 matching is only used within a batch.* `match_clusters()` will happily assign three clusters
   to three clusters between two libraries that share no bird -- the trap dev-full1's incomplete
   design fell into, and eight of the fifteen library pairs here are exactly that case. Those pairs
   are reported separately, where a match at different-bird distance is the correct outcome rather
   than a disagreement. Connected components remain the primary method.

## Packages

```{r}
#| label: packages
#| message: false
library(tidyverse)
library(qs2)
library(parallel)      # mclapply, for the VCF read
library(ComplexHeatmap) # Heatmap / gpar
library(knitr)         # kable

## VariantAnnotation is deliberately *not* attached. Every call into it is qualified and lives in
## the sourced helpers (VariantAnnotation::readVcf / ScanVcfParam / geno), and a qualified call
## loads the namespace, which is all S4 dispatch needs. Attaching it here bought nothing except the
## masking it forces on tidyverse's verbs -- which is what the re-binding block below exists to
## undo. That block is kept anyway: ComplexHeatmap still pulls MatrixGenerics into the search path.

## Loading, depth masking, distance and matching primitives, plus the run-quality metrics, come from
## the stable project rather than being copied here: SAME_BIRD_MAX and the distance definition are
## the bar a reprocessing run has to clear, and a second copy that drifted would invalidate the
## comparison rather than improve it. The cost is a hard path dependency; fail loudly rather than
## half-run if it moves.
STABLE_R <- path.expand("~/ssd/rstudio/stable/R")
for (f in c("souporcell_helpers.R", "souporcell_qc.R")) {
  p <- file.path(STABLE_R, f)
  if (!file.exists(p)) {
    stop("Missing helper: ", p, "\nSet STABLE_R to the stable project's R/ directory.", call. = FALSE)
  }
  source(p)
}

## tidyverse is attached first, so its verbs are masked by the Bioconductor packages above. Re-bind
## rather than reorder the library() calls, so the Bioconductor versions stay reachable under their
## namespace. `count` is the one that bites silently -- MatrixGenerics::count (via ComplexHeatmap ->
## matrixStats) returns something that is not a vector, so the failure surfaces several calls later
## inside sort()/unique() rather than at the count() itself.
select <- dplyr::select
rename <- dplyr::rename
first <- dplyr::first
desc <- dplyr::desc
count <- dplyr::count
filter <- dplyr::filter
slice <- dplyr::slice
reduce <- purrr::reduce # GenomicRanges::reduce otherwise
intersect <- base::intersect # BiocGenerics::intersect otherwise
setdiff <- base::setdiff
union <- base::union
paste <- base::paste # BiocGenerics::paste dispatches S4 and mangles collapse=
sort <- base::sort
unique <- base::unique
table <- base::table
```

## Parameters

```{r}
#| label: params
## Same base_dir as snrna_seurat_cellbender_preprocess.qmd -- keep them equal or the cell names this
## notebook writes will not join onto the object.
base_dir <- "~/nest/snRNA/210607_snrna"

## The six libraries the preprocessing notebook uses, in the same order: the cell-name suffix is the
## 1-based position in this vector (via aggr.csv, below), so reordering it silently relabels every
## cell. nc_run2 is in aggr.csv but is deliberately not processed.
samples_to_process <- c(
  "ra_run1", "arco_run1", "hvc_run1", "nc_run1",
  "lman_run1", "nr_run1"
)

## souporcell was run once per library at k=3 (see run_souporcell.sh), output straight into
## <sample>/souporcell rather than into a k-suffixed directory, so there is nothing to glob over --
## unlike dev-full1, where `souporcell_<k>` had to be discovered. If a reprocessing run at another k
## is ever added, point this at that directory name.
souporcell_subdir <- "souporcell"

## Genome label for readVcf. Only a label; it is not used to validate coordinates. Note that
## souporcell was run against GCF_005870125.1_lonStrDom2_genomic_only.fna, i.e. **NCBI** contig
## names (NC_/NW_), not the UCSC names the rest of this project uses. Variant ids are only ever
## compared between these VCFs, so no renaming is needed -- but do not expect the ids written below
## to join against anything UCSC-named without translation.
genome_label <- "lonStrDom2"

## The design, as measured rather than as assumed. This notebook was written expecting three birds
## dissected into all six regions (3 x 6 = 18 slots), which would put every bird in six libraries.
## The genotypes say otherwise, and the tables below are what say it:
##
##   ra / arco / hvc / nc  share three birds, matching each other at 0.03-0.08 against a
##                         different-bird background of 0.55-0.70;
##   lman / nr             share three *different* birds, matching each other at 0.04-0.13 and
##                         sitting at 0.55-0.65 against every cluster in the other four libraries --
##                         i.e. no bird in common with them at all.
##
## So there are six birds in two batches, each bird in four or two libraries. Nothing below depends
## on that being true: the threshold is chosen from a structural criterion that holds for any design
## (see the sweep chunk), and these numbers are only used to *report* whether the result matches
## expectation. Set them if a future run has a known design worth checking against.
n_libraries_expected <- length(samples_to_process)
MAX_COMPONENT <- n_libraries_expected # outer bound: a bird cannot be in more libraries than exist
N_BIRDS_EXPECTED <- 6L
K_EXPECTED <- 3L # souporcell was run at k=3 per library; asserted against the VCFs below

## Reads (AO + RO, from that cluster's own cells) required to believe a genotype call; below this the
## call becomes NA. souporcell emits a genotype wherever a cluster had any coverage at all and most
## calls are thin, so this is the parameter that decides whether there is individual structure in the
## matrix at all. stable settled on 10 for RNA and dev-full1 re-derived it; the depth sweep below
## derives it here rather than inheriting either. Set `min_depth_fixed` to override.
depth_grid <- c(0L, 3L, 5L, 10L, 20L, 30L)
min_depth_fixed <- NULL

## A depth cut that leaves too few comparable sites buys sharpness with noise: a distance over a
## handful of sites is not evidence. Floor the sweep's choice at this many shared called sites per
## library pair (median over pairs).
min_sites_floor <- 2000

## Candidate same-bird distance thresholds. stable's SAME_BIRD_MAX (0.15) and
## SAME_BIRD_MAX_COMPARE (0.22) sit inside this range but are not assumed.
## 0.01 steps rather than 0.02: the usable window here turned out to be only ~0.05 wide (bounded
## below by the last cluster to link and above by the related-birds band in batch 2), and a coarse
## grid reports that as three points, which understates how well resolved the boundaries are.
threshold_grid <- seq(0.02, 0.40, by = 0.01)

## Only 3 clusters x 6 libraries here, and each VCF is 380-680 MB. stable measured ~12 min for a
## 350 MB file, so budget ~15-25 min per library. load_gt_matrix(slim = TRUE) keeps a worker near
## 0.5 GB rather than several GB, so with only six libraries all of them fit at once and the read
## costs one library's wall time rather than six.
n_workers <- length(samples_to_process)
seed <- 10
set.seed(seed)

script_name <- "snrna_souporcell_clustering"
out_dir <- here::here("snrna/clustering", script_name)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

gt_fname <- file.path(out_dir, "gt_matrices.qs2")
depth_sweep_fname <- file.path(out_dir, "depth_sweep.qs2")
components_fname <- file.path(out_dir, "bird_components.qs2")
assignments_fname <- file.path(out_dir, "souporcell_assignments.csv")

## Per-stage redo flags: reading the VCFs costs ~1 h, the depth sweep a few minutes on top.
redo <- list(gt = FALSE, sweep = FALSE)
```

## Libraries and cell-name indices

```{r}
#| label: discover
## The cell-name suffix. snrna_seurat_cellbender_preprocess.qmd builds it as
## `which(aggr_csv$sample_id == sample_name)` *after* reordering aggr.csv to samples_to_process, so
## it is just the position in samples_to_process -- but read aggr.csv and reproduce the expression
## rather than assuming, so the two notebooks cannot drift apart silently.
aggr_csv <- read_csv(file.path(base_dir, "aggr.csv"), show_col_types = FALSE)
aggr_csv <- aggr_csv[match(samples_to_process, aggr_csv$sample_id), ]
stopifnot(
  "samples_to_process contains an id absent from aggr.csv" = !any(is.na(aggr_csv$sample_id))
)

samples <- tibble(
  sample_id = samples_to_process,
  sample_index = map_int(samples_to_process, ~ which(aggr_csv$sample_id == .x)),
  run_dir = file.path(base_dir, samples_to_process, souporcell_subdir)
) %>%
  mutate(
    vcf = file.path(run_dir, "cluster_genotypes.vcf"),
    tsv = file.path(run_dir, "clusters.tsv"),
    complete = file.exists(vcf) & file.exists(tsv)
  )

samples %>%
  select(sample_id, sample_index, complete, run_dir) %>%
  kable(caption = "souporcell runs. sample_index is the cell-name suffix in the Seurat object.")

stopifnot(
  "some souporcell runs are missing cluster_genotypes.vcf or clusters.tsv" = all(samples$complete)
)
run_dirs <- set_names(samples$run_dir, samples$sample_id)
```

## Run quality, before any genotype work

```{r}
#| label: run-qc
## Ambient RNA is the single best predictor of how well a library demultiplexes (stable measured
## Spearman -0.92 against its matching score across 8 libraries; doublet rate correlates the wrong
## way). Read it before trusting any library's genotypes. These six sit at 6.5-11.4%, i.e. all
## comfortably better than stable's worst (41.6%).
ambient <- souporcell_ambient(run_dirs)
sizes <- souporcell_cluster_sizes(run_dirs)

left_join(ambient, sizes, by = "sample_short") %>%
  rename(sample_id = sample_short) %>%
  arrange(desc(ambient_pct)) %>%
  kable(digits = 3, caption = paste(
    "Per-library run quality. n_clusters_seen below 3 would mean souporcell left a cluster empty;",
    "size_ratio above ~20 flags a badly unbalanced run."
  ))
```

## Load genotype matrices

```{r}
#| label: load-gt
## list(gt, depth) per library. Depth is cached rather than applied, so min_depth can be retuned
## without re-reading 6 x ~500 MB of VCF. Caches predating AO/RO hold a bare matrix -- treat those
## as stale rather than letting `$gt` return NULL downstream.
gt_raw <- NULL
if (!redo$gt && file.exists(gt_fname)) {
  gt_raw <- qs_read(gt_fname, nthreads = 8)
  if (!all(map_lgl(gt_raw, ~ is.list(.x) && all(c("gt", "depth") %in% names(.x))))) {
    message("Cached genotypes predate depth filtering (no AO/RO) -- re-reading the VCFs.")
    gt_raw <- NULL
  }
}
if (is.null(gt_raw)) {
  ## mclapply (fork), not furrr/future: `load_gt_matrix` is sourced into the global environment, so
  ## future's globals scan resolves its free variables against that environment and priced the
  ## export at 1.2 GiB per worker -- it reported base functions like `t` and `nrow` at 230 MiB each,
  ## i.e. it was measuring the enclosing environment rather than the function. Forking copies the
  ## session lazily instead of serialising anything, which is both correct here and free. Reading a
  ## VCF is deterministic, so nothing is lost by giving up furrr's RNG handling.
  gt_raw <- mclapply(
    set_names(samples$vcf, samples$sample_id),
    load_gt_matrix,
    genome = genome_label,
    mc.cores = n_workers, mc.preschedule = FALSE
  )
  failed <- map_lgl(gt_raw, ~ inherits(.x, "try-error") || is.null(.x))
  if (any(failed)) {
    stop(
      "Failed to read ", sum(failed), " VCF(s): ", paste(names(gt_raw)[failed], collapse = ", "),
      "\n", paste(map_chr(gt_raw[failed], as.character), collapse = "\n"),
      call. = FALSE
    )
  }
  qs_save(gt_raw, gt_fname, nthreads = 8)
}

## First hard check, and it costs nothing: k genotype rows per library. souporcell emits a row even
## for a cluster it left empty, so a shortfall here means the VCF is truncated, not that a cluster
## was unused.
imap_dfr(gt_raw, ~ tibble(sample_id = .y, n_clusters = nrow(.x$gt), n_variants = ncol(.x$gt))) %>%
  kable(caption = "Raw genotype matrices (clusters x variants)")

stopifnot(
  "a library does not have K_EXPECTED genotype rows" =
    all(map_int(gt_raw, ~ nrow(.x$gt)) == K_EXPECTED)
)

depth_summary(gt_raw, thresholds = c(1, 3, 5, 10, 20, 30)) %>%
  rename(sample_id = sample_short) %>%
  kable(digits = 3, caption = paste(
    "Read support behind each cluster's genotype calls.",
    "If frac_ge_10 is far below stable's 13-26%, min_depth needs lowering here."
  ))
```

## Choose min_depth

```{r}
#| label: depth-sweep
## dev-full1 could sweep min_depth against labelled same-bird pairs. With no truth sheet, use the
## label-free criterion stable used instead: the ratio of the *matched* cluster distance to the
## all-pairs distance. Matching is 1:1 and reference-free, and a depth filter is paying its way
## while that ratio keeps falling -- the matched distance is dominated by same-bird comparisons and
## the all-pairs median by different-bird ones, so the ratio is a separation measure that needs no
## labels. Read it against n_sites: sharpening distances by discarding the sites they rest on is
## not progress, which is what min_sites_floor guards.
##
## The pairwise pass is cheap at this scale (15 library pairs x 3 x 3 cells), so unlike dev-full1
## this sweep recomputes the distances at every depth rather than only counting surviving sites.
## Masking is one pass over the cached depth matrix, so only the distances are cached and the
## masked matrices are rebuilt for whichever depth is chosen. Keeping all six depths' matrices
## would cost ~1 GB in memory and on disk to save a few seconds of recomputation.
masked_at <- function(d) {
  imap(gt_raw, function(x, nm) {
    m <- mask_low_depth(x$gt, x$depth, min_depth = d)
    rownames(m) <- paste0(nm, "_cl", rownames(m))
    m
  })
}

sweep <- NULL
if (!redo$sweep && file.exists(depth_sweep_fname)) sweep <- qs_read(depth_sweep_fname, nthreads = 8)
if (is.null(sweep)) {
  sweep <- map(set_names(depth_grid, depth_grid), function(d) {
    masked <- masked_at(d)
    list(
      min_depth = d,
      frac_called = mean(map_dbl(masked, ~ mean(!is.na(.x)))),
      pair_dists = cluster_pair_distances(masked)
    )
  })
  qs_save(sweep, depth_sweep_fname, nthreads = 8)
}

## The ratio has to be computed over library pairs that actually *share* birds. Eight of the fifteen
## pairs here share none, and on those the Hungarian "matched" distance is just another
## different-bird distance, so including them pins the ratio near 1 (0.99 -> 0.95 across the whole
## depth grid) and hides the real effect. Which pairs share birds is not known before the threshold
## is chosen, so identify them scale-free instead: a pair with genuine structure has a minimum
## distance far below its own median, and a pair without has min ~ median. The `sharing` count is
## worth reading in its own right -- it is 7 at every depth, which is the batch structure showing up
## independently of any threshold.
sharing_pairs <- function(pair_dists, factor = 0.5) {
  keep <- map_lgl(pair_dists, function(x) {
    d <- x$dist[!is.na(x$dist)]
    length(d) > 0 && min(d) < factor * median(d)
  })
  pair_dists[keep]
}

depth_trade <- map_dfr(sweep, function(s) {
  shared_pd <- sharing_pairs(s$pair_dists)
  ms_all <- pair_match_scores(s$pair_dists, threshold = SAME_BIRD_MAX)
  ms <- if (length(shared_pd) > 0) pair_match_scores(shared_pd, threshold = SAME_BIRD_MAX) else ms_all
  tibble(
    min_depth = s$min_depth,
    frac_called = round(s$frac_called, 4),
    median_sites = median(map_dbl(s$pair_dists, ~ median(.x$n_sites, na.rm = TRUE))),
    n_sharing_pairs = length(shared_pd),
    matched_median = median(ms$matched_median),
    all_pairs_median = median(ms$all_pairs_median),
    ratio = round(median(ms$matched_median) / median(ms$all_pairs_median), 3),
    ratio_all_pairs = round(median(ms_all$matched_median) / median(ms_all$all_pairs_median), 3)
  )
})
kable(depth_trade, digits = 3, caption = paste(
  "What each depth cut buys. `ratio` is matched / all-pairs genotype distance over the library",
  "pairs that share birds -- lower is better separated. stable measured 0.668 unfiltered falling to",
  "0.427 at depth 20. `ratio_all_pairs` is the same quantity over all 15 pairs, shown to make the",
  "dilution visible rather than to be used."
))
```

```{r}
#| label: choose-depth
## Deepest cut that still leaves enough comparable sites, then the best separation among those.
## Derived rather than typed in, so it moves if the run flavour or reference changes; override with
## min_depth_fixed if a manual choice is ever wanted.
usable_depth <- depth_trade %>% filter(median_sites >= min_sites_floor)
min_depth <- if (!is.null(min_depth_fixed)) {
  as.integer(min_depth_fixed)
} else if (nrow(usable_depth) > 0) {
  as.integer(usable_depth$min_depth[which.min(usable_depth$ratio)])
} else {
  warning(
    "No depth cut leaves >= ", min_sites_floor, " median shared sites; falling back to min_depth = 0.",
    " The genotypes are thin -- read the separation table before trusting anything below.",
    call. = FALSE
  )
  0L
}
min_depth
```

```{r}
#| label: mask-depth
## Cluster ids carry their provenance: <sample_id>_cl<cluster>. cluster_pair_distances() and
## bird_components() both key on these, and they are what joins back to clusters.tsv.
gt_mats <- masked_at(min_depth)
## Reuse the sweep's distances when the chosen depth is one it visited, which it always is unless
## min_depth_fixed was set off-grid.
pair_dists <- if (!is.null(sweep[[as.character(min_depth)]])) {
  sweep[[as.character(min_depth)]]$pair_dists
} else {
  cluster_pair_distances(gt_mats)
}

imap_dfr(gt_mats, ~ tibble(
  sample_id = .y, n_clusters = nrow(.x), n_variants = ncol(.x),
  frac_called_raw = round(mean(!is.na(gt_raw[[.y]]$gt)), 3),
  frac_called = round(mean(!is.na(.x)), 3)
)) %>%
  kable(caption = paste0(
    "Genotype matrices, before and after the depth >= ", min_depth, " filter"
  ))

length(pair_dists) # 15 library pairs
```

## Calibrate the same-bird threshold

```{r}
#| label: threshold-sweep
## With no labelled pairs, the structure of the partition is the calibration. The criterion is
## `disjoint`: **no component may contain two clusters from the same library.** A bird is one
## individual and souporcell assigns each cell to one cluster, so a bird cannot be two clusters of
## the same library -- if a component holds two, single linkage has chained two birds together
## through an intermediate distance. This replaces the size bound (`max_size <= n_libraries`) as the
## primary check, for two reasons:
##
##   - it needs no design. The size bound needs to know how many libraries each bird is in, and the
##     first version of this notebook asserted six, which the genotypes then contradicted;
##   - it is far tighter. Here `max_size <= 6` holds all the way to the top of the grid, so choosing
##     the largest "sane" threshold picked 0.40 -- deep inside the different-bird range -- and
##     produced a four-cluster component containing two lman and two nr clusters. Disjointness
##     rejects exactly that.
##
## `n_confident` from the 1:1 matching is shown alongside as an independent view of the same
## threshold, but note that the *ceiling* is not 45: a library pair can only match the birds it
## actually shares, and the two batches share none, so 3 per within-batch pair (7 pairs = 21) is
## the most that can legitimately be confirmed.
component_libraries_disjoint <- function(bc) {
  tab <- bc$membership %>%
    mutate(sample_id = sub("_cl[0-9]+$", "", cluster)) %>%
    count(component, sample_id)
  all(tab$n == 1)
}

threshold_table <- map_dfr(threshold_grid, function(t) {
  bc <- bird_components(pair_dists, threshold = t)
  ms <- pair_match_scores(pair_dists, threshold = t)
  tibble(
    threshold = t,
    n_components = bc$components,
    n_multi = sum(bc$sizes > 1),
    n_singletons = sum(bc$sizes == 1),
    max_size = max(bc$sizes),
    sizes = paste(bc$sizes, collapse = ","),
    n_confident = sum(ms$n_confident),
    disjoint = component_libraries_disjoint(bc),
    within_bound = max(bc$sizes) <= MAX_COMPONENT,
    usable = component_libraries_disjoint(bc) & sum(bc$sizes == 1) == 0
  )
})
kable(threshold_table, caption = paste0(
  "Threshold sweep. `disjoint`: no component holds two clusters from one library. `within_bound`:",
  " no component exceeds ", MAX_COMPONENT, " clusters, the number of libraries -- an outer bound",
  " only. `usable` is `disjoint` plus every cluster placed in some component (no singletons)."
))
```

```{r}
#| label: choose-threshold
## Largest usable threshold, and the plateau it sits in. Largest, not smallest: a lower value leaves
## clusters unlinked and penalises whichever method has the wider usable range. The *width* of the
## plateau is the real result -- a single qualifying threshold is a coincidence, a wide plateau over
## which the partition does not change at all is structure. So report both the range of usable
## thresholds and whether the partition is identical across it.
usable <- threshold_table %>% filter(usable)
plateau <- if (nrow(usable) > 0) range(usable$threshold) else c(NA_real_, NA_real_)

same_bird_max <- if (nrow(usable) > 0) {
  max(usable$threshold)
} else {
  ## Nothing is both disjoint and singleton-free. Either some library's genotypes are too thin to
  ## link at any threshold (singletons persist across the whole sweep -- lower min_depth or suspect
  ## that library) or birds are related closely enough that chaining begins before the last cluster
  ## links (disjoint fails early). Prefer disjointness over completeness in that case: a chained
  ## component is a wrong individual, an unlinked one is only a missing individual.
  disjoint_only <- threshold_table %>% filter(disjoint, n_multi > 0)
  warning(
    "No threshold both keeps components library-disjoint and places every cluster. Falling back to ",
    "the largest disjoint threshold; some clusters will carry no individual. Read the sweep table.",
    call. = FALSE
  )
  if (nrow(disjoint_only) > 0) max(disjoint_only$threshold) else SAME_BIRD_MAX
}

## Is the partition constant over the plateau, or merely the summary statistics?
partition_key <- function(t) {
  bird_components(pair_dists, threshold = t)$membership %>%
    group_by(component) %>%
    summarise(g = paste(sort(cluster), collapse = "+"), .groups = "drop") %>%
    pull(g) %>%
    sort() %>%
    paste(collapse = " | ")
}
plateau_stable <- if (nrow(usable) > 1) {
  length(unique(map_chr(usable$threshold, partition_key))) == 1
} else {
  NA
}

tibble(
  same_bird_max = same_bird_max,
  plateau_min = plateau[1], plateau_max = plateau[2],
  n_usable_thresholds = nrow(usable),
  partition_identical_across_plateau = plateau_stable,
  stable_same_bird_max = SAME_BIRD_MAX, stable_compare = SAME_BIRD_MAX_COMPARE
) %>%
  kable(caption = paste(
    "Chosen threshold and the range over which the partition is usable.",
    "stable's constants are shown for orientation only -- a threshold is not portable between",
    "datasets, let alone between methods."
  ))
```

## Genotype-derived individuals

```{r}
#| label: components
components <- bird_components(pair_dists, threshold = same_bird_max)

## Name the components. With no truth sheet there are no bird IDs to recover, so the labels are
## arbitrary but must be *stable*: order components by the library order of their first cluster, so
## re-running produces the same bird1/bird2/bird3 and any metadata attached by hand downstream keeps
## meaning something. Singletons (only present in the fallback path) are named unassigned_<n>.
lib_order <- set_names(seq_along(samples_to_process), samples_to_process)
comp_named <- components$membership %>%
  mutate(
    sample_id = sub("_cl[0-9]+$", "", cluster),
    sc_cluster = sub(".*_cl", "", cluster)
  ) %>%
  group_by(component) %>%
  mutate(
    n_in_component = n(),
    first_lib = min(lib_order[sample_id]),
    first_cl = first(sc_cluster[order(lib_order[sample_id], sc_cluster)])
  ) %>%
  ungroup() %>%
  mutate(
    rank = dense_rank(paste(sprintf("%02d", first_lib), first_cl)),
    individual = if_else(n_in_component > 1, paste0("bird", rank), paste0("unassigned_", rank))
  )

qs_save(list(components = components, named = comp_named, same_bird_max = same_bird_max),
  components_fname,
  nthreads = 8
)

tibble(
  n_components = components$components,
  n_multi = sum(components$sizes > 1),
  n_singletons = sum(components$sizes == 1),
  max_size = max(components$sizes),
  n_birds_expected = N_BIRDS_EXPECTED
) %>%
  kable(caption = paste0("Components at ", same_bird_max))

comp_named %>%
  select(individual, cluster, sample_id, sc_cluster, n_in_component) %>%
  arrange(individual, lib_order[sample_id]) %>%
  kable(caption = "Which souporcell cluster in which library is which bird")

## The same thing as a grid: one row per library, one column per bird, holding that bird's souporcell
## cluster index in that library. This is the table to paste into notes -- it is the whole point of
## the notebook. Each row should be a permutation of 0/1/2 over the birds that library contains, and
## an empty cell means that bird is not in that library.
cluster_grid <- comp_named %>%
  filter(n_in_component > 1) %>%
  select(sample_id, individual, sc_cluster) %>%
  pivot_wider(names_from = individual, values_from = sc_cluster) %>%
  arrange(lib_order[sample_id])

kable(cluster_grid, caption = "souporcell cluster index per library per bird (blank = bird absent)")

stopifnot(
  "a bird appears twice in one library -- the partition is chained, not individual" =
    !any(map_lgl(cluster_grid, ~ any(lengths(.x) > 1)))
)
```

```{r}
#| label: batch-structure
## Which libraries share birds? This is the design, read off the genotypes rather than assumed, and
## it is where this dataset turned out not to be what the notebook was first written for. Entry
## (i, j) is the number of birds libraries i and j have in common; the diagonal is how many birds
## each library holds.
birds_by_lib <- comp_named %>%
  filter(n_in_component > 1) %>%
  select(sample_id, individual)

overlap <- outer(samples_to_process, samples_to_process, Vectorize(function(a, b) {
  length(intersect(
    birds_by_lib$individual[birds_by_lib$sample_id == a],
    birds_by_lib$individual[birds_by_lib$sample_id == b]
  ))
}))
dimnames(overlap) <- list(samples_to_process, samples_to_process)
kable(overlap, caption = "Birds shared between each pair of libraries (diagonal = birds per library)")

## Libraries that share no bird with each other partition into batches. Reported rather than
## assumed: a zero off-diagonal block is a statement about the dissection, and it means no
## individual-level comparison can be made across that block -- any downstream model with a bird
## effect can only be fit within a batch.
batch_of <- {
  seen <- set_names(rep(NA_integer_, length(samples_to_process)), samples_to_process)
  b <- 0L
  for (lib in samples_to_process) {
    if (!is.na(seen[[lib]])) next
    b <- b + 1L
    grp <- lib
    repeat {
      more <- samples_to_process[rowSums(overlap[, grp, drop = FALSE] > 0) > 0]
      if (setequal(more, grp)) break
      grp <- union(grp, more)
    }
    seen[grp] <- b
  }
  seen
}

tibble(sample_id = names(batch_of), batch = unname(batch_of)) %>%
  left_join(count(birds_by_lib, sample_id, name = "n_birds"), by = "sample_id") %>%
  arrange(batch, lib_order[sample_id]) %>%
  kable(caption = paste(
    "Batches of libraries that share individuals. Libraries in different batches have no bird in",
    "common, so `individual` is only comparable within a batch."
  ))

tibble(
  n_batches = max(batch_of),
  n_birds_total = n_distinct(birds_by_lib$individual),
  n_birds_expected = N_BIRDS_EXPECTED,
  birds_per_library = paste(sort(unique(count(birds_by_lib, sample_id)$n)), collapse = ","),
  libraries_per_bird = paste(sort(unique(count(birds_by_lib, individual)$n)), collapse = ",")
) %>%
  kable(caption = "Design as measured")
```

```{r}
#| label: separation
## How well separated is the partition that was chosen? Within-component distances are same-bird
## comparisons and between-component ones are different-bird, so this is the same-bird /
## different-bird distance distribution the truth sheet would have given -- derived from the
## partition instead of from labels. It is not independent evidence for the partition (it was
## thresholded into existence), but a *narrow* gap here means the threshold sat in noise, and the
## plateau width above is then the thing to trust.
comp_of <- set_names(comp_named$component, comp_named$cluster)

pair_long <- map_dfr(pair_dists, function(x) {
  d <- as.data.frame(as.table(x$dist), stringsAsFactors = FALSE)
  names(d) <- c("cluster_a", "cluster_b", "distance")
  d$n_sites <- as.data.frame(as.table(x$n_sites), stringsAsFactors = FALSE)$Freq
  d
}) %>%
  filter(!is.na(distance)) %>%
  mutate(same_bird = comp_of[cluster_a] == comp_of[cluster_b])

pair_long %>%
  group_by(same_bird) %>%
  summarise(
    n = n(), min = round(min(distance), 3), q05 = round(quantile(distance, .05), 3),
    median = round(median(distance), 3), q95 = round(quantile(distance, .95), 3),
    max = round(max(distance), 3), median_sites = median(n_sites), .groups = "drop"
  ) %>%
  kable(caption = paste(
    "Cross-library genotype distance within vs between the derived components.",
    "A usable partition needs a clear gap between the same-bird max and the different-bird min."
  ))

tibble(
  same_bird_max_observed = round(max(pair_long$distance[pair_long$same_bird]), 3),
  diff_bird_min_observed = round(min(pair_long$distance[!pair_long$same_bird]), 3)
) %>%
  mutate(gap = round(diff_bird_min_observed - same_bird_max_observed, 3)) %>%
  kable(caption = "The gap the threshold sits in. Negative means the two distributions overlap.")
```

```{r}
#| label: matching-crosscheck
## Independent cross-check: match each library pair's three clusters 1:1 by total genotype distance
## (Hungarian) and ask whether that agrees with the component partition. Two methods with different
## failure modes -- matching always returns something and can be fooled by a bird that is absent;
## single linkage can chain two birds together -- so agreement is a stronger statement than either
## alone.
##
## **Only within-batch pairs are meaningful.** solve_LSAP will happily 1:1 match two libraries that
## share no bird at all, which is precisely the trap dev-full1's incomplete design fell into, and
## this dataset has 8 such pairs (every ra/arco/hvc/nc x lman/nr combination). Those matchings are
## reported separately rather than counted as failures: a cross-batch match landing at
## different-bird distance is the correct outcome, not a disagreement.
match_tbl <- map_dfr(pair_dists, function(x) {
  oriented_match(x$dist)$match %>%
    mutate(
      pair = paste(x$libs, collapse = "-"),
      lib_a = x$libs[1], lib_b = x$libs[2],
      same_batch = batch_of[[x$libs[1]]] == batch_of[[x$libs[2]]],
      component_agrees = comp_of[row] == comp_of[matched]
    )
})

match_tbl %>%
  group_by(same_batch) %>%
  summarise(
    n_matchings = n(),
    n_agree_with_components = sum(component_agrees, na.rm = TRUE),
    n_below_threshold = sum(distance < same_bird_max, na.rm = TRUE),
    matched_median = round(median(distance, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  kable(caption = paste(
    "1:1 matching vs the component partition, split by whether the two libraries share birds.",
    "Within batch: all three columns should equal n_matchings. Across batch: n_below_threshold",
    "should be 0 and matched_median should sit at the different-bird distance -- those matchings",
    "are artefacts of forcing a 1:1 assignment where there is nothing to assign."
  ))

match_tbl %>%
  filter(same_batch, !component_agrees | distance >= same_bird_max) %>%
  select(pair, row, matched, distance, margin, outbid, component_agrees) %>%
  kable(digits = 3, caption = paste(
    "Within-batch matchings that disagree with the partition or sit above the threshold.",
    "Read `distance` before `margin`: solve_LSAP minimises the total, so a row can legitimately",
    "lose its closest column to a better claimant (negative margin, outbid TRUE) with nothing wrong."
  ))

pair_batch <- match_tbl %>%
  distinct(pair, same_batch) %>%
  deframe()

pair_match_scores(pair_dists, threshold = same_bird_max) %>%
  mutate(same_batch = pair_batch[pair]) %>%
  arrange(desc(same_batch), pair) %>%
  kable(digits = 3, caption = paste(
    "Per library pair. n_confident should be 3 of 3 for the 7 within-batch pairs and 0 for the 8",
    "across-batch pairs -- n_possible is min(rows, cols) and so overstates the ceiling for those."
  ))

## Not library_match_scores(): it sums n_confident over all five of a library's pairings against a
## ceiling of 5 x 3, which is unreachable here because two or four of those pairings are with
## libraries that share no bird. Sum over within-batch pairings only, so the number is comparable
## between libraries and against the ambient column in the run-quality table (stable measured
## Spearman -0.92 between the two).
## pivot_longer duplicates each matching once per participating library, which is what is wanted:
## a pair contributes its confident matches to both of its libraries' totals.
match_tbl %>%
  filter(same_batch) %>%
  pivot_longer(c(lib_a, lib_b), values_to = "sample_id") %>%
  group_by(sample_id) %>%
  summarise(
    n_within_batch_pairings = n_distinct(pair),
    confident = sum(distance < same_bird_max, na.rm = TRUE),
    of = K_EXPECTED * n_distinct(pair),
    .groups = "drop"
  ) %>%
  left_join(ambient %>% rename(sample_id = sample_short), by = "sample_id") %>%
  arrange(desc(confident)) %>%
  kable(digits = 2, caption = paste(
    "Confident matches per library, over its within-batch pairings only, beside its ambient-RNA",
    "estimate. The ceiling is 3 x n_within_batch_pairings."
  ))
```

```{r}
#| label: cluster-support
## Per cluster: how many other libraries hold a counterpart at same-bird distance, and how close the
## best one is. This is where the width of the usable threshold window comes from -- the window is
## bounded below by whichever cluster links last, so a cluster with a conspicuously large
## `median_best` (or a `best_distance` near the chosen threshold rather than far below it) is the
## one making the calibration tight, and the one to look at if the window ever closes entirely.
best_per_cluster <- map_dfr(pair_dists, function(x) {
  bind_rows(
    tibble(cluster = rownames(x$dist), best = suppressWarnings(apply(x$dist, 1, min, na.rm = TRUE))),
    tibble(cluster = colnames(x$dist), best = suppressWarnings(apply(x$dist, 2, min, na.rm = TRUE)))
  )
}) %>%
  filter(is.finite(best)) %>%
  group_by(cluster) %>%
  summarise(best_distance = round(min(best), 3), .groups = "drop")

## cluster_support()'s `row` is already the full `<library>_cl<n>` id, because the matrices were
## labelled before cluster_pair_distances(); `lib` is that cluster's own library either way round.
cluster_support(pair_dists, threshold = same_bird_max) %>%
  rename(cluster = row) %>%
  left_join(best_per_cluster, by = "cluster") %>%
  left_join(comp_named %>% select(cluster, individual), by = "cluster") %>%
  select(cluster, individual, n_confident, best_distance, median_best) %>%
  arrange(desc(best_distance)) %>%
  kable(digits = 3, caption = paste0(
    "Cross-library support per cluster, worst first. n_confident is how many other libraries hold a",
    " counterpart below ", same_bird_max, "; for this design the ceiling is 3 for a batch-1 cluster",
    " and 1 for a batch-2 cluster."
  ))
```

```{r}
#| label: within-library
## Corroboration for k=3: if souporcell split one bird into two clusters, those two should sit at
## same-bird distance. Weak evidence on its own -- souporcell's model pushes its own clusters apart
## by construction -- so read it beside the component sizes, which are the real check on k.
within_library_distances(gt_mats) %>%
  rename(sample_id = sample_short) %>%
  kable(caption = paste0(
    "Distances between a library's own clusters. Values near the same-bird range (< ",
    same_bird_max, ") would suggest k = 3 oversplit that library."
  ))
```

```{r}
#| label: heatmap-gt
#| fig-width: 10
#| fig-height: 6
## Genotypes of all 18 clusters at the most variable shared sites, split by derived bird. Clusters
## of the same bird should sit together; if they do not, the matching is wrong. Not "sites with no
## missing call": no variant is called in every cluster row, so that selection is empty and the
## heatmap draws nothing.
shared <- reduce(map(gt_mats, colnames), intersect)
gt_all <- do.call(rbind, map(gt_mats, ~ .x[, shared, drop = FALSE]))
keep <- colMeans(is.na(gt_all)) <= 0.05
gt_hm <- gt_all[, keep, drop = FALSE]
vars <- apply(gt_hm, 2, var, na.rm = TRUE)
gt_hm <- gt_hm[, order(vars, decreasing = TRUE)[seq_len(min(2000, sum(!is.na(vars))))], drop = FALSE]

Heatmap(gt_hm,
  name = "genotype",
  row_split = comp_named$individual[match(rownames(gt_hm), comp_named$cluster)],
  cluster_row_slices = FALSE, show_column_names = FALSE, row_names_gp = gpar(fontsize = 8),
  column_title = sprintf("%s shared sites, max 5%% missing", ncol(gt_hm))
)
```

## Cell-level assignments

```{r}
#| label: assign-cells
## Attach individual labels to barcodes. Cell names are built exactly as in
## snrna_seurat_cellbender_preprocess.qmd: CellRanger's `-1` suffix replaced by the library's
## sample_index, then an underscore prefixed (Merge_Seurat_List leaves that prefix on every cell).
## Get this wrong and the join in the preprocessing notebook silently produces all-NA metadata, so
## it is asserted against a sanity check below rather than trusted.
assignments <- samples %>%
  select(sample_id, sample_index, tsv) %>%
  mutate(cl = map(tsv, ~ read_tsv(.x, show_col_types = FALSE, progress = FALSE))) %>%
  select(-tsv) %>%
  unnest(cl) %>%
  mutate(
    sc_cluster = as.character(assignment),
    cluster = paste0(sample_id, "_cl", sc_cluster),
    ## Strip the suffix and paste the index on, rather than sub()-ing one for the other: sub() is
    ## not vectorised over `replacement`, so `sub("-1$", paste0("-", sample_index), barcode)` warns
    ## and silently uses the *first* library's index for all six, giving six copies of every cell
    ## name. The preprocessing notebook's own str_interp() version is safe only because it runs
    ## inside a per-sample function where sample_index is scalar.
    cell = paste0("_", sub("-1$", "", barcode), "-", sample_index),
    doublet_status = status
  ) %>%
  left_join(comp_named %>% select(cluster, component, individual), by = "cluster") %>%
  ## `batch` travels with the cells because it constrains every downstream comparison: birds are
  ## only comparable within a batch, so a model with a bird term needs it. It is a property of the
  ## library, not of the cell, but carrying it here saves the preprocessing notebook from
  ## re-deriving it.
  left_join(tibble(sample_id = names(batch_of), batch = unname(batch_of)), by = "sample_id") %>%
  ## souporcell's own doublet/unassigned calls: only singlets get an individual. `assignment` for a
  ## doublet is a pair of cluster indices ("0/2"), which is why it must not be parsed as an
  ## individual -- the old preprocessing block kept those strings as a level of `assignment`.
  mutate(individual = if_else(doublet_status == "singlet", individual, NA_character_)) %>%
  select(
    cell, barcode, sample_id, sample_index, sc_cluster, cluster,
    doublet_status, component, individual, batch
  )

stopifnot(
  "duplicate cell names -- the library index did not make it into every cell name" =
    !any(duplicated(assignments$cell))
)

assignments %>%
  count(sample_id, doublet_status) %>%
  pivot_wider(names_from = doublet_status, values_from = n, values_fill = 0) %>%
  kable(caption = "souporcell status per library")

assignments %>%
  filter(doublet_status == "singlet") %>%
  count(sample_id, individual) %>%
  pivot_wider(names_from = individual, values_from = n, values_fill = 0) %>%
  arrange(lib_order[sample_id]) %>%
  kable(caption = "Singlet nuclei per bird per library -- the table the old `assignment` could not give")

assignments %>%
  summarise(
    n_cells = n(),
    n_singlet = sum(doublet_status == "singlet"),
    n_with_individual = sum(!is.na(individual)),
    pct_singlets_assigned = round(
      100 * sum(!is.na(individual)) / sum(doublet_status == "singlet"), 1
    )
  ) %>%
  kable(caption = "Coverage of the genotype-derived assignment")

write_csv(assignments, assignments_fname)
message("Wrote ", assignments_fname)
```

`snrna_seurat_cellbender_preprocess.qmd` reads this file in its *Load souporcell* chunk and adds
`individual`, `batch`, `doublet_status`, `souporcell_cluster` and `souporcell_component` to the
object, and uses the per-library doublet rates below to set DoubletFinder's `nExp`. Check the per-bird
cell counts and the matching cross-check above before re-running it: everything downstream that
grouped by `assignment` now groups by `individual`, and any comparison across birds has to respect
`batch`.

## Session info

```{r}
#| label: session-info
sessionInfo()
```
