---
title: "Xenium ROI polygons — segmentation-independent region definitions"
format: html
editor: source
---

The collaborator's Xenium Explorer selections are **lists of cell IDs**, not
geometry. Cell IDs are a property of a particular segmentation run, so any
resegmentation (Proseg, Baysor, 10x multimodal) invalidates them — this is
exactly why the `1um_*` selection files, which came from a 1 µm expansion run,
match essentially no cells in the delivered 5 µm nucleus-expansion outputs.

This script converts each usable selection into a **polygon** in micron
coordinates, by taking the concave hull of the selected cells' centroids. The
polygon is segmentation-independent: it can be applied to any future
segmentation by point-in-polygon test on the new centroids.

Run this **before** resegmenting.

```{r}
library(tidyverse)
library(qs2)
library(sp)          # point.in.polygon; avoids an sf/GDAL system dependency
library(cowplot)
library(here)

source(here::here("config/paths.R"))
select = dplyr::select
```

```{r}
script_name = "roi_polygons"
out_dir = here::here("xenium/preprocessing", script_name)
dir.create(out_dir, recursive = T, showWarnings = F)

## The selections are convex by construction — hand-drawn outlines around a
## single nucleus, and 1 mm rectangles — so a convex hull is sufficient and
## avoids an sf/GDAL dependency. The recovery check below is what tells us
## whether that assumption holds: `extra` counts cells the hull admits that the
## original selection did not, which is where a convex hull would show strain.

## The two usable selection families. `1um_*` files are excluded: different
## segmentation, cell IDs do not exist in this object.
roi_patterns = c(HVC = "^(orpu|oryw)_HVC_.*cells_stats.csv$",
                 box = "^HVC&Surround.*cells_stats.csv$")

seed = 10
set.seed(seed)
```

## Cell coordinates

```{r}
md = read_csv(here::here("xenium/preprocessing/xenium_preprocess/cell_metadata.csv.gz"),
              show_col_types = FALSE) %>%
  mutate(cell_id = str_remove(cell, "^.*_")) %>%
  select(cell, cell_id, section_id, x_centroid, y_centroid)

nrow(md)
```

## Selections

```{r}
read_selection = function(path) {
  hdr = readLines(path, n = 5)
  skip = max(c(0, which(str_starts(hdr, "#"))))
  d = suppressWarnings(read_csv(path, skip = skip, show_col_types = FALSE))
  if (!"Cell ID" %in% colnames(d)) return(character())
  d$`Cell ID`
}

roi_files = list.files(XENIUM_SELECTIONS_DIR, pattern = "cells_stats.csv$",
                       full.names = TRUE)

roi = imap_dfr(roi_patterns, function(pat, type) {
  f = roi_files[str_detect(basename(roi_files), pat)]
  map_dfr(f, ~ tibble(file = basename(.x), roi_type = type,
                      cell_id = read_selection(.x)))
})

roi %>% count(roi_type, file)
```

Each selection is assigned to the section that actually contains its cell IDs,
not to the section named in the file or in the `#Selection name` header — those
disagree with each other for several files.

```{r}
roi_xy = roi %>%
  inner_join(md, by = "cell_id", relationship = "many-to-many")

roi_section = roi_xy %>%
  count(file, roi_type, section_id) %>%
  group_by(file) %>%
  slice_max(n, n = 1) %>%
  ungroup()

roi_section %>% arrange(section_id, roi_type)
```

```{r}
roi_xy = roi_xy %>% semi_join(roi_section, by = c("file", "section_id"))
```

## Hulls

```{r}
hull_of = function(d) {
  pts = as.matrix(d[, c("x_centroid", "y_centroid")])
  i = grDevices::chull(pts)
  tibble(x = pts[i, 1], y = pts[i, 2])
}

polys = roi_xy %>%
  group_by(file, roi_type, section_id) %>%
  group_modify(~ hull_of(.x)) %>%
  ungroup()

polys %>% count(file, roi_type, section_id, name = "n_vertices")
write_csv(polys, file.path(out_dir, "roi_polygons.csv"))
```

Verification: re-select cells by point-in-polygon and compare with the original
cell ID list. Recovery should be near 1 — a low value means the hull is
distorted, usually because the selection is not simply connected.

```{r}
## Reusable: assign cells to an ROI polygon by point-in-polygon. This is the
## function to apply to any future segmentation's centroids.
cells_in_polygon = function(cells, poly) {
  inside = sp::point.in.polygon(cells$x_centroid, cells$y_centroid,
                                poly$x, poly$y) > 0
  cells$cell_id[inside]
}

recovery = polys %>%
  group_by(file, roi_type, section_id) %>%
  group_split() %>%
  map_dfr(function(p) {
    sid = p$section_id[1]; f = p$file[1]
    cells = md %>% filter(section_id == sid)
    inside = cells_in_polygon(cells, p)
    orig = roi_xy %>% filter(file == f) %>% pull(cell_id)
    tibble(file = f, section_id = sid,
           n_original = length(orig), n_in_polygon = length(inside),
           recovered = mean(orig %in% inside),
           extra = mean(!inside %in% orig))
  })

recovery %>% arrange(recovered)
write_csv(recovery, file.path(out_dir, "roi_polygon_recovery.csv"))
```

```{r}
for (sid in unique(polys$section_id)) {
  p = ggplot() +
    geom_point(data = md %>% filter(section_id == sid) %>% slice_sample(n = 3e4),
               aes(x_centroid, y_centroid), size = 0.1, colour = "grey85") +
    geom_polygon(data = polys %>% filter(section_id == sid),
                 aes(x, y, colour = roi_type, group = file),
                 fill = NA, linewidth = 0.6) +
    coord_equal() + labs(title = sid) + theme_cowplot()
  ggsave(file.path(out_dir, paste0("roi_polygons_", sid, ".png")), p,
         width = 9, height = 9, dpi = 150, bg = "white")
}
```

```{r}
sessionInfo()
```
