---
title: "Xenium preprocessing — ten-section merged object"
format: html
editor: source
---

Builds the analysis object for the Xenium run `Brainard_KL_052905`: ten sections
(two birds × five coronal levels) loaded from the Xenium Ranger outputs, QC
filtered, merged into a single Seurat object with one field of view per section,
and normalized.

Objects are rebuilt from the raw outputs rather than taken from
`Xenium Objects/*.RData`, because the collaborator's section labels disagree
with the run metadata (see `xenium/README.md`); `region_name` from each
region's `experiment.xenium` is treated as authoritative.

Normalization matches `xenium/label_transfer/panel_resolution_benchmark.qmd`
(log-normalization scaled to the median transcript count) so that reference and
query are treated identically at label transfer.

```{r}
library(tidyverse)
library(Seurat)
library(qs2)
library(data.table)
library(harmony)
library(cowplot)
library(jsonlite)
library(future)
library(here)

source(here::here("config/paths.R"))
select = dplyr::select
```

```{r}
## Threading. ScaleData is future-aware; PCA and Harmony go through OpenBLAS.
## FindNeighbors (annoy) and FindClusters (Louvain) are single-threaded in
## Seurat with no thread argument — which is why the embedding runs on a
## subsample rather than on all ~1.6M cells.
n_threads = 24
RhpcBLASctl::blas_set_num_threads(n_threads)
plan("multicore", workers = n_threads)
```

```{r}
script_name = "xenium_preprocess"
out_dir = here::here("xenium/preprocessing", script_name)          # tables and figures
obj_dir = file.path(path.expand(XENIUM_PROC_DIR), script_name)     # large objects
dir.create(out_dir, recursive = T, showWarnings = F)
dir.create(obj_dir, recursive = T, showWarnings = F)

## QC thresholds. 20 counts follows the collaborator's script and the 10x
## guidance for a panel this size; cells below it cannot be typed reliably.
min_counts = 20
min_features = 10

## Embedding / clustering QC. Run on a stratified subsample: the graph steps are
## single-threaded and scale badly (>2 h on the full 1.6M cells), and these
## clusters are only a sanity check — cell identity comes from label transfer.
run_embedding = TRUE
embedding_n_cells = 1e5      # total, split evenly across the ten sections
npcs = 30
cluster_res = 0.5

## Anatomy-defining panel genes, plotted on *all* cells as the spatial check
## that does not depend on the subsample
## AQP4 is not on this panel; SLC1A3 is the astrocyte marker that is
marker_genes = c("SLC17A6", "GAD1", "PLP1", "SLC1A3", "PVALB", "UCHL1")

out_obj_fname = file.path(obj_dir, "obj_xenium.qs2")
out_qc_obj_fname = file.path(obj_dir, "obj_xenium_qc-subsample.qs2")

seed = 10
set.seed(seed)
options(future.globals.maxSize = Inf)
```

## Sections

```{r}
region_dirs = list.dirs(XENIUM_RAW_DIR, recursive = FALSE)
region_dirs = region_dirs[grepl("^output-XETG", basename(region_dirs))]

sections = map_dfr(region_dirs, function(d) {
  e = jsonlite::fromJSON(file.path(d, "experiment.xenium"))
  tibble(dir = d,
         bird = e$cassette_name,
         slide = e$slide_id,
         section = e$region_name,
         run = e$run_name,
         panel = e$panel_name,
         segmentation = if_else(e$segmented_cell_nuc_expansion_frac == 1,
                                "nucleus expansion", "mixed"),
         n_cells_reported = e$num_cells,
         analysis_sw = e$analysis_sw_version)
}) %>%
  mutate(section_id = paste(bird, section, sep = "_")) %>%
  arrange(bird, section)

sections %>% select(section_id, bird, slide, section, segmentation, n_cells_reported)
write_csv(sections %>% select(-dir), file.path(out_dir, "sections.csv"))
```

## Load

Molecule coordinates are not loaded — `transcripts.parquet` is ~2 GB per section
and is only needed for resegmentation or subcellular work. Segmentation polygons
are likewise left on disk; centroids carry the spatial information every planned
analysis uses.

```{r}
load_section = function(d, section_id) {
  x = LoadXenium(d, fov = section_id,
                 molecule.coordinates = FALSE,
                 cell.centroids = TRUE)

  ## Per-cell covariates LoadXenium does not import: areas and control counts.
  ## Piped through zcat so the read does not depend on R.utils.
  cells = fread(cmd = paste("zcat", shQuote(file.path(d, "cells.csv.gz"))))
  cells = as.data.frame(cells)
  rownames(cells) = cells$cell_id
  cells = cells[colnames(x), ]

  x$cell_area = cells$cell_area
  x$nucleus_area = cells$nucleus_area
  x$nucleus_count = cells$nucleus_count
  x$control_counts = cells$control_probe_counts + cells$control_codeword_counts +
    cells$genomic_control_counts + cells$unassigned_codeword_counts +
    cells$deprecated_codeword_counts
  x$control_frac = x$control_counts / pmax(cells$total_counts, 1)
  x$x_centroid = cells$x_centroid
  x$y_centroid = cells$y_centroid
  x
}

obj_list = pmap(list(sections$dir, sections$section_id), function(d, sid) {
  message("loading ", sid)
  x = load_section(d, sid)
  x$section_id = sid
  x$bird = sections$bird[sections$section_id == sid]
  x$slide = sections$slide[sections$section_id == sid]
  x$section = sections$section[sections$section_id == sid]
  x
})
names(obj_list) = sections$section_id

map_int(obj_list, ncol)
```

## QC

```{r}
qc_tbl = imap_dfr(obj_list, function(x, sid) {
  x@meta.data %>% as_tibble() %>% mutate(section_id = sid)
})

qc_summary = qc_tbl %>%
  group_by(section_id, bird, section) %>%
  summarize(n_cells = n(),
            median_counts = median(nCount_Xenium),
            median_features = median(nFeature_Xenium),
            median_cell_area = median(cell_area),
            median_nucleus_area = median(nucleus_area),
            mean_control_frac = mean(control_frac),
            n_zero_count = sum(nCount_Xenium == 0),
            n_missing_area = sum(is.na(cell_area)),
            frac_below_min_counts = mean(nCount_Xenium < min_counts),
            .groups = "drop")

qc_summary
write_csv(qc_summary, file.path(out_dir, "qc_summary_prefilter.csv"))
```

The control fraction is the readout that matters most here: it bounds how much
of each cell's signal is background rather than panel transcripts.

```{r}
## Empty cells (0 transcripts) are counted in qc_summary but excluded from the
## log-scaled panels, where they would become -Inf
qc_plot_tbl = qc_tbl %>% filter(nCount_Xenium > 0, !is.na(cell_area), cell_area > 0)

p1 = ggplot(qc_plot_tbl, aes(nCount_Xenium, section_id, fill = bird)) +
  ggridges::geom_density_ridges(alpha = 0.6, scale = 1.2) +
  scale_x_log10() + geom_vline(xintercept = min_counts, linetype = 2) +
  labs(x = "transcripts per cell", y = NULL) + theme_cowplot()

p2 = ggplot(qc_plot_tbl, aes(nFeature_Xenium, section_id, fill = bird)) +
  ggridges::geom_density_ridges(alpha = 0.6, scale = 1.2) +
  geom_vline(xintercept = min_features, linetype = 2) +
  labs(x = "genes per cell", y = NULL) + theme_cowplot()

p3 = ggplot(qc_plot_tbl, aes(cell_area, section_id, fill = bird)) +
  ggridges::geom_density_ridges(alpha = 0.6, scale = 1.2) +
  scale_x_log10() + labs(x = "cell area (um^2)", y = NULL) + theme_cowplot()

p4 = ggplot(qc_summary, aes(mean_control_frac, section_id, fill = bird)) +
  geom_col() + labs(x = "mean control fraction", y = NULL) + theme_cowplot()

p = plot_grid(p1, p2, p3, p4, nrow = 2)
ggsave(file.path(out_dir, "qc_distributions.pdf"), p, width = 16, height = 10)
p
```

```{r}
obj_list = map(obj_list, function(x) {
  keep = colnames(x)[x$nCount_Xenium >= min_counts & x$nFeature_Xenium >= min_features]
  subset(x, cells = keep)
})

retention = tibble(section_id = names(obj_list),
                   n_cells_prefilter = qc_summary$n_cells[match(names(obj_list), qc_summary$section_id)],
                   n_cells_postfilter = map_int(obj_list, ncol)) %>%
  mutate(frac_retained = n_cells_postfilter / n_cells_prefilter)

retention
write_csv(retention, file.path(out_dir, "qc_retention.csv"))
```

## Merge

One field of view per section, so each section keeps its own coordinate frame.

```{r}
obj = merge(obj_list[[1]], obj_list[-1], add.cell.ids = names(obj_list))

## Control assays are retained for QC but are not used downstream
DefaultAssay(obj) = "Xenium"

obj$section_id = factor(obj$section_id, levels = sections$section_id)
obj

rm(obj_list); gc()
```

## Normalize

```{r}
obj = JoinLayers(obj, assay = "Xenium")
scale_factor = median(obj$nCount_Xenium)
scale_factor

obj = NormalizeData(obj, assay = "Xenium", scale.factor = scale_factor, verbose = FALSE)

## All 425 panel genes are informative by design; no variable feature selection
VariableFeatures(obj) = rownames(obj[["Xenium"]])
```

## Embedding QC

Run on a stratified subsample. `FindNeighbors` and `FindClusters` are
single-threaded with no thread argument in Seurat, and take over two hours on
the full object; the clusters they produce are a sanity check that gets replaced
by transferred labels, so paying that cost buys nothing. `RunUMAP` is threaded
through uwot — note that `n_sgd_threads > 1` makes the layout non-reproducible
even under a fixed seed, which is acceptable for a QC figure.

```{r}
if (run_embedding) {
  set.seed(seed)
  qc_cells = obj@meta.data %>%
    rownames_to_column("cell") %>%
    group_by(section_id) %>%
    slice_sample(n = ceiling(embedding_n_cells / n_distinct(obj$section_id))) %>%
    pull(cell)

  obj_qc = subset(obj, cells = qc_cells)
  obj_qc = ScaleData(obj_qc, verbose = FALSE)
  obj_qc = RunPCA(obj_qc, npcs = npcs, verbose = FALSE)
  obj_qc = RunHarmony(obj_qc, group.by.vars = "section_id",
                      reduction.use = "pca", reduction.save = "harmony")
  obj_qc = RunUMAP(obj_qc, reduction = "harmony", dims = 1:npcs,
                   reduction.name = "umap", verbose = FALSE,
                   n_threads = n_threads, n_sgd_threads = n_threads)
  obj_qc = FindNeighbors(obj_qc, reduction = "harmony", dims = 1:npcs, verbose = FALSE)
  obj_qc = FindClusters(obj_qc, resolution = cluster_res, verbose = FALSE)
  obj_qc$cluster_qc = obj_qc$seurat_clusters

  ## Carry the QC clusters back onto the full object (NA outside the subsample)
  cl = setNames(rep(NA_character_, ncol(obj)), colnames(obj))
  cl[colnames(obj_qc)] = as.character(obj_qc$cluster_qc)
  obj$cluster_qc = cl

  ncol(obj_qc)
}
```

Clusters here are a QC readout only — cell type identity comes from label
transfer in `xenium/label_transfer/`, at the resolution the panel benchmark
showed to be supportable.

```{r}
if (run_embedding) {
  p = DimPlot(obj_qc, group.by = c("cluster_qc", "section_id", "bird"),
              raster = TRUE, raster.dpi = c(1200, 1200), ncol = 3) &
    theme(legend.position = "none")
  ggsave(file.path(out_dir, "umap_qc.png"), p, width = 18, height = 6, dpi = 200)
  p
}
```

Spatial coherence of the QC clusters is the check on the merge: anatomical
structure should be visible without position having been used in the clustering.

```{r}
if (run_embedding) {
  for (sid in levels(obj_qc$section_id)) {
    p = ImageDimPlot(obj_qc, fov = sid, group.by = "cluster_qc",
                     size = 0.8, cols = "polychrome", dark.background = FALSE) +
      ggtitle(sid)
    ggsave(file.path(out_dir, paste0("spatial_cluster_qc_", sid, ".png")),
           p, width = 10, height = 10, dpi = 200)
  }
}
```

The stronger anatomical check does not depend on the subsample: marker genes
plotted on every cell. `PLP1` should trace fibre tracts and delineate song
nucleus boundaries, and the glutamatergic/GABAergic split should follow known
telencephalic structure.

```{r}
marker_genes_present = intersect(marker_genes, rownames(obj[["Xenium"]]))
setdiff(marker_genes, marker_genes_present)

## ImageFeaturePlot has no ncol argument; combine = FALSE returns the panels
## and plot_grid lays them out
for (sid in levels(obj$section_id)) {
  ps = ImageFeaturePlot(obj, fov = sid, features = marker_genes_present,
                        size = 0.3, max.cutoff = "q95",
                        dark.background = FALSE, combine = FALSE)
  ps = map2(ps, marker_genes_present, ~ .x + ggtitle(.y))
  p = plot_grid(plotlist = ps, ncol = 3)
  ggsave(file.path(out_dir, paste0("spatial_markers_", sid, ".png")),
         p, width = 18, height = 12, dpi = 200, bg = "white")
}
```

```{r}
if (run_embedding) {
  comp = obj_qc@meta.data %>%
    count(section_id, cluster_qc) %>%
    group_by(section_id) %>%
    mutate(frac = n / sum(n)) %>%
    ungroup()

  ggplot(comp, aes(section_id, frac, fill = cluster_qc)) +
    geom_col() +
    labs(x = NULL, y = "fraction of cells", fill = "QC cluster") +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  ggsave(file.path(out_dir, "qc_cluster_composition.pdf"), width = 10, height = 6)
  write_csv(comp, file.path(out_dir, "qc_cluster_composition.csv"))
}
```

## Save

The full object carries counts, normalized data, spatial coordinates, and the QC
cluster labels for the subsampled cells; the subsample object additionally
carries the PCA/Harmony/UMAP reductions. `scale.data` is dropped — a dense
425 × n matrix, needed only for the PCA computed above, and cheap to regenerate.

```{r}
if (run_embedding) {
  obj_qc[["Xenium"]]$scale.data = NULL
  qs2::qs_save(obj_qc, out_qc_obj_fname)
}

qs2::qs_save(obj, out_obj_fname)
c(out_obj_fname, out_qc_obj_fname)

obj@meta.data %>%
  rownames_to_column("cell") %>%
  write_csv(file.path(out_dir, "cell_metadata.csv.gz"))
```

```{r}
sessionInfo()
```
