---
title: "Untitled"
format: html
---
```{r}
library(Seurat)
library(SeuratDisk)
library(scCustomize)
library(ArchR)
library(tidyverse)
library(qs)
library(cowplot)
library(viridis)
library(ComplexHeatmap)
library(circlize)
library(scales)
library(rhdf5)
library(hdf5r)
library(anndata)
select = dplyr::select

source("../../scripts/stats.R")
```

```{r}
scplus_dir = '/hdd/jupyter/brad/scenicplus/motor-pathway_multiome/motor-pathway_multiome_seurat_cellbender.0.05_preprocess_cr/'

scplus_subdirs <- list(glut_ra = "glut-ra_glut-arco-1_run2_seurat-snrna-clustering",
                       gaba = "ra-arco-hvc-nc_gaba_run2_seurat-snrna-clustering",
                       nn = "ra-arco-hvc-nc_astro-oligo_seurat-snrna-clustering",
                       glut = "ra-arco_glut_run2_seurat-snrna-clustering")

ereg_files <- map(scplus_subdirs, function(x) {
  ereg_file <-  list.files(file.path(scplus_dir, x, "scenicplus"), pattern = "scplusmdata.h5mu", recursive = T, full.names = T)
  ereg_file <- grep("motif_anno", ereg_file, invert = T, value=T)
  configs = str_extract(ereg_file, "config[0-9]+")
  names(ereg_file) = configs
  ereg_file
})


## snRNA
snrna_dir = "~/hdd/rstudio/snRNA/snrna_cellranger/snrna_seurat_cellbender.0.01_preprocess/"
#multiome_dir = "~/nest/rstudio/multiome/archr/motor-pathway"
snrna_file = file.path(snrna_dir, "obj_clustered.qs")

## Seurat object
in_dir = "~/ssd/rstudio/multiome/motor-pathway/seurat/motor-pathway_multiome_seurat_cellbender.0.05_preprocess_cr/"
obj_fname = file.path(in_dir, "obj_clustered.qs")


out_base = "~/ssd/rstudio/multiome/motor-pathway/scenicplus/"
script_name = "motor-pathway_scenicplus_v2_cluster-snrna-cr_run2_intersections"
out_dir = file.path(out_base, script_name)
dir.create(out_dir)

scRNA_file = file.path(out_dir, "scRNA.h5ad")
```

## Read anndata
```{r}
obj_mult = qread(obj_fname)
```

```{r}
labels_order = unique(obj_mult$cluster)
labels_order = c(sort(grep("HVC", labels_order, value=T)),
                 sort(grep("NC", labels_order, value=T)),
                 sort(grep("Nido", labels_order, value=T)),
                 sort(grep("RA", labels_order, value=T)),
                 sort(grep("Arco", labels_order, value=T)),
                 sort(grep("Glut-Pre", labels_order, value=T)),
                 sort(grep("GABA", labels_order, value=T)),
                 sort(grep("Astro", labels_order, value=T)),
                 sort(grep("Epen", labels_order, value=T)),
                 sort(grep("Oligo", labels_order, value=T)),
                 sort(grep("OPC", labels_order, value=T)),
                 sort(grep("Micro", labels_order, value=T)),
                 sort(grep("Endo", labels_order, value=T)))
labels_order = unique(labels_order)
```

## Average expression
```{r}
obj_mult_avg = log1p(AggregateExpression(obj_mult, group.by="cluster", assays = "CB")[[1]])
```


## Extract other data
```{r}
convert_category_to_factor = function(x) {
  factor(x$codes, labels = x$categories, levels=0:(length(x$categories)-1))
}

extract_matrix = function(ereg_file, obs_name) {
  cur = rhdf5::h5read(ereg_file, name = str_interp(obs_name)) 
  rnames = cur[["var"]]$`_index`
  cnames = cur[["obs"]]$Cell
  mat = data.frame(t(cur[["X"]]))
  colnames(mat) = rnames
  rownames(mat) = cnames
  mat
}
```

```{r}
cis_types = c("direct", "extended")

ereg = map(ereg_files, function(ereg_file) {
  map(ereg_file, function(ereg_file_config) {
    x = map(cis_types, function(cis_type) {
      ereg_cur = rhdf5::h5read(ereg_file_config, name = str_interp("/uns/${cis_type}_e_regulon_metadata"))
      ereg_cur %>% do.call(cbind.data.frame, .)
    }) %>% bind_rows()
  }) %>% bind_rows(.id="config") 
}) %>% bind_rows(.id="class")

ereg = ereg %>%
  rename(importance_x_abs_rho_R2G = importance_x_abs_rho,
         importance_x_rho_R2G = importance_x_rho)

ereg = ereg %>% 
  mutate(importance_x_abs_rho_TF2G = importance_TF2G * abs(rho_TF2G),
         importance_x_rho_TF2G = importance_TF2G * rho_TF2G) %>%
  mutate(importance_x_rho_log_TF2G = log(importance_x_rho_TF2G),
         importance_x_abs_rho_log_TF2G = log(importance_x_abs_rho_TF2G),
         importance_log_TF2G = log(importance_TF2G))

gene_aucs =  map(ereg_files, function(ereg_file) {
  map(ereg_file, function(ereg_file_config) {
    gene_auc = map(cis_types, \(cis_type) {
      extract_matrix(ereg_file_config, str_interp("/mod/${cis_type}_gene_based_AUC"))
    }) %>%  bind_cols() %>% rownames_to_column("_index")
    gene_auc
  })
})

region_aucs = map(ereg_files, function(ereg_file) {
  map(ereg_file, function(ereg_file_config) {
    region_auc = map(cis_types, \(cis_type) {
      extract_matrix(ereg_file_config, str_interp("/mod/${cis_type}_region_based_AUC"))
    }) %>%  bind_cols() %>% rownames_to_column("_index")
    region_auc
  })
})


mds = map(ereg_files, function(ereg_file) {
  map(ereg_file, function(ereg_file_config) {
    group_set =  rhdf5::h5read(ereg_file_config, name = str_interp("/mod/scRNA_counts/obs"))
    group_set = group_set[c("_index", "cluster")]
    factor_ind = which(map_lgl(group_set, ~length(.x)==2))
    
    for (fi in factor_ind) {
      print(fi)
      group_set[[fi]] = convert_category_to_factor(group_set[[fi]]) 
    }
    group_df = cbind.data.frame(group_set)
    group_df
  })
})

gene_aucs = imap(gene_aucs, function(gene_auc, i) imap(gene_auc, function(gene_auc_config, j) gene_auc_config %>% left_join(mds[[i]][[j]], by="_index")))
region_aucs = imap(region_aucs, function(region_auc, i) imap(region_auc, function(region_auc_config, j) region_auc_config %>% left_join(mds[[i]][[j]], by="_index")))
```

## Load TFs
```{r}
tfs = readr::read_csv("../../data/lambert_tfs/tfs.csv")
tfs_dbd = tfs %>% filter(DBD!="Unknown")
```

```{r}
eregs = split(ereg, ereg$class)
common_tfs = intersect(unique(eregs[["nn"]]$TF), unique(eregs[["glut"]]$TF))
sort(common_tfs)
```

## Filter eRegulons
```{r}
ereg_sane = ereg %>%
  #filter(grepl("\\+\\/\\+", eRegulon_name) | grepl("-\\/-", eRegulon_name)) %>%
  filter(grepl("\\+\\/\\+", eRegulon_name)) %>%
  rename(gene = Gene)
ereg_tf = ereg_sane %>% 
  filter(gene %in% tfs_dbd$Name)
ereg_tftf = ereg_sane %>% filter(gene %in% .$TF)
```

## Dot plot
```{r}
dot_dir = file.path(out_dir, "dotplots")
dir.create(dot_dir)
```

```{r}
obj_mult$cluster = factor(obj_mult$cluster, levels=labels_order)
```

```{r}
to_plot = unique(ereg_sane$TF)
ereg_select = list("MAFB_.*_\\+/\\+" = c("glut_ra", "gaba"),
                   "EMX2_.*" = c("glut_ra", "nn"))
DefaultAssay(obj_mult) = "SCT"
#reduction_cur = "umap_rna_int"glut


nplot = 10
iwalk(ereg_select, function(value_cur, name_cur) {
  
  #ereg_select_cur_form = sub("/", ":", name_cur)
  
  ereg_sane_cur = ereg_sane %>% filter(grepl(name_cur,  eRegulon_name)) %>%
    filter(class %in% value_cur) 
    #distinct(gene, .keep_all=T) %>% 
  name_out = str_extract(name_cur, "^[A-Z]+[0-9]*")
  
  ereg_sane_cur_split = split(ereg_sane_cur, ereg_sane_cur$class)
  ereg_sane_cur_split_genes = map(ereg_sane_cur_split, function(x) sort(x$gene))
  to_plot = Reduce(intersect, ereg_sane_cur_split_genes)
  to_plot = intersect(to_plot, Features(obj_mult))
  
  if (length(to_plot) == 0)
    return()
  ereg_sane_cur_filt = ereg_sane_cur %>%
    filter(gene %in% to_plot) %>%    
    distinct(gene, .keep_all=T) %>% 
    arrange(desc(importance_x_rho_TF2G)) 
  
  to_plot = ereg_sane_cur_filt$gene[1:nplot]


  obj_mult$cluster = factor(obj_mult$cluster, levels=labels_order)
  gg = DotPlot_scCustom(obj_mult, features=to_plot,  group.by = "cluster", flip_axes = T, x_lab_rotate = T)
  print(gg)
  save_plot(file.path(dot_dir, str_interp("${name_out}.pdf")), gg, base_height = 3 * nplot/10, base_width = 10)
  
  obj_mult$cluster = factor(obj_mult$cluster, levels=rev(labels_order))
  gg = DotPlot_scCustom(obj_mult, features=to_plot,  group.by = "cluster", flip_axes = F, x_lab_rotate = T)
  print(gg)
  save_plot(file.path(dot_dir, str_interp("${name_out}_flip.pdf")), gg, base_width = 3 * nplot/6, base_height = 8)
})

```



## eReg intersection
```{r}
library(GeneOverlap)
```

```{r}
ereg_sane_gaba = ereg_sane %>% filter(class=="gaba") 
ereg_sane_gaba_un = ereg_sane_gaba %>% distinct(TF, gene, .keep_all=T)

ereg_sane_glut = ereg_sane %>% filter(class=="glut") 
ereg_sane_glut_un = ereg_sane_glut %>% distinct(TF, gene, .keep_all=T)

ereg_sane_glut_ra = ereg_sane %>% filter(class=="glut_ra") 
ereg_sane_glut_ra_un = ereg_sane_glut_ra %>% distinct(TF, gene, .keep_all=T)

ereg_sane_glut_ra_pm_un = ereg_sane_glut_ra %>% 
  filter(grepl("\\+\\/\\-", eRegulon_name)) %>%
  distinct(TF, gene, .keep_all=T)

ereg_sane_all  = ereg_sane %>% filter(class=="all") 
ereg_sane_all_un = ereg_sane_all %>% distinct(TF, gene, .keep_all=T)

ereg_sane_nn  = ereg_sane %>% filter(class=="nn") 
ereg_sane_nn_un = ereg_sane_nn %>% 
  distinct(TF, gene, .keep_all=T)

ereg_sane_nn_pp_un = ereg_sane_nn %>% 
  filter(grepl("\\+\\/\\+", eRegulon_name)) %>%
  distinct(TF, gene, .keep_all=T)

total_genes = rownames(obj_mult)

```


## eReg intersection functions
```{r}
compute_overlaps <- function(ereg_sane, to_compare, total_features, feature_col = "gene") {
  ereg_sane_cur = ereg_sane %>% filter(class %in% to_compare)
  
 # browser()
  configs_common <- ereg_sane_cur %>%
    distinct(class, config) %>%
    count(config) %>%
    filter(n > 1) %>%
    pull(config)

  overs = map(configs_common, function(config_cur) {
    #browser()
    ereg_sane_cur1 <- ereg_sane_cur %>% filter(config==config_cur)
    to_compare_data = split(ereg_sane_cur1, ereg_sane_cur1$class)
    features = map(to_compare_data, function(x) {
      features_split = split(x[[feature_col]], x$TF) 
      map(features_split, unique)
      })
    #total_features = unique(unlist(features))
    tfs = map(to_compare_data, ~ .x$TF)
    common_tfs = Reduce(intersect, tfs)
    features = map(features, ~.x[common_tfs])

    over = newGOM(features[[1]], features[[2]], length(total_features))
    mat = getMatrix(over, name="pval")
    mat = -1 * log10(mat)

    ncolors = 9
    colors_to_use = brewer_pal(palette = "Greys", direction = 1)(ncolors)
    max_val = max(mat)
    if (max_val == 0) max_val = 1
    cols = colorRamp2(breaks=seq(0, max_val, length.out=ncolors), colors = colors_to_use)
    hm = Heatmap(mat, cluster_columns=F, cluster_rows=F, col=cols)
    print(hm)

    return(over)
  }) %>% set_names(configs_common)
  overs
}

plot_overlaps <- function(overs, compare_label, out_dir, feature_col = "gene") {
  feature_suffix = if (feature_col == "gene") "" else str_interp("_${feature_col}")
  iwalk(overs, function(over, config_cur) {
    nlist = getNestedList(over, "cont.tbl")

    pval = getMatrix(over, name="pval")
    pval = -1 * log10(pval)
    pval_diag = diag(pval)

    odds = getMatrix(over, name="odds.ratio")
    odds_diag = diag(odds)

    mat_df = data.frame(pval = pval_diag, odds = odds_diag, tf = names(pval_diag)) %>%
        filter(pval>1)
    nlist_filt = nlist[as.character(mat_df$tf)]
    nlist_filt = imap(nlist_filt, function(x,i) x[[i]])
    ninter = map(nlist_filt, function(x) {
      data.frame(A = sum(x[1,2], x[2,2]),
                 B = sum(x[2,1], x[2,2]),
                 int = x[2,2])
    }) %>% bind_rows(.id="tf") %>%
      mutate(perc_b = round(int / B * 100),
             frac_b_text = sprintf("%s/%s", int, B))

    mat_df = mat_df %>%
      left_join(ninter) %>%
      arrange(odds) %>%
      mutate(tf = factor(tf, levels=tf))

    gg = ggplot(mat_df, aes(odds, tf, size=pval)) +
      geom_point(stat="identity") +
      theme_bw() +
      labs(y = "odds ratio", x="", title=config_cur) +
      scale_size_continuous("-log10(p-value)", range=c(.5, 3)) +
      theme(
        plot.margin = margin(10,200,10,10, unit="pt"),
        axis.text = element_text(size=6),
            axis.title = element_text(size=6),
            panel.grid = element_blank(),
            legend.background = element_blank(),
            legend.title = element_text(size=6, hjust = .5),
            legend.position = "inside",
            legend.position.inside = c(.75, .3),
            legend.key.size = unit(.1, "in"),
            legend.text = element_text(size=5))

    for (i in 1:nrow(mat_df)) {
      gg = gg + annotate("text", y=i,x= 16, label=mat_df$frac_b_text[i], hjust = 0, size=2)
    }

    gg = gg + coord_cartesian(xlim = c(0,25), clip = "off")
    print(gg)

    save_plot(file.path(out_dir, str_interp("${compare_label}_eRegulon_overlaps_odds-pval${feature_suffix}_${config_cur}.pdf")), gg, base_height=1.5, base_width=1.5)

    mat_df = data.frame(pval = pval_diag, odds = odds_diag, tf = names(pval_diag)) %>%
      left_join(ninter) %>%
      arrange(pval) %>%
      mutate(tf = factor(tf, levels=tf)) %>%
      filter(pval > 1)

    gg = ggplot(mat_df, aes(pval, tf, size=odds)) +
      geom_point(stat="identity") +
      theme_bw() +
      coord_cartesian(xlim = c(0,15), clip="off") +
      labs(x = "-log10(p-value)", y="", title=config_cur) +
      scale_size_continuous("odds ratio", range=c(1, 3)) +
      theme(
        plot.margin = margin(5,20,5,5, unit="pt"),
        axis.text = element_text(size=6),
            axis.title = element_text(size=6), panel.grid = element_blank(),
            legend.background = element_blank(), legend.title = element_text(size=6, hjust = .5),
            legend.key.size = unit(.1, "in"),
            legend.text = element_text(size=5))

    for (i in 1:nrow(mat_df)) {
      gg = gg + annotate("text", y=i,x= 16, label=mat_df$frac_b_text[i], hjust = 0, size=2)
    }

    print(gg)

    save_plot(file.path(out_dir, str_interp("${compare_label}_eRegulon_overlaps_pval-odds${feature_suffix}_${config_cur}.pdf")), gg, base_height=1.5, base_width=3)
  })
}
```

### Non-neuron vs. Glut
```{r fig.height=10, fig.width=10}
overs_nn_glut = compute_overlaps(ereg_sane, c("nn", "glut"), total_genes)
```

```{r}
plot_overlaps(overs_nn_glut, "non-neuron_vs_glut", out_dir)
```


### Non-neuron vs. Glut-RA

```{r fig.height=10, fig.width=10}

overs = compute_overlaps(ereg_sane, c("nn", "glut_ra"), total_genes)
```

```{r}
plot_overlaps(overs, "non-neuron_vs_glut-ra", out_dir)
```

### GABA vs. Glut
```{r fig.height=10, fig.width=10}
overs_gaba_glut = compute_overlaps(ereg_sane, c("gaba", "glut"), total_genes)
```

```{r}
plot_overlaps(overs_gaba_glut, "gaba_vs_glut", out_dir)
```


### GABA vs. Glut-RA
```{r fig.height=10, fig.width=10}
overs_gaba_glut_ra = compute_overlaps(ereg_sane, c("gaba", "glut_ra"), total_genes)
plot_overlaps(overs_gaba_glut_ra, "gaba_vs_glut-ra", out_dir)
```

### Region intersection (plyranges)
```{r}
library(plyranges)

regions_to_granges <- function(regions) {
  parts = str_match(regions, "^(.+):([0-9]+)-([0-9]+)$")
  GRanges(seqnames = parts[,2],
          ranges = IRanges(start = as.integer(parts[,3]), end = as.integer(parts[,4])))
}

overlap_fraction <- function(gr_a, gr_b, min_overlap = 0.2) {
  hits = findOverlaps(gr_a, gr_b)
  if (length(hits) == 0) return(integer(0))

  widths_a = width(gr_a[queryHits(hits)])
  widths_b = width(gr_b[subjectHits(hits)])
  inter = pintersect(gr_a[queryHits(hits)], gr_b[subjectHits(hits)])
  inter_width = width(inter)

  frac_a = inter_width / widths_a
  frac_b = inter_width / widths_b
  keep = frac_a >= min_overlap & frac_b >= min_overlap
  unique(queryHits(hits)[keep])
}

compute_region_overlaps <- function(ereg_sane, to_compare, min_overlap = 0.2) {
  total_regions_gr = regions_to_granges(unique(ereg_sane$Region))
  n_total = length(total_regions_gr)

  ereg_sane_cur = ereg_sane %>% filter(class %in% to_compare)

  configs_common <- ereg_sane_cur %>%
    distinct(class, config) %>%
    count(config) %>%
    filter(n > 1) %>%
    pull(config)

  results = map(configs_common, function(config_cur) {
    ereg_sane_cur1 <- ereg_sane_cur %>% filter(config==config_cur)
    to_compare_data = split(ereg_sane_cur1, ereg_sane_cur1$class)

    tfs_per_class = map(to_compare_data, ~ unique(.x$TF))
    common_tfs = Reduce(intersect, tfs_per_class)

    regions_per_class = map(to_compare_data, function(x) {
      split(x$Region, x$TF) %>% map(unique)
    })

    hyper_results = map_dfr(common_tfs, function(tf_cur) {
      regs_a = regions_per_class[[1]][[tf_cur]]
      regs_b = regions_per_class[[2]][[tf_cur]]

      gr_a = regions_to_granges(regs_a)
      gr_b = regions_to_granges(regs_b)

      overlap_idx = overlap_fraction(gr_a, gr_b, min_overlap = min_overlap)
      n_overlap = length(overlap_idx)
      n_a = length(gr_a)
      n_b = length(gr_b)

      pval = phyper(n_overlap - 1, n_b, n_total - n_b, n_a, lower.tail = FALSE)

      data.frame(tf = tf_cur, n_a = n_a, n_b = n_b,
                 n_overlap = n_overlap, n_total = n_total,
                 pval = pval,
                 frac_a = n_overlap / n_a,
                 frac_b = n_overlap / n_b)
    }) %>%
      mutate(pval_adj = p.adjust(pval, method = "BH"),
             mlog10_pval = -log10(pval))

    hyper_results
  }) %>% set_names(configs_common)

  results
}

plot_region_overlaps <- function(results, compare_label, out_dir) {
  iwalk(results, function(hyper_df, config_cur) {
    hyper_sig = hyper_df %>%
      filter(mlog10_pval > 1) %>%
      arrange(mlog10_pval) %>%
      mutate(tf = factor(tf, levels=tf),
             frac_b_text = sprintf("%s/%s", n_overlap, n_a))

    if (nrow(hyper_sig) == 0) return()

    gg = ggplot(hyper_sig, aes(mlog10_pval, tf, size=frac_a)) +
      geom_point() +
      theme_bw() +
      coord_cartesian(xlim = c(0, max(hyper_sig$mlog10_pval) * 1.3), clip="off") +
      labs(x = "-log10(p-value)", y="", title=config_cur) +
      scale_size_continuous("fraction overlap", range=c(1, 3)) +
      theme(
        plot.margin = margin(5,100,5,5, unit="pt"),
        axis.text = element_text(size=6),
        axis.title = element_text(size=6),
        panel.grid = element_blank(),
        legend.background = element_blank(),
        legend.title = element_text(size=6, hjust = .5),
        legend.key.size = unit(.1, "in"),
        legend.text = element_text(size=5))

    for (i in 1:nrow(hyper_sig)) {
      gg = gg + annotate("text", y=i, x=max(hyper_sig$mlog10_pval) * 1.1,
                          label=hyper_sig$frac_b_text[i], hjust = 0, size=2)
    }

    print(gg)

    save_plot(file.path(out_dir, str_interp("${compare_label}_eRegulon_region-overlaps_pval_${config_cur}.pdf")),
              gg, base_height=1.5, base_width=3)
  })
}
```

```{r}
region_results_nn_glut = compute_region_overlaps(ereg_sane, c("nn", "glut"))
plot_region_overlaps(region_results_nn_glut, "non-neuron_vs_glut", out_dir)
```

```{r}
region_results_nn_glut_ra = compute_region_overlaps(ereg_sane, c("nn", "glut_ra"))
plot_region_overlaps(region_results_nn_glut_ra, "non-neuron_vs_glut-ra", out_dir)
```

```{r}
region_results_gaba_glut = compute_region_overlaps(ereg_sane, c("gaba", "glut"))
plot_region_overlaps(region_results_gaba_glut, "gaba_vs_glut", out_dir)
```

```{r}
region_results_gaba_glut_ra = compute_region_overlaps(ereg_sane, c("gaba", "glut_ra"), min_overlap = .1)
plot_region_overlaps(region_results_gaba_glut_ra, "gaba_vs_glut-ra", out_dir)
```

## Load snRNA data
```{r}
obj = qread(snrna_file)
#cells = Cells(obj)[obj$position %in% c("RA", "ARCO")]
#obj = subset(obj, cells=cells)
```

## Find DEG
```{r}
pairs = list(ra = c("Glut-RA", "Glut-Arco-1"),
             gaba4 = c("GABA-4-1", "GABA-1-1"),
             astro = c("Astro-1", "Oligo"))

obj_agg = AggregateExpression(obj_mult, assays="SCT", group.by = c("assignment", "cluster"), return.seurat = T)

deg = map(pairs, function(pair_cur) {
  deg = FindMarkers(obj_agg, ident.1 = pair_cur[1], ident.2=pair_cur[2], test.use = "DESeq2", assay="SCT", 
                    group.by = "cluster", logfc.threshold = 0, min.pct =0 )
  deg = deg %>% rownames_to_column("gene")
  deg
}) %>% bind_rows(.id="set")
```

### EMX2
```{r}
ereg_all = ereg %>% 
  filter(class=="all")
#filter(class %in% c("nn", "glut_ra"))
ereg_all_emx2 = ereg_all %>% filter(grepl("EMX2_direct_\\+\\/\\+",  eRegulon_name)) %>%
  rename(gene = Gene)

deg_emx2 = deg %>% 
  filter(set %in% c("ra", "astro")) %>%
  inner_join(ereg_all_emx2)

deg_emx2_w = deg_emx2 %>% 
  #filter(p_val_adj < 0.1) %>% 
  select(-p_val, -pct.1, -pct.2, -p_val_adj) %>% pivot_wider(names_from = "set", values_from = "avg_log2FC", values_fill = 0)
gg = ggplot(deg_emx2_w, aes(ra, astro, color = importance_log_TF2G)) + 
    geom_hline(yintercept=0, color="gray20") + 
  geom_vline(xintercept=0, color="gray20") + 
  geom_point() +
  scale_color_viridis() + 
  theme(legend.position = c(0.8, 0.8))
gg

save_plot(file.path(out_dir, "scatter_deg-ra-astro_ereg-emx2.pdf"), gg, base_height = 3, base_width = 3)
```





### MAFB
```{r}
ereg_all = ereg %>% 
  filter(class=="all") 
  #filter(class %in% c("glut_ra", "gaba"))
ereg_all_mafb = ereg_all %>% filter(grepl("MAFB_.+_\\+\\/\\+",  eRegulon_name)) %>%
  rename(gene = Gene)

deg_mafb = deg %>% 
  filter(set %in% c("ra", "gaba4")) %>%
  inner_join(ereg_all_mafb %>% distinct(gene, .keep_all=T))

deg_mafb_w = deg_mafb %>% 
  #filter(p_val_adj < 0.1) %>% 
  select(-p_val, -pct.1, -pct.2, -p_val_adj) %>% pivot_wider(names_from = "set", values_from = "avg_log2FC", values_fill = NA) %>%
  mutate(mult = ra * gaba4)

gg = ggplot(deg_mafb_w, aes(ra, gaba4, color = importance_log_TF2G)) + 
    geom_hline(yintercept=0, color="gray20") + 
  geom_vline(xintercept=0, color="gray20") + 
  geom_point() +
  scale_color_viridis() + 
  theme(legend.position = c(0.8, 0.8))
gg

save_plot(file.path(out_dir, "scatter_deg-ra-gaba4_ereg-mafb.pdf"), gg, base_height = 3, base_width = 3)
```

```{r}

```



### Glut
```{r}
ereg_sane_glut = ereg_sane %>% filter(class=="glut")
ereg_sane_glut_ra = ereg_sane %>% filter(class=="glut_ra")
```

```{r}
cells = Cells(obj)[grep("Glut" , obj$cluster)]
obj_glut = subset(obj, cells=cells )

md = obj_glut@meta.data %>% rownames_to_column() 
md = md %>% mutate(cluster2 = if_else(grepl("RA", cluster), "RA-Glut", cluster)) %>%
  column_to_rownames()
obj_glut = AddMetaData(obj_glut, md)
```


```{r}
pairs = list(ra = c("RA-Glut", "Arco-Glut-1"),
             hvc_1 = c("HVC-Glut-1", "Nido-Glut-1"),
             hvc_3 = c("HVC-Glut-1", "Nido-Glut-3"))

deg_glut = map(pairs, function(pair_cur) {
  deg = FindMarkers(obj_glut, ident.1 = pair_cur[1], ident.2=pair_cur[2],
                    group.by = "cluster", logfc.threshold = 0)
  deg = deg %>% rownames_to_column("gene")
  deg
}) %>% bind_rows(.id="set")

#deg_ra_aid$MAFB_ereg = deg_ra_aid$gene %in% ereg1_un_glut$gene
```

Density plot of DE of genes in or out of eregulon
```{r}

current_tfs = unique(ereg_sane_glut$TF)

walk(names(pairs), function(set_cur) {
  deg_glut_cur = deg_glut %>%
    filter(set==set_cur)
  
  x_axis = str_interp("Averge logFC ${pairs[[set_cur]][1]} vs. ${pairs[[set_cur]][2]}")
  
  walk(current_tfs, function(current_tf) {
    ereg_cur = ereg_sane_glut %>% filter(TF==current_tf)
    deg_glut_cur_1 = deg_glut_cur %>%
      mutate(in_ereg = gene %in% ereg_cur$gene)
    gg = ggplot(deg_glut_cur_1, aes(avg_log2FC, color=in_ereg)) + 
      geom_density() + 
      theme_bw() + scale_color_brewer("in eRegulon", palette="Dark2") + 
      xlim(-2, 2) + 
      labs(x=x_axis)
    print(gg)
    save_plot(file.path(out_dir, str_interp("avgr_logFC_${set_cur}_${current_tf}-regulon.pdf")), gg)
  })
})
```

```{r}
ereg_sane_glut_un = ereg_sane_glut %>% distinct(gene, TF, .keep_all=T)
```

```{r}

variables = c("importance_x_rho_log_TF2G", "importance_log_TF2G", "rho_TF2G")

walk(names(pairs), function(set_cur) {
  deg_glut_cur = deg_glut %>% filter(set==set_cur)
  walk(current_tfs, function(current_tf) {
    ereg_sane_glut_un_cur = ereg_sane_glut_un %>% filter(TF==current_tf) %>%
      left_join(deg_glut_cur)
  ereg_sane_glut_un_cur_sig = ereg_sane_glut_un_cur %>% filter(p_val_adj < .01)
  for (var_cur in variables) {
    gg = ggplot(ereg_sane_glut_un_cur_sig, aes(!!as.symbol(var_cur), avg_log2FC)) + 
      geom_point() + 
      stat_smooth(method="lm", se = F) + 
      theme_bw()
    #print(gg)
    save_plot(file.path(out_dir, str_interp("avg_logFC_${set_cur}_vs_${var_cur}_${current_tf}.pdf")), gg)
  }
  })
})
```

### GABA
```{r}
gaba_dir = file.path(out_dir, "gaba")
dir.create(gaba_dir)
```

```{r}
ereg_sane_gaba = ereg_sane %>% filter(class=="gaba")
ereg_sane_gaba_un = ereg_sane_gaba %>% distinct(gene, .keep_all=T)
```

```{r}
cells = Cells(obj)[grep("GABA" , obj$cluster)]
obj_gaba = subset(obj, cells=cells )

md = obj_gaba@meta.data %>% rownames_to_column() 
md = md %>% 
  mutate(cluster2 = case_when(grepl("GABA-1", cluster) ~ "GABA-1",
                                        grepl("GABA-5", cluster) ~ "GABA-5",
                                        TRUE ~ cluster)) %>%
  mutate(cluster_mafb = case_when(cluster %in% c("GABA-2-1", "GABA-4-1") ~ "MAFB_high_expr",
                                  TRUE ~ "MAFB_low_expr")) %>% 
  mutate(cluster_fsi = case_when(cluster %in% c("GABA-4-1") ~ "FSI",
                                 cluster %in% c("GABA-2-1", "GABA-3") ~ "exclude",
                                 TRUE ~ "non-FSI")) %>% 
  column_to_rownames()
obj_gaba = AddMetaData(obj_gaba, md)

```

```{r}
DotPlot(obj_gaba, features= c("MAFB", "GAD1", "NPXH1", "PVALB", "LHX6", "KCNC1"), group.by="cluster")
```

```{r}
sets = list(cluster_mafb = c("MAFB_high_expr", "MAFB_low_expr"),
            cluster_fsi = c("FSI", "non-FSI"))

deg_gaba = imap(sets, function(set_cur, set_name) {
  deg_gaba = FindMarkers(obj_gaba, ident.1 = set_cur[1], ident.2 = set_cur[2], group.by = set_name, logfc.threshold = 0)
  deg_gaba = deg_gaba %>% rownames_to_column("gene")
  deg_gaba
}) %>% bind_rows(.id="set")
#deg_gaba$MAFB_ereg = deg_gaba$gene %in% ereg1_un_gaba$gene
```

```{r}
current_tfs = unique(ereg_sane_gaba$TF)

walk(names(sets), function(set_cur) {
  deg_gaba_cur = deg_gaba %>%
    filter(set==set_cur)
  
  x_axis = str_interp("Averge logFC ${pairs[[set_cur]][1]} vs. ${pairs[[set_cur]][2]}")
  
  walk(current_tfs, function(current_tf) {
    ereg_cur = ereg_sane_gaba %>% filter(TF==current_tf)
    deg_gaba_cur_1 = deg_gaba_cur %>%
      mutate(in_ereg = gene %in% ereg_cur$gene)
    gg = ggplot(deg_gaba_cur_1, aes(avg_log2FC, color=in_ereg)) + 
      geom_density() + 
      theme_bw() + scale_color_brewer("in eRegulon", palette="Dark2") + 
      xlim(-2, 2) + 
      labs(x=x_axis)
    #print(gg)
    save_plot(file.path(gaba_dir, str_interp("avgr_logFC_${set_cur}_${current_tf}-regulon.pdf")), gg)
  })
})
```

```{r}
deg_glut_gaba = deg_ra_aid %>% 
  select(gene, avg_log2FC, p_val_adj, MAFB_ereg ) %>%
  rename_with(~ paste(.x, "glut", sep="_"), !contains("gene")) %>% 
  left_join(deg_gaba %>% select(avg_log2FC, gene, p_val_adj, MAFB_ereg ) %>% rename_with(~ paste(.x, "gaba", sep="_"), !contains("gene")) ) %>%
  left_join(ereg1_un_glut %>% select(gene, contains("R2G"), contains("TF2G")) %>% rename_with(~ paste(.x, "glut", sep="_"), !contains("gene"))) %>%
  left_join(ereg1_un_gaba %>% select(gene, contains("R2G"), contains("TF2G")) %>% rename_with(~ paste(.x, "gaba", sep="_"), !contains("gene")))
```

```{r}
deg_glut_gaba_mafb = deg_glut_gaba %>% filter(MAFB_ereg_glut, MAFB_ereg_gaba) %>%
  filter(gene !="MAFB")
gg = ggplot(deg_glut_gaba_mafb, aes(avg_log2FC_glut, avg_log2FC_gaba, color=log(TF2G_importance_x_rho_glut)) )+ 
  scale_color_viridis() + 
  geom_hline(yintercept=0) + 
  geom_vline(xintercept=0) + 
  geom_point() +
  theme_bw()
print(gg)

#ggsave(file.path(out_dir, "avg_log2FC_glut_vs_gaba-MAFB-high-low.pdf"), gg, height=6, width=6)
save_plot(file.path(out_dir, "avg_log2FC_glut_vs_gaba-MAFB-high-low.pdf"), gg, base_height=4, base_width=7)
```

## GO
```{r}
deg_glut_gaba = deg_glut %>% 
  select(gene, avg_log2FC, p_val_adj, set) %>%
  rename_with(~ paste(.x, "glut", sep="_"), !(contains("gene"))) %>% 
  left_join(deg_gaba %>% select(avg_log2FC, gene, p_val_adj, set) %>% 
               rename_with(~ paste(.x, "gaba", sep="_"), !(contains("gene"))))
# 
# %>%
#   left_join(ereg_sane_glut_un %>% select(gene, contains("TF2G")) %>% rename_with(~ paste(.x, "glut", sep="_"), !contains("gene"))) %>%
#   left_join(ereg_sane_gaba_un %>% select(gene, contains("TF2G")) %>% rename_with(~ paste(.x, "gaba", sep="_"), !contains("gene")))
```
```{r}
library(clusterProfiler)
```

```{r}
gene = bitr(deg_glut_gaba_mafb$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
gene_all = bitr(deg_ra_aid$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
library(org.Hs.eg.db)
ego <- enrichGO(gene          = gene$ENTREZID,
                universe      = gene_all$ENTREZID,
                OrgDb         = org.Hs.eg.db,
                ont           = "ALL",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.1,
        readable      = TRUE)
head(ego)
```

```{r}
ekegg <- enrichKEGG(
  gene          = gene$ENTREZID,
  universe      = gene_all$ENTREZID,
  organism = "hsa", 
  keyType = "ncbi-geneid",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.01,
  qvalueCutoff  = 0.1)
head(ekegg)
```

```{r}
library(msigdbr)
m_df = msigdbr(species = "Homo sapiens")
m_df_sub = m_df %>% 
  filter(gs_cat == "C5") %>%
  dplyr::select("gs_name", "human_entrez_gene")
```

```{r}
deg_glut_gaba = deg_glut_gaba %>% mutate(TF2G_rho_glut_gaba = TF2G_rho_glut * TF2G_rho_gaba,
                                         TF2G_importance_glut_gaba = TF2G_importance_glut + TF2G_importance_gaba) %>%
  filter(gene!="MAFB")
```

```{r}
gene = bitr(deg_glut_gaba$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
genelist = deg_glut_gaba$TF2G_importance_glut_gaba
names(genelist) = gene$ENTREZID
genelist = sort(genelist, decreasing = T)
egsea = GSEA(genelist, TERM2GENE = m_df_sub, pvalueCutoff=.1)
egsea
```


#### FSI
```{r}
deg_fsi = FindMarkers(obj_gaba, ident.1 = "FSI", ident.2 = "non-FSI", group.by = "cluster_fsi", logfc.threshold = 0)
deg_fsi = deg_fsi %>% rownames_to_column("gene")
deg_fsi$MAFB_ereg = deg_fsi$gene %in% ereg1_un_gaba$gene
```

```{r}
gg = ggplot(deg_fsi, aes(avg_log2FC, color=MAFB_ereg)) + 
  geom_density() + 
  theme_bw() + scale_color_brewer("in MAFB eRegulon", palette="Dark2") + 
  xlim(-2, 2) + 
  labs(x="Average logFC GABA High vs. Low MAFB")
gg
save_plot(file.path(out_dir, "avgr_logFC_gaba-fsi_mafb-regulon.pdf"), gg)
```

```{r}
deg_glut_gaba = deg_ra_aid %>% 
  select(gene, avg_log2FC, p_val_adj, MAFB_ereg ) %>%
  rename_with(~ paste(.x, "glut", sep="_"), !contains("gene")) %>% 
  left_join(deg_fsi %>% select(avg_log2FC, gene, p_val_adj, MAFB_ereg ) %>% rename_with(~ paste(.x, "gaba", sep="_"), !contains("gene")) ) %>%
  left_join(ereg1_un_glut %>% select(gene, contains("R2G"), contains("TF2G")) %>% rename_with(~ paste(.x, "glut", sep="_"), !contains("gene"))) %>%
  left_join(ereg1_un_gaba %>% select(gene, contains("R2G"), contains("TF2G")) %>% rename_with(~ paste(.x, "gaba", sep="_"), !contains("gene")))
```

```{r}
deg_glut_gaba_mafb = deg_glut_gaba %>% filter(MAFB_ereg_glut, MAFB_ereg_gaba)
gg = ggplot(deg_glut_gaba_mafb, aes(avg_log2FC_glut, avg_log2FC_gaba, color=log(TF2G_importance_x_rho_glut)) )+ 
  scale_color_viridis() + 
  geom_hline(yintercept=0) + 
  geom_vline(xintercept=0) + 
  geom_point() +
  theme_bw()
gg

#ggsave(file.path(out_dir, "avg_log2FC_glut_vs_gaba-MAFB-high-low.pdf"), gg, height=6, width=6)
save_plot(file.path(out_dir, "avg_log2FC_glut_vs_gaba-FSI.pdf"), gg, base_height=4, base_width=7)
```


## Heatmap
```{r}
ereg_pattern = "\\+\\/\\+"
gene_auc_cur = gene_auc %>% select(one_of( "cluster"), matches(ereg_pattern))
colnames(gene_auc_cur) = sub("_(direct|extended)_\\+\\/\\+_\\([0-9]+[a-z]\\)", "", colnames(gene_auc_cur))
gene_auc_cur = gene_auc_cur %>% pivot_longer(-cluster, names_to="tf")
region_auc_cur = region_auc %>% select(one_of( "cluster"), matches(ereg_pattern))
colnames(region_auc_cur) = sub("_\\+\\/\\+_\\([0-9]+[a-z]\\)", "", colnames(region_auc_cur))

gene_auc_cur_mean = gene_auc_cur %>%
  group_by(cluster, tf) %>%
  summarize(value_mean = mean(value)) %>%
  group_by(tf) %>%
  mutate(value_mean_scale = scale_01(value_mean)) %>%
  pivot_wider(id_cols=!contains("value"), names_from="tf", values_from="value_mean_scale") %>% 
  column_to_rownames("cluster") %>%
  as.matrix() %>%
  t()
  
region_auc_cur_mean = region_auc_cur %>%
  group_by(cluster) %>%
  summarize_all(mean) %>%
  column_to_rownames("cluster") %>%
  mutate_all(scale) %>% 
  as.matrix() %>%
  t()

expr_mat = obj_avg[intersect(rownames(gene_auc_cur_mean), rownames(obj_avg)), intersect(colnames(gene_auc_cur_mean), colnames(obj_avg))]
expr_mat = t(scale(t(expr_mat)))
```

### Cluster
```{r}
hc = hclust(dist(gene_auc_cur_mean), method="ward.D")
```

### Plot
```{r}
height = 4 
width = 3

mat = gene_auc_cur_mean
colors = scales::brewer_pal(palette = "BuGn", direction = 1 )(9)
#colors = viridis_pal(option = "C")(9)
cols = colorRamp2(breaks = seq(min(mat), max(mat), length.out=9), 
                  colors = colors )

hm = Heatmap(gene_auc_cur_mean, cluster_columns = F, cluster_rows = hc,
             show_row_names = F,
             col = cols,
             height = unit(height, "in"),
             width = unit(width, "in"))
hm
pdf(file.path(out_dir, "gene_auc_heatmap_no-names.pdf"), height =height + 2, width = width + 2)
print(hm)
dev.off()

height = 10 
width = height * ncol(mat) / nrow(mat)
hm = Heatmap(gene_auc_cur_mean, cluster_columns = F, cluster_rows = hc,
             show_row_names = T,
             col = cols, row_names_gp = gpar(fontsize=6),
             height = unit(height, "in"),
             width = unit(width, "in"))
hm
pdf(file.path(out_dir, "gene_auc_heatmap_names.pdf"), height =height + 2, width = width + 2)
print(hm)
dev.off()
```

```{r}
height = 4 
width = 3

mat = expr_mat
cols = colorRamp2(breaks = seq(min(mat), max(mat), length.out=9), 
                  colors = scales::brewer_pal(palette = "PuOr", direction = -1 )(9))
hm = Heatmap(mat, cluster_columns = F, cluster_rows = hc,
             show_row_names = F,
             col = cols,
             height = unit(height, "in"),
             width = unit(width, "in"))
hm
pdf(file.path(out_dir, "expr_heatmap_no-names.pdf"), height =height + 2, width = width + 2)
print(hm)
dev.off()

height = 10 
width = 3
hm = Heatmap(mat, cluster_columns = F, cluster_rows = hc,
             show_row_names = T,
             col = cols, row_names_gp = gpar(fontsize=6),
             height = unit(height, "in"),
             width = unit(width, "in"))
hm
pdf(file.path(out_dir, "epxr_heatmap_names.pdf"), height =height + 2, width = width + 2)
print(hm)
dev.off()
```

## UMAP
```{r}
umap_dir = file.path(out_dir, "umaps")
dir.create(umap_dir)
```

### Filter for just glut
```{r}
obj_glut_fname = file.path(in_dir, "reduction_viz", "combined_glut_umap", "obj_clustered.qs")
obj_glut = qread(obj_glut_fname)
```



```{r}
to_plot = unique(ereg_sane$TF)
tp = "AR"
DefaultAssay(obj_glut) = "CB"
reduction_cur = "dims20nn30mindist0.3"

for (tp in to_plot) {
  gg = FeaturePlot_scCustom(obj_glut, features = tp, figure_plot = T, 
                            aspect_ratio = 1, reduction = reduction_cur,
                            raster = T, 
                            raster.dpi=c(1024,1024),
                            pt.size = 2, 
                            alpha_exp = 1,
                            alpha_na_exp = .1,
                            #pt.size = .1
  )
  save_plot(file.path(umap_dir, str_interp("${tp}.pdf")), gg, base_width=7, base_asp = 1)
  
  gg = Nebulosa::plot_density(obj_glut, features = tp, size = .1,
                              reduction = reduction_cur) + 
    coord_equal()
  save_plot(file.path(umap_dir, str_interp("${tp}_density.pdf")), gg, base_width=7, base_asp = 1)
  
}
```

## GO analysis
```{r}
library(clusterProfiler)
```

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
```

```{r}
tfs = unique(ereg_sane$TF)
pair_cur = "ra"

Log2FC_thresh = 1
FDR_thresh = 1E-2

all_genes = rownames(obj)
all_genes_id = bitr(all_genes, "SYMBOL", "ENTREZID", OrgDb = org.Hs.eg.db )
  
gos = map(tfs, function(tf) {
  print(tf)
  ereg_sane_cur = ereg_sane %>% 
    filter(TF == tf) 
  
  regs = unique(ereg_sane_cur$regulation)
  
  go_cur = map(regs, function(reg) {
    print(reg)
    ereg_sane_cur1 = ereg_sane_cur %>% filter(regulation==reg)
    sel_genes = unique(ereg_sane_cur1$Gene)
    sel_genes_id = bitr(sel_genes, "SYMBOL", "ENTREZID", OrgDb = org.Hs.eg.db )
    go <- enrichGO(gene     = sel_genes_id$ENTREZID, 
                   universe = all_genes_id$ENTREZID,
                   OrgDb    = org.Hs.eg.db,
                   ont      = "ALL", 
                   maxGSSize = 200,
                   readable = TRUE)
    
    kegg <- enrichMKEGG(gene     = sel_genes_id$ENTREZID,
                       universe = all_genes_id$ENTREZID,
                       organism     = 'hsa')
    
    wp <- enrichWP(gene     = sel_genes_id$ENTREZID,
                        universe = all_genes_id$ENTREZID,
                        organism     = 'Homo sapiens')
    go 
  }) %>% set_names(regs)
}) %>% set_names(tfs)
```

```{r}
marker_peaks_glut_df_cur = marker_peaks_glut_df_cur %>%
  mutate(mlog10FDR = -1 * log10(FDR))
gg = ggplot(marker_peaks_glut_df_cur, aes(Log2FC, mlog10FDR)) + 
  geom_point()
gg
```

```{r}
go_df = map(gos, function(go1) {
  map(go1, as.data.frame) %>% bind_rows(.id = "sign")
}) %>% bind_rows(.id="pair")
```

## TF networks

```{r}
ereg_sane_out = ereg_sane %>%
  distinct(TF, Gene, .keep_all=T) %>%
  select(TF, Gene, regulation, contains("importance"), contains("rho"))
out_fname = file.path(out_dir, "scenicplus_sane.txt")
write_delim(ereg_sane_out, out_fname, delim="\t")

ereg_tf_out = ereg_tf %>%
  distinct(TF, Gene, .keep_all=T) %>%
  select(TF, Gene, regulation, contains("importance"), contains("rho"))
out_fname = file.path(out_dir, "scenicplus_tf-tf.txt")
write_delim(ereg_tf_out, out_fname, delim="\t")

ereg_tftf_out = ereg_tftf %>%
  distinct(TF, Gene, .keep_all=T) %>%
  select(TF, Gene, regulation, contains("importance"), contains("rho"))
out_fname = file.path(out_dir, "scenicplus_tf-tf-only.txt")
write_delim(ereg_tftf_out, out_fname, delim="\t")
```



```{r}
gene_cur = "IRF2"
ereg_glut_cur = ereg_glut %>%
  filter(Gene==gene_cur)
```

```{r}
tf_cur = "BARX2"
ereg_glut_tf_cur = ereg_glut %>%
  filter(TF==tf_cur)
ereg_glut_tf_cur
```

## Export for Cytoscape
```{r}
out_fname = file.path(out_dir, "scenicplus_ra-arco-glut_eRegulons.txt")
write_delim(ereg_glut, out_fname, delim="\t")

ereg_glut_un = ereg_glut %>%
  distinct(TF, Gene, .keep_all=T)
out_fname = file.path(out_dir, "scenicplus_ra-arco-glut_eRegulons_unique-gene.txt")
write_delim(ereg_glut_un, out_fname, delim="\t")
```

