---
title: "Bidirectional label transfer: Colquitt 2021 HVC/RA <-> hybrid-labelled snRNA"
format:
  html:
    toc: true
    df-print: paged
---

Seurat label transfer (`FindTransferAnchors` / `TransferData`) between the published Colquitt 2021
HVC/RA dataset and this project's newly annotated dataset (the hybrid division-based labels from
`snrna/naming/hybrid_division_naming.qmd`), run **in both directions**: published-as-reference onto
the new data, and new-data-as-reference onto the published cells. Independent of the naming notebook
by the same convention used throughout this project: reads its output read-only, writes its own copy.

Running it both ways is what makes the result interpretable rather than suggestive. A one-way
transfer always returns *some* label for every cell, including for populations the reference never
sampled, so a confident-looking assignment can be an artifact of the reference having nowhere else to
put a cell. A correspondence that holds in both directions -- a **reciprocal best match** -- cannot be
produced that way, so the reciprocal table at the end is the actual result; the two per-direction
heatmaps are the supporting detail.

The question: do the newly derived hybrid labels -- in particular the song-nucleus labels
(`Glut-DACH2-HVCra`, `Glut-DACH2-HVCx`, `Glut-CACNA1H-RA`) -- correspond to the populations the
published dataset already defined (`HVC_Glut-1..5`, `RA_Glut-1..3`)?

## Reference dataset, checked before use

`/mnt/nest/sc_datasets/colquitt_2021/HVC_RA_paper.qs`. Facts below were confirmed directly against
the object rather than assumed, because several change what the transfer has to do:

- **Format.** The original is `.qs` (old format); this project's environment has `qs2` but not `qs`.
  Converted once to `.qs2` in place with `~/ssd/rstudio/multiome/convert_qs_to_qs2.R` under the
  `qs-convert` micromamba env (the same script/env already used for 131 other Seurat objects,
  0 failures). This notebook reads the `.qs2`.
- **Seurat version.** The object is version 3.1.2 with old-style `Assay` objects; this project runs
  Seurat 5.5.0. Passed through `UpdateSeuratObject()` before anything else.
- **The `RNA` `data` layer holds raw counts, not normalised values.** Confirmed by direct comparison:
  `identical(data, counts)` was `TRUE`, every value integer, per-cell sums ~1000. Both datasets are
  therefore explicitly re-normalised below -- taking the stored `data` layer at face value would have
  fed raw counts into anchor-finding, which fails quietly rather than loudly.
- **No SCT model.** The `SCT` assay is an old `Assay` without the `SCTModel` that
  `normalization.method = "SCT"` requires, so the transfer is `LogNormalize`-style on the `RNA` assay
  of both datasets in both directions.
- **Contents.** 29,497 cells, both species (`zf` 25,392 / `bf` 4,105), positions `hvc` (23,017) and
  `ra` (6,480), 31 harmonised cluster labels in `cluster_int_sub2`. All reference cells are used:
  `cluster_int_sub2` is the cross-species-harmonised labelling, and several clusters have almost no
  `bf` cells (`RA_Glut-3` has 1), so restricting to the query's own species would gut the very
  populations of interest.

**The published dataset sampled only HVC and RA.** The new dataset also covers NC, NR, LMAN and
arcopallium. So the non-song clusters have no true counterpart in the published data, and the
published data has no counterpart for much of the new dataset -- in *both* directions there are
populations that must be mismatched. This is the expected, informative outcome for them, and the
reason the reciprocal criterion below is the one worth reading.

## Query-side gotcha, also checked

The new dataset's `RNA` counts layer is a **BPCells on-disk `IterableMatrix`** (class `RenameDims`),
not an in-memory sparse matrix -- the preprocessing pipeline wrote it that way. Generic numeric ops
are undefined on it (the first run of this notebook died on a plain `max()` with
`invalid 'type' (S4) of argument`), and more importantly `FindTransferAnchors` would otherwise be
operating against a lazy on-disk matrix. Materialised to a `dgCMatrix` below (~8 s, ~0.7 GB).

## Gene space

The two datasets are annotated to different genomes (zebra finch `taeGut` vs. Bengalese finch
`lonStrDom`) whose `LOC######` placeholder IDs are **disjoint ID spaces** -- the same `LOC` number
means unrelated genes across the two, so any `LOC` "match" would be a coincidence of numbering rather
than homology. Checked directly: the 12,213-gene overlap between the two `RNA` assays contains
**zero** `LOC` genes already, i.e. it is entirely real gene symbols. `LOC*` is nevertheless filtered
explicitly (per direct instruction) so this is guaranteed by the code rather than left as a property
that happened to hold, and asserted below.

The feature set is deliberately **symmetric**: the union of each dataset's own informative genes (the
published `integrated` assay's 3000 features, and the new dataset's 3000 `SCT` variable features),
restricted to the shared non-`LOC` space. Using one dataset's variable genes for both directions
would privilege its notion of what varies, and any asymmetry in the result could then be an artifact
of feature choice rather than of direction -- which is precisely what the reciprocal comparison is
supposed to test.

```{r setup}
#| message: false
library(Seurat)
library(tidyverse)
library(qs2)
library(Matrix)
library(ComplexHeatmap)
library(circlize)
library(cowplot)
library(here)
set.seed(10)
theme_set(theme_cowplot())
options(future.globals.maxSize = 50 * 1e9)
```

# Parameters

```{r params}
pub_fname = "/mnt/nest/sc_datasets/colquitt_2021/HVC_RA_paper.qs2"
new_fname = here::here("snrna/naming/hybrid_division_naming", "obj_hybrid_labels.qs2")

new_label = "celltype_hybrid"    # this project's new labels
pub_label = "cluster_int_sub2"   # published harmonised cluster labels
pub_position = "position2"       # published dissection (hvc / ra)

assay_to_use = "RNA"
n_dims = 30

script_name = "colquitt2021_label_transfer"
out_dir = here::here("snrna/integration", script_name)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

## Anchor-finding is the expensive step, once per direction. Cached manually rather than via knitr's
## `cache: true`, which invalidates on chunk text rather than on the upstream objects actually
## changing -- the same silent-staleness reasoning applied to the caches in snrna/naming/exploratory/apply_composite_labels.qmd.
redo = list(anchors = FALSE)

## Cluster order for figures, identical to celltypes_hclust_all_hybrid.qmd's ct_order so the figures
## can be read against each other. Excluded/artifact clusters are absent by construction
## (celltype_hybrid is NA for them and those cells are dropped below).
ct_order = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCra-Pre", "Glut-DACH2-HVCx",
             "Glut-DACH2-1", "Glut-DACH2-2", "Glut-DACH2-3", "Glut-DACH2-4",
             "Glut-DACH2-LMANsh", "Glut-DACH2-LMANco",
             "Glut-DACH2-5", "Glut-DACH2-6", "Glut-DACH2-7", "Glut-DACH2-8", "Glut-SATB2-1",
             "Glut-CACNA1H-RA", "Glut-CACNA1H-1", "Glut-CACNA1H-2", "Glut-CACNA1H-3", "Glut-CACNA1H-4",
             "Glut-GABA",
             "Glut-Im", "Glut-NB", "Glut-NSC",
             "GABA-1-1", "GABA-1-2", "GABA-2-1", "GABA-2-2", "GABA-2-3",
             "GABA-3", "GABA-4-1", "GABA-4-2",
             "GABA-5-1", "GABA-5-2",
             "GABA-6", "GABA-7", "GABA-8",
             "GABA-Im",
             "Astro-1", "Astro-2", "Astro-3", "Epen",
             "Oligo-1", "Oligo-2", "Oligo-3", "OPC",
             "Micro", "Endo")

## Published labels grouped Glut (HVC then RA) -> GABA -> precursor -> non-neuronal, so the block
## structure is legible rather than alphabetical.
pub_order = c(paste0("HVC_Glut-", 1:5), paste0("RA_Glut-", 1:3),
              "GABA-1-1", "GABA-1-2", "GABA-2", "GABA-3", "GABA-4",
              "GABA-5-1", "GABA-5-2", "GABA-5-3", "GABA-6", "GABA-7", "GABA-8", "GABA-Pre",
              paste0("Pre-", 1:4),
              "Astro", "Epen", "Oligo", "OPC", "Micro", "Endo", "Mural", "VLMC", "RBC")

song_hybrid = c("Glut-DACH2-HVCra", "Glut-DACH2-HVCra-Pre", "Glut-DACH2-HVCx", "Glut-CACNA1H-RA",
                "Glut-DACH2-LMANsh", "Glut-DACH2-LMANco")
```

# Load

```{r load}
pub = qs_read(pub_fname, nthreads = 8)
pub = UpdateSeuratObject(pub)

## Published informative gene list captured from the `integrated` assay before anything is dropped.
## Used only as a feature list -- the integrated assay's batch-corrected values have no counterpart
## in the new dataset, so they are never used as expression input.
pub_int_features = rownames(pub[["integrated"]])
DefaultAssay(pub) = assay_to_use

new = qs_read(new_fname, nthreads = 8)
new_sct_features = VariableFeatures(new, assay = "SCT")
## Cells whose cluster was excluded under the hybrid scheme (Glut-Arco-4, Glut-Nido-3: confirmed
## artifacts) carry NA here -- dropped, matching every other hybrid-labels notebook, rather than
## transferred onto and then silently reported as an "NA" row.
new = subset(new, cells = colnames(new)[!is.na(new@meta.data[[new_label]])])
DefaultAssay(new) = assay_to_use

## BPCells -> in-memory, see the note above.
counts_mem = as(SeuratObject::LayerData(new, assay = assay_to_use, layer = "counts"), "dgCMatrix")
SeuratObject::LayerData(new, assay = assay_to_use, layer = "counts") = counts_mem
rm(counts_mem); gc(verbose = FALSE)
stopifnot(
  "new-dataset RNA counts did not materialise to an in-memory sparse matrix" =
    inherits(SeuratObject::LayerData(new, assay = assay_to_use, layer = "counts"), "dgCMatrix"),
  "ct_order does not match the hybrid labels present in the new dataset" =
    setequal(ct_order, unique(new@meta.data[[new_label]])),
  "pub_order does not cover the published labels present" =
    setequal(pub_order, unique(pub@meta.data[[pub_label]]))
)

c(published_cells = ncol(pub), new_cells = ncol(new))
```

# Shared, symmetric feature space

```{r features}
shared = intersect(rownames(pub[[assay_to_use]]), rownames(new[[assay_to_use]]))
shared_noloc = shared[!grepl("^LOC", shared)]

## Union of both datasets' own informative genes -- symmetric by construction, see the note above.
transfer_features = intersect(union(pub_int_features, new_sct_features), shared_noloc)

tibble(
  published_genes = nrow(pub[[assay_to_use]]),
  new_genes = nrow(new[[assay_to_use]]),
  shared = length(shared),
  loc_dropped = length(shared) - length(shared_noloc),
  pub_informative = length(pub_int_features),
  new_informative = length(new_sct_features),
  transfer_features = length(transfer_features)
)
```

```{r feature-checks}
stopifnot(
  "no shared non-LOC genes -- gene symbols are probably not comparable between the two annotations" =
    length(shared_noloc) > 1000,
  "too few informative features survive in both datasets to anchor on" =
    length(transfer_features) > 500,
  "a LOC gene survived into the transfer feature set" =
    !any(grepl("^LOC", transfer_features))
)
cat(length(transfer_features), "features used for anchor-finding (both directions)\n")
```

# Normalise

The published `data` layer arrived holding raw counts (confirmed above), so both datasets are
normalised here with the same method rather than trusting either object's stored state.

```{r normalize}
pub = NormalizeData(pub, assay = assay_to_use, verbose = FALSE)
new = NormalizeData(new, assay = assay_to_use, verbose = FALSE)

## Checked rather than assumed, precisely because "data looks populated" turned out to mean nothing
## here: a log-normalised matrix is non-integer and far smaller in magnitude than its source counts.
norm_check = function(obj, nm) {
  d = as(SeuratObject::LayerData(obj, assay = assay_to_use, layer = "data")[, 1:200], "dgCMatrix")
  cnt = as(SeuratObject::LayerData(obj, assay = assay_to_use, layer = "counts")[, 1:200], "dgCMatrix")
  tibble(object = nm, data_max = max(d@x), counts_max = max(cnt@x),
         data_all_integer = all(d@x == round(d@x)))
}
norm_summary = bind_rows(norm_check(pub, "published"), norm_check(new, "new"))
stopifnot(
  "a `data` layer still looks like raw counts after NormalizeData -- transfer input would be wrong" =
    !any(norm_summary$data_all_integer)
)
norm_summary
```

# Transfer, both directions

One helper, called twice with the roles swapped, so the two directions cannot drift apart in method
-- only in which dataset is the reference.

```{r transfer-fn}
#| message: false
## Returns a tibble of per-cell predictions for `query_obj`, transferring each column named in
## `refdata` from `ref_obj`. Anchors cached per direction, with the cache validated against the
## actual cell sets and feature set rather than just checked for existence -- the failure mode a bare
## file.exists() would hand back silently.
run_transfer = function(ref_obj, query_obj, refdata, direction) {
  cache_fname = file.path(out_dir, sprintf("anchors_%s.qs2", direction))

  cache_ok = FALSE
  if (!redo$anchors && file.exists(cache_fname)) {
    cached = qs_read(cache_fname, nthreads = 8)
    cache_ok = setequal(cached$ref_cells, colnames(ref_obj)) &&
      setequal(cached$query_cells, colnames(query_obj)) &&
      setequal(cached$features, transfer_features)
    if (!cache_ok) {
      warning("Anchor cache for direction '", direction,
              "' does not match the current cells / features -- recomputing.", call. = FALSE)
    }
  }

  if (cache_ok) {
    anchors = cached$anchors
  } else {
    anchors = FindTransferAnchors(
      reference = ref_obj, query = query_obj,
      normalization.method = "LogNormalize",
      reference.assay = assay_to_use, query.assay = assay_to_use,
      features = transfer_features,
      reduction = "pcaproject", dims = 1:n_dims, verbose = FALSE
    )
    qs_save(list(anchors = anchors, ref_cells = colnames(ref_obj),
                 query_cells = colnames(query_obj), features = transfer_features),
            cache_fname, nthreads = 8)
  }

  out = tibble(cell = colnames(query_obj))
  for (nm in names(refdata)) {
    pred = TransferData(anchorset = anchors,
                        refdata = ref_obj@meta.data[[refdata[[nm]]]],
                        weight.reduction = "pcaproject", dims = 1:n_dims, verbose = FALSE)
    out[[nm]] = pred$predicted.id
    out[[paste0(nm, "_score")]] = pred$prediction.score.max
  }
  out
}
```

## Direction 1: published -> new

```{r transfer-fwd}
#| message: false
fwd = run_transfer(pub, new,
                   refdata = list(pred_pub_cluster = pub_label, pred_pub_position = pub_position),
                   direction = "pub_to_new") %>%
  mutate(own_label = new@meta.data[[new_label]][match(cell, colnames(new))],
         position = new$position[match(cell, colnames(new))])
write_csv(fwd, file.path(out_dir, "predictions_pub_to_new.csv"))
head(fwd)
```

## Direction 2: new -> published

```{r transfer-rev}
#| message: false
rev_ = run_transfer(new, pub,
                    refdata = list(pred_new_label = new_label),
                    direction = "new_to_pub") %>%
  mutate(own_label = pub@meta.data[[pub_label]][match(cell, colnames(pub))],
         species = pub$species[match(cell, colnames(pub))],
         position = pub@meta.data[[pub_position]][match(cell, colnames(pub))])
write_csv(rev_, file.path(out_dir, "predictions_new_to_pub.csv"))
head(rev_)
```

```{r save-objects}
new$pred_pub_cluster = fwd$pred_pub_cluster[match(colnames(new), fwd$cell)]
new$pred_pub_cluster_score = fwd$pred_pub_cluster_score[match(colnames(new), fwd$cell)]
new$pred_pub_position = fwd$pred_pub_position[match(colnames(new), fwd$cell)]
pub$pred_new_label = rev_$pred_new_label[match(colnames(pub), rev_$cell)]
pub$pred_new_label_score = rev_$pred_new_label_score[match(colnames(pub), rev_$cell)]

qs_save(new, file.path(out_dir, "obj_new_predicted.qs2"), nthreads = 8)
qs_save(pub, file.path(out_dir, "obj_pub_predicted.qs2"), nthreads = 8)
```

# Per-direction composition

```{r composition-fn}
## Row-normalised composition matrix: each row is one source cluster's cells distributed over the
## labels they were predicted as, so rows sum to 1 and read as "what this cluster looks like",
## independent of cluster size.
composition = function(df, from_col, to_col, from_order, to_order) {
  m = df %>%
    count(.data[[from_col]], .data[[to_col]]) %>%
    pivot_wider(names_from = all_of(to_col), values_from = n, values_fill = 0) %>%
    column_to_rownames(from_col) %>%
    as.matrix()
  m = m[intersect(from_order, rownames(m)), , drop = FALSE]
  m = m[, intersect(to_order, colnames(m)), drop = FALSE]
  sweep(m, 1, rowSums(m), FUN = "/")
}

comp_heatmap = function(m, row_title, column_title, fname) {
  scale_factor = 0.16
  hm = Heatmap(m, name = "frac of\nsource cluster",
               cluster_rows = FALSE, cluster_columns = FALSE,
               col = colorRamp2(seq(0, 1, length.out = 9),
                                scales::brewer_pal(palette = "Greys")(9)),
               border = TRUE, border_gp = gpar(lwd = 0.5),
               height = unit(nrow(m) * scale_factor, "in"),
               width = unit(ncol(m) * scale_factor, "in"),
               row_names_gp = gpar(fontsize = 6), column_names_gp = gpar(fontsize = 6),
               row_title = row_title, row_title_gp = gpar(fontsize = 8),
               column_title = column_title, column_title_gp = gpar(fontsize = 8))
  draw(hm)
  pdf(file.path(out_dir, fname),
      height = nrow(m) * scale_factor + 3, width = ncol(m) * scale_factor + 3)
  draw(hm)
  dev.off()
  invisible(hm)
}
```

```{r composition-fwd}
#| fig-width: 10
#| fig-height: 9
m_fwd = composition(fwd, "own_label", "pred_pub_cluster", ct_order, pub_order)
comp_heatmap(m_fwd, "hybrid label (new dataset)", "predicted Colquitt 2021 label",
             "composition_pub_to_new.pdf")
```

```{r composition-rev}
#| fig-width: 11
#| fig-height: 8
m_rev = composition(rev_, "own_label", "pred_new_label", pub_order, ct_order)
comp_heatmap(m_rev, "Colquitt 2021 label", "predicted hybrid label (new dataset)",
             "composition_new_to_pub.pdf")
```

# Reciprocal best match

The actual result. For each hybrid cluster, its top predicted published label; for each published
cluster, its top predicted hybrid label; and whether the two agree. `reciprocal = TRUE` means the
pair is each other's top match in both directions -- a correspondence a one-way transfer cannot
manufacture, since every cell gets assigned something regardless.

```{r reciprocal}
top_fwd = fwd %>%
  count(own_label, pred_pub_cluster) %>%
  group_by(own_label) %>%
  mutate(frac_fwd = n / sum(n)) %>%
  slice_max(frac_fwd, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(hybrid_label = own_label, pub_match = pred_pub_cluster, frac_fwd,
            n_new_cells = n)

top_rev = rev_ %>%
  count(own_label, pred_new_label) %>%
  group_by(own_label) %>%
  mutate(frac_rev = n / sum(n)) %>%
  slice_max(frac_rev, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(pub_label = own_label, hybrid_match = pred_new_label, frac_rev,
            n_pub_cells = n)

mean_scores = fwd %>%
  group_by(own_label) %>%
  summarise(mean_score_fwd = mean(pred_pub_cluster_score), .groups = "drop")

reciprocal = top_fwd %>%
  left_join(top_rev, by = c("pub_match" = "pub_label")) %>%
  left_join(mean_scores, by = c("hybrid_label" = "own_label")) %>%
  mutate(reciprocal = !is.na(hybrid_match) & hybrid_match == hybrid_label,
         hybrid_label = factor(hybrid_label, levels = ct_order)) %>%
  arrange(desc(reciprocal), hybrid_label) %>%
  select(hybrid_label, pub_match, frac_fwd, hybrid_match, frac_rev, reciprocal,
         mean_score_fwd, n_new_cells, n_pub_cells)

write_csv(reciprocal, file.path(out_dir, "reciprocal_best_match.csv"))
reciprocal
```

```{r reciprocal-summary}
reciprocal %>%
  mutate(class = case_when(hybrid_label %in% song_hybrid ~ "song nucleus",
                           grepl("^Glut", hybrid_label) ~ "other Glut",
                           grepl("^GABA", hybrid_label) ~ "GABA",
                           TRUE ~ "non-neuronal")) %>%
  group_by(class) %>%
  summarise(n_clusters = n(), n_reciprocal = sum(reciprocal),
            frac_reciprocal = mean(reciprocal),
            mean_frac_fwd = mean(frac_fwd), .groups = "drop")
```

# Song-nucleus clusters specifically

```{r song-focus}
song_detail = fwd %>%
  filter(own_label %in% song_hybrid) %>%
  count(own_label, pred_pub_cluster) %>%
  group_by(own_label) %>%
  mutate(frac = n / sum(n)) %>%
  slice_max(frac, n = 4) %>%
  ungroup() %>%
  mutate(own_label = factor(own_label, levels = song_hybrid)) %>%
  arrange(own_label, desc(frac))
write_csv(song_detail, file.path(out_dir, "song_cluster_predictions.csv"))
song_detail
```

```{r song-reverse}
## The same populations from the other side: which hybrid label do the published HVC/RA glutamatergic
## clusters map onto?
pub_glut_detail = rev_ %>%
  filter(grepl("_Glut-", own_label)) %>%
  count(own_label, pred_new_label) %>%
  group_by(own_label) %>%
  mutate(frac = n / sum(n)) %>%
  slice_max(frac, n = 3) %>%
  ungroup() %>%
  mutate(own_label = factor(own_label, levels = pub_order)) %>%
  arrange(own_label, desc(frac))
write_csv(pub_glut_detail, file.path(out_dir, "pub_glut_predictions.csv"))
pub_glut_detail
```

```{r score-plot}
#| fig-width: 9
#| fig-height: 6
## Prediction score per hybrid cluster (forward direction). A cluster with a confident, consistent
## counterpart sits high and tight; a cluster the published data has no counterpart for -- which is
## most of the non-song set, since it sampled only HVC and RA -- sits low and/or wide. That is the
## expected outcome for them, not a transfer failure.
p_score = fwd %>%
  mutate(own_label = factor(own_label, levels = rev(ct_order)),
         class = case_when(grepl("^Glut", own_label) ~ "Glut",
                           grepl("^GABA", own_label) ~ "GABA",
                           TRUE ~ "Non-neuron")) %>%
  ggplot(aes(own_label, pred_pub_cluster_score, color = class)) +
  geom_hline(yintercept = 0.5, linetype = 2, color = "grey50", linewidth = 0.25) +
  geom_boxplot(outlier.size = 0.2, outlier.alpha = 0.3, linewidth = 0.25) +
  coord_flip() + ylim(0, 1) +
  labs(x = NULL, y = "max prediction score (published -> new)") +
  theme_cowplot(font_size = 8) +
  theme(legend.position = "bottom", legend.title = element_blank())
print(p_score)
save_plot(file.path(out_dir, "prediction_scores_pub_to_new.pdf"), p_score,
          base_height = length(ct_order) * 0.14 + 1.5, base_width = 6)
```

# Session info

```{r session}
sessionInfo()
```
